### 2016-03-14 tests/hgf_benchmarking/doc/tea_main_text.tex
Added benchmarking info
Benchmarks added.

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%          <8 June 2012>              %%
%%                                     %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

%%% loading packages, author definitions

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass[linenumbers, doublespacing]{bmcart}

%%% Load packages
\usepackage{amsthm,amsmath}
\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage{lmodern}% http://ctan.org/pkg/lm

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\def\includegraphic{}
\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Methodology}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Tissue Enrichment Analysis: TEA}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   % corref={aff1},                       % id of corresponding address, if any
   email={dangeles@caltech.edu}   % email address
]{\inits{DA}\fnm{David} \snm{Angeles-Albores}}
\author[
   addressref={aff2},                   % id's of addresses, e.g. {aff1,aff2}
   email={raymond@caltech.edu}   % email address
]{\inits{RYL}\fnm{Raymond Y} \snm{Lee}}
\author[
   addressref={aff2},                   % id's of addresses, e.g. {aff1,aff2}
   email={azurebrd@caltech.edu}   % email address
]{\inits{JC}\fnm{Juancarlos} \snm{Chan}}
\author[
   addressref={aff1},
   corref= {aff1},
   email={pws@caltech.edu}
]{\inits{PWS}\fnm{Paul W} \snm{Sternberg}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%                           % unique id
  \orgname{California Institute of Technology, Division of Biology and Biological Engineering}, % university, etc
  \street{1200 E California Blvd},                     %
  \postcode{91125}                                % post or zip code
  \city{Pasadena},                              % city
  \cny{US}                                    % country
}

% raymond's address?
\address[id=aff2]{%
  \orgname{California Institute of Technology},
  \street{1200 E California Blvd},                     %
  \postcode{91125}                                % post or zip code
  \city{Pasadena},                              % city
  \cny{US}                                    % country
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{fmbox}% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract} % abstract
Over the last ten years, there has been an explosive development in tools 
capable of measuring gene expression. These tools generate a large number
of gene targets, but understanding these datasets, and forming hypotheses 
based on them remains challenging.

We present a method for detecting tissue enrichment in \emph{C. elegans} using the tissue ontology for this organism. We also present an efficient method for trimming the ontology that results in concise yet useful output.

Our tool, Tissue Enrichment Analysis (TEA),
can be found at www.wormbase.org/tea
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{Gene Ontology}
\kwd{Tissue Ontology}
\kwd{Wormbase}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
	RNA-seq and other high-throughput methods in biology have the ability to identify thousands of genes that are altered between conditions. These genes are often correlated in their biological characteristics or functions, but identifying these functions remains challenging. In order to interpret these long lists of genes, biologists need to abstract genes into fewer terms that are biologically relevant in order to form hypotheses about what is happening in the data. One such abstraction method relies on Gene Ontology (GO). GO provides a controlled set of hierarchically ordered terms in the form of an directed acyclic graph\cite{TheGeneOntologyConsortium2000, Ontology2009, TheGeneOntologyConsortium2015} that provide detailed information about the molecular, cellular or biochemical functions of the gene among others. For a given gene list, certain software programs can query whether a particular gene is enriched\cite{Mi2009, McLean2010, Huang2009}. However, GO is often difficult to interpret due to the large number of terms associated with a given gene. There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.  A common tool for GO analysis, DAVID, clusters terms into broad categories that are amenable to exploration by researchers \cite{Huang2007}, whereas PANTHER, a different software package \cite{Mi2009, Mi2013}, attempts to solve this issue by employing a manually reduced ontology, GOslim (pers. comm.).

	Here we provide a new framework that analyses user-input list for enrichment of specific tissues. We believe that tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation. As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition. Our analysis also cuts down on result verbosity by filtering the ontology before testing using a small set of well-defined criteria to remove terms that don’t contribute extra information. To our knowledge, such filtering has never been performed in an algorithmic fashion for an ontology before — indeed, tools such as DAVID do not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms. We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.

	We built our software using a pre-established tissue ontology for the worm, C. elegans \cite{Lee2003}. The C. elegans  database, Wormbase\cite{Harris2014}, maintains a carefully curated list of gene expression data from GFP-reporters. We use this gold-standard list to develop a tissue enrichment analysis that reliably identifies even small tissues and show that we can reliably discriminate between embryonic and larval tissues. Our tool is available in Wormbase at the address http://mangolassi.caltech.edu/~azurebrd/cgi-bin/testing/amigo/getWithPost.cgi and provides users with a text-based file of the enrichment results as well as a simple and clear graph of the results that exhibit the largest fold-change enrichment. Although we present results here for the worm, we note that our software is species agnostic, and we are working to integrate tissue ontologies from other databases to provide a broader service to the community.

\section*{Methods}
\subsection*{Generating a Useful Dictionary}
\subsubsection*{Reducing term redundancy through a similarity metric}
As a first step to generate our tissue enrichment software, we wished to select tissue terms that were reasonably well-annotated, yet specific enough to provide insight and not redundant with other terms. We also wanted to avoid testing tissues at levels where redundancy becomes problematic. For example, several left and right neurons have at least 25 annotating genes and we may want to include them for enrichment testing. However, many left/right neuronal sisters have almost entirely the same annotations, with at most one or two gene differences between them. We reasoned that when two tissues have almost identical annotations, we cannot have statistical confidence in differentiating between them. As a result, testing these sister tissues provides no additional information compared with testing only the parent node to these sisters. We refer to such sisters as ‘redundant’. In order to identify redundancy, we defined a similarity metric

\begin{eqnarray}\label{similarity def}
	s_i = \frac{|g_i|}{|\bigcup_{i= 0}^k g_i|}
\end{eqnarray}

Where $s_i$ is the similarity for a tissue $i$ eith $k$ sisters; $g_i$ refers to the set of tissues associated with tissue $i$ and $|g|$ refers to the cardinality of set $g$. For a given set of sisters, we called them redundant if they exceeded a given similarity threshold. We envisioned two possible criteria and built different dictionaries using each one. Under a threshold criteron `any' with parameter $S$ between $(0, 1)$, a given set of sisters $j$ was considered redundant if the condition

\begin{eqnarray}\label{any threshold}
	s_{i, j} > S
\end{eqnarray}

was true for any sister $i$ in set $j$. Under a threshold criterion `avg' with parameter $S$, a given set of sisters $j$ was considered redundant if the condition

\begin{eqnarray}\label{avg threshold}
	\mathrm{E}[s_i]_j > S
\end{eqnarray}

was true for the set of sisters $j$ (see figure~\ref{fig:simdiagram}).

\subsubsection*{Terminal branch terms and parent terms can be safely removed in an algorithmic fashion }
Another problem arises from the fact that the tissue ontology is scarcely populated at this point in time. Many nodes have 0-10 annotations, which we consider too few to accurately test. To solve this issue, we implemented a straightforward trimming algorithm. For a given terminal node, we test whether the node has more than a threshold number of annotations. If it does not, the node is removed. The next node in the branch is tested and removed recursively until a node which satisfies the condition is found. At that point, no more nodes can be removed from that branch. This is guaranteed by the structure of the ontology: Parent nodes inherit all of the annotations of all of their descendants, so the number of annotating terms monotonically increases with increasing term hierarchy (see figure~\ref{fig:trim_ends}). In this way, we ensure that our term dictionary includes only those tissues that are considered sufficiently well annotated for statistical purposes.

	Finally, we also wanted to remove as many terms as possible from the dictionary with the goals of reducing covariance between terms, decreasing multiple testing and removing as many non-informative terms as possible. Decreasing covariance between terms is important because we employ a frequentist approach that assumes all terms are independent. Large covariation coefficients between some terms means that if one of these tissues tests significant, the other terms are much more likely to pass significance testing as well. This makes adequate correction for false positive rates considerably more difficult. Moreover, from a data analysis perspective, we reasoned that, for any parent node, if all its daughters were selected for testing, there was no additional benefit to test the parent. In other words, if all the daughter nodes are tested, there is little additional information to be gained by including the parent node.  To address this issue we removed parent nodes from the analysis if all their daughter nodes passed the annotation threshold (see figure~\ref{fig:trim_roots}).

\subsubsection*{Filtering greatly reduces the number of nodes used for analysis}
	By itself, each of these filters can reduce the number of nodes employed for analysis. Notably, these filters are not all commutative -- while trimming and redundancy filtering are commutative, applying the ceiling filter is not commutative with either the trimming or the redundancy filter. If the ceiling filter is  applied before any other filter, only terminal nodes will remain, since all the parents have complete daughter sets. Since terminal nodes are the most poorly annotated, after applying the remaining filters very few nodes will be left behind if any. On the other hand, applying the ceiling operator after trimming and redundancy filtering  will result in greater numbers of nodes. We always applied the ceiling at the end. For validation (see below) we made a number of different dictionaries. The original ontology has 1675 terms with more than 5 gene annotations. After filtering, dictionary sizes ranged from 21 to a maximum of 400 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by tenfold by application of a few simple filters.

	These filters were used to compile a static dictionary that we employ for all analyses. Because we have integrated our scripts to draw on the WormBase databases, our dictionary will remain up to date as tissue expression data improves. Our completed static trimmed dictionary is available for download at the following ftp URL: XX. The final dictionary includes XX tissues for testing, and has XX annotating genes. All code was implemented in Python.

\subsection*{Tissue enrichment testing via a hypergeometric model}
	Having built a static dictionary, we generated a Python script that implements a significance testing algorithm based on the hypergeometric model. Briefly, the hypergeometric model assumes the existence of an urn with a pre-determined number of balls inside it. The balls can be painted one of several colors. The hypergeometric model provides an answer to the question: If an individual removes $N$ balls, what is the probability of observing $n_i$ balls of color $i$, if the balls are selected without replacement? Mathematically, this is expressed as:

\begin{eqnarray}\label{hypergeometric}
	\mathrm{P}(n_i | N, m_1,\ldots, m_k, M) = \frac{\dbinom{m_i}{n_i} \dbinom{M-m_i}{N - n_i}}{\dbinom{N}{n_i}}
\end{eqnarray}

	Here, $n_i$ is the number of balls of type $i$ drawn, $N$ is the total number of draws, $m_i$ is tissue $i$ and $M = \sum_i m_i$ is the total number of balls in the urn. In our specific case, $M_i$ is equal to the total number of annotations in our dictionary. $N$ is found by taking the user-input list and removing any genes that are not in our annotation dictionary. The remaining genes are then associated with their annotation profiles — if a tissue is associated with $s$ tissues, it generates $s$ balls of $s$ colors. Our program counts the number of times each tissue appears in the user list, and calculates the probability of having withdrawn as many or more balls for each tissue in the user list. Due to the discrete nature of the hypergeometric distribution, this algorithm can generate artifacts when the list is small. To avoid spurious results, a tissue is never considered significant if there are no annotations for it in the user-provided list.

	Once the probability of drawing the labels has been quantified, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm\cite{Benjamini1995}. Genes that have a q-value less than a given alpha are considered significant. Our default setting is to set the alpha threshold at 0.1, but users will be able to modify this value either in batch or in our web application. The program returns a text-based table showing the tissues that tested significant, along with their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change (observed hits / expected hits). Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the largest enrichment fold change. Our software relies heavily on the Pandas, Numpy, Seaborn and SciPy modules to perform all statistical testing and data handling\cite{McKinney2011, VanDerWalt2011, Oliphant2007}.

	Our software is implemented in an easy to use GUI within WormBase. Users input a gene-list (see figure~\ref{fig:GUI}) using any valid gene name for \emph{C. elegans}. These names are processed into standard WBIDs and the result is displayed in the same window in an easy to read format containing all the relevant information, and a graph of the results is also displayed (see figure~\ref{fig:GUIresults}).

\subsection*{Validation of the algorithm and parameter selection}
	In order to select an appropriate dictionary and validate our tool, we found a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues\cite{}. Some of these studies went on to use GFP to identify expression patterns and for this reason we generated a clean Since the expression data is curated from GFP expression at this time and does not include RNA-seq data, these gold standards are statistically independent from the dataset. We wanted to select a dictionary which included enough terms to be specific beyond the largest C. elegans tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change. Selection of a dictionary based only on minimization of spurious results would result in a dictionary with a large number of annotations per tissue, and would therefore include only the major tissues. On the other hand, selecting a dictionary that can detect smaller tissues will bias us towards tissues with lesser annotations. To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.

	As a first attempt to select a good dictionary, we generated all the possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using ‘average’ or ‘any’ thresholding criteria for the latter (see table ~\ref{tab:DictionarySpecs}). For these dictionaries, the number of tissues tested ranged from 97 to 676. The number of tissues was inversely correlated to the minimum annotation, as expected, and was largely insensitive to the redundancy threshold, at least in the range we explored (0.9-1). Next, we analyzed all 30 datasets using each dictionary. Because of the large number of results, instead of analyzing each set of terms individually, we pooled all results for a given dictionary into histograms. When we analyzed the distribution of significant q-values for the dictionaries, we found that the similarity threshold mattered relatively little for any dictionary. We also noticed that the `any' thresholding method resulted in tighter histograms with a mode closer to 0 (data not shown). For this reason, we chose the ‘any’ method for dictionary generation. The average q-value increased with decreasing annotation cut-off (see figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant the trade-off between power and term specificity is. Based on these observations, we ruled out the dictionary with the 100 annotation cut-off - it had the fewest terms and its q-values were not low enough to compensate the trade-off in specificity.

	%dictionaries that work well
	To select between dictionaries generated between 50, 33 and 25 annotation cut-offs, and also to ensure the terms that are selected as enriched by our algorithm are reasonable, we looked in detail at the enrichment analysis results. 
	%Work well
	Most results were highly comparable and in line with what was expected. For some sets, all dictionaries seemed to perform well. For example, in our `all neuron enriched set'~\ref{WBPaper00037950} the result was an amalgamation of neuron related terms including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays regardless of the dictionary used. On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~\ref{WBPaper00037950}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells -- at the 33-cutoff, our tool identified `Z2' and `Z3' as being five-fold enriched; whereas at the 25 gene-cutoff the terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
	%Differences between dictionaries
	We queried an embryonic stage intestine precursor associate geneset~\ref{WBPaper00037950}. Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary. However, the 33 cutoff dictionary suggested, probably correctly, that the E lineage was heavily enriched in this set. 
	%Bad stuff.
	Not all queries worked equally well. For example, a number of intestinal enriched genes sets~\ref{WBPaper00026980, WBPaper00037950} were not enriched in intestine in any dictionary, but they were enriched for pharynx- and hypodermis-related terms. We were somewhat surprised that intestinal gene sets performed poorly, since the intestine is a relatively well-annotated tissue.
	%Dictionary intra-agreement
	We also assessed the internal agreement of our tool by using independent gene-sets that we expected to be enriched in the same tissues. We had two independent pan-neuronal sets~\ref{}; two independent PVD enriched sets~\ref{}; two independent GABAergic gene sets~\ref{}; two independent pharyngeal gene sets; and two independent intestinal gene sets~\ref{}. Overall, the tool seems to have good internal agreement. On most sets, the same terms were enriched, although order was somewhat variable. However, most high-scoring terms were preserved between gene sets. The intestinal gene-sets and pharyngeal gene sets comparisons were exceptions, since at least one gene set was missing each for intestine and pharynx in every dictionary, so we didn't consider them as informative for assessing internal agreement.

	 All comparisons can be found online at: www.XXX.com. Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95 redundancy threshold using the `any' criterion. seemed to perform well, with a good balance between specificity, verbosity and accuracy, so we selected this parameter set to generate our static dictionary.

\section*{Results}
	We applied our tool to the RNA-seq datasets developed by Engelmann et al. \cite{Engelmann2010} in order to attempt to gain further understanding of the biology underlying these datasets. Engelmann et al.  exposed young adult worms to 5 different pathogenic bacteria or fungi for 24 hours, after which mRNA was extracted from the worms for sequencing. We obtained the genes that Engelmann et al identified as up- or down- regulated in their assay, and ran TEA using these lists. Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade. Strikingly, 4 out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues (in the case of Harposporium sp) amongst the down-regulated genes. A possible explanation for this might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioural response against the pathogen. Indeed, several studies\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C. elegans} uses chemosensory neurons to identify pathogens Interestingly, one bacterium did not exhibit the same pattern of down-regulation of neuronal-associated genes. \emph{E. faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E. faecalis} may have a different pathogenic profile. Up-regulated tissues, when detected, included the hypodermis and excretory duct. 
	Our results highlight the involvement of various \emph{C. elegans} neuronal tissues in pathogen defense and/or illness.

\section*{Discussion}

We have presented a tissue enrichment analysis tool that employs a standard hypergeometric model to test the \emph{C. elegans} tissue ontology. We have also presented the first, to our knowledge, ontology trimming algorithm. This algorithm, which is very easy to execute, places strong limits on the number of terms selected for testing. Due to the nature of all ontologies as hierarchical, acyclical graphs with term inheritance, term annotations are correlated along any given branch. This correlation reduces the benefits of including all terms for statistical analysis - for any given term along a branch, if that term passes significance, there is a high probability that many other terms along that branch will also pass significant. If the branch is enriched by random chance, error propagation along a branch means that many more false positives will follow. Thus, a researcher might be misled by the number of terms of correlated function and assign importance to this finding; the fact that the branching structure of GO amplifies false positive signals is a powerful argument for either reducing branch length or branch intracorrelation, or both. On the other hand, if a term is actually enriched, we argue that there is little benefit to presenting the user with additional terms along that branch. Instead, a user will benefit most from testing sparsely along the tree at a suitable specificity for hypothesis formation. Related terms of the same level should only be tested when there is sufficient annotation to differentiate, with statistical confidence, whether one term is enriched above the other (see SI for a back-of-the-envelope calculation of when this can be the case). Our algorithm reduces branch length by identifying and removing nodes that are insufficiently annotated and parents that are likely to include sparse information.

It is important to note that our tool is not the first tissue enrichment model for the worm that has been reported. Chikina \emph{et al} \cite{Chikina2009} report a tissue enrichment model based on an SVM classifier that has been trained on microarray studies. SVM classifiers are powerful tools capable of great sensitivity, but they require continuous retraining as tissue expression data widens. Our tool benefits from the fact that it will be integrated in WormBase and will therefore be updated continuously as new data is integrated.

We have tried hard to benchmark our tool well. However, our analysis suffers from the drawback that is very hard to benchmark negative controls. Even for our set of positive controls, the statistical analysis sometimes throws out unexpected results. For example, the embryonic germline precursor gene set had the term `AB' as the most enriched term in the dictionaries with cut off of 25 and 33. Is this an error, or does this hint at new biology? Although we were unable to determine false-positive and false-negative rates, we don't believe this should deter scientists from using our tool. Rather, we encourage researchers to use our tool carefully as a guide, integrating evidence from multiple sources to inform the most likely hypotheses. As with any other tool based on statistical sampling, our analysis is most vulnerable to bias in the data collection stage. For example, we know that tissue expression reports are negatively biased against germline expression due to the difficulty associated with extrachromosomal array expression in that tissue. Support from the community will be crucial in correcting these flaws going forward; indeed, without the community reports of tissue expression this tool would not be possible.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{backmatter}

\section*{Competing interests}
 	The authors declare that they have no competing interests.
\section*{Author's contributions}
    DA and PWS conceived of the project; DA developed algorithm; 
	RYL made intellectual contributions to the project; RYL and JC 
	developed the web GUI.
\section*{Acknowledgements}
	We would like to acknowledge all members of the Sternberg
	lab for helpful discussion. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
%\nocite{label}

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}

% Figures describing algorithm
  \begin{figure}[h!]
  \caption{\csentence{Schematic diagram of annotations for two sisters.}
      The parent node (green) contains at least as many annotations as the union of the two sisters.
	  These two sisters share annotations extensively. Therefore they are too similar and should be removed.
	  }
	  \label{fig:simdiagram}
  \end{figure}

\begin{figure}[h!]
  \caption{\csentence{Schematic showing terminal node removal.}
      Nodes with less than a threshold number of genes are trimmed (light red) and discarded from the dictionary. Here, the threshold is 25 genes.}
	  \label{fig:trim_ends}
      \end{figure}

\begin{figure}[h!]
  \caption{\csentence{Schematic showing root node removal.}
      We trim parent nodes (light red) if all their daughter nodes have more than the threshold number of annotations. Here, the threshold is 25 genes.}
	  \label{fig:trim_roots}
      \end{figure}

%gui
\begin{figure}[h!]
  \caption{\csentence{Screenshot of the web GUI.}
      }
	  \label{fig:GUI}

      \end{figure}
\begin{figure}[h!]
  \caption{\csentence{Screenshot of results from web GUI.}
      }
	  \label{fig:GUIresults}
      \end{figure}

% statistical testing and validation figures
%q plot
\begin{figure}[h!]
  \caption{\csentence{Kernel density estimates for 30 gold standard datasets.}
      We ran TEA on 30 datasets we believed to be enriched in particulae tissues and pooled all the results to observe the distribution of q-values. The mode of the distribution for dictionaries with annotation cut-offs of 100 and 50 genes are very similar; however, when the cut-off is lowered to 25 genes, the mode of the distribution shifts to the left, potentially signalling a decrease in measurement power.}
	  \label{fig:qvals}
      \end{figure}

% %ovd-oll
\begin{figure}[h!]
  \caption{\csentence{Comparison of Enrichment Results for dictionary size 50 (left) and 25 (right) for a PVD-OLL enriched gene set.}
      Left, at 50 annotation cut-off, TEA singles PVD as highly enriched. Other mechanosensory neurons are also enriched . Right, when the dictionary cut-off is set to 25, TEA shows embryonic tissues that are unrelated to the PVD and OLL lineages.}
	  \label{fig:ovdoll}
      \end{figure}
%
%
%
%
%results figures - engelmann et al reanalysis
\begin{figure}[h!]
  \caption{\csentence{Genes altered in \emph{C. elegans} after 24hr exposure to \emph{D. coniospora} (fungus)}
      Figure legend text.}
      \end{figure}
\begin{figure}[h!]
  \caption{\csentence{Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Harposporium sp.} (fungus)}
      Figure legend text.}
      \end{figure}
\begin{figure}[h!]
  \caption{\csentence{Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Serratia marcescens} (bacteria)}
      Figure legend text.}
      \end{figure}
\begin{figure}[h!]
  \caption{\csentence{Genes altered in \emph{C. elegans} after 24hr exposure to \emph{E. faecalis} (bacteria)}
      Figure legend text.}
      \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Parameter specifications and number of tissues for all dictionaries.}
      \begin{tabular}{cccc}
        \hline
      \end{tabular}
	  \label{tab:DictionarySpecs}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Supplementary Information}
	Complete results from benchmarking analysis
  \subsection*{Additional file 2 --- Supplementary Information}
    Complete results from re-analysis of Engelmann et al
  \subsection*{Additional file 3 --- IPython Notebook}
    Tutorial for users interested in batch script generation using our software.

\end{backmatter}
\end{document}

### 2016-03-14 build/scripts-3.5/tea.py

@@ -1,99 +1,0 @@
-#!/Users/dangeles/anaconda3/envs/testenv/bin/python
-
-import hypergeometricTessts as hgt
-
-
-if __name__ == '__main__':
-    
-    import re
-    
-    path= './'
-    os.chdir(path)
-    
-    import argparse
-    import matplotlib
-    matplotlib.use('Agg')
-#    from matplotlib.pylab import * 
-    import matplotlib.pyplot as plt
-    import seaborn as sns
-    sns.set_context('paper')
-    sns.set_style('whitegrid')
-#    sns.despine(left= True)
-#    sns.despine(trim= True)
-
-    defQ= 0.1    
-    
-    parser = argparse.ArgumentParser(description='Run TEA.')
-    parser = argparse.ArgumentParser()
-    parser.add_argument("tissue_dictionary", help= 'The full path to the tissue dictionary provided by WormBase')
-    parser.add_argument("gene_list", help= 'The full path to the gene list (WBIDs) you would like to analyse in .csv format')
-    parser.add_argument('-t', "--title", help= 'Title for your analysis (shouldn\'t include file extension', type= str)
-    parser.add_argument("-q", help= 'Qvalue threshold for significance. Default is {0} if not provided'.format(defQ), type= float)
-    parser.add_argument('-p','--print', help= 'Indicate whether you would like to print results', action= 'store_true')    
-    parser.add_argument('-pl', "--plot", help= 'Indicate whether you would like to plot results.', action= 'store_true')
-    parser.add_argument('-s', "--save", help= 'Boolean variable indicating whether to save your plot or not.', action= 'store_true')
-    args = parser.parse_args()
-    
-    tdf_name= args.tissue_dictionary
-    gl_name= args.gene_list
-    title= args.title
-    
-    if args.q:   
-        q= args.q
-    else:
-        q= defQ
-    
-    if args.print:
-        show= True
-    else:
-        show= False
-        
-    if args.plot:
-        plot= True
-    else:
-        plot= False
-    
-    if args.save:
-        save= True
-                    
-    else:
-        save= False
-            
-    tissue_df= pd.read_csv(tdf_name)
-    
-    gene_list= []
-    with open(gl_name, 'r') as f:
-        gene_list= [x.strip() for x in f.readlines()]
-    
-    df_results, unused= enrichment_analysis(gene_list, tissue_df, alpha= q, show= False)
-    
-    dfname= title+'.csv'
[…]

### 2016-03-14 setup.py

@@ -10,7 +10,7 @@ from setuptools import setup, find_packages
 import os
 import sys
 
-version= '0.13.2'
+version= '0.13.4'
 
 #just type in python setup.py publish and this takes care of publishing to to pypi!
 
@@ -56,7 +56,8 @@ def readme():
 
 setup(
   name = 'tissue_enrichment_analysis',
-  packages = ['tissue_enrichment_analysis'], # this must be the same as the name above
+#  packages = ['tissue_enrichment_analysis'], # this must be the same as the name above
+  packages=find_packages(exclude=("tests",)),
   version = version,
   description = 'This package contains all the software used to implement\
   TEA in WormBase and remotely',

### 2016-03-14 tests/hgf_benchmarking/src/clean_dictionary_maker.py

@@ -1,0 +1,118 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Mar  2 14:24:04 2016
+
+A script to generate 'clean' dictionaries that exclude
+golden standards. 
+
+This script 
+
+@author: dangeles
+"""
+import dictionary_generator as dg
+from dictionary_generator import ontology
+import sys
+x_array= [100, 50, 33, 25]
+thresh_array= [1, .95, .9]
+methods= ['avg', 'any']
+path= '../input/WS252AnatomyDictionary/'
+
+
+ #main solr url
+solr_url = 'http://caprica.caltech.edu:8080/solr/anatomy/'
+#queries must be lambda functions
+#query for terms. Finds terms that have x or more annotating genes
+query_terms= lambda x: 'select?qt=standard&indent=on&wt=json&version=2.2&fl=id&start=0&rows=0&q=document_category:bioentity&facet=true&facet.field=regulates_closure&facet.limit=-1&facet.mincount={0}&facet.sort=count&fq=source:%22WB%22&fq=-qualifier:%22not%22'.format(x)
+#query for relationships. given a wbbt ID, find the nodes connected to it. 
+query_relation= lambda x: "select?qt=standard&fl=topology_graph_json&version=2.2&wt=json&indent=on&rows=1&q=id:%22{0}%22&fq=document_category:%22ontology_class%22".format(x)
+#query for number of genes. given a wbbt ID, find genes assoc. with it. 
+query_genes= lambda x: "select?qt=standard&indent=on&wt=json&version=2.2&fl=id&start=0&rows=10000&q=document_category:bioentity&fq=source:%22WB%22&fq=-qualifier:%22not%22&fq=regulates_closure:%22{0}%22".format(x)
+#query for readable names
+query_readable= "select?qt=standard&fl=id,annotation_class_label&version=2.2&wt=json&indent=on&rows=100000&q=id:*&fq=document_category:ontology_class&fq=-is_obsolete:true"
+queries= [query_terms, query_relation, query_genes, query_readable]
+min_annot= 2
+
+for x in x_array:
+    for threshold in thresh_array:
+        for method in methods:    
+            print(x, threshold, method)
+            sys.stdout.flush()
+            print(solr_url)
+            trial1= ontology('tissue_ontology', x, threshold, method, solr_url)
+            trial1.set_min_cutoff(5)
+            trial1.add_nodes(query_terms, query_readable)
+            trial1.find_node_annotations(query_genes)
[…]

### 2016-03-14 tests/hgf_benchmarking/src/hgf_benchmark_script.py

@@ -1,0 +1,276 @@
+# -*- coding: utf-8 -*-
+"""
+Spyder Editor
+
+
+@david angeles
+dangeles@caltech.edu
+"""
+
+import tissue_enrichment_analysis as tea #the library to be used
+import pandas as pd
+import os
+import importlib as imp
+import numpy as np
+import seaborn as sns
+import matplotlib.pyplot as plt
+import re
+import matplotlib as mpl
+
+sns.set_context('paper')
+
+pd.set_option('display.float_format', lambda x:'%f'%x) 
+
+
+#this script generates a few directories. 
+dirOutput= '../output/'
+dirSummaries= '../output/SummaryInformation/'
+dirHGT25_any= '../output/HGT25_any_Results/'
+dirHGT33_avg= '../output/HGT33_avg_Results/'
+dirHGT33_any= '../output/HGT33_any_Results/'
+dirHGT50_any= '../output/HGT50_any_Results/'
+dirHGT100_any= '../output/HGT100_any_Results/'
+
+DIRS= [dirOutput, dirSummaries, dirHGT25_any, dirHGT33_avg,
+       dirHGT50_any, dirHGT100_any]
+#open the relevant file
+path_sets= '../input/genesets_golden/'
+path_dicts= '../input/WS252AnatomyDictionary/'
+
+
+#Make all the necessary dirs if they don't already exist
+for d in DIRS:
+    if not os.path.exists(d):
+        os.makedirs(d)
+
+#Make the file that will hold the summaries and make the columns. 
+with open(dirSummaries+'ExecutiveSummary.csv', 'w') as fSum:
+    fSum.write('#Summary of results from all benchmarks\n')
+    fSum.write('NoAnnotations,Threshold,Method,EnrichmentSetUsed,TissuesTested,GenesSubmitted,TissuesReturned,GenesUsed,AvgFold,AvgQ,GenesInDict\n')
+
+#==============================================================================
+#==============================================================================
+# # Perform the bulk of the analysis, run every single dictionary on every set
+#==============================================================================
+#==============================================================================
+i= 0
+#look in the dictionaries
+for folder in os.walk(path_dicts):
+    #open each one
+    for f_dict in folder[2]:
+        if f_dict == '.DS_Store':
+            continue
+        tissue_df= pd.read_csv(path_dicts+f_dict)
+        
+        #tobedropped when tissue dictionary is corrected
[…]

### 2016-03-14 tests/hgf_benchmarking/doc/tea_main_text.tex
added citations to text

@@
 \subsection*{Validation of the algorithm and parameter selection}
-	In order to select an appropriate dictionary and validate our tool, we found a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues\cite{}.
+	In order to select an appropriate dictionary and validate our tool, we found a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues\cite{Gaudet2004, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}.
 Some of these studies went on to use GFP to identify expression patterns and for this reason we generated a clean Since the expression data is curated from GFP expression at this time and does not include RNA-seq data, these gold standards are statistically independent from the dataset.
@@
 For some sets, all dictionaries seemed to perform well.
-For example, in our `all neuron enriched set'~\ref{WBPaper00037950} the result was an amalgamation of neuron related terms including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays regardless of the dictionary used.
-On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~\ref{WBPaper00037950}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells -- at the 33-cutoff, our tool identified `Z2' and `Z3' as being five-fold enriched; whereas at the 25 gene-cutoff the terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
+For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} the results were an amalgamation of neuron related terms including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays regardless of the dictionary used.
+On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells -- at the 33-cutoff, our tool identified `Z2' and `Z3' as being five-fold enriched; whereas at the 25 gene-cutoff the terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
 	%Differences between dictionaries
-	We queried an embryonic stage intestine precursor associate geneset~\ref{WBPaper00037950}.
+	We queried an embryonic stage intestine precursor geneset~\cite{Spencer2011}.
 Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary.
@@
 	Not all queries worked equally well.
-For example, a number of intestinal enriched genes sets~\ref{WBPaper00026980, WBPaper00037950} were not enriched in intestine in any dictionary, but they were enriched for pharynx- and hypodermis-related terms.
+For example, a number of intestinal enriched genes sets~\cite{Spencer2011, Pauli2006} were not enriched in intestine in any dictionary, but they were enriched for pharynx- and hypodermis-related terms.
 We were somewhat surprised that intestinal gene sets performed poorly, since the intestine is a relatively well-annotated tissue.
@@
 	We also assessed the internal agreement of our tool by using independent gene-sets that we expected to be enriched in the same tissues.
-We had two independent pan-neuronal sets~\ref{}; two independent PVD enriched sets~\ref{}; two independent GABAergic gene sets~\ref{}; two independent pharyngeal gene sets; and two independent intestinal gene sets~\ref{}.
+We had two independent pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two independent PVD enriched sets~\cite{Spencer2011, Smith2010}; two independent GABAergic gene sets~\cite{Spencer2011, Cinar2005}; two independent pharyngeal gene sets~\cite{Spencer2011, Gaudet2004}; and two independent intestinal gene sets~\cite{Spencer2011, Pauli2006}.
 Overall, the tool seems to have good internal agreement.
@@
 The intestinal gene-sets and pharyngeal gene sets comparisons were exceptions, since at least one gene set was missing each for intestine and pharynx in every dictionary, so we didn't consider them as informative for assessing internal agreement.
-
-	 All comparisons can be found online at: www.XXX.com.
+	 All comparisons can be found online in our Github repository (see Availability of data and materials).
 Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95 redundancy threshold using the `any' criterion.
@@
 Support from the community will be crucial in correcting these flaws going forward; indeed, without the community reports of tissue expression this tool would not be possible.
+
+\section*{Availability of data and materials}
+Our web implementation is available at https://www.wormbase.org/tea. Our software can also be downloaded using Python's pip installer via the command
+
+\texttt{pip install tissue\_enrichment\_tool}
+
+Alternatively, our software is available for download at:
+
+http://dangeles.github.io/TissueEnrichmentAnalysis.
+
+All benchmark gene sets, benchmarking code and figures can also be found at the same address, under the `tests' folder.
 

### 2016-03-14 tests/hgf_benchmarking/doc/tea_main_text.tex
wrote a little bit

@@
 \subsection*{Validation of the algorithm and parameter selection}
-	In order to select an appropriate dictionary and validate our tool, we found a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues\cite{Gaudet2004, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}.
+	In order to select an appropriate dictionary and validate our tool, we found a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}.
 Some of these studies went on to use GFP to identify expression patterns and for this reason we generated a clean Since the expression data is curated from GFP expression at this time and does not include RNA-seq data, these gold standards are statistically independent from the dataset.
@@
 To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.
+>>>>>>> master
 
@@
 	We also assessed the internal agreement of our tool by using independent gene-sets that we expected to be enriched in the same tissues.
-We had two independent pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two independent PVD enriched sets~\cite{Spencer2011, Smith2010}; two independent GABAergic gene sets~\cite{Spencer2011, Cinar2005}; two independent pharyngeal gene sets~\cite{Spencer2011, Gaudet2004}; and two independent intestinal gene sets~\cite{Spencer2011, Pauli2006}.
+We had two independent pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two independent PVD enriched sets~\cite{Spencer2011, Smith2010}; two independent GABAergic gene sets~\cite{Spencer2011, Cinar2005}; two independent pharyngeal gene sets~\cite{Spencer2011, Gaudet2004a}; and two independent intestinal gene sets~\cite{Spencer2011, Pauli2006}.
 Overall, the tool seems to have good internal agreement.
@@
 All benchmark gene sets, benchmarking code and figures can also be found at the same address, under the `tests' folder. 
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{figure}[h!]
-  \caption{\csentence{Genes altered in \emph{C. elegans} after 24hr exposure to \emph{D. coniospora} (fungus)}
-      Figure legend text.}
+  \caption{\csentence{\emph{D. coniospora} enrichment results}
+      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{D. coniospora} (fungus).}
       \end{figure}
 \begin{figure}[h!]
-  \caption{\csentence{Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Harposporium sp.} (fungus)}
-      Figure legend text.}
+  \caption{\csentence{\emph{Harposporium sp.} enrichment results}
+      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Harposporium sp.} (fungus).}
       \end{figure}
 \begin{figure}[h!]
-  \caption{\csentence{Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Serratia marcescens} (bacteria)}
-      Figure legend text.}
+  \caption{\csentence{\emph{Serratia marcescens} enrichment results}
+      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Serratia marcescens} (bacteria).}
       \end{figure}
 \begin{figure}[h!]
-  \caption{\csentence{Genes altered in \emph{C. elegans} after 24hr exposure to \emph{E. faecalis} (bacteria)}
-      Figure legend text.}
+  \caption{\csentence{\emph{E. faecalis} enrichment results}
+      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{E. faecalis} (bacteria)}
       \end{figure}

### 2016-03-15 tests/hgf_benchmarking/doc/tea_main_text.tex
More writing
yay….

@@
 %%                                                                 %%
-%% All additional figures and files should be attached             %%
+%% All additional Figures and files should be attached             %%
 %% separately and not embedded in the \TeX\ document itself.       %%
@@
 %\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass[linenumbers, doublespacing]{bmcart}
+% \documentclass[linenumbers, doublespacing]{bmcart}
+\documentclass{bmcart}
 
@@
 \RequirePackage{natbib}
-%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
-%\RequirePackage{hyperref}
-\usepackage[utf8]{inputenc} %unicode support
+% \RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
+\RequirePackage{hyperref}
+% \usepackage[utf8]{inputenc} %unicode support
 %\usepackage[applemac]{inputenc} %applemac support if unicode package fails
@@
 \usepackage{lmodern}% http://ctan.org/pkg/lm
+\usepackage{subcaption} %for subfigures
+\usepackage{graphicx} %graphics support to specify path
+\usepackage{csvsimple} %insert tables using .csv files
+\usepackage{caption}
 
@@
 %%  submitting to BMC.                         %%
-%%  All figure files must be submitted as      %%
+%%  All Figure files must be submitted as      %%
 %%  separate graphics through the BMC          %%
@@
 
-\def\includegraphic{}
-\def\includegraphics{}
+% \def\includegraphic{}
+% \def\includegraphics{}
+
+\graphicspath {{figures/}}
 
@@
    addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
-   % corref={aff1},                       % id of corresponding address, if any
    email={dangeles@caltech.edu}   % email address
@@
 \author[
-   addressref={aff1},
-   corref= {aff1},
+   addressref={aff1, aff3},
+   corref={aff1},
    email={pws@caltech.edu}
@@
   \street{1200 E California Blvd},                     %
-  \postcode{91125}                                % post or zip code
+  \postcode{91125},                               % post or zip code
   \city{Pasadena},                              % city
   \cny{US}                                    % country
+}
+
+\address[id=aff3]{%                           % unique id
+  \orgname{HHMI}, % university, etc
 }
@@
   \street{1200 E California Blvd},                     %
-  \postcode{91125}                                % post or zip code
+  \postcode{91125},                                % post or zip code
   \city{Pasadena},                              % city
@@
 \begin{abstract} % abstract
+\parttitle{Background} %if any
 Over the last ten years, there has been an explosive development in tools
@@
 of gene targets, but understanding these datasets, and forming hypotheses
-based on them remains challenging.
-
-We present a method for detecting tissue enrichment in \emph{C.
-elegans} using the tissue ontology for this organism.
-We also present an efficient method for trimming the ontology that results in concise yet useful output.
-
-Our tool, Tissue Enrichment Analysis (TEA),
-can be found at www.wormbase.org/tea
+based on them remains challenging. One way to analyze these datasets is to
+associate ontologies, which are controlled, hierarchical, descriptive vocabularies
+with genes and to look for enrichment of specific terms. Although gene ontology
+(GO) is available for \emph{C. elegans}, this ontology does not include
+anatomy or physiology information.
+\parttitle{Results} %if any
+We have developed an enrichment analysis tool for the \emph{C. elegans} tissue
+ontology, is available on the web via WormBase and is available for download 
+using Python's standard pip installer. In order to cut down on verbosity, we
+have come up with three straightforward filtering criteria that slim the ontology
+by almost tenfold. 
+\parttitle{Conclusions} %if any
+Our Tissue Enrichment Analysis (TEA), which can be found at www.wormbase.org/tea, 
+uses a standard hypergeometric function to test a slimmed-down \emph{C. elegans} tissue 
+ontology and provides users with a text and graphic representation of the results. 
 \end{abstract}
@@
 \kwd{Tissue Ontology}
-\kwd{Wormbase}
+\kwd{WormBase}
+\kwd{RNA-seq}
+\kwd{High-throughput biology}
 \end{keyword}
@@
 These genes are often correlated in their biological characteristics or functions, but identifying these functions remains challenging.
-In order to interpret these long lists of genes, biologists need to abstract genes into fewer terms that are biologically relevant in order to form hypotheses about what is happening in the data.
+To interpret these long lists of genes, biologists need to abstract genes into fewer terms that are biologically relevant to form hypotheses about what is happening in the data.
 One such abstraction method relies on Gene Ontology (GO).
-GO provides a controlled set of hierarchically ordered terms in the form of an directed acyclic graph\cite{TheGeneOntologyConsortium2000, Ontology2009, TheGeneOntologyConsortium2015} that provide detailed information about the molecular, cellular or biochemical functions of the gene among others.
-For a given gene list, certain software programs can query whether a particular gene is enriched\cite{Mi2009, McLean2010, Huang2009}.
-However, GO is often difficult to interpret due to the large number of terms associated with a given gene.
-There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.
-A common tool for GO analysis, DAVID, clusters terms into broad categories that are amenable to exploration by researchers \cite{Huang2007}, whereas PANTHER, a different software package \cite{Mi2009, Mi2013}, attempts to solve this issue by employing a manually reduced ontology, GOslim (pers.
-comm.).
-
+GO provides a controlled set of hierarchically ordered terms in the form of an directed acyclic graph  \cite{TheGeneOntologyConsortium2000, Ontology2009, TheGeneOntologyConsortium2015} that provide detailed information about the molecular, cellular or biochemical functions of the gene among others.
+For a given gene list, certain software programs can query whether a particular gene is enriched  \cite{Mi2009, McLean2010, Huang2009}.
+One area of biological significance that GO does not include is physiology and anatomy.
+One way to address this shortcoming is to generate a `tissue ontology' that provides a complete anatomical description for an organism or sets of organisms, such as `tissue', `organ' or `neuronal cell', for example.
+Such a tissue ontology has been developed previously \cite{Lee2003}.
+The C.
+elegans  database, Wormbase \cite{Harris2014}, maintains a carefully curated list of gene expression data from GFP-reporters.
 	Here we provide a new framework that analyses user-input list for enrichment of specific tissues.
@@
 As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.
-Our analysis also cuts down on result verbosity by filtering the ontology before testing using a small set of well-defined criteria to remove terms that don’t contribute extra information.
-To our knowledge, such filtering has never been performed in an algorithmic fashion for an ontology before — indeed, tools such as DAVID do not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
+
+	Another problem frequently associated with GO analaysis is that it is often difficult to interpret due to the large number of terms associated with a given gene.
+There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.
+A common tool for GO analysis, DAVID, clusters terms into broad categories that are amenable to exploration by researchers  \cite{Huang2007}, whereas PANTHER, a different software package  \cite{Mi2009, Mi2013}, attempts to solve this issue by employing a manually reduced ontology, GOslim (pers.
+comm.
+H.
+Yu and P.
+Thomas).
+
+	To prevent our tool from suffering from the same drawbacks, we have cut down on result verbosity by filtering the ontology by using a small set of well-defined criteria to remove terms that do not contribute extra information.
+To our knowledge, such filtering has never been performed in an algorithmic fashion for an ontology before---indeed, tools such as DAVID do not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
 We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.
 
-	We built our software using a pre-established tissue ontology for the worm, C.
-elegans \cite{Lee2003}.
-The C.
-elegans  database, Wormbase\cite{Harris2014}, maintains a carefully curated list of gene expression data from GFP-reporters.
-We use this gold-standard list to develop a tissue enrichment analysis that reliably identifies even small tissues and show that we can reliably discriminate between embryonic and larval tissues.
-Our tool is available in Wormbase at the address http://mangolassi.caltech.edu/~azurebrd/cgi-bin/testing/amigo/getWithPost.cgi and provides users with a text-based file of the enrichment results as well as a simple and clear graph of the results that exhibit the largest fold-change enrichment.
-Although we present results here for the worm, we note that our software is species agnostic, and we are working to integrate tissue ontologies from other databases to provide a broader service to the community.
+	Our tool is available within WormBase and provides users with a text-based file of the enrichment results as well as a simple and clear graph of the results that exhibit the largest fold-change enrichment.
 
-\section*{Methods}
+\section*{Results}
 \subsection*{Generating a Useful Dictionary}
@@
 For example, several left and right neurons have at least 25 annotating genes and we may want to include them for enrichment testing.
-However, many left/right neuronal sisters have almost entirely the same annotations, with at most one or two gene differences between them.
+However, many left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them.
 We reasoned that when two tissues have almost identical annotations, we cannot have statistical confidence in differentiating between them.
 As a result, testing these sister tissues provides no additional information compared with testing only the parent node to these sisters.
-We refer to such sisters as ‘redundant’.
-In order to identify redundancy, we defined a similarity metric
-
-\begin{eqnarray}\label{similarity def}
-	s_i = \frac{|g_i|}{|\bigcup_{i= 0}^k g_i|}
-\end{eqnarray}
-
-Where $s_i$ is the similarity for a tissue $i$ eith $k$ sisters; $g_i$ refers to the set of tissues associated with tissue $i$ and $|g|$ refers to the cardinality of set $g$.
-For a given set of sisters, we called them redundant if they exceeded a given similarity threshold.
-We envisioned two possible criteria and built different dictionaries using each one.
-Under a threshold criteron `any' with parameter $S$ between $(0, 1)$, a given set of sisters $j$ was considered redundant if the condition
-
-\begin{eqnarray}\label{any threshold}
-	s_{i, j} > S
-\end{eqnarray}
-
-was true for any sister $i$ in set $j$. Under a threshold criterion `avg' with parameter $S$, a given set of sisters $j$ was considered redundant if the condition
-
-\begin{eqnarray}\label{avg threshold}
-	\mathrm{E}[s_i]_j > S
-\end{eqnarray}
-
-was true for the set of sisters $j$ (see figure~\ref{fig:simdiagram}).
+We refer to such sisters as `redundant'.
+To identify redundancy, we defined a similarity metric (see \emph{Methods} section and Figure~\ref{fig:simdiagram}).
+Our similarity metric can be used to identify sisters that have very high similarity between them; alternatively, redundant sisters could be identified if a single sister had a very high similarity score.
+We referred to these two scoring criteria as `avg' and `any' respectively.
 
@@
 At that point, no more nodes can be removed from that branch.
-This is guaranteed by the structure of the ontology: Parent nodes inherit all of the annotations of all of their descendants, so the number of annotating terms monotonically increases with increasing term hierarchy (see figure~\ref{fig:trim_ends}).
+This is guaranteed by the structure of the ontology: Parent nodes inherit all of the annotations of all of their descendants, so the number of annotating terms monotonically increases with increasing term hierarchy (see Figure~\ref{fig:trim_ends}).
 In this way, we ensure that our term dictionary includes only those tissues that are considered sufficiently well annotated for statistical purposes.
@@
 In other words, if all the daughter nodes are tested, there is little additional information to be gained by including the parent node.
-To address this issue we removed parent nodes from the analysis if all their daughter nodes passed the annotation threshold (see figure~\ref{fig:trim_roots}).
+To address this issue we removed parent nodes from the analysis if all their daughter nodes passed the annotation threshold (see Figure~\ref{fig:trim_roots}).
 
@@
 	By itself, each of these filters can reduce the number of nodes employed for analysis.
-Notably, these filters are not all commutative -- while trimming and redundancy filtering are commutative, applying the ceiling filter is not commutative with either the trimming or the redundancy filter.
+Notably, these filters are not all commutative: while trimming and redundancy filtering are commutative, applying the ceiling filter is not commutative with either the trimming or the redundancy filter.
 If the ceiling filter is  applied before any other filter, only terminal nodes will remain, since all the parents have complete daughter sets.
@@
 The original ontology has 1675 terms with more than 5 gene annotations.
-After filtering, dictionary sizes ranged from 21 to a maximum of 400 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by tenfold by application of a few simple filters.
-
+After filtering, dictionary sizes ranged from 21 to a maximum of 400 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by tenfold by application of a few simple filters (see Table ~\ref{tab:DictionarySpecs}).
 These filters were used to compile a static dictionary that we employ for all analyses.
-Because we have integrated our scripts to draw on the WormBase databases, our dictionary will remain up to date as tissue expression data improves.
-Our completed static trimmed dictionary is available for download at the following ftp URL: XX.
-The final dictionary includes XX tissues for testing, and has XX annotating genes.
-All code was implemented in Python.
 
@@
 Mathematically, this is expressed as:
-
 \begin{eqnarray}\label{hypergeometric}
-	\mathrm{P}(n_i | N, m_1,\ldots, m_k, M) = \frac{\dbinom{m_i}{n_i} \dbinom{M-m_i}{N - n_i}}{\dbinom{N}{n_i}}
+	\mathrm{P}(n_i | N, m_i, M) = \frac{\dbinom{m_i}{n_i} \dbinom{M-m_i}{N - n_i}}{\dbinom{N}{n_i}}
 \end{eqnarray}
-
-	Here, $n_i$ is the number of balls of type $i$ drawn, $N$ is the total number of draws, $m_i$ is tissue $i$ and $M = \sum_i m_i$ is the total number of balls in the urn.
+	Here, $n_i$ is the number of balls of type $i$ drawn, $N$ is the total number of draws, $m_i$ is tissue $i$ and $M$ is the total number of balls in the urn.
 In our specific case, $M_i$ is equal to the total number of annotations in our dictionary.
 $N$ is found by taking the user-input list and removing any genes that are not in our annotation dictionary.
-The remaining genes are then associated with their annotation profiles — if a tissue is associated with $s$ tissues, it generates $s$ balls of $s$ colors.
+The remaining genes are then associated with their annotation profiles---if a tissue is associated with $s$ tissues, it generates $s$ balls of $s$ colors.
 Our program counts the number of times each tissue appears in the user list, and calculates the probability of having withdrawn as many or more balls for each tissue in the user list.
@@
 
-	Once the probability of drawing the labels has been quantified, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm\cite{Benjamini1995}.
+	Once the probability of drawing the labels has been quantified, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm \cite{Benjamini1995}.
 Genes that have a q-value less than a given alpha are considered significant.
@@
 The program returns a text-based table showing the tissues that tested significant, along with their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change (observed hits / expected hits).
-Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the largest enrichment fold change.
-Our software relies heavily on the Pandas, Numpy, Seaborn and SciPy modules to perform all statistical testing and data handling\cite{McKinney2011, VanDerWalt2011, Oliphant2007}.
-
+Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values.
 	Our software is implemented in an easy to use GUI within WormBase.
-Users input a gene-list (see figure~\ref{fig:GUI}) using any valid gene name for \emph{C.
+Users input a gene-list using any valid gene name for \emph{C.
 elegans}.
-These names are processed into standard WBIDs and the result is displayed in the same window in an easy to read format containing all the relevant information, and a graph of the results is also displayed (see figure~\ref{fig:GUIresults}).
+These names are processed into standard WBIDs and the result is displayed in the same window in an easy to read format containing all the relevant information, and a graph of the results is also displayed (see Figure ~\ref{fig:GUIresults}).
 
 \subsection*{Validation of the algorithm and parameter selection}
-	In order to select an appropriate dictionary and validate our tool, we found a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}.
-Some of these studies went on to use GFP to identify expression patterns and for this reason we generated a clean Since the expression data is curated from GFP expression at this time and does not include RNA-seq data, these gold standards are statistically independent from the dataset.
 	We wanted to select a dictionary which included enough terms to be specific beyond the largest C.
@@
 On the other hand, selecting a dictionary that can detect smaller tissues will bias us towards tissues with lesser annotations.
-To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.
->>>>>>> master
+To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.To help us select an appropriate dictionary and validate our tool, we found a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues \cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}.
+Some of these studies have been used to annotate gene expression, and so they did not constitute an independent testing set.
+To correct this flaw, we built a clean dictionary that specifically excluded all annotation evidence that came from these studies.
 
-	As a first attempt to select a good dictionary, we generated all the possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using ‘average’ or ‘any’ thresholding criteria for the latter (see table ~\ref{tab:DictionarySpecs}).
-For these dictionaries, the number of tissues tested ranged from 97 to 676.
+	As a first attempt to select a good dictionary, we generated all the possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any' thresholding criteria for the latter (see Table ~\ref{tab:DictionarySpecs}).
+For these dictionaries, the number of tissues tested ranged from 21 to 460.
 The number of tissues was inversely correlated to the minimum annotation, as expected, and was largely insensitive to the redundancy threshold, at least in the range we explored (0.9-1).
@@
 When we analyzed the distribution of significant q-values for the dictionaries, we found that the similarity threshold mattered relatively little for any dictionary.
-We also noticed that the `any' thresholding method resulted in tighter histograms with a mode closer to 0 (data not shown).
-For this reason, we chose the ‘any’ method for dictionary generation.
-The average q-value increased with decreasing annotation cut-off (see figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant the trade-off between power and term specificity is.
-Based on these observations, we ruled out the dictionary with the 100 annotation cut-off - it had the fewest terms and its q-values were not low enough to compensate the trade-off in specificity.
+We also noticed that the `any' thresholding method resulted in tighter histograms with a mode closer to 0.
+For this reason, we chose the `any' method for dictionary generation.
+The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant the trade-off between power and term specificity is.
+Based on these observations, we ruled out the dictionary with the 100 gene annotation cut-off: it had the fewest terms and its q-values were not low enough to compensate the trade-off in specificity.
 
@@
 For some sets, all dictionaries seemed to perform well.
-For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} the results were an amalgamation of neuron related terms including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays regardless of the dictionary used.
-On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells -- at the 33-cutoff, our tool identified `Z2' and `Z3' as being five-fold enriched; whereas at the 25 gene-cutoff the terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
+For example, in our `all neuron enriched sets'~ \cite{Spencer2011, Watson2008a} the results were an amalgamation of neuron related terms including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays regardless of the dictionary used (see~\ref{fig:interagree}).
+On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~ \cite{Spencer2011}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells ---at the 33-cutoff, our tool identified `Z2' and `Z3' as being five-fold enriched; whereas at the 25 gene-cutoff the terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
 %Differences between dictionaries
-	We queried an embryonic stage intestine precursor geneset~\cite{Spencer2011}.
+We queried an embryonic stage intestine precursor geneset~ \cite{Spencer2011}.
 Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary.
@@
 Not all queries worked equally well.
-For example, a number of intestinal enriched genes sets~\cite{Spencer2011, Pauli2006} were not enriched in intestine in any dictionary, but they were enriched for pharynx- and hypodermis-related terms.
+For example, a number of intestinal enriched genes sets~ \cite{Spencer2011, Pauli2006} were not enriched in intestine in any dictionary, but they were enriched for pharynx- and hypodermis-related terms.
 We were somewhat surprised that intestinal gene sets performed poorly, since the intestine is a relatively well-annotated tissue.
@@
 We also assessed the internal agreement of our tool by using independent gene-sets that we expected to be enriched in the same tissues.
-We had two independent pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two independent PVD enriched sets~\cite{Spencer2011, Smith2010}; two independent GABAergic gene sets~\cite{Spencer2011, Cinar2005}; two independent pharyngeal gene sets~\cite{Spencer2011, Gaudet2004a}; and two independent intestinal gene sets~\cite{Spencer2011, Pauli2006}.
+We had two independent pan-neuronal sets~ \cite{Spencer2011, Watson2008a}; two independent PVD enriched sets~ \cite{Spencer2011, Smith2010}; two independent GABAergic gene sets~ \cite{Spencer2011, Cinar2005}; two independent pharyngeal gene sets~ \cite{Spencer2011, Gaudet2004a}; and two independent intestinal gene sets~ \cite{Spencer2011, Pauli2006}.
 Overall, the tool seems to have good internal agreement.
-On most sets, the same terms were enriched, although order was somewhat variable.
+On most sets, the same terms were enriched, although order was somewhat variable see ~\ref{fig:intragree}.
 However, most high-scoring terms were preserved between gene sets.
@@
 
-\section*{Results}
+\subsection*{A brief example}
 We applied our tool to the RNA-seq datasets developed by Engelmann et al.
-\cite{Engelmann2010} in order to attempt to gain further understanding of the biology underlying these datasets.
+\cite{Engelmann2011} to gain further understanding of the biology underlying these datasets.
 Engelmann et al.
 exposed young adult worms to 5 different pathogenic bacteria or fungi for 24 hours, after which mRNA was extracted from the worms for sequencing.
-We obtained the genes that Engelmann et al identified as up- or down- regulated in their assay, and ran TEA using these lists.
+We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated.
 Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade.
-Strikingly, 4 out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues (in the case of Harposporium sp) amongst the down-regulated genes.
+Strikingly, three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes (see \ref{fig:Dcon, fig:Hsp, fig:Smar}).
 A possible explanation for this might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioural response against the pathogen.
-Indeed, several studies\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.
-elegans} uses chemosensory neurons to identify pathogens Interestingly, one bacterium did not exhibit the same pattern of down-regulation of neuronal-associated genes.
+Indeed, several studies \cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.
+elegans} uses chemosensory neurons to identify pathogens.
+Interestingly, one bacterium did not exhibit the same pattern of down-regulation of neuronal-associated genes.
 \emph{E.
@@
 faecalis} may have a different pathogenic profile.
-Up-regulated tissues, when detected, included the hypodermis and excretory duct.
+Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct.
 	Our results highlight the involvement of various \emph{C. elegans} neuronal tissues in pathogen defense and/or illness.
@@
 \section*{Discussion}
-
 We have presented a tissue enrichment analysis tool that employs a standard hypergeometric model to test the \emph{C.
@@
 Due to the nature of all ontologies as hierarchical, acyclical graphs with term inheritance, term annotations are correlated along any given branch.
-This correlation reduces the benefits of including all terms for statistical analysis - for any given term along a branch, if that term passes significance, there is a high probability that many other terms along that branch will also pass significant.
+This correlation reduces the benefits of including all terms for statistical analysis: for any given term along a branch, if that term passes significance, there is a high probability that many other terms along that branch will also pass significant.
 If the branch is enriched by random chance, error propagation along a branch means that many more false positives will follow.
@@
 Instead, a user will benefit most from testing sparsely along the tree at a suitable specificity for hypothesis formation.
-Related terms of the same level should only be tested when there is sufficient annotation to differentiate, with statistical confidence, whether one term is enriched above the other (see SI for a back-of-the-envelope calculation of when this can be the case).
+Related terms of the same level should only be tested when there is sufficient annotation to differentiate, with statistical confidence, whether one term is enriched above the other.
 Our algorithm reduces branch length by identifying and removing nodes that are insufficiently annotated and parents that are likely to include sparse information.
 
-It is important to note that our tool is not the first tissue enrichment model for the worm that has been reported.
 Chikina \emph{et al}  \cite{Chikina2009} report a tissue enrichment model based on an SVM classifier that has been trained on microarray studies.
-SVM classifiers are powerful tools capable of great sensitivity, but they require continuous retraining as tissue expression data widens.
-Our tool benefits from the fact that it will be integrated in WormBase and will therefore be updated continuously as new data is integrated.
+SVM classifiers are powerful tools capable of great sensitivity, but they require continuous retraining as tissue expression data becomes more available.
+Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for some genome-wide studies.
+Our tool relies on an annotation dictionary that is continuously updated, does not require retraining and does not require ranked genes.
+To our knowledge, there are no other tissue ontology enrichment tools in \emph{C.
+elegans}, but similar projects exist for humans and zebrafish ~\cite{Lee2013, Prykhozhij2013}, highlighting the relevance of our tool for high-dimensionality biology.
 
@@
 Is this an error, or does this hint at new biology?
-Although we were unable to determine false-positive and false-negative rates, we don't believe this should deter scientists from using our tool.
+Although we were unable to determine false-positive and false-negative rates, we  do not believe this should deter scientists from using our tool.
 Rather, we encourage researchers to use our tool carefully as a guide, integrating evidence from multiple sources to inform the most likely hypotheses.
@@
 Support from the community will be crucial in correcting these flaws going forward; indeed, without the community reports of tissue expression this tool would not be possible.
+
+\section*{Methods}
+\subsection*{Filtering nodes}
+\subsubsection*{Defining a Similarity Metric}
+In order to identify redundant sisters, we defined the following similarity metric:
+\begin{eqnarray}\label{similarity def}
+	s_i = \frac{|g_i|}{|\bigcup_{i= 0}^k g_i|}
+\end{eqnarray}
+Where $s_i$ is the similarity for a tissue $i$ with $k$ sisters; $g_i$ refers to the set of tissues associated with tissue $i$ and $|g|$ refers to the cardinality of set $g$.
+For a given set of sisters, we called them redundant if they exceeded a given similarity threshold.
+We envisioned two possible criteria and built different dictionaries using each one.
+Under a threshold criteron `any' with parameter $S$ between $(0, 1)$, a given set of sisters $j$ was considered redundant if the condition
+\begin{eqnarray}\label{any threshold}
+	s_{i, j} > S
+\end{eqnarray}
+was true for any sister $i$ in set $j$. Under a threshold criterion `avg' with parameter $S$, a given set of sisters $j$ was considered redundant if the condition
+\begin{eqnarray}\label{avg threshold}
+	\mathrm{E}[s_i]_j > S
+\end{eqnarray}
+was true for the set of sisters $j$ (see Figure~\ref{fig:simdiagram}).
+
+\subsubsection*{Implementation}
+All scripts were written in Python.
+Our software relies on the Pandas, Numpy, Seaborn and SciPy modules to perform all statistical testing and data handling \cite{McKinney2011, VanDerWalt2011, Oliphant2007}.
 
@@
 
-http://dangeles.github.io/TissueEnrichmentAnalysis.
+http://dangeles.github.io/TissueEnrichmentAnalysis
 
-All benchmark gene sets, benchmarking code and figures can also be found at the same address, under the `tests' folder. 
+All benchmark gene sets, benchmarking code and Figures can also be found at the same address, under the `tests' folder.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % b) uncomment next line
-%\nocite{label}
+\nocite{label}
 
@@
 % Figures describing algorithm
+
 \begin{figure}[h!]
-  \caption{\csentence{Schematic diagram of annotations for two sisters.}
-      The parent node (green) contains at least as many annotations as the union of the two sisters.
-	  These two sisters share annotations extensively. Therefore they are too similar and should be removed.
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{figures/RedundancyTrimmingOntogeny.png}
+   	 	\caption{}
+    \label{fig:simdiagram}
+  \end{subfigure}
+  %
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{figures/FloorTrimmingOntogeny.png}
+	      \caption{}
+    \label{fig:trim_ends}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \includegraphics[width=\textwidth]{figures/CeilingTrimmingOntogeny.png}
+    	\caption{}
+    \label{fig:trim_roots}
+  \end{subfigure}
+
+  \captionsetup{width= 0.95\textwidth}
+  \caption{
+  	\csentence{Schematic representation of trimming filters for an acyclical ontology.}
+    \csentence{\textbf{a)}} The parent node (green) contains at least as many annotations as the union of the two sisters.
+These two sisters share annotations extensively.
+Therefore they are too similar and should be removed.
+	\csentence{\textbf{b)}} Nodes with less than a threshold number of genes are trimmed (light red) and discarded from the dictionary. Here, the example threshold is 25 genes.
+	\csentence{\textbf{c)}} We trim parent nodes (light red) if all their daughter nodes have more than the threshold number of annotations.
   }
-	  \label{fig:simdiagram}
 \end{figure}
 
-\begin{figure}[h!]
-  \caption{\csentence{Schematic showing terminal node removal.}
-      Nodes with less than a threshold number of genes are trimmed (light red) and discarded from the dictionary. Here, the threshold is 25 genes.}
-	  \label{fig:trim_ends}
-      \end{figure}
-
-\begin{figure}[h!]
-  \caption{\csentence{Schematic showing root node removal.}
-      We trim parent nodes (light red) if all their daughter nodes have more than the threshold number of annotations. Here, the threshold is 25 genes.}
-	  \label{fig:trim_roots}
-      \end{figure}
+%   \begin{figure}[h!]
+%   \caption{\csentence{Schematic diagram of annotations for two sisters.}
+%       The parent node (green) contains at least as many annotations as the union of the two sisters.
+% 	  These two sisters share annotations extensively. Therefore they are too similar and should be removed.
+% 	  }
+% 	  % \label{fig:simdiagram}
+%   \end{figure}
+%
+% \begin{figure}[h!]
+%   \caption{\csentence{Schematic showing terminal node removal.}
+%       Nodes with less than a threshold number of genes are trimmed (light red) and discarded from the dictionary. Here, the threshold is 25 genes.}
+% 	  \label{fig:trim_ends}
+%       \end{figure}
+%
+% \begin{figure}[h!]
+%   \caption{\csentence{Schematic showing root node removal.}
+%       We trim parent nodes (light red) if all their daughter nodes have more than the threshold number of annotations. Here, the threshold is 25 genes.}
+% 	  \label{fig:trim_roots}
+%       \end{figure}
 
 %gui
-\begin{figure}[h!]
-  \caption{\csentence{Screenshot of the web GUI.}
-      }
-	  \label{fig:GUI}
 
-      \end{figure}
-\begin{figure}[h!]
-  \caption{\csentence{Screenshot of results from web GUI.}
+\begin{figure}
+  % \begin{subfigure}[b]{0.4\textwidth}
+  %   \includegraphics[width=\textwidth]{figures/gui.png}
+  %   \caption{}
+  %   \label{fig:GUI}
+  % \end{subfigure}
+  % %
+  % \begin{subfigure}[b]{0.4\textwidth}
+    \includegraphics[width=\textwidth]{figures/guiresults.png}
+    % \caption{}
+  % \end{subfigure}
+  \captionsetup{width= 0.95\textwidth}
+  \caption{\csentence{Screenshot of results from the web GUI.}
+  After inputting a gene-list, the user is provided with the results. 
+  An HTML table is output with hyperlinks to the ontology terms. A 
+  publication-ready graph is provided below, which can be saved by 
+  dragging to the desktop. Finally, lists of the genes used and discarded
+  for the analysis are also presented. 
       }
@@
 
-% statistical testing and validation figures
+% \begin{figure}[h!]
+%   \caption{\csentence{Screenshot of the web GUI.}
+%       }
+% 	  \label{fig:GUI}
+%
+%       \end{figure}
+% \begin{figure}[h!]
+%   \caption{\csentence{Screenshot of results from web GUI.}
+%       }
+% 	  \label{fig:GUIresults}
+      % \end{figure}
+
+% statistical testing and validation Figures
 %q plot
 \begin{figure}[h!]
+	\includegraphics[width=\textwidth]{avgQKDE_method=any.png}
+  \captionsetup{width= 0.95\textwidth}
   \caption{\csentence{Kernel density estimates for 30 gold standard datasets.}
@@
 
-% %ovd-oll
+%pan neuronal
+\begin{figure}
+  \begin{subfigure}[b]{0.6\textwidth}
+    \includegraphics[width=\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_33cutoff.png}
+    \caption{}
+    \label{fig:PanNeuronalWatson33}
+  \end{subfigure}
+  %
+  \begin{subfigure}[b]{0.6\textwidth}
+    \includegraphics[width=\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_50cutoff}
+    \caption{}
+    \label{fig:PanNeuronalWatson50}
+  \end{subfigure}
+  \captionsetup{width= 0.95\textwidth}
+  \caption{\csentence{Comparison of Enrichment Results for two dictionaries for a pan-neuronal enriched gene set from ~ \cite{Watson2008a}.}
+   \textbf{a)} Dictionary with cutoff: 33; threshold: 0.95; method: `any'
+   \textbf{b)} Dictionary with cutoff: 50; threshold: 0.95; method: `any'
+   }
+   \label{fig:interagree}
+\end{figure}
+
+%agreements
+\begin{figure}
+  \begin{subfigure}[b]{0.6\textwidth}
+    \includegraphics[width=\textwidth]{WBPaper00024970_GABAergic_neuron_specific_WBbt_0005190_247_33cutoff.png}
+    \caption{}
+    \label{fig:agreement1}
+  \end{subfigure}
+  %
+  \begin{subfigure}[b]{0.6\textwidth} 	\includegraphics[width=\textwidth]{WBPaper00037950_GABAergic-motor-neurons_larva_enriched_WBbt_0005190_132_33cutoff.png}
+    \caption{}
+    \label{fig:agreement2}
+  \end{subfigure}
+  \captionsetup{width= 0.95\textwidth}
+  \caption{\csentence{Independently derived genesets show similar results within a given dictionary}
+  \textbf{a)} Gene set from  \cite{Watson2008a}. Dictionary with cutoff: 33; threshold: 0.95; method: `any'
+  \textbf{b)} Gene set from  \cite{Spencer2011}. Dictionary with cutoff: 50; threshold: 0.95; method: `any'
+   }
+   \label{fig:intragree}
+\end{figure}
+
+%results Figures - engelmann et al reanalysis
+%labels are missing.
 \begin{figure}[h!]
-  \caption{\csentence{Comparison of Enrichment Results for dictionary size 50 (left) and 25 (right) for a PVD-OLL enriched gene set.}
-      Left, at 50 annotation cut-off, TEA singles PVD as highly enriched.
-Other mechanosensory neurons are also enriched .
-Right, when the dictionary cut-off is set to 25, TEA shows embryonic tissues that are unrelated to the PVD and OLL lineages.}
-	  \label{fig:ovdoll}
+    \includegraphics[width=\textwidth]{figures/DrechmeriaconiosporaEnrichment.png}
+	\captionsetup{width= 0.95\textwidth}
+  	\caption{\csentence{\emph{D. coniospora} enrichment results}
+      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{D. coniospora}, a fungus, show enrichment
+	  in many neuronal tissues.}
+	\label{fig:Dcon}
 \end{figure}
-%
-%
-%
-%
-%results figures - engelmann et al reanalysis
+
 \begin{figure}[h!]
-  \caption{\csentence{\emph{D. coniospora} enrichment results}
-      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{D. coniospora} (fungus).}
+    \includegraphics[width=\textwidth]{figures/HarposporiumspEnrichment.png}
+	\captionsetup{width= 0.95\textwidth}
+	\caption{\csentence{\emph{Harposporium sp.} enrichment results}
+      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Harposporium sp.} a fungus, shows enrichment
+	  in the posterior lateral ganglion, similarly to \emph{D. coniospora}. This particular fungus seems to affect
+	  genes normally associated with neuronal precursor tissues as well. 
+	  }
+	\label{fig:Hsp}
 \end{figure}
+
 \begin{figure}[h!]
-  \caption{\csentence{\emph{Harposporium sp.} enrichment results}
-      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Harposporium sp.} (fungus).}
+    \includegraphics[width=\textwidth]{figures/SerratiamarcescensEnrichment.png}
+	\captionsetup{width= 0.95\textwidth}
+  	\caption{ \csentence{\emph{Serratia marcescens} enrichment results}
+      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Serratia marcescens}, a bacterium, shows 
+	  down-regulation of chemosensory neurons notably ASER, ASEL, ASJ and ASK. The lateral ganglions are also 
+	  affected. Up-regulated genes are overrepresented for terms involving the excretory duct cell and hypodermis.
+	}  
+	\label{fig:Smar}
 \end{figure}
+
 \begin{figure}[h!]
-  \caption{\csentence{\emph{Serratia marcescens} enrichment results}
-      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Serratia marcescens} (bacteria).}
+    \includegraphics[width=\textwidth]{figures/EnterococcusfaecalisEnrichment.png}
+	\captionsetup{width= 0.95\textwidth}
+  	\caption{\csentence{\emph{E. faecalis} enrichment results}
+      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{E. faecalis}, a bacterium, shows an unusual pattern:
+	  No terms were identified as enriched in the down-regulated category. Surprisingly, up-regulated genes show enrichment
+	  of various head neurons that were down-regulated with exposure to other pathogens. Interestingly, some pharyngeal
+	  muscle terms are also enriched. 
+	  }
+	\label{fig:Efaec}
 \end{figure}
+
 \begin{figure}[h!]
-  \caption{\csentence{\emph{E. faecalis} enrichment results}
-      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{E. faecalis} (bacteria)}
+    \includegraphics[width=\textwidth]{figures/OtorhabdusluminescensEnrichment.png}
+	\captionsetup{width= 0.95\textwidth}
+  	\caption{\csentence{\emph{O. luminescens} enrichment results}
+      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{O. luminescens}, a bacterium, shows no
+	  significantly enriched terms in down-regulated genes. However, up-regulated genes suggest that the hypodermis
+	  and excretory duct cell may be affected by infection. 
+	  }
+	\label{fig:Olum}
 \end{figure}
@@
 \caption{Parameter specifications and number of tissues for all dictionaries.}
-      \begin{tabular}{cccc}
-        \hline
-      \end{tabular}
+	\csvautotabular{figures/TissueNumbers.csv}
 	\label{tab:DictionarySpecs}
@@
 \section*{Additional Files}
-  \subsection*{Additional file 1 --- Supplementary Information}
-	Complete results from benchmarking analysis
-  \subsection*{Additional file 2 --- Supplementary Information}
-    Complete results from re-analysis of Engelmann et al
-  \subsection*{Additional file 3 --- IPython Notebook}
-    Tutorial for users interested in batch script generation using our software.
+  \subsection*{Additional file 1 --- IPython Notebook}
+    Tutorial for users interested in using our software within a python script
 

### 2016-03-15 tests/hgf_benchmarking/doc/tea_main_text.tex
more writing

@@
 %\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-% \documentclass[linenumbers, doublespacing]{bmcart}
-\documentclass{bmcart}
+\documentclass[linenumbers, doublespacing]{bmcart}
+% \documentclass{bmcart}
 
@@
 Overall, the tool seems to have good internal agreement.
-On most sets, the same terms were enriched, although order was somewhat variable see ~\ref{fig:intragree}.
+On most sets, the same terms were enriched, although order was somewhat variable (see Figure ~\ref{fig:intragree}).
 However, most high-scoring terms were preserved between gene sets.
@@
 	\csentence{\textbf{b)}} Nodes with less than a threshold number of genes are trimmed (light red) and discarded from the dictionary. Here, the example threshold is 25 genes.
-	\csentence{\textbf{c)}} We trim parent nodes (light red) if all their daughter nodes have more than the threshold number of annotations.
+	\csentence{\textbf{c)}} We trim parent nodes recursively, starting from the root, if all their daughter nodes have more than the threshold number of annotations (light red).
   }
 \end{figure}
-
-%   \begin{figure}[h!]
-%   \caption{\csentence{Schematic diagram of annotations for two sisters.}
-%       The parent node (green) contains at least as many annotations as the union of the two sisters.
-% 	  These two sisters share annotations extensively. Therefore they are too similar and should be removed.
-% 	  }
-% 	  % \label{fig:simdiagram}
-%   \end{figure}
-%
-% \begin{figure}[h!]
-%   \caption{\csentence{Schematic showing terminal node removal.}
-%       Nodes with less than a threshold number of genes are trimmed (light red) and discarded from the dictionary. Here, the threshold is 25 genes.}
-% 	  \label{fig:trim_ends}
-%       \end{figure}
-%
-% \begin{figure}[h!]
-%   \caption{\csentence{Schematic showing root node removal.}
-%       We trim parent nodes (light red) if all their daughter nodes have more than the threshold number of annotations. Here, the threshold is 25 genes.}
-% 	  \label{fig:trim_roots}
-%       \end{figure}
 
@@
 \begin{figure}
-  % \begin{subfigure}[b]{0.4\textwidth}
-  %   \includegraphics[width=\textwidth]{figures/gui.png}
-  %   \caption{}
-  %   \label{fig:GUI}
-  % \end{subfigure}
-  % %
-  % \begin{subfigure}[b]{0.4\textwidth}
-    \includegraphics[width=\textwidth]{figures/guiresults.png}
-    % \caption{}
-  % \end{subfigure}
+	\centering
+    \includegraphics[width=0.95\textwidth]{figures/guiresults.png}
   	\captionsetup{width= 0.95\textwidth}
@@
 \begin{figure}
-  \begin{subfigure}[b]{0.6\textwidth}
-    \includegraphics[width=\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_33cutoff.png}
+  \begin{subfigure}[b]{0.95\textwidth}
+    \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_33cutoff.png}
     \caption{}
@@
   %
-  \begin{subfigure}[b]{0.6\textwidth}
-    \includegraphics[width=\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_50cutoff}
+  \begin{subfigure}[b]{0.95\textwidth}
+    \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_50cutoff}
     \caption{}
@@
   \captionsetup{width= 0.95\textwidth}
-  \caption{\csentence{Comparison of Enrichment Results for two dictionaries for a pan-neuronal enriched gene set from ~ \cite{Watson2008a}.}
+  \caption{\csentence{Comparison of Enrichment Results for two dictionaries for a pan-neuronal enriched gene set from ~\cite{Watson2008a}.}
    \textbf{a)} Dictionary with cutoff: 33; threshold: 0.95; method: `any'
@@
 \begin{figure}
-  \begin{subfigure}[b]{0.6\textwidth}
-    \includegraphics[width=\textwidth]{WBPaper00024970_GABAergic_neuron_specific_WBbt_0005190_247_33cutoff.png}
+  \begin{subfigure}[b]{0.95\textwidth}
+    \includegraphics[width=0.95\textwidth]{WBPaper00024970_GABAergic_neuron_specific_WBbt_0005190_247_33cutoff.png}
     \caption{}
@@
   %
-  \begin{subfigure}[b]{0.6\textwidth} 	\includegraphics[width=\textwidth]{WBPaper00037950_GABAergic-motor-neurons_larva_enriched_WBbt_0005190_132_33cutoff.png}
+  \begin{subfigure}[b]{0.95\textwidth} 	\includegraphics[width=0.95\textwidth]{WBPaper00037950_GABAergic-motor-neurons_larva_enriched_WBbt_0005190_132_33cutoff.png}
     \caption{}
@@
   \caption{\csentence{Independently derived genesets show similar results within a given dictionary}
-  \textbf{a)} Gene set from  \cite{Watson2008a}. Dictionary with cutoff: 33; threshold: 0.95; method: `any'
-  \textbf{b)} Gene set from  \cite{Spencer2011}. Dictionary with cutoff: 50; threshold: 0.95; method: `any'
+  \textbf{a)} Gene set from~\cite{Watson2008a}. Dictionary with cutoff: 33; threshold: 0.95; method: `any'
+  \textbf{b)} Gene set from~\cite{Spencer2011}. Dictionary with cutoff: 50; threshold: 0.95; method: `any'
    }
@@
 \begin{figure}[h!]
-    \includegraphics[width=\textwidth]{figures/DrechmeriaconiosporaEnrichment.png}
+    \includegraphics[width=0.95\textwidth]{figures/DrechmeriaconiosporaEnrichment.png}
 	\captionsetup{width= 0.95\textwidth}
@@
 \begin{figure}[h!]
-    \includegraphics[width=\textwidth]{figures/HarposporiumspEnrichment.png}
+    \includegraphics[width=0.95\textwidth]{figures/HarposporiumspEnrichment.png}
 	\captionsetup{width= 0.95\textwidth}
@@
 \begin{figure}[h!]
-    \includegraphics[width=\textwidth]{figures/SerratiamarcescensEnrichment.png}
+    \includegraphics[width=0.95\textwidth]{figures/SerratiamarcescensEnrichment.png}
 	\captionsetup{width= 0.95\textwidth}
@@
 \begin{figure}[h!]
-    \includegraphics[width=\textwidth]{figures/EnterococcusfaecalisEnrichment.png}
+    \includegraphics[width=0.95\textwidth]{figures/EnterococcusfaecalisEnrichment.png}
 	\captionsetup{width= 0.95\textwidth}
@@
 \begin{figure}[h!]
-    \includegraphics[width=\textwidth]{figures/OtorhabdusluminescensEnrichment.png}
+    \includegraphics[width=0.95\textwidth]{figures/OtorhabdusluminescensEnrichment.png}
 	\captionsetup{width= 0.95\textwidth}

### 2016-03-21 tests/hgf_benchmarking/doc/tea_main_text.tex
Added fetch function and writing
fetch_dictionary will now fetch the tissue dictionary.

Writing.

@@
    addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
-   email={dangeles@caltech.edu}   % email address
+   % email={dangeles@caltech.edu}   % email address
 ]{\inits{DA}\fnm{David} \snm{Angeles-Albores}}
 \author[
-   addressref={aff2},                   % id's of addresses, e.g. {aff1,aff2}
-   email={raymond@caltech.edu}   % email address
+   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
+   % email={raymond@caltech.edu}   % email address
 ]{\inits{RYL}\fnm{Raymond Y} \snm{Lee}}
 \author[
-   addressref={aff2},                   % id's of addresses, e.g. {aff1,aff2}
-   email={azurebrd@caltech.edu}   % email address
+   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
+   % email={azurebrd@caltech.edu}   % email address
 ]{\inits{JC}\fnm{Juancarlos} \snm{Chan}}
 \author[
-   addressref={aff1, aff3},
+   addressref={aff1},
    corref={aff1},
@@
 \address[id=aff1]{%                           % unique id
-  \orgname{California Institute of Technology, Division of Biology and Biological Engineering}, % university, etc
-  \street{1200 E California Blvd},                     %
-  \postcode{91125},                               % post or zip code
-  \city{Pasadena},                              % city
-  \cny{US}                                    % country
-}
-
-\address[id=aff3]{%                           % unique id
-  \orgname{HHMI}, % university, etc
-}
-
-% raymond's address?
-\address[id=aff2]{%
-  \orgname{California Institute of Technology},
+  \orgname{HHMI and California Institute of Technology, Division of Biology and Biological Engineering}, % university, etc
   \street{1200 E California Blvd},                     %
@@
 \parttitle{Conclusions} %if any
-Our Tissue Enrichment Analysis (TEA), which can be found at www.wormbase.org/tea, 
+Our Tissue Enrichment Analysis (TEA), which can be found at www.WormBase.org/tea, 
 uses a standard hypergeometric function to test a slimmed-down   \emph{C. elegans} tissue 
@@
 For a given gene list, certain software programs can query whether a particular gene is enriched  \cite{Mi2009, McLean2010, Huang2009}.
-One area of biological significance that GO does not include is physiology and anatomy.
+One area of biological significance that GO does not include is anatomy.
 One way to address this shortcoming is to generate a `tissue ontology' that provides a complete anatomical description for an organism or sets of organisms, such as `tissue', `organ' or `neuronal cell', for example.
 Such a tissue ontology has been developed previously \cite{Lee2003}.
-The C.
-elegans  database, Wormbase \cite{Harris2014}, maintains a carefully curated list of gene expression data from GFP-reporters.
+The  \emph{C.
+elegans} database, WormBase \cite{Harris2014}, maintains a carefully curated list of gene expression data from GFP-reporters.
 	Here we provide a new framework that analyses user-input list for enrichment of specific tissues.
@@
 
-	Another problem frequently associated with GO analaysis is that it is often difficult to interpret due to the large number of terms associated with a given gene.
+	Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene.
 There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.
-A common tool for GO analysis, DAVID, clusters terms into broad categories that are amenable to exploration by researchers  \cite{Huang2007}, whereas PANTHER, a different software package  \cite{Mi2009, Mi2013}, attempts to solve this issue by employing a manually reduced ontology, GOslim (pers.
+A common tool for GO analysis, DAVID, clusters terms into broad categories that are amenable to exploration by researchers  \cite{Huang2007}, whereas PANTHER, a different software package  \cite{Mi2009, Mi2013}, attempts to solve this issue by employing a manually reduced ontology, GOslim (H.
+Yu and P.
+Thomas, pers.
 comm.
-H.
-Yu and P.
-Thomas).
+).
 
@@
 We also wanted to avoid testing tissues at levels where redundancy becomes problematic.
-For example, several left and right neurons have at least 25 annotating genes and we may want to include them for enrichment testing.
+For example, several left and right neurons have at least 25 annotated genes and we may want to include them for enrichment testing.
 However, many left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them.
@@
 At that point, no more nodes can be removed from that branch.
-This is guaranteed by the structure of the ontology: Parent nodes inherit all of the annotations of all of their descendants, so the number of annotating terms monotonically increases with increasing term hierarchy (see Figure~\ref{fig:trim_ends}).
+This is guaranteed by the structure of the ontology: Parent nodes inherit all of the annotations of all of their descendants, so the number of annotated terms monotonically increases with increasing term hierarchy (see Figure~\ref{fig:trim_ends}).
 In this way, we ensure that our term dictionary includes only those tissues that are considered sufficiently well annotated for statistical purposes.
@@
 \subsection*{Validation of the algorithm and parameter selection}
-	We wanted to select a dictionary which included enough terms to be specific beyond the largest C.
-elegans tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change.
+	We wanted to select a dictionary which included enough terms to be specific beyond the largest  \emph{C.
+elegans}tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change.
 Selection of a dictionary based only on minimization of spurious results would result in a dictionary with a large number of annotations per tissue, and would therefore include only the major tissues.
@@
 For this reason, we chose the `any' method for dictionary generation.
-The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant the trade-off between power and term specificity is.
+The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant is the trade-off between power and term specificity.
 Based on these observations, we ruled out the dictionary with the 100 gene annotation cut-off: it had the fewest terms and its q-values were not low enough to compensate the trade-off in specificity.
@@
 Strikingly, three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes (see \ref{fig:Dcon, fig:Hsp, fig:Smar}).
-A possible explanation for this might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioural response against the pathogen.
+A possible explanation for this might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen.
 Indeed, several studies \cite{Meisel2014, Zhang2005} have provided evidence that   \emph{C.
@@
 elegans} tissue ontology.
-We have also presented the first, to our knowledge, ontology trimming algorithm.
+We have also presented the first, to our knowledge, ontology trimming algorithm for biomedical ontologies.
 This algorithm, which is very easy to execute, places strong limits on the number of terms selected for testing.
@@
 
-Chikina \emph{et al}  \cite{Chikina2009} report a tissue enrichment model based on an SVM classifier that has been trained on microarray studies.
+Chikina \emph{et al}  \cite{Chikina2009} report a tissue enrichment model based on a Support Vector Machine classifier that has been trained on microarray studies.
 SVM classifiers are powerful tools capable of great sensitivity, but they require continuous retraining as tissue expression data becomes more available.
@@
 \section*{Availability of data and materials}
-Our web implementation is available at https://www.wormbase.org/tea. Our software can also be downloaded using Python's pip installer via the command
+Our web implementation is available at https://www.WormBase.org/tea. Our software can also be downloaded using Python's pip installer via the command
 
@@
 \section*{Acknowledgements}
+	We thank Justin Bois for his help and support. 
 	We would like to acknowledge all members of the Sternberg

### 2016-03-21 setup.py

@@ -10,25 +10,34 @@ from setuptools import setup, find_packages
 import os
 import sys
 
-version= '0.13.4'
+version= '0.13.9'
 
 #just type in python setup.py publish and this takes care of publishing to to pypi!
 
 
 #tag with git
-if sys.argv[-1] == 'tag':
-    os.system("git tag -a %s -m 'version %s'" % (version, version))
+if sys.argv[1] == 'tag':
+    
+    if len(sys.argv) > 2:   
+        if sys.argv[2] == '-m':
+            print(sys.argv[2])
+            print('please input your commit message')
+            message= sys.stdin.readline().strip()
+            print("git tag -a %s -m '%s'" % (version, message))
+            os.system("git tag -a %s -m '%s'" % (version, message))
+    else:    
+        os.system("git tag -a %s -m 'version %s'" % (version, version))
+        
     os.system("git push --tags")
     sys.exit()
-
 
 #publish to pypi
 if sys.argv[-1] == 'publish':
     os.system("python setup.py sdist upload")
     os.system("python setup.py bdist_wheel upload")
-    print("You probably want to also tag the version now:")
-    print("  git tag -a %s -m 'version %s'" % (version, version))
-    print("  git push --tags")
+#    print("You probably want to also tag the version now:")
+#    print("  git tag -a %s -m 'version %s'" % (version, version))
+#    print("  git push --tags")
     sys.exit()
 
 

### 2016-03-21 tests/hgf_benchmarking/doc/tea_main_text.tex
More writing.

@@
 with genes and to look for enrichment of specific terms. Although gene ontology
-(GO) is available for   \emph{C. elegans}, this ontology does not include
+(GO) is available for   \emph{Caenorhabditis elegans}, this ontology does not include
 anatomy or physiology information.
@@
 These names are processed into standard WBIDs and the result is displayed in the same window in an easy to read format containing all the relevant information, and a graph of the results is also displayed (see Figure ~\ref{fig:GUIresults}).
+Anatomy terms are displayed in human-readable format followed by their unique WBbt ID.
 
@@
 On the other hand, selecting a dictionary that can detect smaller tissues will bias us towards tissues with lesser annotations.
-To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.To help us select an appropriate dictionary and validate our tool, we found a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues \cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}.
+To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.
+	% Gold-standard data sets were annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
+	To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues \cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}.
+These data sets are annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
 Some of these studies have been used to annotate gene expression, and so they did not constitute an independent testing set.
@@
 Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade.
-Strikingly, three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes (see \ref{fig:Dcon, fig:Hsp, fig:Smar}).
+Strikingly, three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes (see \ref{fig:Dcon}, \ref{fig:Hsp},\ref{fig:Smar}).
 A possible explanation for this might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen.
@@
 \section*{Methods}
+\subsection*{Fetching annotation terms}
+We used the WormBase gene expression data, which includes annotated descriptions of spatial-temporal expression patterns of genes, to build our dictionary.
+Gene lists per anatomy term were extracted from a solr document store of gene expression data from the WS252 database provided by WormBase (PMID: 26578572).
+We used the solr document store because it provided a convenient access to expression data that included inferred annotations.
+That is, for each anatomy term, the expression gene list includes genes that were directly annotated to the term, as well as those that were annotated to the term's descendant terms (if there were any).
+Descendant terms were those connected with the focus term by is\_a/part\_of relationship chains defined in the anatomy term ontology hierarchy.
 \subsection*{Filtering nodes}
@@
 Therefore they are too similar and should be removed.
-	\csentence{\textbf{b)}} Nodes with less than a threshold number of genes are trimmed (light red) and discarded from the dictionary. Here, the example threshold is 25 genes.
-	\csentence{\textbf{c)}} We trim parent nodes recursively, starting from the root, if all their daughter nodes have more than the threshold number of annotations (light red).
+	\csentence{\textbf{b)}} Nodes with less than a threshold number of genes are trimmed (red) and discarded from the dictionary.
+Here, the example threshold is 25 genes.
+Nodes $\epsilon, \zeta, \eta$, shown in red will be removed.
+	\csentence{\textbf{c)}} We trim parent nodes recursively, starting from the root, if all their daughter nodes have more than the threshold number of annotations.
+Nodes in grey ($\epsilon, \zeta, \eta$) were removed in the previous step.
+Nodes $\alpha, \beta$ shown in red will be trimmed because each one has a complete daughter set.
+Only nodes $\gamma$ and $\delta$ will be used to generate the static dictionary.
   }
@@
   publication-ready graph is provided below, which can be saved by 
-  dragging to the desktop. Finally, lists of the genes used and discarded
-  for the analysis are also presented.
+  dragging to the desktop. The graph and the table show anatomy terms in
+  human-readable format, followed by their unique WBbt ID. Finally, lists 
+  of the genes used and discarded for the analysis are also presented.
   }
@@
   \captionsetup{width= 0.95\textwidth}
-  \caption{\csentence{Comparison of Enrichment Results for two dictionaries for a pan-neuronal enriched gene set from ~\cite{Watson2008a}.}
+  \caption{\csentence{Comparison of Enrichment Results for two dictionaries for a pan-neuronal enriched gene set from Watson, 2008~\cite{Watson2008a}.}
+	Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID.
    \textbf{a)} Dictionary with cutoff: 33; threshold: 0.95; method: `any'
@@
   \caption{\csentence{Independently derived genesets show similar results within a given dictionary}
-  \textbf{a)} Gene set from~\cite{Watson2008a}. Dictionary with cutoff: 33; threshold: 0.95; method: `any'
-  \textbf{b)} Gene set from~\cite{Spencer2011}. Dictionary with cutoff: 50; threshold: 0.95; method: `any'
+  Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID.
+  \textbf{a)} Gene set from Watson, 2008~\cite{Watson2008a}. Dictionary with cutoff: 33; threshold: 0.95; method: `any'
+  \textbf{b)} Gene set from Spencer, 2011~\cite{Spencer2011}. Dictionary with cutoff: 50; threshold: 0.95; method: `any'
    }
@@
 \begin{table}[h!]
-\caption{Parameter specifications and number of tissues for all dictionaries.}
+\caption{Parameter specifications and number of tissues for all dictionaries.
+The `Method' column refers to the trimming criterion for the similarity metric. We 
+used two such criteria, `any' and `avg'.
+`any': For a given sister set, if any sister had a similarity exceeding the corresponding
+threshold, all sisters were removed from the final dictionary.
+`avg': For a given sister set, if the average similarity across all the sisters in the set
+was greater than the corresponding threshold, all sisters were removed from the final dictionary.}
 	\csvautotabular{figures/TissueNumbers.csv}

### 2016-03-24 hypergeometricTests.py
Under some circumstances, not setting str() type causes error.

@@ -393,7 +393,7 @@ if __name__ == '__main__':
     
     tdf_name= args.tissue_dictionary
     gl_name= args.gene_list
-    title= str(args.title)
+    title= args.title
 
     if args.q:   
         q= args.q

### 2016-03-25 tests/hgf_benchmarking/doc/tea_main_text.tex
writing

@@
 %\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass[linenumbers, doublespacing]{bmcart}
-% \documentclass{bmcart}
+% \documentclass[linenumbers, doublespacing]{bmcart}
+\documentclass{bmcart}
 
@@
 Over the last ten years, there has been an explosive development in tools
-capable of measuring gene expression. These tools generate a large number
-of gene targets, but understanding these datasets, and forming hypotheses
+capable of measuring gene expression. These tools can identify thousands of altered 
+genes between conditions, but understanding these datasets and forming hypotheses
 based on them remains challenging. One way to analyze these datasets is to
-associate ontologies, which are controlled, hierarchical, descriptive vocabularies
+associate ontologies (which are controlled, hierarchical, descriptive vocabularies
+with controlled relations)
 with genes and to look for enrichment of specific terms. Although gene ontology
-(GO) is available for   \emph{Caenorhabditis elegans}, this ontology does not include
-anatomy or physiology information.
+(GO) is available for   \emph{Caenorhabditis elegans}, it does not include
+anatomy information.
 \parttitle{Results} %if any
-We have developed an enrichment analysis tool for the   \emph{C. elegans} tissue
-ontology, is available on the web via WormBase and is available for download 
-using Python's standard pip installer. In order to cut down on verbosity, we
+We have developed an  tool for identifying  enrichment of \emph{C. elegans} tissues. We
+applied a hypergeometric model to a pre-existing anatomy ontology for the worm in order to
+identify enriched tissues and generated a website GUI. To cut down on verbosity, we
 have come up with three straightforward filtering criteria that slim the ontology
-by almost tenfold. 
+by almost tenfold. One filter removes all terms with less than an arbitrary number
+of annotating genes; a second filter removes all sister terms that have identical
+gene annotations; and the third filter removes all parents whose daughters all survived
+the previous two filters. We adjusted these filters and validated our tool using a set of 30 
+gold standards from Expression Cluster data in WormBase. 
 \parttitle{Conclusions} %if any
-Our Tissue Enrichment Analysis (TEA), which can be found at www.WormBase.org/tea, 
-uses a standard hypergeometric function to test a slimmed-down   \emph{C. elegans} tissue 
-ontology and provides users with a text and graphic representation of the results. 
+Our Tissue Enrichment Analysis (TEA), which can be found at www.WormBase.org/tea, and
+can be downloaded using Python's standard pip installer. It employs a standard hypergeometric 
+function to test a slimmed-down   \emph{C. elegans} tissue ontology and provides users with a 
+text and graphic representation of the results. Our tool can discriminate between 
+embryonic and larval tissues and can even identify tissues down to the single-cell level. 
 \end{abstract}
@@
 \kwd{Gene Ontology}
-\kwd{Tissue Ontology}
+\kwd{Anatomy Ontology}
 \kwd{WormBase}
@@
 One such abstraction method relies on Gene Ontology (GO).
-GO provides a controlled set of hierarchically ordered terms in the form of an directed acyclic graph  \cite{TheGeneOntologyConsortium2000, Ontology2009, TheGeneOntologyConsortium2015} that provide detailed information about the molecular, cellular or biochemical functions of the gene among others.
-For a given gene list, certain software programs can query whether a particular gene is enriched  \cite{Mi2009, McLean2010, Huang2009}.
+GO provides a controlled set of hierarchically ordered terms  \cite{TheGeneOntologyConsortium2000, Ontology2009, TheGeneOntologyConsortium2015} that provide detailed information about the molecular, cellular or biochemical functions of any gene.
+For a given gene list, certain software programs can query whether a particular term is enriched  \cite{Mi2009, McLean2010, Huang2009}.
 One area of biological significance that GO does not include is anatomy.
-One way to address this shortcoming is to generate a `tissue ontology' that provides a complete anatomical description for an organism or sets of organisms, such as `tissue', `organ' or `neuronal cell', for example.
-Such a tissue ontology has been developed previously \cite{Lee2003}.
+One way to address this shortcoming is to use test a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `neuronal cell'), in this case for \emph{C.
+elegans} \cite{Lee2003}.
 The  \emph{C.
 elegans} database, WormBase \cite{Harris2014}, maintains a carefully curated list of gene expression data from GFP-reporters.
-	Here we provide a new framework that analyses user-input list for enrichment of specific tissues.
-We believe that tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation.
+Here we provide a new framework that analyses a user-input list for enrichment of specific tissues.
+Tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation.
 As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.
@@
 There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.
-A common tool for GO analysis, DAVID, clusters terms into broad categories that are amenable to exploration by researchers  \cite{Huang2007}, whereas PANTHER, a different software package  \cite{Mi2009, Mi2013}, attempts to solve this issue by employing a manually reduced ontology, GOslim (H.
+DAVID, a common tool for GO enrichment analysis, clusters terms into broad categories  \cite{Huang2007}, whereas PANTHER \cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (H.
 Yu and P.
@@
 
-	To prevent our tool from suffering from the same drawbacks, we have cut down on result verbosity by filtering the ontology by using a small set of well-defined criteria to remove terms that do not contribute extra information.
+	To prevent our tool from suffering from the same drawbacks, we have cut down on result verbosity by filtering the ontology by using a small set of well-defined criteria to remove terms that do not contribute additional information.
 To our knowledge, such filtering has never been performed in an algorithmic fashion for an ontology before---indeed, tools such as DAVID do not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
@@
 For validation (see below) we made a number of different dictionaries.
-The original ontology has 1675 terms with more than 5 gene annotations.
-After filtering, dictionary sizes ranged from 21 to a maximum of 400 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by tenfold by application of a few simple filters (see Table ~\ref{tab:DictionarySpecs}).
+The original ontology has almost 6,000 terms of which 1675 have at least 5 gene annotations.
+After filtering, dictionary sizes ranged from 21 to a maximum of 400 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by an order of magnitude through application of a few simple filters (see Table ~\ref{tab:DictionarySpecs}).
 These filters were used to compile a static dictionary that we employ for all analyses.
@@
 
-% \begin{figure}[h!]
-%   \caption{\csentence{Screenshot of the web GUI.}
-%       }
-% 	  \label{fig:GUI}
-%
-%       \end{figure}
-% \begin{figure}[h!]
-%   \caption{\csentence{Screenshot of results from web GUI.}
-%       }
-% 	  \label{fig:GUIresults}
-      % \end{figure}
-
 % statistical testing and validation Figures

### 2016-03-25 tissue_enrichment_analysis/hypergeometricTests.py

@@ -30,32 +30,32 @@ import contextlib
 
 def pass_list(user_provided, tissue_dictionary):
     """
-    A function to check which genes are in the 
+    A function to check which genes are in the
     provided list of user provided names and to
     return a dataframe of presence or fail
-    
+
     Takes two vectors, user_provided and tissue_dictionary in that order
     tissue_dictionary is a pandas dataframe that you should know about
     user_provided is a list of gene names
     """
     length= tissue_dictionary.shape[0] #how many genes in the dict
-    
+
     #make an empty dataframe
     present= pd.DataFrame(index= range(length), columns= ['wbid','provided'])
-    
+
     #fill wbid column with all the gene names
     present.wbid= tissue_dictionary.wbid
-    
+
     #go through and pass attendance -- 1 if present, 0 otherwise
     for item in user_provided:
         present.provided[present.wbid==item]= 1
-        
+
     #remove NA's and make them 0
     present.provided= present.provided.fillna(0)
-    
+
     #return df
     return present
-    
+
 #==============================================================================
 #hgf is short for hypergeometric function
 #==============================================================================
@@ -65,44 +65,44 @@ def hgf(gene_list, tissue_df):
     returns a p-dictionary for the enrichment of every tissue
     (a p-dictionary is a vector of length equal to the number
     of tissues in the tissue_dictionary, sorted by value)
-    
+
     The entries of the p-vector are p-values not corrected
-    for multiple hypothesis testing. 
-    
+    for multiple hypothesis testing.
+
     gene_list should be a list or list-like
     tissue_dictionary should be a pandas df
-    """    
+    """
     #figure out what genes are in the user provided list
-    present= pass_list(gene_list, tissue_df)  
-    
+    present= pass_list(gene_list, tissue_df)
+
     unused= present[present.provided == 0].wbid
-    
-    #slice out only the genes that were present from the user-provided list    
+
+    #slice out only the genes that were present from the user-provided list
     wanted= present.wbid[present.provided==1]
 
     #re-index the dictionary s.t. the wbid is the index
     tissue_df= tissue_df.set_index('wbid')
-    
+
     #number of tissues in the dictionary
-    sums_of_tissues= tissue_df.sum()[1:] #this object can be identified by 
[…]

### 2016-03-30 tests/hgf_benchmarking/doc/tea_main_text.tex
Minor modifications
Modified hgf_benchmark to add some analyses.

Writing the article.

@@
 \graphicspath {{figures/}}
+\newcommand{\ignore}[2]{\hspace{0in}#2}
 
@@
 Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade.
-Strikingly, three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes (see \ref{fig:Dcon}, \ref{fig:Hsp},\ref{fig:Smar}).
+Strikingly, three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes (see \ref{fig:Dcon},\ignore{ \ref{fig:Hsp},\ref{fig:Smar}}).
 A possible explanation for this might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen.
@@
   \caption{\csentence{Independently derived genesets show similar results within a given dictionary}
-  Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID.
-  \textbf{a)} Gene set from Watson, 2008~\cite{Watson2008a}. Dictionary with cutoff: 33; threshold: 0.95; method: `any'
-  \textbf{b)} Gene set from Spencer, 2011~\cite{Spencer2011}. Dictionary with cutoff: 50; threshold: 0.95; method: `any'
+  Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID. Dictionary with cutoff: 33; threshold: 0.95; method: `any'.
+  \textbf{a)} GABAergic gene set from Watson, 2008~\cite{Watson2008a}.
+  \textbf{b)} GABAergic gene set from Spencer, 2011~\cite{Spencer2011}.
    }
@@
 
-\begin{figure}[h!]
-    \includegraphics[width=0.95\textwidth]{figures/HarposporiumspEnrichment.png}
-	\captionsetup{width= 0.95\textwidth}
-	\caption{\csentence{\emph{Harposporium sp.} enrichment results}
-      Genes altered in   \emph{C. elegans} after 24hr exposure to \emph{Harposporium sp.} a fungus, shows enrichment
-	  in the posterior lateral ganglion, similarly to \emph{D. coniospora}. This particular fungus seems to affect
-	  genes normally associated with neuronal precursor tissues as well. 
-	  }
-	\label{fig:Hsp}
-\end{figure}
+% \begin{figure}[h!]
+%     \includegraphics[width=0.95\textwidth]{figures/HarposporiumspEnrichment.png}
+% 	\captionsetup{width= 0.95\textwidth}
+% 	\caption{\csentence{\emph{Harposporium sp.} enrichment results}
+%       Genes altered in   \emph{C. elegans} after 24hr exposure to \emph{Harposporium sp.} a fungus, shows enrichment
+% 	  in the posterior lateral ganglion, similarly to \emph{D. coniospora}. This particular fungus seems to affect
+% 	  genes normally associated with neuronal precursor tissues as well.
+% 	  }
+% 	\label{fig:Hsp}
+% \end{figure}
 
@@
 \end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=0.95\textwidth]{figures/EnterococcusfaecalisEnrichment.png}
-	\captionsetup{width= 0.95\textwidth}
-  	\caption{\csentence{\emph{E. faecalis} enrichment results}
-      Genes altered in   \emph{C. elegans} after 24hr exposure to \emph{E. faecalis}, a bacterium, shows an unusual pattern:
-	  No terms were identified as enriched in the down-regulated category. Surprisingly, up-regulated genes show enrichment
-	  of various head neurons that were down-regulated with exposure to other pathogens. Interestingly, some pharyngeal
-	  muscle terms are also enriched. 
-	  }
-	\label{fig:Efaec}
-\end{figure}
-
-\begin{figure}[h!]
-    \includegraphics[width=0.95\textwidth]{figures/OtorhabdusluminescensEnrichment.png}
-	\captionsetup{width= 0.95\textwidth}
-  	\caption{\csentence{\emph{O. luminescens} enrichment results}
-      Genes altered in   \emph{C. elegans} after 24hr exposure to \emph{O. luminescens}, a bacterium, shows no
-	  significantly enriched terms in down-regulated genes. However, up-regulated genes suggest that the hypodermis
-	  and excretory duct cell may be affected by infection. 
-	  }
-	\label{fig:Olum}
-\end{figure}
+%
+% \begin{figure}[h!]
+%     \includegraphics[width=0.95\textwidth]{figures/EnterococcusfaecalisEnrichment.png}
+% 	\captionsetup{width= 0.95\textwidth}
+%   	\caption{\csentence{\emph{E. faecalis} enrichment results}
+%       Genes altered in   \emph{C. elegans} after 24hr exposure to \emph{E. faecalis}, a bacterium, shows an unusual pattern:
+% 	  No terms were identified as enriched in the down-regulated category. Surprisingly, up-regulated genes show enrichment
+% 	  of various head neurons that were down-regulated with exposure to other pathogens. Interestingly, some pharyngeal
+% 	  muscle terms are also enriched.
+% 	  }
+% 	\label{fig:Efaec}
+% \end{figure}
+%
+% \begin{figure}[h!]
+%     \includegraphics[width=0.95\textwidth]{figures/OtorhabdusluminescensEnrichment.png}
+% 	\captionsetup{width= 0.95\textwidth}
+%   	\caption{\csentence{\emph{O. luminescens} enrichment results}
+%       Genes altered in   \emph{C. elegans} after 24hr exposure to \emph{O. luminescens}, a bacterium, shows no
+% 	  significantly enriched terms in down-regulated genes. However, up-regulated genes suggest that the hypodermis
+% 	  and excretory duct cell may be affected by infection.
+% 	  }
+% 	\label{fig:Olum}
+% \end{figure}
 

### 2016-03-30 tests/hgf_benchmarking/src/hgf_benchmark_script.py

@@ -1,100 +1,102 @@
 # -*- coding: utf-8 -*-
 """
-Spyder Editor
-
+A script to benchmark TEA.
 
 @david angeles
 dangeles@caltech.edu
 """
 
-import tissue_enrichment_analysis as tea #the library to be used
+import tissue_enrichment_analysis as tea  # the library to be used
 import pandas as pd
 import os
-import importlib as imp
 import numpy as np
 import seaborn as sns
 import matplotlib.pyplot as plt
 import re
 import matplotlib as mpl
 
+
 sns.set_context('paper')
 
-pd.set_option('display.float_format', lambda x:'%f'%x) 
+# pd.set_option('display.float_format', lambda x:'%f'%x)
+pd.set_option('precision', 3)
+# this script generates a few directories.
+
+dirOutput = '../output/'
+dirSummaries = '../output/SummaryInformation/'
+dirHGT25_any = '../output/HGT25_any_Results/'
+dirHGT33_avg = '../output/HGT33_any_Results/'
+dirHGT33_any = '../output/HGT33_any_Results/'
+dirHGT50_any = '../output/HGT50_any_Results/'
+dirHGT100_any = '../output/HGT100_any_Results/'
+
+DIRS = [dirOutput, dirSummaries, dirHGT25_any, dirHGT33_avg,
+        dirHGT50_any, dirHGT100_any]
+# open the relevant file
+path_sets = '../input/genesets_golden/'
+path_dicts = '../input/WS252AnatomyDictionary/'
 
 
-#this script generates a few directories. 
-dirOutput= '../output/'
-dirSummaries= '../output/SummaryInformation/'
-dirHGT25_any= '../output/HGT25_any_Results/'
-dirHGT33_avg= '../output/HGT33_avg_Results/'
-dirHGT33_any= '../output/HGT33_any_Results/'
-dirHGT50_any= '../output/HGT50_any_Results/'
-dirHGT100_any= '../output/HGT100_any_Results/'
-
-DIRS= [dirOutput, dirSummaries, dirHGT25_any, dirHGT33_avg,
-       dirHGT50_any, dirHGT100_any]
-#open the relevant file
-path_sets= '../input/genesets_golden/'
-path_dicts= '../input/WS252AnatomyDictionary/'
-
-
-#Make all the necessary dirs if they don't already exist
+# Make all the necessary dirs if they don't already exist
 for d in DIRS:
     if not os.path.exists(d):
         os.makedirs(d)
 
[…]

### 2016-04-04 tests/hgf_benchmarking/doc/tea_main_text.tex
writing.

@@
 
-%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
+% \documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
 % \documentclass[linenumbers, doublespacing]{bmcart}
 \documentclass{bmcart}
+% \bibliographystyle{unsrtnat}
 
@@
 \usepackage{caption}
+\usepackage{adjustbox}
 
@@
 % \def\includegraphics{}
+\bibliographystyle{unsrt}
 
 \graphicspath {{figures/}}
-\newcommand{\ignore}[2]{\hspace{0in}#2}
 
@@
 \startlocaldefs
+
+% ignore command 
+\newcommand{\ignore}[2]{\hspace{0in}#2}
+
 \endlocaldefs
@@
   \city{Pasadena},                              % city
-  \cny{US}                                    % country
+  \cny{USA}                                    % country
 }
@@
 \parttitle{Background} %if any
-Over the last ten years, there has been an explosive development in tools
-capable of measuring gene expression. These tools can identify thousands of altered 
-genes between conditions, but understanding these datasets and forming hypotheses
-based on them remains challenging. One way to analyze these datasets is to
-associate ontologies (which are controlled, hierarchical, descriptive vocabularies
-with controlled relations)
-with genes and to look for enrichment of specific terms. Although gene ontology
-(GO) is available for   \emph{Caenorhabditis elegans}, it does not include
-anatomy information.
+Over the last ten years, there has been explosive development in tools
+for measuring gene expression. These tools can identify thousands of altered 
+genes between conditions, but understanding these datasets and forming hypotheses based on them remains challenging.
+One way to analyze these datasets is to associate ontologies (hierarchical, descriptive vocabularies
+with controlled relations between terms) with genes and to look for enrichment of specific terms.
+Although gene ontology (GO) is available for   \emph{Caenorhabditis~elegans}, it does not include anatomy information.
 \parttitle{Results} %if any 
-We have developed an  tool for identifying  enrichment of \emph{C. elegans} tissues. We
-applied a hypergeometric model to a pre-existing anatomy ontology for the worm in order to
-identify enriched tissues and generated a website GUI. To cut down on verbosity, we
-have come up with three straightforward filtering criteria that slim the ontology
-by almost tenfold. One filter removes all terms with less than an arbitrary number
-of annotating genes; a second filter removes all sister terms that have identical
-gene annotations; and the third filter removes all parents whose daughters all survived
-the previous two filters. We adjusted these filters and validated our tool using a set of 30 
-gold standards from Expression Cluster data in WormBase. 
+We have developed an  tool for identifying  enrichment of \emph{C.~elegans} tissues among gene sets.
+We applied a hypergeometric model to a pre-existing anatomy ontology for the worm in order to identify enriched tissues and generated a website GUI.
+To cut down on verbosity, we have come up with three straightforward filtering criteria that slim the ontology by almost tenfold.
+One filter removes all terms with fewer than an arbitrary number of annotating genes; the second filter removes all sister terms that have identical gene annotations; and the third filter removes all parents whose daughters all survived the previous two filters.
+We adjusted these filters and validated our tool using a set of 30  gold standards from Expression Cluster data in WormBase.
 \parttitle{Conclusions} %if any
-Our Tissue Enrichment Analysis (TEA), which can be found at www.WormBase.org/tea, and
-can be downloaded using Python's standard pip installer. It employs a standard hypergeometric 
-function to test a slimmed-down   \emph{C. elegans} tissue ontology and provides users with a 
-text and graphic representation of the results. Our tool can discriminate between 
-embryonic and larval tissues and can even identify tissues down to the single-cell level. 
+Our Tissue Enrichment Analysis (TEA) can be found at www.WormBase.org/tea, and can be downloaded using Python's standard pip installer.
+It tests a slimmed-down \emph{C.~elegans} tissue ontology for enrichment of specific terms and provides users with a  text and graphic representation of the results.
+Our tool can discriminate between embryonic and larval tissues and can even identify tissues down to the single-cell level.
 \end{abstract}
@@
 One such abstraction method relies on Gene Ontology (GO).
-GO provides a controlled set of hierarchically ordered terms  \cite{TheGeneOntologyConsortium2000, Ontology2009, TheGeneOntologyConsortium2015} that provide detailed information about the molecular, cellular or biochemical functions of any gene.
-For a given gene list, certain software programs can query whether a particular term is enriched  \cite{Mi2009, McLean2010, Huang2009}.
+GO provides a controlled set of hierarchically ordered terms~\cite{TheGeneOntologyConsortium2000a, TheGeneOntologyConsortium2015} that provide detailed information about the molecular, cellular or biochemical functions of any gene.
+For a given gene list, certain software programs can query whether a particular term is enriched~\cite{Mi2009, McLean2010, Huang2009}.
 One area of biological significance that GO does not include is anatomy.
-One way to address this shortcoming is to use test a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `neuronal cell'), in this case for \emph{C.
-elegans} \cite{Lee2003}.
-The  \emph{C.
-elegans} database, WormBase \cite{Harris2014}, maintains a carefully curated list of gene expression data from GFP-reporters.
+One way to address this shortcoming is to use test a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `neuronal cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}.
+The \emph{C.~elegans} database, WormBase~\cite{Harris2014}, maintains a carefully curated list of gene expression data from GFP-reporters.
 Here we provide a new framework that analyses a user-input list for enrichment of specific tissues.
@@
 There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.
-DAVID, a common tool for GO enrichment analysis, clusters terms into broad categories  \cite{Huang2007}, whereas PANTHER \cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (H.
+DAVID, a common tool for GO enrichment analysis, clusters enriched terms into broad categories~\cite{Huang2007}, whereas PANTHER~\cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (H.
 Yu and P.
@@
 
-	To prevent our tool from suffering from the same drawbacks, we have cut down on result verbosity by filtering the ontology by using a small set of well-defined criteria to remove terms that do not contribute additional information.
-To our knowledge, such filtering has never been performed in an algorithmic fashion for an ontology before---indeed, tools such as DAVID do not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
+ To prevent our tool from suffering from verbosity problems, we have filtered our ontology using a small set of well-defined criteria to remove terms that do not contribute additional information.
+To our knowledge, such filtering has never been performed in an algorithmic fashion for a biomedical ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
+Other pruning tools do exist (see for example~\cite{Kim2007}), but the pruning is query-dependent.
 We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.
@@
 As a first step to generate our tissue enrichment software, we wished to select tissue terms that were reasonably well-annotated, yet specific enough to provide insight and not redundant with other terms.
-We also wanted to avoid testing tissues at levels where redundancy becomes problematic.
+We also wanted to avoid
+testing tissues at levels where redundancy becomes problematic.
 For example, several left and right neurons have at least 25 annotated genes and we may want to include them for enrichment testing.
-However, many left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them.
+However, many left/right neuronal
+pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them.
 We reasoned that when two tissues have almost identical annotations, we cannot have statistical confidence in differentiating between them.
@@
 Many nodes have 0-10 annotations, which we consider too few to accurately test.
-To solve this issue, we implemented a straightforward trimming algorithm.
+To solve this issue, we implemented a
+straightforward trimming algorithm.
 For a given terminal node, we test whether the node has more than a threshold number of annotations.
@@
 To address this issue we removed parent nodes from the analysis if all their daughter nodes passed the annotation threshold (see Figure~\ref{fig:trim_roots}).
+We called this a ceiling filter.
 
@@
 By itself, each of these filters can reduce the number of nodes employed for analysis.
-Notably, these filters are not all commutative: while trimming and redundancy filtering are commutative, applying the ceiling filter is not commutative with either the trimming or the redundancy filter.
+These filters are not all commutative: while trimming and redundancy filtering are commutative, applying the ceiling filter is not commutative with either the trimming or the redundancy filter.
 If the ceiling filter is applied before any other filter, only terminal nodes will remain, since all the parents have complete daughter sets.
@@
 The original ontology has almost 6,000 terms of which 1675 have at least 5 gene annotations.
-After filtering, dictionary sizes ranged from 21 to a maximum of 400 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by an order of magnitude through application of a few simple filters (see Table ~\ref{tab:DictionarySpecs}).
-These filters were used to compile a static dictionary that we employ for all analyses.
+After filtering, dictionary sizes ranged from 21 to a maximum of 460 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by an order of magnitude through application of a few simple filters (see Table~\ref{tab:DictionarySpecs}).
+These filters were used to compile a static dictionary that we employ for all analyses (see \emph{Validation of the algorithm and parameter selection} section for details).
 
@@
 \end{eqnarray}
+
 Here, $n_i$ is the number of balls of type $i$ drawn, $N$ is the total number  of draws, $m_i$ is tissue $i$ and $M$ is the total number of balls in the urn.
@@
 
-	Once the probability of drawing the labels has been quantified, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm \cite{Benjamini1995}.
+Once the probability of drawing the labels has been quantified, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm~\cite{Benjamini1995}.
 Genes that have a q-value less than a given alpha are considered significant.
-Our default setting is to set the alpha threshold at 0.1, but users will be able to modify this value either in batch or in our web application.
+Our default setting is to set alpha at 0.1.
+
+Users input a gene-list using any valid gene name for \emph{C.~elegans}.
+These names are processed into standard WBIDs.
 The program returns a text-based table showing the tissues that tested significant, along with their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change (observed hits / expected hits).
 Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values.
-	Our software is implemented in an easy to use GUI within WormBase.
-Users input a gene-list using any valid gene name for  \emph{C.
-elegans}.
-These names are processed into standard WBIDs and the result is displayed in the same window in an easy to read format containing all the relevant information, and a graph of the results is also displayed (see Figure ~\ref{fig:GUIresults}).
+Our software is implemented in an easy to use GUI within WormBase (see Figure~\ref{fig:GUIresults}).
 Anatomy terms are displayed in human-readable format followed by their unique WBbt ID.
@@
 \subsection*{Validation of the algorithm and parameter selection}
-	We wanted to select a dictionary which included enough terms to be specific beyond the largest  \emph{C.
-elegans}tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change.
+We wanted to select a dictionary which included enough terms to be specific beyond the largest \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change.
 Selection of a dictionary based only on minimization of spurious results would result in a dictionary with a large number of annotations per tissue, and would therefore include only the major tissues.
 On the other hand, selecting a dictionary that can detect smaller tissues will bias us towards tissues with lesser annotations.
-To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.
+To our knowledge there is no good method for assessing false-positive or false-negative results for
+annotations.
 % Gold-standard data sets were annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
-	To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues \cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}.
+To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues~\cite{Gaudet2004a, Spencer2011,
+Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}.
 These data sets are annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
-Some of these studies have been used to annotate gene expression, and so they did not constitute an independent testing set.
+Some of these studies have
+been used to annotate gene expression, and so they did not constitute an independent testing set.
 To correct this flaw, we built a clean dictionary that specifically excluded all annotation evidence that came from these studies.
 
-	As a first attempt to select a good dictionary, we generated all the possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any' thresholding criteria for the latter (see Table ~\ref{tab:DictionarySpecs}).
+As a first attempt to select a good dictionary, we generated all the possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any'
+thresholding criteria for the latter (see Table~\ref{tab:DictionarySpecs}).
 For these dictionaries, the number of tissues tested ranged from 21 to 460.
-The number of tissues was inversely correlated to the minimum annotation, as expected, and was largely insensitive to the redundancy threshold, at least in the range we explored (0.9-1).
+The number of tissues was inversely correlated to the minimum annotation, as
+expected, and was largely insensitive to the redundancy threshold, at least in the range we explored (0.9-1).
 Next, we analyzed all 30 datasets using each dictionary.
@@
 When we analyzed the distribution of significant q-values for the dictionaries, we found that the similarity threshold mattered relatively little for any dictionary.
-We also noticed that the `any' thresholding method resulted in tighter histograms with a mode closer to 0.
+We also noticed that the `any' thresholding method resulted in tighter histograms with fa mode closer to 0.
 For this reason, we chose the `any' method for dictionary generation.
-The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant is the trade-off between power and term specificity.
+The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant is the trade-off between power and term
+specificity.
 Based on these observations, we ruled out the dictionary with the 100 gene annotation cut-off: it had the fewest terms and its q-values were not low enough to compensate the trade-off in specificity.
@@
 %Work well
+% TODO add WBbt codes!
 Most results were highly comparable and in line with what was expected.
 For some sets, all dictionaries seemed to perform well.
-For example, in our `all neuron enriched sets'~ \cite{Spencer2011, Watson2008a} the results were an amalgamation of neuron related terms including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays regardless of the dictionary used (see~\ref{fig:interagree}).
-On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~ \cite{Spencer2011}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells ---at the 33-cutoff, our tool identified `Z2' and `Z3' as being five-fold enriched; whereas at the 25 gene-cutoff the terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
+For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} the results were an amalgamation of neuron related terms including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays regardless of the dictionary used (see~\ref{fig:interagree}).
+On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells ---at the 33-cutoff, our tool identified `Z2' and `Z3' as being five-fold enriched; whereas at the 25 gene-cutoff the terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
 %Differences between dictionaries
-We queried an embryonic stage intestine precursor geneset~ \cite{Spencer2011}.
+We queried an embryonic stage intestine precursor geneset~\cite{Spencer2011}.
 Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary.
@@
 Not all queries worked equally well.
-For example, a number of intestinal enriched genes sets~ \cite{Spencer2011, Pauli2006} were not enriched in intestine in any dictionary, but they were enriched for pharynx- and hypodermis-related terms.
+For example, a number of intestinal enriched genes sets~\cite{Spencer2011, Pauli2006} were not enriched in intestine in any dictionary, but they were enriched for pharynx- and hypodermis-related terms.
 We were somewhat surprised that intestinal gene sets performed poorly, since the intestine is a relatively well-annotated tissue.
@@
 We also assessed the internal agreement of our tool by using independent gene-sets that we expected to be enriched in the same tissues.
-We had two independent pan-neuronal sets~ \cite{Spencer2011, Watson2008a}; two independent PVD enriched sets~ \cite{Spencer2011, Smith2010}; two independent GABAergic gene sets~ \cite{Spencer2011, Cinar2005}; two independent pharyngeal gene sets~ \cite{Spencer2011, Gaudet2004a}; and two independent intestinal gene sets~ \cite{Spencer2011, Pauli2006}.
+We had two independent pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two independent PVD enriched sets~\cite{Spencer2011, Smith2010}; two independent GABAergic gene sets~\cite{Spencer2011, Cinar2005}; two independent pharyngeal gene sets~\cite{Spencer2011, Gaudet2004a}; and two independent intestinal gene sets~\cite{Spencer2011, Pauli2006}.
 Overall, the tool seems to have good internal agreement.
-On most sets, the same terms were enriched, although order was somewhat variable (see Figure ~\ref{fig:intragree}).
+On most sets, the same terms were enriched, although order was somewhat variable (see Table~\ref{tab:intraagree}).
 However, most high-scoring terms were preserved between gene sets.
@@
 All comparisons can be found online in our Github repository (see Availability of data and materials).
-Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95 redundancy threshold using the `any' criterion.
+Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95
+redundancy threshold using the `any' criterion.
 seemed to perform well, with a good balance between specificity, verbosity and accuracy, so we selected this parameter set to generate our static dictionary.
@@
 \subsection*{A brief example}
-We applied our tool to the RNA-seq datasets developed by Engelmann et al.
-\cite{Engelmann2011} to gain further understanding of the biology underlying these datasets.
+We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of the biology underlying these datasets.
 Engelmann et al.
@@
 We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated.
-Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade.
-Strikingly, three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes (see \ref{fig:Dcon},\ignore{ \ref{fig:Hsp},\ref{fig:Smar}}).
+Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade.Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct.
+Three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes (for an example, see~\ref{fig:Dcon}\ignore{,~\ref{fig:Hsp},\ref{fig:Smar}}).
 A possible explanation for this might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen.
-Indeed, several studies \cite{Meisel2014, Zhang2005} have provided evidence that   \emph{C.
-elegans} uses chemosensory neurons to identify pathogens.
-Interestingly, one bacterium did not exhibit the same pattern of down-regulation of neuronal-associated genes.
-\emph{E.
-faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E.
-faecalis} may have a different pathogenic profile.
-Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct.
-	Our results highlight the involvement of various   \emph{C. elegans} neuronal tissues in pathogen defense and/or illness.
+Indeed, several studies~\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.~elegans} uses chemosensory neurons to identify pathogens.
+One bacterium did not exhibit the same pattern of down-regulation of neuronal-associated genes.
+\emph{E.~faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E.~faecalis} may have a different pathogenic profile.
+Our results highlight the involvement of various \emph{C.~elegans} neuronal tissues in pathogen defense.
 
 \section*{Discussion}
-We have presented a tissue enrichment analysis tool that employs a standard hypergeometric model to test the   \emph{C.
-elegans} tissue ontology.
+We have presented a tissue enrichment analysis tool that employs a standard hypergeometric model to test the \emph{C.~elegans} tissue ontology.
 We have also presented the first, to our knowledge, ontology trimming algorithm for biomedical ontologies.
@@
 
-Chikina \emph{et al}  \cite{Chikina2009} report a tissue enrichment model based on a Support Vector Machine classifier that has been trained on microarray studies.
+Chikina \emph{et al}~\cite{Chikina2009} report a tissue enrichment model based on a Support Vector Machine classifier that has been trained on microarray studies.
 SVM classifiers are powerful tools capable of great sensitivity, but they require continuous retraining as tissue expression data becomes more available.
@@
 Our tool relies on an annotation dictionary that is continuously updated, does not require retraining and does not require ranked genes.
-To our knowledge, there are no other tissue ontology enrichment tools in   \emph{C.
-elegans}, but similar projects exist for humans and zebrafish ~\cite{Lee2013, Prykhozhij2013}, highlighting the relevance of our tool for high-dimensionality biology.
+To our knowledge, there are no other tissue ontology enrichment tools in \emph{C.~elegans}, but similar projects exist for humans and zebrafish~\cite{Lee2013, Prykhozhij2013}, highlighting the relevance of our tool for high-dimensionality biology.
 
@@
 All scripts were written in Python.
-Our software relies on the Pandas, Numpy, Seaborn and SciPy modules to perform all statistical testing and data handling \cite{McKinney2011, VanDerWalt2011, Oliphant2007}.
+Our software relies on the Pandas, Numpy, Seaborn and SciPy modules to perform all statistical testing and data handling~\cite{McKinney2011, VanDerWalt2011, Oliphant2007}.
 
@@
   Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID. Dictionary with cutoff: 33; threshold: 0.95; method: `any'.
-  \textbf{a)} GABAergic gene set from Watson, 2008~\cite{Watson2008a}.
-  \textbf{b)} GABAergic gene set from Spencer, 2011~\cite{Spencer2011}.
+  \textbf{a)} GABAergic gene set from Watson~\cite{Watson2008a}.
+  \textbf{b)} GABAergic gene set from Spencer~\cite{Spencer2011}.
    }
@@
   	\caption{\csentence{\emph{D. coniospora} enrichment results}
-      Genes altered in   \emph{C. elegans} after 24hr exposure to \emph{D. coniospora}, a fungus, show enrichment
+      Genes altered in   \emph{C.~elegans} after 24hr exposure to \emph{D. coniospora}, a fungus, show enrichment
 	  in many neuronal tissues.}
@@
 % 	\caption{\csentence{\emph{Harposporium sp.} enrichment results}
-%       Genes altered in   \emph{C. elegans} after 24hr exposure to \emph{Harposporium sp.} a fungus, shows enrichment
+%       Genes altered in   \emph{C.~elegans} after 24hr exposure to \emph{Harposporium sp.} a fungus, shows enrichment
 % 	  in the posterior lateral ganglion, similarly to \emph{D. coniospora}. This particular fungus seems to affect
@@
 \begin{figure}[h!]
+	% \begin{centering}
     \includegraphics[width=0.95\textwidth]{figures/SerratiamarcescensEnrichment.png}
@@
   	\caption{ \csentence{\emph{Serratia marcescens} enrichment results}
-      Genes altered in   \emph{C. elegans} after 24hr exposure to \emph{Serratia marcescens}, a bacterium, shows 
+      Genes altered in   \emph{C.~elegans} after 24hr exposure to \emph{Serratia marcescens}, a bacterium, shows 
 	  down-regulation of chemosensory neurons notably ASER, ASEL, ASJ and ASK. The lateral ganglions are also 
@@
 	}  
+	% \end{centering}
 	\label{fig:Smar}
@@
 %   	\caption{\csentence{\emph{E. faecalis} enrichment results}
-%       Genes altered in   \emph{C. elegans} after 24hr exposure to \emph{E. faecalis}, a bacterium, shows an unusual pattern:
+%       Genes altered in   \emph{C.~elegans} after 24hr exposure to \emph{E. faecalis}, a bacterium, shows an unusual pattern:
 % 	  No terms were identified as enriched in the down-regulated category. Surprisingly, up-regulated genes show enrichment
@@
 %   	\caption{\csentence{\emph{O. luminescens} enrichment results}
-%       Genes altered in   \emph{C. elegans} after 24hr exposure to \emph{O. luminescens}, a bacterium, shows no
+%       Genes altered in   \emph{C.~elegans} after 24hr exposure to \emph{O. luminescens}, a bacterium, shows no
 % 	  significantly enriched terms in down-regulated genes. However, up-regulated genes suggest that the hypodermis
@@
 \section*{Tables}
+\begin{table}[h!]
+	\caption{Comparison of results for a neuronal-enriched geneset from Watson~\cite{Watson2008a}. 
+	We ran the same genelist on a dictionary with a minimum annotation cutoff of 50, similarity threshold of 0.95
+	and similarity method `any' versus another with a minimum annotation cutoff of 33, similarity threshold of 0.95
+	and similarity method `any'. In the table, columns are labeled with their significance value (Q-value) or enrichment
+	fold change followed by a hyphen and a number which indicates which dictionary generated that value. 
+	Not all tissues are present in either dictionary. Hyphens denote not-applicable values, which occurs when a particular
+	tissue is not present in both dictionaries. Full table is available on github
+	}
+	\begin{adjustbox}{width=1\textwidth}
+		\csvautotabular{figures/dict-comparison-50-33.csv}
+		\label{tab:intraagree}
+	\end{adjustbox}
+\end{table}
+
 \begin{table}[h!]

### 2016-04-04 tests/hgf_benchmarking/src/hgf_benchmark_script.py

@@ -15,7 +15,6 @@ import matplotlib.pyplot as plt
 import re
 import matplotlib as mpl
 
-
 sns.set_context('paper')
 
 # pd.set_option('display.float_format', lambda x:'%f'%x)
@@ -29,9 +28,10 @@ dirHGT33_avg = '../output/HGT33_any_Results/'
 dirHGT33_any = '../output/HGT33_any_Results/'
 dirHGT50_any = '../output/HGT50_any_Results/'
 dirHGT100_any = '../output/HGT100_any_Results/'
+dirComp = '../output/comparisons/'
 
 DIRS = [dirOutput, dirSummaries, dirHGT25_any, dirHGT33_avg,
-        dirHGT50_any, dirHGT100_any]
+        dirHGT50_any, dirHGT100_any, dirComp]
 # open the relevant file
 path_sets = '../input/genesets_golden/'
 path_dicts = '../input/WS252AnatomyDictionary/'
@@ -41,6 +41,7 @@ path_dicts = '../input/WS252AnatomyDictionary/'
 for d in DIRS:
     if not os.path.exists(d):
         os.makedirs(d)
+
 
 # Make the file that will hold the summaries and make the columns.
 with open(dirSummaries+'ExecutiveSummary.csv', 'w') as fSum:
@@ -111,7 +112,7 @@ df_summary['fracTissues'] = df_summary['TissuesReturned']/df_summary[
     'TissuesTested']
 
 df_summary.sort_values(['NoAnnotations', 'Threshold', 'Method'], inplace=True)
-#==============================================================================
+# ==============================================================================
 # ==============================================================================
 # # Plot summary graphs
 # ==============================================================================
@@ -183,11 +184,10 @@ plt.close()
 
 
 def line_prepender(filename, line):
-    """
-    Given a filename, opens it and prepends the line 'line'
+    """Given a filename, opens it and prepends the line 'line'
     at the beginning of the file
-    """
 
+    """
     with open(filename, 'r+') as f:
         content = f.read()
         f.seek(0, 0)
@@ -261,6 +261,9 @@ def compare(resA, resB, l, r):
     df1 = pd.read_csv(resA, comment='#')
     df2 = pd.read_csv(resB, comment='#')
 
+    # drop observed column from df1
[…]

### 2016-04-06 tests/hgf_benchmarking/doc/tea_main_text.tex
writing!!!

@@
 % \usepackage[utf8]{inputenc} %unicode support
-%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
-%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+% \usepackage[applemac]{inputenc} %applemac support if unicode package fails
+% \usepackage[latin1]{inputenc} %UNIX support if unicode package fails
 \usepackage{lmodern}% http://ctan.org/pkg/lm
@@
 
-\title{Tissue Enrichment Analysis: TEA}
+\title{Tissue Enrichment Analysis (TEA)}
 
@@
 \parttitle{Results} %if any 
-We have developed an  tool for identifying  enrichment of \emph{C.~elegans} tissues among gene sets.
+We have developed a tool for identifying  enrichment of \emph{C.~elegans} tissues among gene sets.
 We applied a hypergeometric model to a pre-existing anatomy ontology for the worm in order to identify enriched tissues and generated a website GUI.
@@
 \parttitle{Conclusions} %if any
-Our Tissue Enrichment Analysis (TEA) can be found at www.WormBase.org/tea, and can be downloaded using Python's standard pip installer.
+Our Tissue Enrichment Analysis (TEA) can be found at \url{www.wormbase.org/tea}, and can be downloaded using Python's standard pip installer.
 It tests a slimmed-down \emph{C.~elegans} tissue ontology for enrichment of specific terms and provides users with a  text and graphic representation of the results.
@@
 One area of biological significance that GO does not include is anatomy.
-One way to address this shortcoming is to use test a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `neuronal cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}.
-The \emph{C.~elegans} database, WormBase~\cite{Harris2014}, maintains a carefully curated list of gene expression data from GFP-reporters.
+One way to address this shortcoming is to test a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `neuronal cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}.
+The \emph{C.~elegans} database, WormBase~\cite{Howe2016}, maintains a curated list of gene expression data from GFP-reporters.
 Here we provide a new framework that analyses a user-input list for enrichment of specific tissues.
 Tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation.
-As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.
+%As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.
 
@@
  To prevent our tool from suffering from verbosity problems, we have filtered our ontology using a small set of well-defined criteria to remove terms that do not contribute additional information.
-To our knowledge, such filtering has never been performed in an algorithmic fashion for a biomedical ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
+To our knowledge, such filtering has not been performed in an algorithmic fashion for a biological ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
 Other pruning tools do exist (see for example~\cite{Kim2007}), but the pruning is query-dependent.
@@
 
-Our tool is available within WormBase and provides users with a text-based file of the enrichment results as well as a simple and clear graph of the results that exhibit the largest fold-change enrichment.
+%Our tool is available within WormBase and provides users with a text-based file of the enrichment results as well as a simple and clear graph of the results that exhibit the largest fold-change enrichment.
 
@@
 As a first step to generate our tissue enrichment software, we wished to select tissue terms that were reasonably well-annotated, yet specific enough to provide insight and not redundant with other terms.
-We also wanted to avoid
-testing tissues at levels where redundancy becomes problematic.
+We also wanted to avoid testing tissues at levels where annotation redundancy becomes problematic.
 For example, several left and right neurons have at least 25 annotated genes and we may want to include them for enrichment testing.
-However, many left/right neuronal
-pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them.
+However, many left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them.
 We reasoned that when two tissues have almost identical annotations, we cannot have statistical confidence in differentiating between them.
@@
 \subsubsection*{Terminal branch terms and parent terms can be safely removed in an algorithmic fashion }
-Another problem arises from the fact that the tissue ontology is scarcely populated at this point in time.
+Another problem arises from the fact that the tissue ontology is still scarcely populated.
 Many nodes have 0-10 annotations, which we consider too few to accurately test.
-To solve this issue, we implemented a
-straightforward trimming algorithm.
+To solve this issue, we implemented a straightforward trimming strategy.
 For a given terminal node, we test whether the node has more than a threshold number of annotations.
@@
 By itself, each of these filters can reduce the number of nodes employed for analysis.
-These filters are not all commutative: while trimming and redundancy filtering are commutative, applying the ceiling filter is not commutative with either the trimming or the redundancy filter.
+These filters are not all commutative: while trimming and redundancy filtering are commutative, applying the ceiling filter is commutative with neither the trimming nor the redundancy filter.
 If the ceiling filter is applied before any other filter, only terminal nodes will remain, since all the parents have complete daughter sets.
@@
 \begin{eqnarray}\label{hypergeometric}
-	\mathrm{P}(n_i | N, m_i, M) = \frac{\dbinom{m_i}{n_i} \dbinom{M-m_i}{N - n_i}}{\dbinom{N}{n_i}}
+	\mathrm{P}(n_i | N, m_i, M) = \frac{\dbinom{m_i}{n_i} \dbinom{M-m_i}{N - n_i}}{\dbinom{N}{n_i}}\text{.}
 \end{eqnarray}
@@
 $N$ is found by taking the user-input list and removing any genes that are not in our annotation dictionary.
-The remaining genes are then associated with their annotation profiles---if a tissue is associated with $s$ tissues, it generates $s$ balls of $s$ colors.
+The remaining genes are associated with their annotation profiles---if a tissue is associated with $s$ tissues, it generates $s$ balls of $s$ colors.
 Our program counts the number of times each tissue appears in the user list, and calculates the probability of having withdrawn as many or more balls for each tissue in the user list.
@@
 
-Once the probability of drawing the labels has been quantified, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm~\cite{Benjamini1995}.
+Once the p-values for each term have been calculated, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm~\cite{Benjamini1995}.
+FDR corrected p-values are called q-values.
 Genes that have a q-value less than a given alpha are considered significant.
@@
 Users input a gene-list using any valid gene name for \emph{C.~elegans}.
-These names are processed into standard WBIDs.
+These names are processed into standard WormBase gene IDs (WBGene IDs).
 The program returns a text-based table showing the tissues that tested significant, along with their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change (observed hits / expected hits).
@@
 Our software is implemented in an easy to use GUI within WormBase (see Figure~\ref{fig:GUIresults}).
-Anatomy terms are displayed in human-readable format followed by their unique WBbt ID.
+Anatomy terms are displayed in human-readable format followed by their unique ontology ID (WBbt ID).
 
 \subsection*{Validation of the algorithm and parameter selection}
-We wanted to select a dictionary which included enough terms to be specific beyond the largest \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change.
-Selection of a dictionary based only on minimization of spurious results would result in a dictionary with a large number of annotations per tissue, and would therefore include only the major tissues.
-On the other hand, selecting a dictionary that can detect smaller tissues will bias us towards tissues with lesser annotations.
-To our knowledge there is no good method for assessing false-positive or false-negative results for
-annotations.
+We wanted to select a dictionary that included enough terms to be specific beyond the largest \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change.
+As expected, larger tissues are correlated with better annotation, so increasing term specificity is associated with losses in statistical power.
+%The phrase below doesn't seem 
+%To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.
 % Gold-standard data sets were annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
-To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues~\cite{Gaudet2004a, Spencer2011,
-Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}.
-These data sets are annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
-Some of these studies have
-been used to annotate gene expression, and so they did not constitute an independent testing set.
+To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}.
+These data sets are annotated gene lists derived from the corresponding Expression Cluster data in WormBase.
+Some of these studies have been used to annotate gene expression, and so they did not constitute an independent testing set.
 To correct this flaw, we built a clean dictionary that specifically excluded all annotation evidence that came from these studies.
 
-As a first attempt to select a good dictionary, we generated all the possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any'
-thresholding criteria for the latter (see Table~\ref{tab:DictionarySpecs}).
-For these dictionaries, the number of tissues tested ranged from 21 to 460.
-The number of tissues was inversely correlated to the minimum annotation, as
-expected, and was largely insensitive to the redundancy threshold, at least in the range we explored (0.9-1).
+As a first attempt to select a good dictionary, we generated all the possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any' similarity thresholding methods (see Table~\ref{tab:DictionarySpecs}).
+The number of tissues was inversely correlated to the minimum annotation cutoff, as expected, and was largely insensitive to the similarity threshold in the range we explored (0.9-1).
 Next, we analyzed all 30 datasets using each dictionary.
-Because of the large number of results, instead of analyzing each set of terms individually, we pooled all results for a given dictionary into histograms.
+Because of the large number of results, instead of analyzing each set of terms individually, we measured the average q-value for significantly enriched terms in each dataset.
 When we analyzed the distribution of significant q-values for the dictionaries, we found that the similarity threshold mattered relatively little for any dictionary.
-We also noticed that the `any' thresholding method resulted in tighter histograms with fa mode closer to 0.
+We also noticed that the `any' thresholding method resulted in tighter histograms with a mode closer to 0.
 For this reason, we chose the `any' method for dictionary generation.
-The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant is the trade-off between power and term
-specificity.
+The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant is the trade-off between power and term specificity.
 Based on these observations, we ruled out the dictionary with the 100 gene annotation cut-off: it had the fewest terms and its q-values were not low enough to compensate the trade-off in specificity.
-
 %dictionaries that work well
@@
 % TODO add WBbt codes!
-Most results were highly comparable and in line with what was expected.
-For some sets, all dictionaries seemed to perform well.
-For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} the results were an amalgamation of neuron related terms including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays regardless of the dictionary used (see~\ref{fig:interagree}).
-On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells ---at the 33-cutoff, our tool identified `Z2' and `Z3' as being five-fold enriched; whereas at the 25 gene-cutoff the terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
+Most results were comparable and in line with expectations.
+For some sets, all dictionaries performed well.
+For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} the results were an amalgamation of neuron-related terms %including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays
+regardless of the dictionary used (see Table~\ref{tab:interagree}).
+On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells---at the 33-cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as being five-fold enriched, and at the 25 gene-cutoff the embryonic germline precursor terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
 %Differences between dictionaries
-We queried an embryonic stage intestine precursor geneset~\cite{Spencer2011}.
+We also queried an embryonic stage intestine precursor geneset~\cite{Spencer2011}.
 Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary.
-However, the 33 cutoff dictionary suggested, probably correctly, that the E lineage was heavily enriched in this set.
+However, the 33 cutoff dictionary identified the E lineage, which is the intestinal precursor lineage in \emph{C.~elegans}, as enriched in this set.
 %Bad stuff.
 Not all queries worked equally well.
-For example, a number of intestinal enriched genes sets~\cite{Spencer2011, Pauli2006} were not enriched in intestine in any dictionary, but they were enriched for pharynx- and hypodermis-related terms.
+For example, a number of intestinal enriched genes sets~\cite{Spencer2011, Pauli2006} were not enriched in intestine related terms in any dictionary, but they were enriched for pharynx- and hypodermis-related terms.
 We were somewhat surprised that intestinal gene sets performed poorly, since the intestine is a relatively well-annotated tissue.
+
 %Dictionary intra-agreement
-We also assessed the internal agreement of our tool by using independent gene-sets that we expected to be enriched in the same tissues.
-We had two independent pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two independent PVD enriched sets~\cite{Spencer2011, Smith2010}; two independent GABAergic gene sets~\cite{Spencer2011, Cinar2005}; two independent pharyngeal gene sets~\cite{Spencer2011, Gaudet2004a}; and two independent intestinal gene sets~\cite{Spencer2011, Pauli2006}.
+We assessed the internal agreement of our tool by using independent gene-sets that we expected to be enriched in the same tissues.
+We used two  pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two PVD enriched sets~\cite{Spencer2011, Smith2010}; two  GABAergic gene sets~\cite{Spencer2011, Cinar2005}; two  pharyngeal gene sets~\cite{Spencer2011, Gaudet2004a}; and two  intestinal gene sets~\cite{Spencer2011, Pauli2006}.
 Overall, the tool seems to have good internal agreement.
-On most sets, the same terms were enriched, although order was somewhat variable (see Table~\ref{tab:intraagree}).
+On most sets, the same terms were enriched, although order was somewhat variable (see Table~\ref{tab:interagree}).
 However, most high-scoring terms were preserved between gene sets.
-The intestinal gene-sets and pharyngeal gene sets comparisons were exceptions, since at least one gene set was missing each for intestine and pharynx in every dictionary, so we didn't consider them as informative for assessing internal agreement.
+%The intestinal gene-sets and pharyngeal gene sets comparisons were exceptions, since at least one gene set showed no enriched terms in all dictionaries, so we did not consider them as informative for assessing internal agreement.
 All comparisons can be found online in our Github repository (see Availability of data and materials).
-Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95
-redundancy threshold using the `any' criterion.
+Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95 redundancy threshold using the `any' criterion.
 seemed to perform well, with a good balance between specificity, verbosity and accuracy, so we selected this parameter set to generate our static dictionary.
@@
 \subsection*{A brief example}
-We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of the biology underlying these datasets.
+We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of their underlying biology.
 Engelmann et al.
@@
 Chikina \emph{et al}~\cite{Chikina2009} report a tissue enrichment model based on a Support Vector Machine classifier that has been trained on microarray studies.
-SVM classifiers are powerful tools capable of great sensitivity, but they require continuous retraining as tissue expression data becomes more available.
-Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for some genome-wide studies.
+SVM classifiers are powerful tools, but they require continuous retraining as more tissue expression data becomes available.
+Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for certain studies.
 Our tool relies on an annotation dictionary that is continuously updated, does not require retraining and does not require ranked genes.
@@
 
-We have tried hard to benchmark our tool well.
-However, our analysis suffers from the drawback that is very hard to benchmark negative controls.
-Even for our set of positive controls, the statistical analysis sometimes throws out unexpected results.
-For example, the embryonic germline precursor gene set had the term `AB' as the most enriched term in the dictionaries with cut off of 25 and 33.
+We have tried hard to benchmark our tool well. However, our analysis suffers from the drawback that is very hard to %identify good negative controls
+  to identify spurious term enrichment.
+% Even for our set of positive controls, the statistical analysis sometimes throws out unexpected results.
+For example, the embryonic germline precursor gene set had the term `AB', which refers to the anterior cell in the 2-cell stage \emph{C.~elegans} embryo, as the most enriched term in the dictionaries with cut off of 25 and 33.
 Is this an error, or does this hint at new biology?
  Although we were unable to determine false-positive and false-negative rates, we do not believe this should deter scientists from using our tool.
-Rather, we encourage researchers to use our tool carefully as a guide, integrating evidence from multiple sources to inform the most likely hypotheses.
-As with any other tool based on statistical sampling, our analysis is most vulnerable to bias in the data collection stage.
-For example, we know that tissue expression reports are negatively biased against germline expression due to the difficulty associated with extrachromosomal array expression in that tissue.
-Support from the community will be crucial in correcting these flaws going forward; indeed, without the community reports of tissue expression this tool would not be possible.
+Rather, we encourage researchers to use our tool as a guide, integrating evidence from multiple sources to inform the most likely hypotheses.
+As with any other tool based on statistical sampling, our analysis is most vulnerable to bias in the data set.
+For example, we know that expression reports are negatively biased against germline expression because of the difficulties associated with expressing DNA in this tissue~\cite{Kelly1997}.
+As time passes, we are certain the accuracy and power of this tool will improve thanks to the efforts of the combined worm community; indeed, without the community reports of tissue expression in the first place, this tool would not be possible.
 
@@
 \subsection*{Fetching annotation terms}
-We used the WormBase gene expression data, which includes annotated descriptions of spatial-temporal expression patterns of genes, to build our dictionary.
-Gene lists per anatomy term were extracted from a solr document store of gene expression data from the WS252 database provided by WormBase (PMID: 26578572).
-We used the solr document store because it provided a convenient access to expression data that included inferred annotations.
+We used WormBase-curated gene expression data, which includes annotated descriptions of spatial-temporal expression patterns of genes, to build our dictionary.
+Gene lists per anatomy term were extracted from a Solr document store of gene expression data from the WS252 database provided by WormBase~\cite{Howe2016}.
+We used the Solr document store because it provided a convenient access to expression data that included inferred annotations.
 That is, for each anatomy term, the expression gene list includes genes that were directly annotated to the term, as well as those that were annotated to the term's descendant terms (if there were any).
@@
 \subsubsection*{Defining a Similarity Metric}
-In order to identify redundant sisters, we defined the following similarity metric:
+To identify redundant sisters, we defined the following similarity metric:
 \begin{eqnarray}\label{similarity def}
@@
 All scripts were written in Python.
-Our software relies on the Pandas, Numpy, Seaborn and SciPy modules to perform all statistical testing and data handling~\cite{McKinney2011, VanDerWalt2011, Oliphant2007}.
+Our software relies on the Pandas, NumPy, Seaborn and SciPy modules to perform all statistical testing and data handling~\cite{McKinney2011, VanDerWalt2011, Oliphant2007}.
 
 \section*{Availability of data and materials}
-Our web implementation is available at https://www.WormBase.org/tea. Our software can also be downloaded using Python's pip installer via the command
+Our web implementation is available at \url{https://www.wormbase.org/tea}. Our software can also be downloaded using Python's pip installer via the command
 
@@
 
-Alternatively, our software is available for download at:
-
-http://dangeles.github.io/TissueEnrichmentAnalysis
+Alternatively, our software is available for download at: \url{http://dangeles.github.io/TissueEnrichmentAnalysis}
 
@@
 % b) uncomment next line
-\nocite{label}
+% \nocite{label}
 
@@
 
-%pan neuronal
-\begin{figure}
-  \begin{subfigure}[b]{0.95\textwidth}
-    \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_33cutoff.png}
-    \caption{}
-    \label{fig:PanNeuronalWatson33}
-  \end{subfigure}
-  %
-  \begin{subfigure}[b]{0.95\textwidth}
-    \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_50cutoff}
-    \caption{}
-    \label{fig:PanNeuronalWatson50}
-  \end{subfigure}
-  \captionsetup{width= 0.95\textwidth}
-  \caption{\csentence{Comparison of Enrichment Results for two dictionaries for a pan-neuronal enriched gene set from Watson, 2008~\cite{Watson2008a}.}
-	Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID.
-   \textbf{a)} Dictionary with cutoff: 33; threshold: 0.95; method: `any'
-   \textbf{b)} Dictionary with cutoff: 50; threshold: 0.95; method: `any'
-   }
-   \label{fig:interagree}
-\end{figure}
+%pan neuronal inter dict agreement
+% \begin{figure}
+%   \begin{subfigure}[b]{0.95\textwidth}
+%     \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_33cutoff.png}
+%     \caption{}
+%     \label{fig:PanNeuronalWatson33}
+%   \end{subfigure}
+%   %
+%   \begin{subfigure}[b]{0.95\textwidth}
+%     \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_50cutoff}
+%     \caption{}
+%     \label{fig:PanNeuronalWatson50}
+%   \end{subfigure}
+%   \captionsetup{width= 0.95\textwidth}
+%   \caption{\csentence{Comparison of Enrichment Results for two dictionaries for a pan-neuronal enriched gene set from Watson, 2008~\cite{Watson2008a}.}
+% 	Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID.
+%    \textbf{a)} Dictionary with cutoff: 33; threshold: 0.95; method: `any'
+%    \textbf{b)} Dictionary with cutoff: 50; threshold: 0.95; method: `any'
+%    }
+%    \label{fig:interagree}
+% \end{figure}
 
@@
 	\captionsetup{width= 0.95\textwidth}
-  	\caption{ \csentence{\emph{Serratia marcescens} enrichment results}
-      Genes altered in   \emph{C.~elegans} after 24hr exposure to \emph{Serratia marcescens}, a bacterium, shows 
-	  down-regulation of chemosensory neurons notably ASER, ASEL, ASJ and ASK. The lateral ganglions are also 
-	  affected. Up-regulated genes are overrepresented for terms involving the excretory duct cell and hypodermis.
-	}  
+  	\caption{ \csentence{\emph{Serratia marcescens} enrichment results}.
+Genes altered in \emph{C.~elegans} after 24hr exposure to \emph{Serratia marcescens}, a bacterium, shows down-regulation of chemosensory neurons notably ASER, ASEL, ASJ and ASK.
+The lateral ganglions are also affected.
+Up-regulated genes are overrepresented for terms involving the excretory duct cell and hypodermis.}
 	% \end{centering}
@@
 \section*{Tables}
+
 \begin{table}[h!]
-	\caption{Comparison of results for a neuronal-enriched geneset from Watson~\cite{Watson2008a}. 
-	We ran the same genelist on a dictionary with a minimum annotation cutoff of 50, similarity threshold of 0.95
-	and similarity method `any' versus another with a minimum annotation cutoff of 33, similarity threshold of 0.95
-	and similarity method `any'. In the table, columns are labeled with their significance value (Q-value) or enrichment
-	fold change followed by a hyphen and a number which indicates which dictionary generated that value. 
-	Not all tissues are present in either dictionary. Hyphens denote not-applicable values, which occurs when a particular
-	tissue is not present in both dictionaries. Full table is available on github
-	}
-	\begin{adjustbox}{width=1\textwidth}
-		\csvautotabular{figures/dict-comparison-50-33.csv}
-		\label{tab:intraagree}
-	\end{adjustbox}
+\caption{Parameter specifications and number of tissues for all dictionaries.
+The `Method' column refers to the trimming criterion for the similarity metric.
+We used two such criteria, `any' and `avg'.`any': For a given sister set, if any sister had a similarity exceeding the corresponding threshold, all sisters were removed from the final dictionary.
+`avg': For a given sister set, if the average similarity across all the sisters in the set was greater than the corresponding threshold, all sisters were removed from the final dictionary.}
+	\csvautotabular{figures/TissueNumbers.csv}
+	\label{tab:DictionarySpecs}
 \end{table}
@@
 \begin{table}[h!]
-\caption{Parameter specifications and number of tissues for all dictionaries.
-The `Method' column refers to the trimming criterion for the similarity metric. We 
-used two such criteria, `any' and `avg'.
-`any': For a given sister set, if any sister had a similarity exceeding the corresponding
-threshold, all sisters were removed from the final dictionary.
-`avg': For a given sister set, if the average similarity across all the sisters in the set
-was greater than the corresponding threshold, all sisters were removed from the final dictionary.}
-	\csvautotabular{figures/TissueNumbers.csv}
-	\label{tab:DictionarySpecs}
+	\caption{Comparison of results for a neuronal-enriched geneset from Watson~\cite{Watson2008a}.
+We ran the same genelist on a dictionary with a minimum annotation cutoff of 50, similarity threshold of 0.95 and similarity method `any' versus another with a minimum annotation cutoff of 33, similarity threshold of 0.95 and similarity method `any'.
+In the table, columns are labeled with their significance value (Q-value) or enrichment fold change followed by a hyphen and a number which indicates which the cutoff for the dictionary that was used for testing.
+Not all tissues are present in either dictionary.
+Hyphens denote not-applicable values, which occurs when a particular tissue is not present in both dictionaries.
+Full table is available on github.}
+	\begin{adjustbox}{width=1\textwidth}
+		\centering
+		\csvautotabular{figures/dict-comparison-50-33.csv}
+		\label{tab:interagree}
+	\end{adjustbox}
 \end{table}

### 2016-04-12 tests/hgf_benchmarking/doc/tea_main_text.tex
Writing

@@
 \parttitle{Results} %if any 
-We have developed a tool for identifying  enrichment of \emph{C.~elegans} tissues among gene sets.
-We applied a hypergeometric model to a pre-existing anatomy ontology for the worm in order to identify enriched tissues and generated a website GUI.
-To cut down on verbosity, we have come up with three straightforward filtering criteria that slim the ontology by almost tenfold.
-One filter removes all terms with fewer than an arbitrary number of annotating genes; the second filter removes all sister terms that have identical gene annotations; and the third filter removes all parents whose daughters all survived the previous two filters.
+We have developed a tool for identifying  enrichment of \emph{C.~elegans} tissues among gene sets and generated a website GUI where users can access this tool.
+Since a common drawback to ontology enrichment analyses is result verbosity, we developed a very simple trimming algorithm to reduce the ontology size by an order of magnitude.
 We adjusted these filters and validated our tool using a set of 30  gold standards from Expression Cluster data in WormBase.
+We used our tool to identify multiple neuronal tissues that are down-regulated due to pathogen infection in \emph{C.~elegans}.
 \parttitle{Conclusions} %if any
@@
 One area of biological significance that GO does not include is anatomy.
-One way to address this shortcoming is to test a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `neuronal cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}.
+One way to address this shortcoming is to use a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `specific cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}.
 The \emph{C.~elegans} database, WormBase~\cite{Howe2016}, maintains a curated list of gene expression data from GFP-reporters.
-Here we provide a new framework that analyses a user-input list for enrichment of specific tissues.
+Here we provide a new framework that analyzes a user-input list for enrichment of specific tissues.
 Tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation.
@@
 
- To prevent our tool from suffering from verbosity problems, we have filtered our ontology using a small set of well-defined criteria to remove terms that do not contribute additional information.
+ To reduce verbosity, we have filtered our ontology using a small set of well-defined criteria to remove terms that do not contribute additional information.
 To our knowledge, such filtering has not been performed in an algorithmic fashion for a biological ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
@@
 As a first step to generate our tissue enrichment software, we wished to select tissue terms that were reasonably well-annotated, yet specific enough to provide insight and not redundant with other terms.
-We also wanted to avoid testing tissues at levels where annotation redundancy becomes problematic.
+%We also wanted to avoid testing tissues at levels where annotation redundancy becomes problematic.
 For example, several left and right neurons have at least 25 annotated genes and we may want to include them for enrichment testing.
-However, many left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them.
-We reasoned that when two tissues have almost identical annotations, we cannot have statistical confidence in differentiating between them.
+For example, since many left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them%.
+We reasoned that when two tissues have almost identical annotations
+, we cannot have statistical confidence in differentiating between them.
 As a result, testing these sister tissues provides no additional information compared with testing only the parent node to these sisters.
-We refer to such sisters as `redundant'.
-To identify redundancy, we defined a similarity metric (see \emph{Methods} section and Figure~\ref{fig:simdiagram}).
-Our similarity metric can be used to identify sisters that have very high similarity between them; alternatively, redundant sisters could be identified if a single sister had a very high similarity score.
+%We refer to such sisters as `redundant'.
+To identify redundancy, we defined a similarity metric (see \emph{Methods} section and Figure~\ref{fig:simdiagram}) that can be used to identify sisters that have very high similarity between them; alternatively, redundant sisters could be identified if a single sister had a very high similarity score.
 We referred to these two scoring criteria as `avg' and `any' respectively.
@@
 \subsubsection*{Terminal branch terms and parent terms can be safely removed in an algorithmic fashion }
-Another problem arises from the fact that the tissue ontology is still scarcely populated.
+Another problem arises from the ontology being scarcely populated.
 Many nodes have 0-10 annotations, which we consider too few to accurately test.
@@
 If it does not, the node is removed.
-The next node in the branch is tested and removed recursively until a node which satisfies the condition is found.
+The next higher node in the branch is tested and removed recursively until a node which satisfies the condition is found.
 At that point, no more nodes can be removed from that branch.
-This is guaranteed by the structure of the ontology: Parent nodes inherit all of the annotations of all of their descendants, so the number of annotated terms monotonically increases with increasing term hierarchy (see Figure~\ref{fig:trim_ends}).
+This is guaranteed by the structure of the ontology: parent nodes inherit all of the annotations of all of their descendants, so the number of annotated terms monotonically increases with increasing term hierarchy (see Figure~\ref{fig:trim_ends}).
 In this way, we ensure that our term dictionary includes only those tissues that are considered sufficiently well annotated for statistical purposes.
 
-Finally, we also wanted to remove as many terms as possible from the dictionary with the goals of reducing covariance between terms, decreasing multiple testing and removing as many non-informative terms as possible.
+Additionally, we reasoned that, for any parent node, if all its daughters were selected for testing, there was no additional benefit to test the parent.
+%In other words, if all the daughter nodes are tested, there is little additional information to be gained by including the parent node.
+To address this issue
+We removed parent nodes from the analysis if all their daughter nodes passed the annotation threshold (see Figure~\ref{fig:trim_roots}).
+We called this a ceiling filter.
+Applying these three filters reduces multiple testing and verbosity.
+
+%Finally, we also wanted to remove as many terms as possible from the dictionary with the goals of reducing covariance between terms and decreasing multiple testing.
 Decreasing covariance between terms is important because we employ a frequentist approach that assumes all terms are independent.
-Large covariation coefficients between some terms means that if one of these tissues tests significant, the other terms are much more likely to pass significance testing as well.
+Covariation  between terms means that if one of these tissues tests significant, the other terms are much more likely to pass significance testing as well.
 This makes adequate correction for false positive rates considerably more difficult.
-Moreover, from a data analysis perspective, we reasoned that, for any parent node, if all its daughters were selected for testing, there was no additional benefit to test the parent.
-In other words, if all the daughter nodes are tested, there is little additional information to be gained by including the parent node.
-To address this issue we removed parent nodes from the analysis if all their daughter nodes passed the annotation threshold (see Figure~\ref{fig:trim_roots}).
-We called this a ceiling filter.
 
 \subsubsection*{Filtering greatly reduces the number of nodes used for analysis}
-By itself, each of these filters can reduce the number of nodes employed for analysis.
-These filters are not all commutative: while trimming and redundancy filtering are commutative, applying the ceiling filter is commutative with neither the trimming nor the redundancy filter.
-If the ceiling filter is applied before any other filter, only terminal nodes will remain, since all the parents have complete daughter sets.
+By itself, each of these filters can reduce the number of nodes employed for analysis, but applying the filters in different orders removes different numbers of nodes (not all the filters are commutative).
+We chose to always execute annotation and similarity thresholding first, followed by the ceiling filter.
+%If the ceiling filter is applied before any other filter, only terminal nodes will remain, since all the parents have complete daughter sets.
 Since terminal nodes are the most poorly annotated, after applying the remaining filters very few nodes will be left behind if any.
@@
 The original ontology has almost 6,000 terms of which 1675 have at least 5 gene annotations.
-After filtering, dictionary sizes ranged from 21 to a maximum of 460 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by an order of magnitude through application of a few simple filters (see Table~\ref{tab:DictionarySpecs}).
+After filtering, dictionary sizes ranged from 21 to a maximum of 460 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by an order of magnitude through the application of a few simple filters (see Table~\ref{tab:DictionarySpecs}).
 These filters were used to compile a static dictionary that we employ for all analyses (see \emph{Validation of the algorithm and parameter selection} section for details).
@@
 Having built a static dictionary, we generated a Python script that implements a significance testing algorithm based on the hypergeometric model.
-Briefly, the hypergeometric model assumes the existence of an urn with a pre-determined number of balls inside it.
-The balls can be painted one of several colors.
-The hypergeometric model provides an answer to the question: If an individual removes $N$ balls, what is the probability of observing $n_i$ balls of color $i$, if the balls are selected without replacement?
+Briefly, the hypergeometric model tests the probability of observing $n_i$ occurences of a tissue $i$ in a list of size $M$ if there are $m_i$ labels for that tissue in a dictionary of total size $N$ that are drawn without replacement.
 Mathematically, this is expressed as:
@@
 
-Here, $n_i$ is the number of balls of type $i$ drawn, $N$ is the total number  of draws, $m_i$ is tissue $i$ and $M$ is the total number of balls in the urn.
-In our specific case, $M_i$ is equal to the total number of annotations in our dictionary.
-$N$ is found by taking the user-input list and removing any genes that are not in our annotation dictionary.
-The remaining genes are associated with their annotation profiles---if a tissue is associated with $s$ tissues, it generates $s$ balls of $s$ colors.
-Our program counts the number of times each tissue appears in the user list, and calculates the probability of having withdrawn as many or more balls for each tissue in the user list.
+Although a user will input gene IDs, we test the number of ocurrences of a term within the gene list, so a single gene can contribute to multiple terms.
 Due to the discrete nature of the hypergeometric distribution, this algorithm can generate artifacts when the list is small.
@@
 These names are processed into standard WormBase gene IDs (WBGene IDs).
-The program returns a text-based table showing the tissues that tested significant, along with their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change (observed hits / expected hits).
+The program returns a table containing all the enriched terms and associated useful information %with significant tissues, their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change ($\text{observed hits} /\text{expected hits}$)
+.
 Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values.
-Our software is implemented in an easy to use GUI within WormBase (see Figure~\ref{fig:GUIresults}).
+Our software is implemented in an easy to use GUI (see Figure~\ref{fig:GUIresults}).
 Anatomy terms are displayed in human-readable format followed by their unique ontology ID (WBbt ID).
@@
 We wanted to select a dictionary that included enough terms to be specific beyond the largest \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change.
-As expected, larger tissues are correlated with better annotation, so increasing term specificity is associated with losses in statistical power.
-%The phrase below doesn't seem 
+Larger tissues are correlated with better annotation, so increasing term specificity is associated with losses in statistical power.
+%The phrase below doesn't seem useful
 %To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.
@@
 
-As a first attempt to select a good dictionary, we generated all the possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any' similarity thresholding methods (see Table~\ref{tab:DictionarySpecs}).
-The number of tissues was inversely correlated to the minimum annotation cutoff, as expected, and was largely insensitive to the similarity threshold in the range we explored (0.9-1).
+As a first attempt to select a dictionary, we generated all possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any' similarity thresholding methods (see Table~\ref{tab:DictionarySpecs}).
+The number of tissues was inversely correlated to the minimum annotation cutoff, and was largely insensitive to the similarity threshold in the range we explored.
 Next, we analyzed all 30 datasets using each dictionary.
@@
 The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant is the trade-off between power and term specificity.
-Based on these observations, we ruled out the dictionary with the 100 gene annotation cut-off: it had the fewest terms and its q-values were not low enough to compensate the trade-off in specificity.
+Based on these observations, we ruled out the dictionary with the 100 gene annotation cut-off: it had the fewest terms and its q-values were not low enough in our opinion to compensate the trade-off in specificity.
+
 %dictionaries that work well
@@
 % TODO add WBbt codes!
-Most results were comparable and in line with expectations.
+Most results were comparable and expected.
 For some sets, all dictionaries performed well.
-For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} the results were an amalgamation of neuron-related terms %including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays
+For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} all terms were neuron-related %including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays
 regardless of the dictionary used (see Table~\ref{tab:interagree}).
-On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells---at the 33-cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as being five-fold enriched, and at the 25 gene-cutoff the embryonic germline precursor terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
+On the other hand, for a set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the 50 cutoff dictionary was only able to identify `oocyte WBbt:006797' which is not a germline precursor, although it is germline related; whereas the two smaller dictionaries singled out actual germline precursor cells---at the 33 cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as enriched, and at the 25 gene cutoff the embryonic germline precursor terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
 %Differences between dictionaries
-We also queried an embryonic stage intestine precursor geneset~\cite{Spencer2011}.
+We also queried an intestine precursor set~\cite{Spencer2011}.
 Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary.
-However, the 33 cutoff dictionary identified the E lineage, which is the intestinal precursor lineage in \emph{C.~elegans}, as enriched in this set.
+However, the 33 cutoff dictionary identified the E lineage, which is the intestinal precursor lineage in \emph{C.~elegans}, as enriched.
+Both of these results capture specific aspects of \emph{C.~elegans} that are well known to developmental biologists.
+
 %Bad stuff.
 Not all queries worked equally well.
-For example, a number of intestinal enriched genes sets~\cite{Spencer2011, Pauli2006} were not enriched in intestine related terms in any dictionary, but they were enriched for pharynx- and hypodermis-related terms.
-We were somewhat surprised that intestinal gene sets performed poorly, since the intestine is a relatively well-annotated tissue.
+For example, a number of intestinal  sets~\cite{Spencer2011, Pauli2006} were not enriched in intestine-related terms in any dictionary, but were enriched for pharynx and hypodermis.
+We were  surprised that intestinal gene sets performed poorly, since the intestine is a relatively well-annotated tissue.
 
 %Dictionary intra-agreement
-We assessed the internal agreement of our tool by using independent gene-sets that we expected to be enriched in the same tissues.
-We used two  pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two PVD enriched sets~\cite{Spencer2011, Smith2010}; two  GABAergic gene sets~\cite{Spencer2011, Cinar2005}; two  pharyngeal gene sets~\cite{Spencer2011, Gaudet2004a}; and two  intestinal gene sets~\cite{Spencer2011, Pauli2006}.
-Overall, the tool seems to have good internal agreement.
-On most sets, the same terms were enriched, although order was somewhat variable (see Table~\ref{tab:interagree}).
-However, most high-scoring terms were preserved between gene sets.
+We assessed the internal agreement of our tool by using independent gene sets that we expected to be enriched in the same tissues.
+We used two  pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two PVD sets~\cite{Spencer2011, Smith2010}; and two  GABAergic sets~\cite{Spencer2011, Cinar2005}.%; two  pharyngeal gene sets~\cite{Spencer2011, Gaudet2004a}; and two  intestinal gene sets~\cite{Spencer2011, Pauli2006}.
+Overall, the tool has good internal agreement.
+On most sets, the same terms were enriched, although order was somewhat variable (see Table~\ref{tab:interagree}), and most high-scoring terms were preserved between sets.
 %The intestinal gene-sets and pharyngeal gene sets comparisons were exceptions, since at least one gene set showed no enriched terms in all dictionaries, so we did not consider them as informative for assessing internal agreement.
 All comparisons can be found online in our Github repository (see Availability of data and materials).
-Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95 redundancy threshold using the `any' criterion.
-seemed to perform well, with a good balance between specificity, verbosity and accuracy, so we selected this parameter set to generate our static dictionary.
+Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95 redundancy threshold using the `any' criterion performed best, with a good balance between specificity, verbosity and accuracy, so we selected this parameter set to generate our static dictionary.
+As of this publication, the testable dictionary contains 261 terms.
 
@@
 We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated.
-Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade.Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct.
-Three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes (for an example, see~\ref{fig:Dcon}\ignore{,~\ref{fig:Hsp},\ref{fig:Smar}}).
-A possible explanation for this might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen.
+Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade.
+Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct.
+Three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes.
+As an independent verification, we also performed GO analysis using PANTHER on the down-regulated genes for \emph{D.~coniospora}.
+These results showed enrichment in terms associated with neurons (see Figure~\ref{fig:Dcon}).
+A possible explanation for this neuronal association might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen.
 Indeed, several studies~\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.~elegans} uses chemosensory neurons to identify pathogens.
-One bacterium did not exhibit the same pattern of down-regulation of neuronal-associated genes.
+One bacterium did not ellicit the same pattern of down-regulation of neuronal-associated genes.
 \emph{E.~faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E.~faecalis} may have a different pathogenic profile.
@@
 \section*{Discussion}
+%TEA Discussion
 We have presented a tissue enrichment analysis tool that employs a standard hypergeometric model to test the \emph{C.~elegans} tissue ontology.
+We use a hypergeometric function to test a user-provided gene list for enrichment of anatomical terms in \emph{C.~elegans}.
+Our hope is that the physiological relevance of anatomical terms will enable researchers to make hypotheses about high-dimensionality data.
+Specifically, we believe an enriched term may broadly suggest one of two hypotheses: if a list is enriched in a particular anatomical region, that anatomical region is affected by the experimental treatment; alternatively, the anatomical regions that are enriched reflect biologically relevant interactions between tissues.
+We believe the first hypothesis is a reasonable hypothesis to make in the case of whole-worm RNA-seq data for example, whereas the second hypothesis may be more plausible in cases where a researcher already knows what tissues a particular gene list came from, as may be the case in single-cell RNA-seq.
+
+%Other tools
+Our tool relies on an annotation dictionary that is continuously updated, does not require retraining and does not require ranked genes.
+To our knowledge, this is the first tool that tests tissue enrichment  in \emph{C.~elegans} via the hypergeometric method, but similar projects exist for humans and zebrafish~\cite{Lee2013, Prykhozhij2013}, highlighting the relevance of our tool for high-dimensionality biology.
+Chikina \emph{et al}~\cite{Chikina2009} have previously reported a tissue enrichment model for \emph{C.~elegans }based on a Support Vector Machine classifier that has been trained on microarray studies.
+SVMs are powerful tools, but they require continuous retraining as more tissue expression data becomes available.
+Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for certain studies.
+Moreover, this tissue enrichment tool generally provides users with enrichment results for 6 large tissues.
+In contrast, our tool routinely tests a much larger number of terms, and we have shown it can even accurately identify enrichment of embryonic precursor lineages for select data sets.
+
+%Trimming discussion
 We have also presented the first, to our knowledge, ontology trimming algorithm for biomedical ontologies.
@@
 
-Chikina \emph{et al}~\cite{Chikina2009} report a tissue enrichment model based on a Support Vector Machine classifier that has been trained on microarray studies.
-SVM classifiers are powerful tools, but they require continuous retraining as more tissue expression data becomes available.
-Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for certain studies.
-Our tool relies on an annotation dictionary that is continuously updated, does not require retraining and does not require ranked genes.
-To our knowledge, there are no other tissue ontology enrichment tools in \emph{C.~elegans}, but similar projects exist for humans and zebrafish~\cite{Lee2013, Prykhozhij2013}, highlighting the relevance of our tool for high-dimensionality biology.
-
-We have tried hard to benchmark our tool well. However, our analysis suffers from the drawback that is very hard to %identify good negative controls
-  to identify spurious term enrichment.
+We have endeavoured to benchmark our tool well. However, our analysis could not identify %identify good negative controls
+ spurious term enrichment.
 % Even for our set of positive controls, the statistical analysis sometimes throws out unexpected results.
@@
 As with any other tool based on statistical sampling, our analysis is most vulnerable to bias in the data set.
-For example, we know that expression reports are negatively biased against germline expression because of the difficulties associated with expressing DNA in this tissue~\cite{Kelly1997}.
+For example, expression reports are negatively biased against germline expression because of the difficulties associated with expressing DNA in this tissue~\cite{Kelly1997}.
 As time passes, we are certain the accuracy and power of this tool will improve thanks to the efforts of the combined worm community; indeed, without the community reports of tissue expression in the first place, this tool would not be possible.
@@
 \subsubsection*{Implementation}
-All scripts were written in Python.
-Our software relies on the Pandas, NumPy, Seaborn and SciPy modules to perform all statistical testing and data handling~\cite{McKinney2011, VanDerWalt2011, Oliphant2007}.
+All scripts were written in Python 3.5.
+Our software relies on the pandas, NumPy, Seaborn and SciPy modules to perform all statistical testing and data handling~\cite{McKinney2011, VanDerWalt2011, Oliphant2007}.
 
@@
 %gui
-
 \begin{figure}
@@
   \caption{\csentence{Kernel density estimates for 30 gold standard datasets.}
-      We ran TEA on 30 datasets we believed to be enriched in particulae tissues and pooled all the results to observe the distribution of q-values.
+      We ran TEA on 30 datasets we believed to be enriched in particular tissues and pooled all the results to observe the distribution of q-values.
 The mode of the distribution for dictionaries with annotation cut-offs of 100 and 50 genes are very similar; however, when the cut-off is lowered to 25 genes, the mode of the distribution shifts to the left, potentially signalling a decrease in measurement power.}
@@
 %agreements
+% \begin{figure}
+%   \begin{subfigure}[b]{0.95\textwidth}
+%     \includegraphics[width=0.95\textwidth]{WBPaper00024970_GABAergic_neuron_specific_WBbt_0005190_247_33cutoff.png}
+%     \caption{}
+%     \label{fig:agreement1}
+%   \end{subfigure}
+%   %
+%   \begin{subfigure}[b]{0.95\textwidth} 	\includegraphics[width=0.95\textwidth]{WBPaper00037950_GABAergic-motor-neurons_larva_enriched_WBbt_0005190_132_33cutoff.png}
+%     \caption{}
+%     \label{fig:agreement2}
+%   \end{subfigure}
+%   \captionsetup{width= 0.95\textwidth}
+%   \caption{\csentence{Independently derived genesets show similar results within a given dictionary}
+%   Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID. Dictionary with cutoff: 33; threshold: 0.95; method: `any'.
+%   \textbf{a)} GABAergic gene set from Watson~\cite{Watson2008a}.
+%   \textbf{b)} GABAergic gene set from Spencer~\cite{Spencer2011}.
+%    }
+%    \label{fig:intragree}
+% \end{figure}
+
+%gaba
 \begin{figure}
-  \begin{subfigure}[b]{0.95\textwidth}
-    \includegraphics[width=0.95\textwidth]{WBPaper00024970_GABAergic_neuron_specific_WBbt_0005190_247_33cutoff.png}
-    \caption{}
-    \label{fig:agreement1}
-  \end{subfigure}
-  %
-  \begin{subfigure}[b]{0.95\textwidth} 	\includegraphics[width=0.95\textwidth]{WBPaper00037950_GABAergic-motor-neurons_larva_enriched_WBbt_0005190_132_33cutoff.png}
-    \caption{}
-    \label{fig:agreement2}
-  \end{subfigure}
+  \includegraphics[width=0.95\textwidth]{GABAcomparison2.png}
   \captionsetup{width= 0.95\textwidth}
-  \caption{\csentence{Independently derived genesets show similar results within a given dictionary}
-  Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID. Dictionary with cutoff: 33; threshold: 0.95; method: `any'.
-  \textbf{a)} GABAergic gene set from Watson~\cite{Watson2008a}.
-  \textbf{b)} GABAergic gene set from Spencer~\cite{Spencer2011}.
+  \caption{\csentence{Independently derived gene sets show similar results 	within a given dictionary.}
+  \textbf{Set 1)} GABAergic gene set from Watson~\cite{Watson2008a}.
+  \textbf{Set 2)} GABAergic gene set from Spencer~\cite{Spencer2011}.
+  Arrowheads highlight identical terms between both analyses. All terms refer to neurons or neuronal tissues and are GABA-associated. Dictionary with cutoff: 33; threshold: 0.95; method: `any'.
    }
@@
 \begin{figure}[h!]
-    \includegraphics[width=0.95\textwidth]{figures/DrechmeriaconiosporaEnrichment.png}
+    \includegraphics[width=0.95\textwidth]{figures/EngelmannGO-TEA.png}
 	\captionsetup{width= 0.95\textwidth}
-  	\caption{\csentence{\emph{D. coniospora} enrichment results}
-      Genes altered in   \emph{C.~elegans} after 24hr exposure to \emph{D. coniospora}, a fungus, show enrichment
-	  in many neuronal tissues.}
+  	\caption{\csentence{\emph{D. coniospora} Gene Enrichment Analysis and Tissue Enrichment Analysis results.}
+	We compared and contrasted the results from a gene enrichment analysis program, pantherDB, with TEA by analyzing genes that were significantly down-regulated when \emph{C.~elegans} was exposed to \emph{D.
+coniospora} in a previously published dataset by Engelmann \emph{et al}~\cite{Engelmann2011} with both tools.
+	\textbf{a)} pantherDB raw results in a tabular format, sorted by p-value. 
+	\textbf{b)} TEA raw results in a histogram format, sorted by q-value (lowest on top). 
+	Both pantherDB and TEA identify terms associated with neurons (red bars). The two analyses provide complementary, not redundant, information.}
 	\label{fig:Dcon}
@@
 
-\begin{figure}[h!]
-	% \begin{centering}
-    \includegraphics[width=0.95\textwidth]{figures/SerratiamarcescensEnrichment.png}
-	\captionsetup{width= 0.95\textwidth}
-  	\caption{ \csentence{\emph{Serratia marcescens} enrichment results}.
+% \begin{figure}[h!]
+% 	% \begin{centering}
+%     \includegraphics[width=0.95\textwidth]{figures/SerratiamarcescensEnrichment.png}
+% 	\captionsetup{width= 0.95\textwidth}
+%   	\caption{ \csentence{\emph{Serratia marcescens} enrichment results}.
 Genes altered in \emph{C.~elegans} after 24hr exposure to \emph{Serratia marcescens}, a bacterium, shows down-regulation of chemosensory neurons notably ASER, ASEL, ASJ and ASK.
@@
 Up-regulated genes are overrepresented for terms involving the excretory duct cell and hypodermis.}
-	% \end{centering}
-	\label{fig:Smar}
-\end{figure}
+% 	% \end{centering}
+% 	\label{fig:Smar}
+% \end{figure}
 %

### 2016-04-13 tissue_enrichment_analysis/hypergeometricTests.py
Fixed problems for commandline excecution.

@@ -389,7 +389,7 @@ if __name__ == '__main__':
 
     #optional args
     if args.dictionary:
-        tdf_name= args.tissue_dictionary
+        tdf_name= args.dictionary
         tissue_df= pd.read_csv(tdf_name)
     else:
         tissue_df= fetch_dictionary()

### 2016-04-14 tests/hgf_benchmarking/doc/tea_main_text.tex
Changed figure format from pdf to png
Version change from 13.9 -> 13.10
Figures are now saved as .pdf

Some files reformatted to be written in PEP8.

Writing.

@@
 Indeed, several studies~\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.~elegans} uses chemosensory neurons to identify pathogens.
-One bacterium did not ellicit the same pattern of down-regulation of neuronal-associated genes.
+%One bacterium did not ellicit the same pattern of down-regulation of neuronal-associated genes.
 \emph{E.~faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E.~faecalis} may have a different pathogenic profile.
@@
 \begin{figure}[h!]
-	\includegraphics[width=\textwidth]{avgQKDE_method=any.png}
+	\includegraphics[width=\textwidth]{avgQKDE_method=any.pdf}
   \captionsetup{width= 0.95\textwidth}
@@
 \begin{figure}
-  \includegraphics[width=0.95\textwidth]{GABAcomparison2.png}
+  \includegraphics[width=0.95\textwidth]{GABAcomparison3.pdf}
   \captionsetup{width= 0.95\textwidth}
-  \caption{\csentence{Independently derived gene sets show similar results 	within a given dictionary.}
+  \caption{\csentence{Independently derived gene sets show similar results 	when tested with the same dictionary.}
   \textbf{Set 1)} GABAergic gene set from Watson~\cite{Watson2008a}.
@@
 \begin{figure}[h!]
-    \includegraphics[width=0.95\textwidth]{figures/EngelmannGO-TEA.png}
+    \includegraphics[width=0.95\textwidth]{figures/engelmann-GO-tea-comparison.png}
 	\captionsetup{width= 0.95\textwidth}
-  	\caption{\csentence{\emph{D. coniospora} Gene Enrichment Analysis and Tissue Enrichment Analysis results.}
-	We compared and contrasted the results from a gene enrichment analysis program, pantherDB, with TEA by analyzing genes that were significantly down-regulated when \emph{C.~elegans} was exposed to \emph{D.
-coniospora} in a previously published dataset by Engelmann \emph{et al}~\cite{Engelmann2011} with both tools.
-	\textbf{a)} pantherDB raw results in a tabular format, sorted by p-value. 
-	\textbf{b)} TEA raw results in a histogram format, sorted by q-value (lowest on top). 
-	Both pantherDB and TEA identify terms associated with neurons (red bars). The two analyses provide complementary, not redundant, information.}
+  	\caption{\csentence{\emph{D.~coniospora} Gene Enrichment Analysis and Tissue Enrichment Analysis results.}
+	We compared and contrasted the results from a gene enrichment analysis program, pantherDB, with TEA by analyzing genes that were significantly down-regulated when \emph{C.~elegans} was exposed to \emph{D.~coniospora} in a previously published dataset by Engelmann \emph{et al}~\cite{Engelmann2011} with both tools.
+	\textbf{a)} pantherDB screenshot of results, sorted by p-value. Only top hits shown.
+	\textbf{b)} TEA results, sorted by q-value (lowest on top). 
+	Both pantherDB and TEA identify terms associated with neurons (red square). The two analyses provide complementary, not redundant, information.}
 	\label{fig:Dcon}
@@
 %       Genes altered in   \emph{C.~elegans} after 24hr exposure to \emph{Harposporium sp.} a fungus, shows enrichment
-% 	  in the posterior lateral ganglion, similarly to \emph{D. coniospora}. This particular fungus seems to affect
+% 	  in the posterior lateral ganglion, similarly to \emph{D.~coniospora}. This particular fungus seems to affect
 % 	  genes normally associated with neuronal precursor tissues as well.

### 2016-04-14 build/lib/tissue_enrichment_analysis/__init__.py

@@ -17,7 +17,7 @@ __author__ = "David Angeles-Albores"
 __copyright__ = "Copyright 2016, WormBase"
 __credits__ = ["David Angeles-Albores", "Raymond Y. Lee", "Juancarlos Chan"]
 __license__ = "MIT"
-__version__ = "0.13.9"
+__version__ = "0.13.10"
 __maintainer__ = "David Angeles-Albores"
 __email__ = "dangeles@caltech.edu"
 __status__ = "Production"

### 2016-04-14 build/lib/tissue_enrichment_analysis/hypergeometricTests.py

@@ -1,4 +1,4 @@
- # -*- coding: utf-8 -*-
+# -*- coding: utf-8 -*-
 """
 Author: David Angeles
 Date: May 26, 2015
@@ -17,275 +17,265 @@ import sys
 from urllib.request import urlopen
 import contextlib
 
-#if __name__ != '__main__':
-#    import matplotlib.pyplot as plt
-#    import seaborn as sns
-#    sns.set_context('paper')
-#    sns.set_style('whitegrid')
-#    sns.despine(left= True)
-#    sns.despine(trim= True)
-
-
-
 
 def pass_list(user_provided, tissue_dictionary):
     """
-    A function to check which genes are in the 
-    provided list of user provided names and to
-    return a dataframe of presence or fail
-    
+    A function to check which genes in provided list are in the dictionary.
+
     Takes two vectors, user_provided and tissue_dictionary in that order
     tissue_dictionary is a pandas dataframe that you should know about
     user_provided is a list of gene names
     """
-    length= tissue_dictionary.shape[0] #how many genes in the dict
-    
-    #make an empty dataframe
-    present= pd.DataFrame(index= range(length), columns= ['wbid','provided'])
-    
-    #fill wbid column with all the gene names
-    present.wbid= tissue_dictionary.wbid
-    
-    #go through and pass attendance -- 1 if present, 0 otherwise
+    length = tissue_dictionary.shape[0]  # how many genes in the dict
+
+    # make an empty dataframe
+    present = pd.DataFrame(index=range(length), columns=['wbid', 'provided'])
+
+    # fill wbid column with all the gene names
+    present.wbid = tissue_dictionary.wbid
+
+    # go through and pass attendance -- 1 if present, 0 otherwise
     for item in user_provided:
-        present.provided[present.wbid==item]= 1
-        
-    #remove NA's and make them 0
-    present.provided= present.provided.fillna(0)
-    
-    #return df
+        present.provided[present.wbid == item] = 1
+
+    # remove NA's and make them 0
+    present.provided = present.provided.fillna(0)
+
+    # return df
     return present
-    
-#==============================================================================
-#hgf is short for hypergeometric function
-#==============================================================================
+
+# ==============================================================================
+# hgf is short for hypergeometric function
+# ==============================================================================
+
+
 def hgf(gene_list, tissue_df):
     """
     Given a list of tissues and a gene-tissue dictionary,
     returns a p-dictionary for the enrichment of every tissue
     (a p-dictionary is a vector of length equal to the number
[…]

### 2016-04-14 setup.py

@@ -10,28 +10,28 @@ from setuptools import setup, find_packages
 import os
 import sys
 
-version= '0.13.9'
+version = '0.13.10'
 
-#just type in python setup.py publish and this takes care of publishing to to pypi!
+# just type in python setup.py publish and this takes care of publishing to pypi!
 
 
-#tag with git
+# tag with git
 if sys.argv[1] == 'tag':
-    
-    if len(sys.argv) > 2:   
+
+    if len(sys.argv) > 2:
         if sys.argv[2] == '-m':
             print(sys.argv[2])
             print('please input your commit message')
-            message= sys.stdin.readline().strip()
+            message = sys.stdin.readline().strip()
             print("git tag -a %s -m '%s'" % (version, message))
             os.system("git tag -a %s -m '%s'" % (version, message))
-    else:    
+    else:
         os.system("git tag -a %s -m 'version %s'" % (version, version))
-        
+
     os.system("git push --tags")
     sys.exit()
 
-#publish to pypi
+# publish to pypi
 if sys.argv[-1] == 'publish':
     os.system("python setup.py sdist upload")
     os.system("python setup.py bdist_wheel upload")
@@ -41,7 +41,7 @@ if sys.argv[-1] == 'publish':
     sys.exit()
 
 
-#for pytest.py purposes -- just type in python setup.py test!
+# for pytest.py purposes -- just type in python setup.py test!
 if sys.argv[-1] == 'test':
     test_requirements = [
         'pytest',
@@ -59,30 +59,30 @@ if sys.argv[-1] == 'test':
 
 # Below this point is the rest of the setup() function
 
+
 def readme():
+    """A function to open a readme.rst file."""
     with open('README.rst') as f:
         return f.read()
 
 setup(
-  name = 'tissue_enrichment_analysis',
-#  packages = ['tissue_enrichment_analysis'], # this must be the same as the name above
+  name='tissue_enrichment_analysis',
   packages=find_packages(exclude=("tests",)),
-  version = version,
-  description = 'This package contains all the software used to implement\
+  version=version,
+  description='This package contains all the software used to implement\
   TEA in WormBase and remotely',
-  author = 'David Angeles-Albores',
-  author_email = 'dangeles@caltech.edu',
-  url = 'https://github.com/dangeles/TissueEnrichmentAnalysis', # use the URL to the github repo
[…]

### 2016-04-14 tests/hgf_benchmarking/src/Engelmann_Reanalysis.py

@@ -7,266 +7,155 @@ and Fungal Infection in C elegans.
 
 @author: dangeles
 """
-import tissue_enrichment_analysis as tea #the library to be used
+import tissue_enrichment_analysis as tea  # the library to be used
 import pandas as pd
 import numpy as np
 import seaborn as sns
 import matplotlib.pyplot as plt
 import os
 
+# tissue dictionary to use
+tissue_df = pd.read_csv('../input/final_cutoff33_threshold0.95_methodany.csv')
 
-#tissue dictionary to use
-tissue_df= pd.read_csv('../input/final_cutoff33_threshold0.95_methodany.csv')
-
-#rna-seq datasets to reanalyze
+# rna-seq datasets to reanalyze
 #Bacterial
-dfLum= pd.read_csv('../input/Engelmann/luminescens_Engelmann_2011.csv')
-dfMarc= pd.read_csv('../input/Engelmann/marcescens_Engelmann_2011.csv')
-dfFaec= pd.read_csv('../input/Engelmann/faecalis_Engelmann_2011.csv')
+dfLum = pd.read_csv('../input/Engelmann/luminescens_Engelmann_2011.csv')
+dfMarc = pd.read_csv('../input/Engelmann/marcescens_Engelmann_2011.csv')
+dfFaec = pd.read_csv('../input/Engelmann/faecalis_Engelmann_2011.csv')
 
-#Fungal
-dfHarp= pd.read_csv('../input/Engelmann/harposporium_Engelmann_2011.csv')
-dfCon= pd.read_csv('../input/Engelmann/coniospora_Engelmann_2011.csv')
+# Fungal
+dfHarp = pd.read_csv('../input/Engelmann/harposporium_Engelmann_2011.csv')
+dfCon = pd.read_csv('../input/Engelmann/coniospora_Engelmann_2011.csv')
+# dfSin= pd.read_csv('../input/Engelmann/sinha_2012.csv')
 
+# dictionary of gene names
+names = pd.read_csv('../input/Engelmann/c_elegans.PRJNA13758.WS241.livegeneIDs.unmaprm.txt',
+                    sep='\t', comment='#')
 
-#dfSin= pd.read_csv('../input/Engelmann/sinha_2012.csv')
+# make the directories to place the analyses
+dirEngelmann = '../output/Engelmann/'
+dirAnalysis = '../output/Engelmann/Analysis'
+dirGraphs = '../output/Engelmann/Graphs'
[…]

### 2016-04-14 tests/hgf_benchmarking/src/hgf_benchmark_script.py

@@ -21,16 +21,16 @@ sns.set_context('paper')
 pd.set_option('precision', 3)
 # this script generates a few directories.
 
+
 dirOutput = '../output/'
 dirSummaries = '../output/SummaryInformation/'
 dirHGT25_any = '../output/HGT25_any_Results/'
-dirHGT33_avg = '../output/HGT33_any_Results/'
 dirHGT33_any = '../output/HGT33_any_Results/'
 dirHGT50_any = '../output/HGT50_any_Results/'
 dirHGT100_any = '../output/HGT100_any_Results/'
 dirComp = '../output/comparisons/'
 
-DIRS = [dirOutput, dirSummaries, dirHGT25_any, dirHGT33_avg,
+DIRS = [dirOutput, dirSummaries, dirHGT25_any,
         dirHGT50_any, dirHGT100_any, dirComp]
 # open the relevant file
 path_sets = '../input/genesets_golden/'
@@ -153,7 +153,7 @@ plt.xlabel('Fraction of all tissues that tested significant')
 plt.xlim(0, 1)
 plt.title('KDE Curves for all dictionaries, benchmarked on all gold standards')
 plt.legend()
-plt.savefig(dirSummaries+'fractissuesKDE_method=any.png')
+plt.savefig(dirSummaries+'fractissuesKDE_method=any.pdf')
 plt.close()
 
 resplot('AvgQ', method='avg')
@@ -161,7 +161,7 @@ plt.xlabel('Fraction of all tissues that tested significant')
 plt.xlim(0, 0.05)
 plt.title('KDE Curves for all dictionaries, benchmarked on all gold standards')
 plt.legend()
-plt.savefig(dirSummaries+'avgQKDE_method=avg.png')
+plt.savefig(dirSummaries+'avgQKDE_method=avg.pdf')
 plt.close()
 
 
@@ -170,7 +170,7 @@ plt.xlabel('AvgQ value')
 plt.xlim(0, .05)
 plt.title('KDE Curves for all dictionaries, benchmarked on all gold standards')
 plt.legend()
-plt.savefig(dirSummaries+'avgQKDE_method=any.png')
+plt.savefig(dirSummaries+'avgQKDE_method=any.pdf')
 plt.close()
 
 # KDE of the fraction of avgFold
@@ -179,15 +179,12 @@ plt.xlabel('Avg Fold Change value')
 plt.xlim(0, 15)
 plt.title('KDE Curves for all dictionaries, benchmarked on all gold standards')
[…]

### 2016-04-14 tissue_enrichment_analysis/hypergeometricTests.py

@@ -1,4 +1,4 @@
- # -*- coding: utf-8 -*-
+# -*- coding: utf-8 -*-
 """
 Author: David Angeles
 Date: May 26, 2015
@@ -17,165 +17,151 @@ import sys
 from urllib.request import urlopen
 import contextlib
 
-#if __name__ != '__main__':
-#    import matplotlib.pyplot as plt
-#    import seaborn as sns
-#    sns.set_context('paper')
-#    sns.set_style('whitegrid')
-#    sns.despine(left= True)
-#    sns.despine(trim= True)
-
-
-
 
 def pass_list(user_provided, tissue_dictionary):
     """
-    A function to check which genes are in the
-    provided list of user provided names and to
-    return a dataframe of presence or fail
+    A function to check which genes in provided list are in the dictionary.
 
     Takes two vectors, user_provided and tissue_dictionary in that order
     tissue_dictionary is a pandas dataframe that you should know about
     user_provided is a list of gene names
     """
-    length= tissue_dictionary.shape[0] #how many genes in the dict
+    length = tissue_dictionary.shape[0]  # how many genes in the dict
 
-    #make an empty dataframe
-    present= pd.DataFrame(index= range(length), columns= ['wbid','provided'])
+    # make an empty dataframe
+    present = pd.DataFrame(index=range(length), columns=['wbid', 'provided'])
 
-    #fill wbid column with all the gene names
-    present.wbid= tissue_dictionary.wbid
+    # fill wbid column with all the gene names
+    present.wbid = tissue_dictionary.wbid
 
-    #go through and pass attendance -- 1 if present, 0 otherwise
+    # go through and pass attendance -- 1 if present, 0 otherwise
     for item in user_provided:
-        present.provided[present.wbid==item]= 1
+        present.provided[present.wbid == item] = 1
 
-    #remove NA's and make them 0
-    present.provided= present.provided.fillna(0)
+    # remove NA's and make them 0
+    present.provided = present.provided.fillna(0)
 
-    #return df
+    # return df
     return present
 
-#==============================================================================
-#hgf is short for hypergeometric function
-#==============================================================================
+# ==============================================================================
+# hgf is short for hypergeometric function
+# ==============================================================================
+
+
 def hgf(gene_list, tissue_df):
     """
     Given a list of tissues and a gene-tissue dictionary,
     returns a p-dictionary for the enrichment of every tissue
     (a p-dictionary is a vector of length equal to the number
-    of tissues in the tissue_dictionary, sorted by value)
[…]

### 2016-04-18 tests/hgf_benchmarking/doc/tea_main_text.tex
Minor changes
Plots are now saved as svg with a dpi=1200

@@
    % email={raymond@caltech.edu}   % email address
-]{\inits{RYL}\fnm{Raymond Y} \snm{Lee}}
+]{\inits{RYL}\fnm{Raymond YN} \snm{Lee}}
 \author[
@@
 These genes are often correlated in their biological characteristics or functions, but identifying these functions remains challenging.
-To interpret these long lists of genes, biologists need to abstract genes into fewer terms that are biologically relevant to form hypotheses about what is happening in the data.
+To interpret these long lists of genes, biologists need to abstract genes into concepts that are biologically relevant to form hypotheses about what is happening in the system.
 One such abstraction method relies on Gene Ontology (GO).
-GO provides a controlled set of hierarchically ordered terms~\cite{TheGeneOntologyConsortium2000a, TheGeneOntologyConsortium2015} that provide detailed information about the molecular, cellular or biochemical functions of any gene.
+GO provides a controlled set of hierarchically ordered terms~\cite{TheGeneOntologyConsortium2000a, TheGeneOntologyConsortium2015} that provide detailed descriptions about the molecular, cellular or biochemical functions of any gene.
 For a given gene list, certain software programs can query whether a particular term is enriched~\cite{Mi2009, McLean2010, Huang2009}.
@@
 One way to address this shortcoming is to use a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `specific cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}.
-The \emph{C.~elegans} database, WormBase~\cite{Howe2016}, maintains a curated list of gene expression data from GFP-reporters.
-Here we provide a new framework that analyzes a user-input list for enrichment of specific tissues.
-Tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation.
+The \emph{C.~elegans} database, WormBase~\cite{Howe2016}, maintains a curated list of gene expression data from the literature.
+Here we provide a new framework that analyzes a user-input list for enrichment of specific cells and tissues.
+They are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation.
 %As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.
@@
 Thomas, pers.
-comm.
-).
+comm.).
 
@@
 \section*{Results}
-\subsection*{Generating a Useful Dictionary}
+\subsection*{Generating a Gene-Tissue Dictionary}
 \subsubsection*{Reducing term redundancy through a similarity metric}
@@
 For example, several left and right neurons have at least 25 annotated genes and we may want to include them for enrichment testing.
-For example, since many left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them%.
+For example, nematodes have a number of neurons that are placed symmetrically along the left/right body axis, and are functionally similar.
+These left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them%.
 We reasoned that when two tissues have almost identical annotations
-, we cannot have statistical confidence in differentiating between them.
+, and therefore we cannot have statistical confidence in differentiating between them.
 As a result, testing these sister tissues provides no additional information compared with testing only the parent node to these sisters.
@@
 
-\subsubsection*{Terminal branch terms and parent terms can be safely removed in an algorithmic fashion }
+\subsubsection*{Terminal branch terms and parent terms can be
+safely removed in an algorithmic fashion }
 Another problem arises from the ontology being scarcely populated.
@@
 We called this a ceiling filter.
-Applying these three filters reduces multiple testing and verbosity.
+Applying these three filters reduced the number of ontology terms by an order of magnitude.
 
@@
 Mathematically, this is expressed as:
+
 \begin{eqnarray}\label{hypergeometric}
@@
 
-Users input a gene-list using any valid gene name for \emph{C.~elegans}.
+Users input a gene list using any valid gene name for \emph{C.~elegans}.
 These names are processed into standard WormBase gene IDs (WBGene IDs).
-The program returns a table containing all the enriched terms and associated useful information %with significant tissues, their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change ($\text{observed hits} /\text{expected hits}$)
+The program returns a table containing all the enriched terms and associated useful information%with significant tissues, their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change ($\text{observed hits} /\text{expected hits}$)
 .
 Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values.
+The bars in the graph are sorted in ascending order of q-value and then in descending order of fold-change.
+Bars are colored for ease of viewing, and color does not convey information.
 Our software is implemented in an easy to use GUI (see Figure~\ref{fig:GUIresults}).
 Anatomy terms are displayed in human-readable format followed by their unique ontology ID (WBbt ID).
+In summary, each time the ontology annotations are updated, a new trimmed ontology is generated using our filters; in parallel, users can submit their gene lists through WormBase for testing, with results output in a number of formats (see Figure~\ref{fig:workflow}).
 
-\subsection*{Validation of the algorithm and parameter selection}
+\subsection*{Validation of the algorithm and optimizing parameter selection}
 We wanted to select a dictionary that included enough terms to be specific beyond the largest \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change.
@@
 Next, we analyzed all 30 datasets using each dictionary.
-Because of the large number of results, instead of analyzing each set of terms individually, we measured the average q-value for significantly enriched terms in each dataset.
-When we analyzed the distribution of significant q-values for the dictionaries, we found that the similarity threshold mattered relatively little for any dictionary.
+Because of the large number of results, instead of analyzing each set of terms individually, we measured the average q-value for significantly enriched terms in each dataset without regard for the perceived accuracy of the terms that tested significant.
+We found that the similarity threshold mattered relatively little for any dictionary.
 We also noticed that the `any' thresholding method resulted in tighter histograms with a mode closer to 0.
@@
 regardless of the dictionary used (see Table~\ref{tab:interagree}).
-On the other hand, for a set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the 50 cutoff dictionary was only able to identify `oocyte WBbt:006797' which is not a germline precursor, although it is germline related; whereas the two smaller dictionaries singled out actual germline precursor cells---at the 33 cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as enriched, and at the 25 gene cutoff the embryonic germline precursor terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
+On the other hand, for a set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the 50 cutoff dictionary was only able to identify `oocyte WBbt:006797' which is not a germline precursor, although it is germline related; whereas the two smaller dictionaries singled out actual germline precursor cells---at the 33 cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as enriched, and at the 25 gene cutoff the embryonic germline precursor terms `P$_4$',`P$_3$' and `P$_2$' were identified in addition to `Z2' and `Z3'.
 %Differences between dictionaries
@@
 
-\subsection*{A brief example}
+\section*{Applying the tool}
 We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of their underlying biology.
@@
 %Other tools
-Our tool relies on an annotation dictionary that is continuously updated, does not require retraining and does not require ranked genes.
+Our tool relies on an annotation dictionary that is continuously updated primarily with data from single gene qualitative analyses, does not require retraining and does not require ranked genes.
 To our knowledge, this is the first tool that tests tissue enrichment  in \emph{C.~elegans} via the hypergeometric method, but similar projects exist for humans and zebrafish~\cite{Lee2013, Prykhozhij2013}, highlighting the relevance of our tool for high-dimensionality biology.
@@
 For example, expression reports are negatively biased against germline expression because of the difficulties associated with expressing DNA in this tissue~\cite{Kelly1997}.
-As time passes, we are certain the accuracy and power of this tool will improve thanks to the efforts of the combined worm community; indeed, without the community reports of tissue expression in the first place, this tool would not be possible.
+As time passes, we are certain the accuracy and power of this tool will improve thanks to the continuing efforts of the  worm research community; indeed, without the community reports of tissue expression in the first place, this tool would not be possible.
 
@@
 
-\texttt{pip install tissue\_enrichment\_tool}
+\texttt{pip install tissue\_enrichment\_analysis}
 
@@
 % Figures describing algorithm
-
 \begin{figure}[h!]
@@
     \csentence{\textbf{a)}} The parent node (green) contains at least as many annotations as the union of the two sisters.
-These two sisters share annotations extensively.
-Therefore they are too similar and should be removed.
+These two sisters share annotations extensively, as expressed by the overlap in the Venn diagram, so they qualify for removal.
 	\csentence{\textbf{b)}} Nodes with less than a threshold number of genes are trimmed (red) and discarded from the dictionary.
 Here, the example threshold is 25 genes.
-Nodes $\epsilon, \zeta, \eta$, shown in red will be removed.
-	\csentence{\textbf{c)}} We trim parent nodes recursively, starting from the root, if all their daughter nodes have more than the threshold number of annotations.
+Nodes $\epsilon, \zeta, \eta$, shown in red are removed.
+	\csentence{\textbf{c)}} Parent nodes are removed recursively, starting from the root, if all their daughter nodes have more than the threshold number of annotations.
 Nodes in grey ($\epsilon, \zeta, \eta$) were removed in the previous step.
-Nodes $\alpha, \beta$ shown in red will be trimmed because each one has a complete daughter set.
+Nodes $\alpha, \beta$ shown in red are trimmed because each one has a complete daughter set.
 Only nodes $\gamma$ and $\delta$ will be used to generate the static dictionary.
@@
   	\captionsetup{width= 0.95\textwidth}
- 	\caption{\csentence{Screenshot of results from the web GUI.}
+ 	\caption{
+	\csentence{Screenshot of results from the web GUI.} 
 	After inputting a gene-list, the user is provided with the results.
-  An HTML table is output with hyperlinks to the ontology terms. A 
-  publication-ready graph is provided below, which can be saved by 
-  dragging to the desktop. The graph and the table show anatomy terms in
-  human-readable format, followed by their unique WBbt ID. Finally, lists 
-  of the genes used and discarded for the analysis are also presented.
+An HTML table is output with hyperlinks to the ontology terms.
+A publication-ready graph is provided below, which can be saved by  dragging to the desktop.
+The graph is colored for better visualization; color is not intended to convey information.
+The graph and the table show anatomy terms in  human-readable format, followed by their unique WBbt ID.
+Finally, lists  of the genes used and discarded for the analysis are also presented.
   }
   \label{fig:GUIresults}
+\end{figure}
+
+% workflow 
+\begin{figure}
+	\centering
+	\includegraphics[width=0.95\textwidth]{figures/workflow.pdf}
+	\captionsetup{width= 0.95\textwidth}
+	\caption{
+	\csentence{TEA Workflow.}
+	The complete ontology is annotated continuously by WormBase curators.
+After each update, the ontology is processed to remove uninformative terms.
+The trimmed ontology is used for statistical testing.
+	Users can select a gene list and input it into our tool using our WormBase portal.
+The gene list is tested for enrichment using the trimmed ontology, and results are output in tabular and graphic formats for analysis.
+	}
+	\label{fig:workflow}
 \end{figure}
@@
   \captionsetup{width= 0.95\textwidth}
-  \caption{\csentence{Kernel density estimates for 30 gold standard datasets.}
+  \caption{
+  \csentence{Kernel density estimates for 30 gold standard datasets.}
       We ran TEA on 30 datasets we believed to be enriched in particular tissues and pooled all the results to observe the distribution of q-values.
-The mode of the distribution for dictionaries with annotation cut-offs of 100 and 50 genes are very similar; however, when the cut-off is lowered to 25 genes, the mode of the distribution shifts to the left, potentially signalling a decrease in measurement power.}
+The mode of the distribution for dictionaries with annotation cut-offs of 100 and 50 genes are very similar; however, when the cut-off is lowered to 25 genes, the mode of the distribution shifts to the left, potentially signalling a decrease in measurement power.
+	  }
 	  \label{fig:qvals}
 \end{figure}
-
-%pan neuronal inter dict agreement
-% \begin{figure}
-%   \begin{subfigure}[b]{0.95\textwidth}
-%     \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_33cutoff.png}
-%     \caption{}
-%     \label{fig:PanNeuronalWatson33}
-%   \end{subfigure}
-%   %
-%   \begin{subfigure}[b]{0.95\textwidth}
-%     \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_50cutoff}
-%     \caption{}
-%     \label{fig:PanNeuronalWatson50}
-%   \end{subfigure}
-%   \captionsetup{width= 0.95\textwidth}
-%   \caption{\csentence{Comparison of Enrichment Results for two dictionaries for a pan-neuronal enriched gene set from Watson, 2008~\cite{Watson2008a}.}
-% 	Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID.
-%    \textbf{a)} Dictionary with cutoff: 33; threshold: 0.95; method: `any'
-%    \textbf{b)} Dictionary with cutoff: 50; threshold: 0.95; method: `any'
-%    }
-%    \label{fig:interagree}
-% \end{figure}
-
-%agreements
-% \begin{figure}
-%   \begin{subfigure}[b]{0.95\textwidth}
-%     \includegraphics[width=0.95\textwidth]{WBPaper00024970_GABAergic_neuron_specific_WBbt_0005190_247_33cutoff.png}
-%     \caption{}
-%     \label{fig:agreement1}
-%   \end{subfigure}
-%   %
-%   \begin{subfigure}[b]{0.95\textwidth} 	\includegraphics[width=0.95\textwidth]{WBPaper00037950_GABAergic-motor-neurons_larva_enriched_WBbt_0005190_132_33cutoff.png}
-%     \caption{}
-%     \label{fig:agreement2}
-%   \end{subfigure}
-%   \captionsetup{width= 0.95\textwidth}
-%   \caption{\csentence{Independently derived genesets show similar results within a given dictionary}
-%   Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID. Dictionary with cutoff: 33; threshold: 0.95; method: `any'.
-%   \textbf{a)} GABAergic gene set from Watson~\cite{Watson2008a}.
-%   \textbf{b)} GABAergic gene set from Spencer~\cite{Spencer2011}.
-%    }
-%    \label{fig:intragree}
-% \end{figure}
 
@@
   \captionsetup{width= 0.95\textwidth}
-  \caption{\csentence{Independently derived gene sets show similar results 	when tested with the same dictionary.}
+  \caption{
+  \csentence{Independently derived gene sets show similar results 	when tested with the same dictionary.}
   \textbf{Set 1)} GABAergic gene set from Watson~\cite{Watson2008a}.
@@
 	\captionsetup{width= 0.95\textwidth}
-  	\caption{\csentence{\emph{D.~coniospora} Gene Enrichment Analysis and Tissue Enrichment Analysis results.}
+  	\caption{
+	\csentence{\emph{D.~coniospora} Gene Enrichment Analysis and Tissue Enrichment Analysis results.}
 	We compared and contrasted the results from a gene enrichment analysis program, pantherDB, with TEA by analyzing genes that were significantly down-regulated when \emph{C.~elegans} was exposed to \emph{D.~coniospora} in a previously published dataset by Engelmann \emph{et al}~\cite{Engelmann2011} with both tools.
 	\textbf{a)} pantherDB screenshot of results, sorted by p-value. Only top hits shown.
-	\textbf{b)} TEA results, sorted by q-value (lowest on top). 
-	Both pantherDB and TEA identify terms associated with neurons (red square). The two analyses provide complementary, not redundant, information.}
+	\textbf{b)} TEA results, sorted by q-value (lowest on top) and fold-change. 
+	Both pantherDB and TEA identify terms associated with neurons (red square). The two analyses provide complementary, not redundant, information.
+	}
 	\label{fig:Dcon}
 \end{figure}
-
-% \begin{figure}[h!]
-%     \includegraphics[width=0.95\textwidth]{figures/HarposporiumspEnrichment.png}
-% 	\captionsetup{width= 0.95\textwidth}
-% 	\caption{\csentence{\emph{Harposporium sp.} enrichment results}
-%       Genes altered in   \emph{C.~elegans} after 24hr exposure to \emph{Harposporium sp.} a fungus, shows enrichment
-% 	  in the posterior lateral ganglion, similarly to \emph{D.~coniospora}. This particular fungus seems to affect
-% 	  genes normally associated with neuronal precursor tissues as well.
-% 	  }
-% 	\label{fig:Hsp}
-% \end{figure}
-
-% \begin{figure}[h!]
-% 	% \begin{centering}
-%     \includegraphics[width=0.95\textwidth]{figures/SerratiamarcescensEnrichment.png}
-% 	\captionsetup{width= 0.95\textwidth}
-%   	\caption{ \csentence{\emph{Serratia marcescens} enrichment results}.
-Genes altered in \emph{C.~elegans} after 24hr exposure to \emph{Serratia marcescens}, a bacterium, shows down-regulation of chemosensory neurons notably ASER, ASEL, ASJ and ASK.
-The lateral ganglions are also affected.
-Up-regulated genes are overrepresented for terms involving the excretory duct cell and hypodermis.}
-% 	% \end{centering}
-% 	\label{fig:Smar}
-% \end{figure}
-%
-% \begin{figure}[h!]
-%     \includegraphics[width=0.95\textwidth]{figures/EnterococcusfaecalisEnrichment.png}
-% 	\captionsetup{width= 0.95\textwidth}
-%   	\caption{\csentence{\emph{E. faecalis} enrichment results}
-%       Genes altered in   \emph{C.~elegans} after 24hr exposure to \emph{E. faecalis}, a bacterium, shows an unusual pattern:
-% 	  No terms were identified as enriched in the down-regulated category. Surprisingly, up-regulated genes show enrichment
-% 	  of various head neurons that were down-regulated with exposure to other pathogens. Interestingly, some pharyngeal
-% 	  muscle terms are also enriched.
-% 	  }
-% 	\label{fig:Efaec}
-% \end{figure}
-%
-% \begin{figure}[h!]
-%     \includegraphics[width=0.95\textwidth]{figures/OtorhabdusluminescensEnrichment.png}
-% 	\captionsetup{width= 0.95\textwidth}
-%   	\caption{\csentence{\emph{O. luminescens} enrichment results}
-%       Genes altered in   \emph{C.~elegans} after 24hr exposure to \emph{O. luminescens}, a bacterium, shows no
-% 	  significantly enriched terms in down-regulated genes. However, up-regulated genes suggest that the hypodermis
-% 	  and excretory duct cell may be affected by infection.
-% 	  }
-% 	\label{fig:Olum}
-% \end{figure}
 
@@
 \begin{table}[h!]
-\caption{Parameter specifications and number of tissues for all dictionaries.
+	\caption{
+	Parameter specifications and number of tissues for all dictionaries.
 The `Method' column refers to the trimming criterion for the similarity metric.
 We used two such criteria, `any' and `avg'.`any': For a given sister set, if any sister had a similarity exceeding the corresponding threshold, all sisters were removed from the final dictionary.
-`avg': For a given sister set, if the average similarity across all the sisters in the set was greater than the corresponding threshold, all sisters were removed from the final dictionary.}
-	\csvautotabular{figures/TissueNumbers.csv}
+`avg': For a given sister set, if the average similarity across all the sisters in the set was greater than the corresponding threshold, all sisters were removed from the final dictionary.
+	}
+	\csvreader[tabular= c | c | c | c,
+	table head= {\textbf{Annotation Cutoff}} & {\textbf{Similarity Threshold}} & {\textbf{Method}} & {\textbf{No. Of Terms in Dictionary}}\\\hline,
+	late after line= \\,
+	late after last line= \\\hline
+	]
+	{figures/TissueNumbers.csv}{}
+	{\csvcoli & \csvcolii & \csvcoliii & \csvcoliv}
+
+	% \csvautotabular{figures/TissueNumbers.csv}
 	\label{tab:DictionarySpecs}
@@
 \begin{table}[h!]
-	\caption{Comparison of results for a neuronal-enriched geneset from Watson~\cite{Watson2008a}.
+	\caption{
+	Comparison of results for a neuronal-enriched gene set from Watson~\cite{Watson2008a}.
 We ran the same genelist on a dictionary with a minimum annotation cutoff of 50, similarity threshold of 0.95 and similarity method `any' versus another with a minimum annotation cutoff of 33, similarity threshold of 0.95 and similarity method `any'.
@@
 Hyphens denote not-applicable values, which occurs when a particular tissue is not present in both dictionaries.
-Full table is available on github.}
+Full table is available on github.
+	}
+
 	\begin{adjustbox}{width=1\textwidth}
 		\centering
-		\csvautotabular{figures/dict-comparison-50-33.csv}
+		% \csvreader{figures/dict-comparison-50-33.csv}
+		\csvreader[tabular={ r | c | c | c | c},
+	    table head ={\textbf{Tissue}} & {\textbf{Q-value-33}} & {\textbf{Q-value-50}} & {\textbf{Enrichment-33}} & {\textbf{Enrichment-50}}\\\hline,
+	    late after line= \\,
+		late after last line=\\\hline
+		]%
+		{figures/dict-comparison-50-33.csv}{}	
+		{\csvcoli & \csvcolii & \csvcoliii & \csvcoliv & \csvcolv}	
 		\label{tab:interagree}

### 2016-04-18 build/lib/tissue_enrichment_analysis/hypergeometricTests.py

@@ -1,9 +1,10 @@
 # -*- coding: utf-8 -*-
 """
+A script to implement a hypergeometric test procedure.
+
 Author: David Angeles
 Date: May 26, 2015
 Requires Python > 3.5
-A script to implement a hypergeometric test
 Needs:
 A tissue dictionary
 A control list of gene names
@@ -280,18 +281,20 @@ def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
     import seaborn as sns
     sns.set_context('paper')
     sns.set_style('whitegrid')
-
+    sns.choose_colorbrewer_palette('sequential', as_cmap=False)
     if df.empty:
         print('dataframe is empty!')
         return
 
     ax = kwargs.pop('ax', None)
+    ftype = kwargs.pop('ftype', 'svg')
 #
     if ax is None:
         ax = plt.gca()
 
     # sort by q value change
-    df.sort_values('Q value', ascending=True, inplace=True)
+    df.sort_values(['Q value', 'Enrichment Fold Change'],
+                   ascending=[True, False], inplace=True)
     # plot first n_bars
     ax = sns.barplot(x=df[y][:n_bars], y=df['Tissue'][:n_bars], ax=ax)
 
@@ -308,11 +311,12 @@ def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
             if not os.path.exists(dirGraphs):
                 os.makedirs(dirGraphs)
             if dirGraphs[len(dirGraphs)-1] != '/':
-                plt.savefig(dirGraphs+'/{0}.pdf'.format(title))
+                plt.savefig(dirGraphs+'/{0}.{1}'.format(title, ftype),
+                            dpi=1200)
             else:
-                plt.savefig(dirGraphs+'{0}.pdf'.format(title))
+                plt.savefig(dirGraphs+'{0}.{1}'.format(title, ftype), dpi=1200)
         else:
-            plt.savefig('{0}.pdf'.format(title))
+            plt.savefig('{0}.{1}'.format(title, ftype), dpi=1200)
 
     return ax
 #    plt.show()

### 2016-04-18 setup.py

@@ -10,7 +10,7 @@ from setuptools import setup, find_packages
 import os
 import sys
 
-version = '0.13.10'
+version = '0.13.12'
 
 # just type in python setup.py publish and this takes care of publishing to pypi!
 

### 2016-04-18 tests/hgf_benchmarking/src/hgf_benchmark_script.py

@@ -132,6 +132,7 @@ NoAnnotations = df_summary.NoAnnotations.unique()
 def resplot(column, method='any'):
     """
     A method to quickly plot all combinations of cutoffs, thresholds.
+
     All cutoffs are same color
     All Thresholds are same line style
 
@@ -232,7 +233,7 @@ def walker(tissue_df, directory, save=True):
                 line_prepender(directory+short_name+'.csv', line)
                 # plot
                 tea.plot_enrichment_results(df_analysis, title=short_name,
-                                            dirGraphs=directory)
+                                            dirGraphs=directory, ftype='pdf')
                 plt.close()
 
             # if it's empty and you want to save, place it in file called empty
@@ -303,7 +304,7 @@ walker(tissue_df, dirHGT100_any)
 tissue_df = pd.read_csv('../input/WS252AnatomyDictionary/cutoff33_threshold0.95_methodany.csv')
 walker(tissue_df, dirHGT33_any)
 
-grouped = df_summary.groupby(['NoAnnotations','Threshold', 'Method'])
+grouped = df_summary.groupby(['NoAnnotations', 'Threshold', 'Method'])
 with open('../doc/figures/TissueNumbers.csv', 'w') as f:
     f.write('Annotation Cutoff,Similarity Threshold,Method')
     f.write(',No. Of Terms in Dictionary\n')
@@ -322,32 +323,34 @@ thresh = df_summary.Threshold.unique()
 NoAnnotations = df_summary.NoAnnotations.unique()
 
 
-def resplot(column, cutoff=25, method='any'):
-    """
-    A method to quickly plot all combinations of cutoffs, thresholds.
-    All cutoffs are same color
-    All Thresholds are same line style
-    """
-    for i, threshold in enumerate(thresh):
-        ax = plt.gca()
-        ax.grid(False)
-        if threshold == 1:
-            continue
-        tissue_data[sel(threshold, method)].plot(x='No. Of Annotations',
-                    y= 'No. Of Tissues in Dictionary', kind='scatter',
-                    color=cols[i], ax=ax, s=50, alpha=.7)
-    ax.set_xlim(20, 110)
-    ax.set_xscale('log')
-    ax.set_xticks([25, 33, 50, 100])
-    ax.get_xaxis().set_major_formatter(mpl.ticker.ScalarFormatter())
-
[…]

### 2016-04-18 tissue_enrichment_analysis/__init__.py

@@ -17,7 +17,7 @@ __author__ = "David Angeles-Albores"
 __copyright__ = "Copyright 2016, WormBase"
 __credits__ = ["David Angeles-Albores", "Raymond Y. Lee", "Juancarlos Chan"]
 __license__ = "MIT"
-__version__ = "0.13.10"
+__version__ = "0.13.12"
 __maintainer__ = "David Angeles-Albores"
 __email__ = "dangeles@caltech.edu"
 __status__ = "Production"

### 2016-04-19 tests/hgf_benchmarking/doc/SI/FolderStructure.tex
writing

--- 2016-04-18 tests/hgf_benchmarking/doc/tea_main_text.tex
@@
-%% BioMed_Central_Tex_Template_v1.06
-%%                                      %
-%  bmc_article.tex            ver: 1.06 %
-%                                       %
-
-%%IMPORTANT: do not delete the first line of this template
-%%It must be present to enable the BMC Submission system to
-%%recognise this template!!
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                     %%
-%%  LaTeX template for BioMed Central  %%
-%%     journal article submissions     %%
-%%                                     %%
-%%          <8 June 2012>              %%
-%%                                     %%
-%%                                     %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                                                 %%
-%% For instructions on how to fill out this Tex template           %%
-%% document please refer to Readme.html and the instructions for   %%
-%% authors page on the biomed central website                      %%
-%% http://www.biomedcentral.com/info/authors/                      %%
-%%                                                                 %%
-%% Please do not use \input{...} to include other tex files.       %%
-%% Submit your LaTeX manuscript as one .tex document.              %%
-%%                                                                 %%
-%% All additional Figures and files should be attached             %%
-%% separately and not embedded in the \TeX\ document itself.       %%
-%%                                                                 %%
-%% BioMed Central currently use the MikTex distribution of         %%
-%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
-%% http://www.miktex.org                                           %%
-%%                                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%% additional documentclass options:
-%  [doublespacing]
-%  [linenumbers]   - put the line numbers on margins
-
-%%% loading packages, author definitions
-
-% \documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-% \documentclass[linenumbers, doublespacing]{bmcart}
-\documentclass{bmcart}
+\documentclass{article}
 % \bibliographystyle{unsrtnat}
@@
 %%% Put your definitions there:
-\startlocaldefs
-
-% ignore command 
-\newcommand{\ignore}[2]{\hspace{0in}#2}
-
-\endlocaldefs
 
@@
 \begin{document}
-
-%%% Start of article front matter
-\begin{frontmatter}
-
-\begin{fmbox}
-\dochead{Methodology}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors here                   %%
-%%                                          %%
-%% Specify information, if available,       %%
-%% in the form:                             %%
-%%   <key>={<id1>,<id2>}                    %%
-%%   <key>=                                 %%
-%% Comment or delete the keys which are     %%
-%% not used. Repeat \author command as much %%
-%% as required.                             %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\author[
-   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
-   % email={dangeles@caltech.edu}   % email address
-]{\inits{DA}\fnm{David} \snm{Angeles-Albores}}
-\author[
-   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
-   % email={raymond@caltech.edu}   % email address
-]{\inits{RYL}\fnm{Raymond YN} \snm{Lee}}
-\author[
-   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
-   % email={azurebrd@caltech.edu}   % email address
-]{\inits{JC}\fnm{Juancarlos} \snm{Chan}}
-\author[
-   addressref={aff1},
-   corref={aff1},
-   email={pws@caltech.edu}
-]{\inits{PWS}\fnm{Paul W} \snm{Sternberg}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors' addresses here        %%
-%%                                          %%
-%% Repeat \address commands as much as      %%
-%% required.                                %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\address[id=aff1]{%                           % unique id
-  \orgname{HHMI and California Institute of Technology, Division of Biology and Biological Engineering}, % university, etc
-  \street{1200 E California Blvd},                     %
-  \postcode{91125},                               % post or zip code
-  \city{Pasadena},                              % city
-  \cny{USA}                                    % country
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter short notes here                   %%
-%%                                          %%
-%% Short notes will be after addresses      %%
-%% on first page.                           %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\end{fmbox}% comment this for two column layout
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Abstract begins here                 %%
-%%                                          %%
-%% Please refer to the Instructions for     %%
-%% authors on http://www.biomedcentral.com  %%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstractbox}
-
-\begin{abstract} % abstract
-\parttitle{Background} %if any
-Over the last ten years, there has been explosive development in tools
-for measuring gene expression. These tools can identify thousands of altered 
-genes between conditions, but understanding these datasets and forming hypotheses based on them remains challenging.
-One way to analyze these datasets is to associate ontologies (hierarchical, descriptive vocabularies
-with controlled relations between terms) with genes and to look for enrichment of specific terms.
-Although gene ontology (GO) is available for   \emph{Caenorhabditis~elegans}, it does not include anatomy information.
-\parttitle{Results} %if any 
-We have developed a tool for identifying  enrichment of \emph{C.~elegans} tissues among gene sets and generated a website GUI where users can access this tool.
-Since a common drawback to ontology enrichment analyses is result verbosity, we developed a very simple trimming algorithm to reduce the ontology size by an order of magnitude.
-We adjusted these filters and validated our tool using a set of 30  gold standards from Expression Cluster data in WormBase.
-We used our tool to identify multiple neuronal tissues that are down-regulated due to pathogen infection in \emph{C.~elegans}.
-\parttitle{Conclusions} %if any
-Our Tissue Enrichment Analysis (TEA) can be found at \url{www.wormbase.org/tea}, and can be downloaded using Python's standard pip installer.
-It tests a slimmed-down \emph{C.~elegans} tissue ontology for enrichment of specific terms and provides users with a  text and graphic representation of the results.
-Our tool can discriminate between embryonic and larval tissues and can even identify tissues down to the single-cell level.
-\end{abstract}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The keywords begin here                  %%
-%%                                          %%
-%% Put each keyword in separate \kwd{}.     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{keyword}
-\kwd{Gene Ontology}
-\kwd{Anatomy Ontology}
-\kwd{WormBase}
-\kwd{RNA-seq}
-\kwd{High-throughput biology}
-\end{keyword}
-
-% MSC classifications codes, if any
-%\begin{keyword}[class=AMS]
-%\kwd[Primary ]{}
-%\kwd{}
-%\kwd[; secondary ]{}
-%\end{keyword}
-
-\end{abstractbox}
-%
-%\end{fmbox}% uncomment this for twcolumn layout
-
-\end{frontmatter}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Main Body begins here                %%
-%%                                          %%
-%% Please refer to the instructions for     %%
-%% authors on:                              %%
-%% http://www.biomedcentral.com/info/authors%%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%% See the Results and Discussion section   %%
-%% for details on how to create sub-sections%%
-%%                                          %%
-%% use  \cite{...} to cite references        %%
-%%   \cite{koon} and                         %%
-%%   \cite{oreg,khar,zvai,xjon,schn,pond}    %%
-%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
-% <put your article body there>
-
 %%%%%%%%%%%%%%%%
@@
 %%
-\section*{Background}
-	RNA-seq and other high-throughput methods in biology have the ability to identify thousands of genes that are altered between conditions.
-These genes are often correlated in their biological characteristics or functions, but identifying these functions remains challenging.
-To interpret these long lists of genes, biologists need to abstract genes into concepts that are biologically relevant to form hypotheses about what is happening in the system.
-One such abstraction method relies on Gene Ontology (GO).
-GO provides a controlled set of hierarchically ordered terms~\cite{TheGeneOntologyConsortium2000a, TheGeneOntologyConsortium2015} that provide detailed descriptions about the molecular, cellular or biochemical functions of any gene.
-For a given gene list, certain software programs can query whether a particular term is enriched~\cite{Mi2009, McLean2010, Huang2009}.
-One area of biological significance that GO does not include is anatomy.
-One way to address this shortcoming is to use a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `specific cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}.
-The \emph{C.~elegans} database, WormBase~\cite{Howe2016}, maintains a curated list of gene expression data from the literature.
-Here we provide a new framework that analyzes a user-input list for enrichment of specific cells and tissues.
-They are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation.
-%As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.
+\section*{Folder Structure}
 
-Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene.
-There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.
-DAVID, a common tool for GO enrichment analysis, clusters enriched terms into broad categories~\cite{Huang2007}, whereas PANTHER~\cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (H.
-Yu and P.
-Thomas, pers.
-comm.).
+In the SI, we provide the entire output of our tests for the interested reader. The folder is organized as follows:
 
- To reduce verbosity, we have filtered our ontology using a small set of well-defined criteria to remove terms that do not contribute additional information.
-To our knowledge, such filtering has not been performed in an algorithmic fashion for a biological ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
-Other pruning tools do exist (see for example~\cite{Kim2007}), but the pruning is query-dependent.
-We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.
+\subsection*{HGTXX\_any\_Results}
+The folders titled HGTXX\_any\_Results contain the enrichment results of all the 30 golden sets tested with a dictionary with cutoff XX, similarity threshold $=0.95$ and thresholding method `any'.
+The folders contain two kinds of files, `.pdf' and `.csv' files.
+The pdf files contain the bar charts for each analysis, whereas the csv files contain the complete output for a particular geneset.
+There is also a file called `empty.txt' that contains the names of any gene sets where no terms were enriched.
 
-%Our tool is available within WormBase and provides users with a text-based file of the enrichment results as well as a simple and clear graph of the results that exhibit the largest fold-change enrichment.
+Files are titled according to a common schema. For example, the file \\
+\texttt{WBPaper00013489\_Ray\_Enriched\_WBbt\_0006941\_25.csv}\\
+refers to the \textbf{WormBase Paper 0013489}, which should be `\textbf{Ray Enriched}', and specifically should be enriched in term `\textbf{WBbt:0006941}'.
+The gene set that was used for this analysis is contained in the SI folder named `\textbf{golden gene sets}' and is contained in a homologously named csv file.
 
-\section*{Results}
-\subsection*{Generating a Gene-Tissue Dictionary}
-\subsubsection*{Reducing term redundancy through a similarity metric}
-As a first step to generate our tissue enrichment software, we wished to select tissue terms that were reasonably well-annotated, yet specific enough to provide insight and not redundant with other terms.
-%We also wanted to avoid testing tissues at levels where annotation redundancy becomes problematic.
-For example, several left and right neurons have at least 25 annotated genes and we may want to include them for enrichment testing.
-For example, nematodes have a number of neurons that are placed symmetrically along the left/right body axis, and are functionally similar.
-These left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them%.
-We reasoned that when two tissues have almost identical annotations
-, and therefore we cannot have statistical confidence in differentiating between them.
-As a result, testing these sister tissues provides no additional information compared with testing only the parent node to these sisters.
-%We refer to such sisters as `redundant'.
-To identify redundancy, we defined a similarity metric (see \emph{Methods} section and Figure~\ref{fig:simdiagram}) that can be used to identify sisters that have very high similarity between them; alternatively, redundant sisters could be identified if a single sister had a very high similarity score.
-We referred to these two scoring criteria as `avg' and `any' respectively.
+\subsection*{Engelmann}
+	Contains all graphs pertaining to the data from Engelmann 2011.
 
-\subsubsection*{Terminal branch terms and parent terms can be
-safely removed in an algorithmic fashion }
-Another problem arises from the ontology being scarcely populated.
-Many nodes have 0-10 annotations, which we consider too few to accurately test.
-To solve this issue, we implemented a straightforward trimming strategy.
-For a given terminal node, we test whether the node has more than a threshold number of annotations.
-If it does not, the node is removed.
-The next higher node in the branch is tested and removed recursively until a node which satisfies the condition is found.
-At that point, no more nodes can be removed from that branch.
-This is guaranteed by the structure of the ontology: parent nodes inherit all of the annotations of all of their descendants, so the number of annotated terms monotonically increases with increasing term hierarchy (see Figure~\ref{fig:trim_ends}).
-In this way, we ensure that our term dictionary includes only those tissues that are considered sufficiently well annotated for statistical purposes.
+\subsection*{Comparisons}
+	Contains csv files of the comparisons between gene sets or within dictionaries. The file nomenclature is:\
+neuronal\_comparison\_33\_WBPaper00024970\_with\_WBPaper0037950\_complete.csv
+Refers to a `neuronal comparison' using dictionary 33, comparing papers 24970 with 37950.
+The word complete at the end of the analysis refers to the fact that this table contains the complete table of results.
 
-Additionally, we reasoned that, for any parent node, if all its daughters were selected for testing, there was no additional benefit to test the parent.
-%In other words, if all the daughter nodes are tested, there is little additional information to be gained by including the parent node.
-To address this issue
-We removed parent nodes from the analysis if all their daughter nodes passed the annotation threshold (see Figure~\ref{fig:trim_roots}).
-We called this a ceiling filter.
-Applying these three filters reduced the number of ontology terms by an order of magnitude.
+Alternatively, when comparing two dictionaries, the nomenclature is:
+neuronal\_comparison\_GABAergic\_33-50_WBPaper0037950_complete.csv
+Which refers to a `GABAergic neuronal' lists, comparing dictionaries with cutoffs of 33-50 (all other parameters are the same) using paper 37950.
 
-%Finally, we also wanted to remove as many terms as possible from the dictionary with the goals of reducing covariance between terms and decreasing multiple testing.
-Decreasing covariance between terms is important because we employ a frequentist approach that assumes all terms are independent.
-Covariation  between terms means that if one of these tissues tests significant, the other terms are much more likely to pass significance testing as well.
-This makes adequate correction for false positive rates considerably more difficult.
+\subsection*{Summary Info}
 
-\subsubsection*{Filtering greatly reduces the number of nodes used for analysis}
-By itself, each of these filters can reduce the number of nodes employed for analysis, but applying the filters in different orders removes different numbers of nodes (not all the filters are commutative).
-We chose to always execute annotation and similarity thresholding first, followed by the ceiling filter.
-%If the ceiling filter is applied before any other filter, only terminal nodes will remain, since all the parents have complete daughter sets.
-Since terminal nodes are the most poorly annotated, after applying the remaining filters very few nodes will be left behind if any.
-On the other hand, applying the ceiling operator after trimming and redundancy filtering will result in greater numbers of nodes.
-We always applied the ceiling at the end.
-For validation (see below) we made a number of different dictionaries.
-The original ontology has almost 6,000 terms of which 1675 have at least 5 gene annotations.
-After filtering, dictionary sizes ranged from 21 to a maximum of 460 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by an order of magnitude through the application of a few simple filters (see Table~\ref{tab:DictionarySpecs}).
-These filters were used to compile a static dictionary that we employ for all analyses (see \emph{Validation of the algorithm and parameter selection} section for details).
+\subsection*{test_list_efaecalis.txt}
 
-\subsection*{Tissue enrichment testing via a hypergeometric model}
-Having built a static dictionary, we generated a Python script that implements a significance testing algorithm based on the hypergeometric model.
-Briefly, the hypergeometric model tests the probability of observing $n_i$ occurences of a tissue $i$ in a list of size $M$ if there are $m_i$ labels for that tissue in a dictionary of total size $N$ that are drawn without replacement.
-Mathematically, this is expressed as:
-
-\begin{eqnarray}\label{hypergeometric}
-	\mathrm{P}(n_i | N, m_i, M) = \frac{\dbinom{m_i}{n_i} \dbinom{M-m_i}{N - n_i}}{\dbinom{N}{n_i}}\text{.}
-\end{eqnarray}
-
-Although a user will input gene IDs, we test the number of ocurrences of a term within the gene list, so a single gene can contribute to multiple terms.
-Due to the discrete nature of the hypergeometric distribution, this algorithm can generate artifacts when the list is small.
-To avoid spurious results, a tissue is never considered significant if there are no annotations for it in the user-provided list.
-
-Once the p-values for each term have been calculated, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm~\cite{Benjamini1995}.
-FDR corrected p-values are called q-values.
-Genes that have a q-value less than a given alpha are considered significant.
-Our default setting is to set alpha at 0.1.
-
-Users input a gene list using any valid gene name for \emph{C.~elegans}.
-These names are processed into standard WormBase gene IDs (WBGene IDs).
-The program returns a table containing all the enriched terms and associated useful information%with significant tissues, their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change ($\text{observed hits} /\text{expected hits}$)
-.
-Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values.
-The bars in the graph are sorted in ascending order of q-value and then in descending order of fold-change.
-Bars are colored for ease of viewing, and color does not convey information.
-Our software is implemented in an easy to use GUI (see Figure~\ref{fig:GUIresults}).
-Anatomy terms are displayed in human-readable format followed by their unique ontology ID (WBbt ID).
-In summary, each time the ontology annotations are updated, a new trimmed ontology is generated using our filters; in parallel, users can submit their gene lists through WormBase for testing, with results output in a number of formats (see Figure~\ref{fig:workflow}).
-
-\subsection*{Validation of the algorithm and optimizing parameter selection}
-We wanted to select a dictionary that included enough terms to be specific beyond the largest \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change.
-Larger tissues are correlated with better annotation, so increasing term specificity is associated with losses in statistical power.
-%The phrase below doesn't seem useful
-%To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.
-% Gold-standard data sets were annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
-To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}.
-These data sets are annotated gene lists derived from the corresponding Expression Cluster data in WormBase.
-Some of these studies have been used to annotate gene expression, and so they did not constitute an independent testing set.
-To correct this flaw, we built a clean dictionary that specifically excluded all annotation evidence that came from these studies.
-
-As a first attempt to select a dictionary, we generated all possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any' similarity thresholding methods (see Table~\ref{tab:DictionarySpecs}).
-The number of tissues was inversely correlated to the minimum annotation cutoff, and was largely insensitive to the similarity threshold in the range we explored.
-Next, we analyzed all 30 datasets using each dictionary.
-Because of the large number of results, instead of analyzing each set of terms individually, we measured the average q-value for significantly enriched terms in each dataset without regard for the perceived accuracy of the terms that tested significant.
-We found that the similarity threshold mattered relatively little for any dictionary.
-We also noticed that the `any' thresholding method resulted in tighter histograms with a mode closer to 0.
-For this reason, we chose the `any' method for dictionary generation.
-The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant is the trade-off between power and term specificity.
-Based on these observations, we ruled out the dictionary with the 100 gene annotation cut-off: it had the fewest terms and its q-values were not low enough in our opinion to compensate the trade-off in specificity.
-
-%dictionaries that work well
-To select between dictionaries generated between 50, 33 and 25 annotation cut-offs, and also to ensure the terms that are selected as enriched by our algorithm are reasonable, we looked in detail at the enrichment analysis results.
-%Work well
-% TODO add WBbt codes!
-Most results were comparable and expected.
-For some sets, all dictionaries performed well.
-For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} all terms were neuron-related %including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays
-regardless of the dictionary used (see Table~\ref{tab:interagree}).
-On the other hand, for a set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the 50 cutoff dictionary was only able to identify `oocyte WBbt:006797' which is not a germline precursor, although it is germline related; whereas the two smaller dictionaries singled out actual germline precursor cells---at the 33 cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as enriched, and at the 25 gene cutoff the embryonic germline precursor terms `P$_4$',`P$_3$' and `P$_2$' were identified in addition to `Z2' and `Z3'.
-%Differences between dictionaries
-We also queried an intestine precursor set~\cite{Spencer2011}.
-Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary.
-However, the 33 cutoff dictionary identified the E lineage, which is the intestinal precursor lineage in \emph{C.~elegans}, as enriched.
-Both of these results capture specific aspects of \emph{C.~elegans} that are well known to developmental biologists.
-
-%Bad stuff.
-Not all queries worked equally well.
-For example, a number of intestinal  sets~\cite{Spencer2011, Pauli2006} were not enriched in intestine-related terms in any dictionary, but were enriched for pharynx and hypodermis.
-We were  surprised that intestinal gene sets performed poorly, since the intestine is a relatively well-annotated tissue.
-
-%Dictionary intra-agreement
-We assessed the internal agreement of our tool by using independent gene sets that we expected to be enriched in the same tissues.
-We used two  pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two PVD sets~\cite{Spencer2011, Smith2010}; and two  GABAergic sets~\cite{Spencer2011, Cinar2005}.%; two  pharyngeal gene sets~\cite{Spencer2011, Gaudet2004a}; and two  intestinal gene sets~\cite{Spencer2011, Pauli2006}.
-Overall, the tool has good internal agreement.
-On most sets, the same terms were enriched, although order was somewhat variable (see Table~\ref{tab:interagree}), and most high-scoring terms were preserved between sets.
-%The intestinal gene-sets and pharyngeal gene sets comparisons were exceptions, since at least one gene set showed no enriched terms in all dictionaries, so we did not consider them as informative for assessing internal agreement.
-All comparisons can be found online in our Github repository (see Availability of data and materials).
-Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95 redundancy threshold using the `any' criterion performed best, with a good balance between specificity, verbosity and accuracy, so we selected this parameter set to generate our static dictionary.
-As of this publication, the testable dictionary contains 261 terms.
-
-\section*{Applying the tool}
-We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of their underlying biology.
-Engelmann et al.
-exposed young adult worms to 5 different pathogenic bacteria or fungi for 24 hours, after which mRNA was extracted from the worms for sequencing.
-We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated.
-Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade.
-Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct.
-Three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes.
-As an independent verification, we also performed GO analysis using PANTHER on the down-regulated genes for \emph{D.~coniospora}.
-These results showed enrichment in terms associated with neurons (see Figure~\ref{fig:Dcon}).
-A possible explanation for this neuronal association might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen.
-Indeed, several studies~\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.~elegans} uses chemosensory neurons to identify pathogens.
-%One bacterium did not ellicit the same pattern of down-regulation of neuronal-associated genes.
-\emph{E.~faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E.~faecalis} may have a different pathogenic profile.
-Our results highlight the involvement of various \emph{C.~elegans} neuronal tissues in pathogen defense.
-
-\section*{Discussion}
-%TEA Discussion
-We have presented a tissue enrichment analysis tool that employs a standard hypergeometric model to test the \emph{C.~elegans} tissue ontology.
-We use a hypergeometric function to test a user-provided gene list for enrichment of anatomical terms in \emph{C.~elegans}.
-Our hope is that the physiological relevance of anatomical terms will enable researchers to make hypotheses about high-dimensionality data.
-Specifically, we believe an enriched term may broadly suggest one of two hypotheses: if a list is enriched in a particular anatomical region, that anatomical region is affected by the experimental treatment; alternatively, the anatomical regions that are enriched reflect biologically relevant interactions between tissues.
-We believe the first hypothesis is a reasonable hypothesis to make in the case of whole-worm RNA-seq data for example, whereas the second hypothesis may be more plausible in cases where a researcher already knows what tissues a particular gene list came from, as may be the case in single-cell RNA-seq.
-
-%Other tools
-Our tool relies on an annotation dictionary that is continuously updated primarily with data from single gene qualitative analyses, does not require retraining and does not require ranked genes.
-To our knowledge, this is the first tool that tests tissue enrichment  in \emph{C.~elegans} via the hypergeometric method, but similar projects exist for humans and zebrafish~\cite{Lee2013, Prykhozhij2013}, highlighting the relevance of our tool for high-dimensionality biology.
-Chikina \emph{et al}~\cite{Chikina2009} have previously reported a tissue enrichment model for \emph{C.~elegans }based on a Support Vector Machine classifier that has been trained on microarray studies.
-SVMs are powerful tools, but they require continuous retraining as more tissue expression data becomes available.
-Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for certain studies.
-Moreover, this tissue enrichment tool generally provides users with enrichment results for 6 large tissues.
-In contrast, our tool routinely tests a much larger number of terms, and we have shown it can even accurately identify enrichment of embryonic precursor lineages for select data sets.
-
-%Trimming discussion
-We have also presented the first, to our knowledge, ontology trimming algorithm for biomedical ontologies.
-This algorithm, which is very easy to execute, places strong limits on the number of terms selected for testing.
-Due to the nature of all ontologies as hierarchical, acyclical graphs with term inheritance, term annotations are correlated along any given branch.
-This correlation reduces the benefits of including all terms for statistical analysis: for any given term along a branch, if that term passes significance, there is a high probability that many other terms along that branch will also pass significant.
-If the branch is enriched by random chance, error propagation along a branch means that many more false positives will follow.
-Thus, a researcher might be misled by the number of terms of correlated function and assign importance to this finding; the fact that the branching structure of GO amplifies false positive signals is a powerful argument for either reducing branch length or branch intracorrelation, or both.
-On the other hand, if a term is actually enriched, we argue that there is little benefit to presenting the user with additional terms along that branch.
-Instead, a user will benefit most from testing sparsely along the tree at a suitable specificity for hypothesis formation.
-Related terms of the same level should only be tested when there is sufficient annotation to differentiate, with statistical confidence, whether one term is enriched above the other.
-Our algorithm reduces branch length by identifying and removing nodes that are insufficiently annotated and parents that are likely to include sparse information.
-
-We have endeavoured to benchmark our tool well. However, our analysis could not identify %identify good negative controls
- spurious term enrichment.
-% Even for our set of positive controls, the statistical analysis sometimes throws out unexpected results.
-For example, the embryonic germline precursor gene set had the term `AB', which refers to the anterior cell in the 2-cell stage \emph{C.~elegans} embryo, as the most enriched term in the dictionaries with cut off of 25 and 33.
-Is this an error, or does this hint at new biology?
- Although we were unable to determine false-positive and false-negative rates, we do not believe this should deter scientists from using our tool.
-Rather, we encourage researchers to use our tool as a guide, integrating evidence from multiple sources to inform the most likely hypotheses.
-As with any other tool based on statistical sampling, our analysis is most vulnerable to bias in the data set.
-For example, expression reports are negatively biased against germline expression because of the difficulties associated with expressing DNA in this tissue~\cite{Kelly1997}.
-As time passes, we are certain the accuracy and power of this tool will improve thanks to the continuing efforts of the  worm research community; indeed, without the community reports of tissue expression in the first place, this tool would not be possible.
-
-\section*{Methods}
-\subsection*{Fetching annotation terms}
-We used WormBase-curated gene expression data, which includes annotated descriptions of spatial-temporal expression patterns of genes, to build our dictionary.
-Gene lists per anatomy term were extracted from a Solr document store of gene expression data from the WS252 database provided by WormBase~\cite{Howe2016}.
-We used the Solr document store because it provided a convenient access to expression data that included inferred annotations.
-That is, for each anatomy term, the expression gene list includes genes that were directly annotated to the term, as well as those that were annotated to the term's descendant terms (if there were any).
-Descendant terms were those connected with the focus term by is\_a/part\_of relationship chains defined in the anatomy term ontology hierarchy.
-\subsection*{Filtering nodes}
-\subsubsection*{Defining a Similarity Metric}
-To identify redundant sisters, we defined the following similarity metric:
-\begin{eqnarray}\label{similarity def}
-	s_i = \frac{|g_i|}{|\bigcup_{i= 0}^k g_i|}
-\end{eqnarray}
-Where $s_i$ is the similarity for a tissue $i$ with $k$ sisters; $g_i$ refers to the set of tissues associated with tissue $i$ and $|g|$ refers to the cardinality of set $g$.
-For a given set of sisters, we called them redundant if they exceeded a given similarity threshold.
-We envisioned two possible criteria and built different dictionaries using each one.
-Under a threshold criteron `any' with parameter $S$ between $(0, 1)$, a given set of sisters $j$ was considered redundant if the condition
-\begin{eqnarray}\label{any threshold}
-	s_{i, j} > S
-\end{eqnarray}
-was true for any sister $i$ in set $j$. Under a threshold criterion `avg' with parameter $S$, a given set of sisters $j$ was considered redundant if the condition
-\begin{eqnarray}\label{avg threshold}
-	\mathrm{E}[s_i]_j > S
-\end{eqnarray}
-was true for the set of sisters $j$ (see Figure~\ref{fig:simdiagram}).
-
-\subsubsection*{Implementation}
-All scripts were written in Python 3.5.
-Our software relies on the pandas, NumPy, Seaborn and SciPy modules to perform all statistical testing and data handling~\cite{McKinney2011, VanDerWalt2011, Oliphant2007}.
-
-\section*{Availability of data and materials}
-Our web implementation is available at \url{https://www.wormbase.org/tea}. Our software can also be downloaded using Python's pip installer via the command
-
-\texttt{pip install tissue\_enrichment\_analysis}
-
-Alternatively, our software is available for download at: \url{http://dangeles.github.io/TissueEnrichmentAnalysis}
-
-All benchmark gene sets, benchmarking code and Figures can also be found at the same address, under the `tests' folder.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\begin{backmatter}
-
-\section*{Competing interests}
- 	The authors declare that they have no competing interests.
-\section*{Author's contributions}
-    DA and PWS conceived of the project; DA developed algorithm;
-	RYL made intellectual contributions to the project; RYL and JC
-	developed the web GUI.
-\section*{Acknowledgements}
-	We thank Justin Bois for his help and support. 
-	We would like to acknowledge all members of the Sternberg
-	lab for helpful discussion.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  Bmc_mathpys.bst  will be used to                       %%
-%%  create a .BBL file for submission.                     %%
-%%  After submission of the .TEX file,                     %%
-%%  you will be prompted to submit your .BBL file.         %%
-%%                                                         %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% if your bibliography is in bibtex format, use those commands:
-\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )
-% for author-year bibliography (bmc-mathphys or spbasic)
-% a) write to bib file (bmc-mathphys only)
-% @settings{label, options="nameyear"}
-% b) uncomment next line
-% \nocite{label}
-
-% or include bibliography directly:
-% \begin{thebibliography}
-% \bibitem{b1}
-% \end{thebibliography}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Figures                       %%
-%%                               %%
-%% NB: this is for captions and  %%
-%% Titles. All graphics must be  %%
-%% submitted separately and NOT  %%
-%% included in the Tex document  %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%
-%% Do not use \listoffigures as most will included as separate files
-
-\section*{Figures}
-
-% Figures describing algorithm
-\begin{figure}[h!]
-  \begin{subfigure}[b]{0.3\textwidth}
-    \includegraphics[width=\textwidth]{figures/RedundancyTrimmingOntogeny.png}
-   	 	\caption{}
-    \label{fig:simdiagram}
-  \end{subfigure}
-  %
-  \begin{subfigure}[b]{0.3\textwidth}
-    \includegraphics[width=\textwidth]{figures/FloorTrimmingOntogeny.png}
-	      \caption{}
-    \label{fig:trim_ends}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.3\textwidth}
-    \includegraphics[width=\textwidth]{figures/CeilingTrimmingOntogeny.png}
-    	\caption{}
-    \label{fig:trim_roots}
-  \end{subfigure}
-
-  \captionsetup{width= 0.95\textwidth}
-  \caption{
-  	\csentence{Schematic representation of trimming filters for an acyclical ontology.}
-    \csentence{\textbf{a)}} The parent node (green) contains at least as many annotations as the union of the two sisters.
-These two sisters share annotations extensively, as expressed by the overlap in the Venn diagram, so they qualify for removal.
-	\csentence{\textbf{b)}} Nodes with less than a threshold number of genes are trimmed (red) and discarded from the dictionary.
-Here, the example threshold is 25 genes.
-Nodes $\epsilon, \zeta, \eta$, shown in red are removed.
-	\csentence{\textbf{c)}} Parent nodes are removed recursively, starting from the root, if all their daughter nodes have more than the threshold number of annotations.
-Nodes in grey ($\epsilon, \zeta, \eta$) were removed in the previous step.
-Nodes $\alpha, \beta$ shown in red are trimmed because each one has a complete daughter set.
-Only nodes $\gamma$ and $\delta$ will be used to generate the static dictionary.
-  }
-\end{figure}
-
-%gui
-\begin{figure}
-	\centering
-    \includegraphics[width=0.95\textwidth]{figures/guiresults.png}
-  	\captionsetup{width= 0.95\textwidth}
- 	\caption{
-	\csentence{Screenshot of results from the web GUI.} 
-	After inputting a gene-list, the user is provided with the results.
-An HTML table is output with hyperlinks to the ontology terms.
-A publication-ready graph is provided below, which can be saved by  dragging to the desktop.
-The graph is colored for better visualization; color is not intended to convey information.
-The graph and the table show anatomy terms in  human-readable format, followed by their unique WBbt ID.
-Finally, lists  of the genes used and discarded for the analysis are also presented.
-  }
-  \label{fig:GUIresults}
-\end{figure}
-
-% workflow 
-\begin{figure}
-	\centering
-	\includegraphics[width=0.95\textwidth]{figures/workflow.pdf}
-	\captionsetup{width= 0.95\textwidth}
-	\caption{
-	\csentence{TEA Workflow.}
-	The complete ontology is annotated continuously by WormBase curators.
-After each update, the ontology is processed to remove uninformative terms.
-The trimmed ontology is used for statistical testing.
-	Users can select a gene list and input it into our tool using our WormBase portal.
-The gene list is tested for enrichment using the trimmed ontology, and results are output in tabular and graphic formats for analysis.
-	}
-	\label{fig:workflow}
-\end{figure}
-
-% statistical testing and validation Figures
-%q plot
-\begin{figure}[h!]
-	\includegraphics[width=\textwidth]{avgQKDE_method=any.pdf}
-  \captionsetup{width= 0.95\textwidth}
-  \caption{
-  \csentence{Kernel density estimates for 30 gold standard datasets.}
-      We ran TEA on 30 datasets we believed to be enriched in particular tissues and pooled all the results to observe the distribution of q-values.
-The mode of the distribution for dictionaries with annotation cut-offs of 100 and 50 genes are very similar; however, when the cut-off is lowered to 25 genes, the mode of the distribution shifts to the left, potentially signalling a decrease in measurement power.
-	  }
-	  \label{fig:qvals}
-\end{figure}
-
-%gaba
-\begin{figure}
-  \includegraphics[width=0.95\textwidth]{GABAcomparison3.pdf}
-  \captionsetup{width= 0.95\textwidth}
-  \caption{
-  \csentence{Independently derived gene sets show similar results 	when tested with the same dictionary.}
-  \textbf{Set 1)} GABAergic gene set from Watson~\cite{Watson2008a}.
-  \textbf{Set 2)} GABAergic gene set from Spencer~\cite{Spencer2011}.
-  Arrowheads highlight identical terms between both analyses. All terms refer to neurons or neuronal tissues and are GABA-associated. Dictionary with cutoff: 33; threshold: 0.95; method: `any'.
-   }
-  \label{fig:intragree}
-\end{figure}
-
-%results Figures - engelmann et al reanalysis
-%labels are missing.
-\begin{figure}[h!]
-    \includegraphics[width=0.95\textwidth]{figures/engelmann-GO-tea-comparison.png}
-	\captionsetup{width= 0.95\textwidth}
-  	\caption{
-	\csentence{\emph{D.~coniospora} Gene Enrichment Analysis and Tissue Enrichment Analysis results.}
-	We compared and contrasted the results from a gene enrichment analysis program, pantherDB, with TEA by analyzing genes that were significantly down-regulated when \emph{C.~elegans} was exposed to \emph{D.~coniospora} in a previously published dataset by Engelmann \emph{et al}~\cite{Engelmann2011} with both tools.
-	\textbf{a)} pantherDB screenshot of results, sorted by p-value. Only top hits shown.
-	\textbf{b)} TEA results, sorted by q-value (lowest on top) and fold-change. 
-	Both pantherDB and TEA identify terms associated with neurons (red square). The two analyses provide complementary, not redundant, information.
-	}
-	\label{fig:Dcon}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Tables                        %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Use of \listoftables is discouraged.
-%%
-\section*{Tables}
-
-\begin{table}[h!]
-	\caption{
-	Parameter specifications and number of tissues for all dictionaries.
-The `Method' column refers to the trimming criterion for the similarity metric.
-We used two such criteria, `any' and `avg'.`any': For a given sister set, if any sister had a similarity exceeding the corresponding threshold, all sisters were removed from the final dictionary.
-`avg': For a given sister set, if the average similarity across all the sisters in the set was greater than the corresponding threshold, all sisters were removed from the final dictionary.
-	}
-	\csvreader[tabular= c | c | c | c,
-	table head= {\textbf{Annotation Cutoff}} & {\textbf{Similarity Threshold}} & {\textbf{Method}} & {\textbf{No. Of Terms in Dictionary}}\\\hline,
-	late after line= \\,
-	late after last line= \\\hline
-	]
-	{figures/TissueNumbers.csv}{}
-	{\csvcoli & \csvcolii & \csvcoliii & \csvcoliv}
-
-	% \csvautotabular{figures/TissueNumbers.csv}
-	\label{tab:DictionarySpecs}
-\end{table}
-
-\begin{table}[h!]
-	\caption{
-	Comparison of results for a neuronal-enriched gene set from Watson~\cite{Watson2008a}.
-We ran the same genelist on a dictionary with a minimum annotation cutoff of 50, similarity threshold of 0.95 and similarity method `any' versus another with a minimum annotation cutoff of 33, similarity threshold of 0.95 and similarity method `any'.
-In the table, columns are labeled with their significance value (Q-value) or enrichment fold change followed by a hyphen and a number which indicates which the cutoff for the dictionary that was used for testing.
-Not all tissues are present in either dictionary.
-Hyphens denote not-applicable values, which occurs when a particular tissue is not present in both dictionaries.
-Full table is available on github.
-	}
-
-	\begin{adjustbox}{width=1\textwidth}
-		\centering
-		% \csvreader{figures/dict-comparison-50-33.csv}
-		\csvreader[tabular={ r | c | c | c | c},
-	    table head ={\textbf{Tissue}} & {\textbf{Q-value-33}} & {\textbf{Q-value-50}} & {\textbf{Enrichment-33}} & {\textbf{Enrichment-50}}\\\hline,
-	    late after line= \\,
-		late after last line=\\\hline
-		]%
-		{figures/dict-comparison-50-33.csv}{}	
-		{\csvcoli & \csvcolii & \csvcoliii & \csvcoliv & \csvcolv}	
-		\label{tab:interagree}
-	\end{adjustbox}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Additional Files              %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Additional Files}
-  \subsection*{Additional file 1 --- IPython Notebook}
-    Tutorial for users interested in using our software within a python script
-
-\end{backmatter}
 \end{document}

### 2016-04-19 tests/hgf_benchmarking/doc/tea_main_text.tex
removed sns.set_palette
wasn’t working, dunno why, no time to troubleshoot.

@@
 with controlled relations between terms) with genes and to look for enrichment of specific terms.
-Although gene ontology (GO) is available for   \emph{Caenorhabditis~elegans}, it does not include anatomy information.
+Although gene ontology (GO) is available for   \emph{Caenorhabditis~elegans}, it does not include anatomical information.
 \parttitle{Results} %if any 
@@
 Since a common drawback to ontology enrichment analyses is result verbosity, we developed a very simple trimming algorithm to reduce the ontology size by an order of magnitude.
-We adjusted these filters and validated our tool using a set of 30  gold standards from Expression Cluster data in WormBase.
+We adjusted these filters and validated our tool using a set of 30  gold standards from Expression Cluster data in WormBase and we show our tool can even discriminate between embryonic and larval tissues and can even identify tissues down to the single-cell level.
 We used our tool to identify multiple neuronal tissues that are down-regulated due to pathogen infection in \emph{C.~elegans}.
@@
 It tests a slimmed-down \emph{C.~elegans} tissue ontology for enrichment of specific terms and provides users with a  text and graphic representation of the results.
-Our tool can discriminate between embryonic and larval tissues and can even identify tissues down to the single-cell level.
 \end{abstract}
@@
 GO provides a controlled set of hierarchically ordered terms~\cite{TheGeneOntologyConsortium2000a, TheGeneOntologyConsortium2015} that provide detailed descriptions about the molecular, cellular or biochemical functions of any gene.
-For a given gene list, certain software programs can query whether a particular term is enriched~\cite{Mi2009, McLean2010, Huang2009}.
+For a given gene list, existing software programs can query whether a particular term is enriched~\cite{Mi2009, McLean2010, Huang2009}.
 One area of biological significance that GO does not include is anatomy.
 One way to address this shortcoming is to use a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `specific cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}.
+Cells and tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation.
 The \emph{C.~elegans} database, WormBase~\cite{Howe2016}, maintains a curated list of gene expression data from the literature.
 Here we provide a new framework that analyzes a user-input list for enrichment of specific cells and tissues.
-They are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation.
 %As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.
@@
 Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene.
-There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.
-DAVID, a common tool for GO enrichment analysis, clusters enriched terms into broad categories~\cite{Huang2007}, whereas PANTHER~\cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (H.
+%There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.
+DAVID, a common tool for GO enrichment analysis, clusters enriched terms into broad categories~\cite{Huang2007}, whereas PANTHER~\cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (pers.
+comm., H.
 Yu and P.
-Thomas, pers.
-comm.).
-
+Thomas).
 To reduce verbosity, we have filtered our ontology using a small set of well-defined criteria to remove terms that do not contribute additional information.
@@
 Genes that have a q-value less than a given alpha are considered significant.
-Our default setting is to set alpha at 0.1.
+Our default setting is an alpha of 0.1.
 
@@
 These names are processed into standard WormBase gene IDs (WBGene IDs).
-The program returns a table containing all the enriched terms and associated useful information%with significant tissues, their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change ($\text{observed hits} /\text{expected hits}$)
-.
+The program returns a table containing all the enriched terms and associated  information such as number of terms in gene list and expected number of terms.%with significant tissues, their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change ($\text{observed hits} /\text{expected hits}$)
  Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values.
@@
 \subsection*{Validation of the algorithm and optimizing parameter selection}
-We wanted to select a dictionary that included enough terms to be specific beyond the largest \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change.
+We wanted to select a dictionary that included enough terms to be specific beyond the most basic %largest -> basic
+ \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change.
 Larger tissues are correlated with better annotation, so increasing term specificity is associated with losses in statistical power.
@@
 
-As a first attempt to select a dictionary, we generated all possible combinations of dictionaries with minimal annotations of 10, 25, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any' similarity thresholding methods (see Table~\ref{tab:DictionarySpecs}).
-The number of tissues was inversely correlated to the minimum annotation cutoff, and was largely insensitive to the similarity threshold in the range we explored.
+As a first attempt to select a dictionary, we generated all possible combinations of dictionaries with minimal annotations of 10, 25, 33, 50 and 100 genes and similarity cutoffs of 0.9, 0.95 and 1, using `avg' or `any' similarity thresholding methods (see Table~\ref{tab:DictionarySpecs}).
+The number of remaining ontology terms was inversely correlated to the minimum annotation cutoff, and was largely insensitive to the similarity threshold in the range we explored.
 Next, we analyzed all 30 datasets using each dictionary.
@@
 The average q-value increased with decreasing annotation cut-off (see Figure~\ref{fig:qvals}), which reflects the decreasing statistical power associated with fewer annotations per term, but we remained agnostic as to how significant is the trade-off between power and term specificity.
-Based on these observations, we ruled out the dictionary with the 100 gene annotation cut-off: it had the fewest terms and its q-values were not low enough in our opinion to compensate the trade-off in specificity.
+Based on these observations, we ruled out the dictionary with the 100 gene annotation cut-off: it had the fewest terms and its q-values were not low enough in our opinion to compensate for the trade-off in specificity.
 
@@
 
-\section*{Applying the tool}
+\subsection*{Applying the tool}
 We applied our tool to the RNA-seq datasets developed by Engelmann et al.~\cite{Engelmann2011} to gain further understanding of their underlying biology.
@@
 As an independent verification, we also performed GO analysis using PANTHER on the down-regulated genes for \emph{D.~coniospora}.
-These results showed enrichment in terms associated with neurons (see Figure~\ref{fig:Dcon}).
+These results also showed enrichment in terms associated with neurons (see Figure~\ref{fig:Dcon}).
 A possible explanation for this neuronal association might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen.
@@
 Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for certain studies.
-Moreover, this tissue enrichment tool generally provides users with enrichment results for 6 large tissues.
+Finally, this tissue enrichment tool provides users with enrichment results for only 6 large tissues.
 In contrast, our tool routinely tests a much larger number of terms, and we have shown it can even accurately identify enrichment of embryonic precursor lineages for select data sets.
@@
 Due to the nature of all ontologies as hierarchical, acyclical graphs with term inheritance, term annotations are correlated along any given branch.
-This correlation reduces the benefits of including all terms for statistical analysis: for any given term along a branch, if that term passes significance, there is a high probability that many other terms along that branch will also pass significant.
+This correlation reduces the benefits of including all terms for statistical analysis: for any given term along a branch, if that term passes significance, there is a high probability that many other terms along that branch will also pass significance.
 If the branch is enriched by random chance, error propagation along a branch means that many more false positives will follow.
@@
 
-We have endeavoured to benchmark our tool well. However, our analysis could not identify %identify good negative controls
+We endeavoured to benchmark our tool well, but our analysis cannot address problems related to %identify good negative controls
  spurious term enrichment.

### 2016-04-19 tissue_enrichment_analysis/hypergeometricTests.py

@@ -281,7 +281,7 @@ def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
     import seaborn as sns
     sns.set_context('paper')
     sns.set_style('whitegrid')
-    sns.choose_colorbrewer_palette('sequential', as_cmap=False)
+    # sns.choose_colorbrewer_palette('sequential', as_cmap=False)
     if df.empty:
         print('dataframe is empty!')
         return

### 2016-04-19 tests/hgf_benchmarking/doc/SI/FolderStructure.tex

@@
 \texttt{WBPaper00013489\_Ray\_Enriched\_WBbt\_0006941\_25.csv}\\
-refers to the \textbf{WormBase Paper 0013489}, which should be `\textbf{Ray Enriched}', and specifically should be enriched in term `\textbf{WBbt:0006941}'.
+refers to the \textbf{WormBase Paper 0013489}, which should be enriched in `\textbf{Ray Enriched (WBbt:0006941)}' and contains \textbf{25} genes.
 The gene set that was used for this analysis is contained in the SI folder named `\textbf{golden gene sets}' and is contained in a homologously named csv file.
@@
 Alternatively, when comparing two dictionaries, the nomenclature is:
-neuronal\_comparison\_GABAergic\_33-50_WBPaper0037950_complete.csv
+neuronal\_comparison\_GABAergic\_33-50\_WBPaper0037950\_complete.csv
 Which refers to a `GABAergic neuronal' lists, comparing dictionaries with cutoffs of 33-50 (all other parameters are the same) using paper 37950.
@@
 
-\subsection*{test_list_efaecalis.txt}
+\subsection*{test\_list\_efaecalis.txt}
 

### 2016-04-20 setup.py
minor changes

@@ -10,7 +10,7 @@ from setuptools import setup, find_packages
 import os
 import sys
 
-version = '0.13.12'
+version = '0.13.13'
 
 # just type in python setup.py publish and this takes care of publishing to pypi!
 

### 2016-04-25 tests/hgf_benchmarking/doc/tea_main_text.tex
uploading paper and updating ipy notebook

@@
 
-\title{Tissue Enrichment Analysis (TEA)}
+\title{Tissue Enrichment Analysis for C. elegans Genomics}
 
@@
 \parttitle{Background} %if any
-Over the last ten years, there has been explosive development in tools
-for measuring gene expression. These tools can identify thousands of altered 
+Over the last ten years, there has been explosive development in methods
+for measuring gene expression. These methods can identify thousands of altered 
 genes between conditions, but understanding these datasets and forming hypotheses based on them remains challenging.
@@
 \parttitle{Conclusions} %if any
-Our Tissue Enrichment Analysis (TEA) can be found at \url{www.wormbase.org/tea}, and can be downloaded using Python's standard pip installer.
+Our Tissue Enrichment Analysis (TEA) can be found within WormBase, and can be downloaded using Python's standard pip installer.
 It tests a slimmed-down \emph{C.~elegans} tissue ontology for enrichment of specific terms and provides users with a  text and graphic representation of the results.
@@
 To our knowledge, such filtering has not been performed in an algorithmic fashion for a biological ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
-Other pruning tools do exist (see for example~\cite{Kim2007}), but the pruning is query-dependent.
+Other pruning methods do exist (see for example~\cite{Kim2007}), but the pruning is query-dependent.
 We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.
@@
 If it does not, the node is removed.
-The next higher node in the branch is tested and removed recursively until a node which satisfies the condition is found.
+The next higher node in the branch is tested and removed recursively until a node that satisfies the condition is found.
 At that point, no more nodes can be removed from that branch.
-This is guaranteed by the structure of the ontology: parent nodes inherit all of the annotations of all of their descendants, so the number of annotated terms monotonically increases with increasing term hierarchy (see Figure~\ref{fig:trim_ends}).
+This completion is guaranteed by the structure of the ontology: parent nodes inherit all of the annotations of all of their descendants, so the number of annotated terms monotonically increases with increasing term hierarchy (see Figure~\ref{fig:trim_ends}).
 In this way, we ensure that our term dictionary includes only those tissues that are considered sufficiently well annotated for statistical purposes.
@@
 Moreover, classifiers require that data be rank-ordered by some metric, something which is not possible for certain studies.
-Finally, this tissue enrichment tool provides users with enrichment results for only 6 large tissues.
+Furthermore, this tissue enrichment tool provides users with enrichment results for only 6 large tissues.
 In contrast, our tool routinely tests a much larger number of terms, and we have shown it can even accurately identify enrichment of embryonic precursor lineages for select data sets.
@@
 \section*{Availability of data and materials}
-Our web implementation is available at \url{https://www.wormbase.org/tea}. Our software can also be downloaded using Python's pip installer via the command
+Our web implementation is available at \url{https://www.wormbase.org/enrichment/tea}. Our software can also be downloaded using Python's pip installer via the command
 
@@
 \bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )
+\bibliography{tea_citations}      % Bibliography file (usually '*.bib' )
 % for author-year bibliography (bmc-mathphys or spbasic)
@@
   \caption{
-  \csentence{Kernel density estimates for 30 gold standard datasets.}
+  \csentence{Kernel density estimates (KDE) for 30 gold standard datasets.}
       We ran TEA on 30 datasets we believed to be enriched in particular tissues and pooled all the results to observe the distribution of q-values.
@@
 Hyphens denote not-applicable values, which occurs when a particular tissue is not present in both dictionaries.
-Full table is available on github.
+Full table is available in Additional File 2.
 	}
@@
     Tutorial for users interested in using our software within a python script
+  \subsection*{Additional file 2 --- Folder Structure for SI files 3 and 4}
+    A file detailing the folder structure of the zipped folders 3 and 4. 
+  \subsection*{Additional file 3 --- Golden Gene Sets}
+    A list of all the genes used for our benchmarking process. 
+  \subsection*{Additional file 4 --- Results}
+    A folder containing a complete version of the results we generated for this paper.
 

### 2016-04-30 tests/hgf_benchmarking/doc/tea_main_text.tex
writing
stupid bmc

@@
 \begin{backmatter}
+\section*{Abbreviations}
+
+\begin{itemize}
+	\item TEA --- Tissue Enrichment Analysis
+	\item GO --- Gene Ontology
+	\item WBbt ID --- A unique ID assigned to reference ontology terms
+	\item WBgene ID -- A unique ID assigned to reference nematode genes
+\end{itemize}
+
+\section*{Ethics Approval and Consent to Participate}
+
+Not applicable
+
+\section*{Consent for Publication}
+
+Not applicable
 
@@
 	lab for helpful discussion.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                               %%
+%% Additional Files              %%
+%%                               %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Additional Files}
+  \subsection*{Additional file 1 --- TEA Tutorial}
+    Tutorial for users interested in using our software within a python script
+  \subsection*{Additional file 2 --- Folder Structure for SI files 3 and 4}
+    A file detailing the folder structure of the zipped folders 3 and 4. 
+  \subsection*{Additional file 3 --- Golden Gene Sets}
+    A list of all the genes used for our benchmarking process. 
+  \subsection*{Additional file 4 --- Results}
+    A folder containing a complete version of the results we generated for this paper.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % Figures describing algorithm
-\begin{figure}[h!]
+\begin{figure}[htbp]
   \begin{subfigure}[b]{0.3\textwidth}
-    \includegraphics[width=\textwidth]{figures/RedundancyTrimmingOntogeny.png}
+    % \includegraphics[width=\textwidth]{figures/RedundancyTrimmingOntogeny.png}
    	 	\caption{}
@@
   \begin{subfigure}[b]{0.3\textwidth}
-    \includegraphics[width=\textwidth]{figures/FloorTrimmingOntogeny.png}
+    % \includegraphics[width=\textwidth]{figures/FloorTrimmingOntogeny.png}
 	      \caption{}
@@
   \begin{subfigure}[b]{0.3\textwidth}
-    \includegraphics[width=\textwidth]{figures/CeilingTrimmingOntogeny.png}
+    % \includegraphics[width=\textwidth]{figures/CeilingTrimmingOntogeny.png}
     	\caption{}
@@
 %gui
-\begin{figure}
+\begin{figure}[htbp]
 	\centering
-    \includegraphics[width=0.95\textwidth]{figures/guiresults.png}
+    % \includegraphics[width=0.95\textwidth]{figures/guiresults.png}
   	\captionsetup{width= 0.95\textwidth}
@@
 % workflow 
-\begin{figure}
+\begin{figure}[htbp]
 	\centering
-	\includegraphics[width=0.95\textwidth]{figures/workflow.pdf}
+	% \includegraphics[width=0.95\textwidth]{figures/workflow.pdf}
 	\captionsetup{width= 0.95\textwidth}
@@
 %q plot
-\begin{figure}[h!]
-	\includegraphics[width=\textwidth]{avgQKDE_method=any.pdf}
+\begin{figure}[htbp]
+	% \includegraphics[width=\textwidth]{avgQKDE_method=any.pdf}
   \captionsetup{width= 0.95\textwidth}
@@
 \begin{figure}
-  \includegraphics[width=0.95\textwidth]{GABAcomparison3.pdf}
+  % \includegraphics[width=0.95\textwidth]{GABAcomparison3.pdf}
   \captionsetup{width= 0.95\textwidth}
@@
 %labels are missing.
-\begin{figure}[h!]
-    \includegraphics[width=0.95\textwidth]{figures/engelmann-GO-tea-comparison.png}
+\begin{figure}[htbp]
+    % \includegraphics[width=0.95\textwidth]{figures/engelmann-GO-tea-comparison.png}
 	\captionsetup{width= 0.95\textwidth}
@@
 %% Use of \listoftables is discouraged.
-%%
+%
+
 \section*{Tables}
 
-\begin{table}[h!]
+\begin{table}[p]
 	\caption{
@@
 	}
-	\csvreader[tabular= c | c | c | c,
-	table head= {\textbf{Annotation Cutoff}} & {\textbf{Similarity Threshold}} & {\textbf{Method}} & {\textbf{No. Of Terms in Dictionary}}\\\hline,
-	late after line= \\,
-	late after last line= \\\hline
-	]
-	{figures/TissueNumbers.csv}{}
-	{\csvcoli & \csvcolii & \csvcoliii & \csvcoliv}
+	% \csvreader[tabular= c | c | c | c,
+	% table head= {\textbf{Annotation Cutoff}} & {\textbf{Similarity Threshold}} & {\textbf{Method}} & {\textbf{No. Of Terms in Dictionary}}\\\hline,
+	% late after line= \\,
+	% late after last line= \\\hline
+	% ]
+	% {figures/TissueNumbers.csv}{}
+	% {\csvcoli & \csvcolii & \csvcoliii & \csvcoliv}
 
-	% \csvautotabular{figures/TissueNumbers.csv}
 	\label{tab:DictionarySpecs}
@@
 
-\begin{table}[h!]
+\begin{table}[p]
 	\caption{
@@
 
-	\begin{adjustbox}{width=1\textwidth}
-		\centering
-		% \csvreader{figures/dict-comparison-50-33.csv}
-		\csvreader[tabular={ r | c | c | c | c},
-	    table head ={\textbf{Tissue}} & {\textbf{Q-value-33}} & {\textbf{Q-value-50}} & {\textbf{Enrichment-33}} & {\textbf{Enrichment-50}}\\\hline,
-	    late after line= \\,
-		late after last line=\\\hline
-		]%
-		{figures/dict-comparison-50-33.csv}{}	
-		{\csvcoli & \csvcolii & \csvcoliii & \csvcoliv & \csvcolv}	
+	% \begin{adjustbox}{width=1\textwidth}
+	% 	\centering
+	% 	% \csvreader{figures/dict-comparison-50-33.csv}
+	%
+	% 	\csvreader[tabular={ r | c | c | c | c},
+	% 		    table head ={\textbf{Tissue}} & {\textbf{Q-value-33}} & {\textbf{Q-value-50}} & {\textbf{Enrichment-33}} & {\textbf{Enrichment-50}}\\\hline,
+	% 		    late after line= \\,
+	% 	late after last line=\\\hline
+	% 	]%
+	% 	{figures/dict-comparison-50-33.csv}{}
+	% 	{\csvcoli & \csvcolii & \csvcoliii & \csvcoliv & \csvcolv}
+	%
+	% \end{adjustbox}
 	\label{tab:interagree}
-	\end{adjustbox}
+
 \end{table}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Additional Files              %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\end{backmatter}
 
-\section*{Additional Files}
-  \subsection*{Additional file 1 --- IPython Notebook}
-    Tutorial for users interested in using our software within a python script
-  \subsection*{Additional file 2 --- Folder Structure for SI files 3 and 4}
-    A file detailing the folder structure of the zipped folders 3 and 4. 
-  \subsection*{Additional file 3 --- Golden Gene Sets}
-    A list of all the genes used for our benchmarking process. 
-  \subsection*{Additional file 4 --- Results}
-    A folder containing a complete version of the results we generated for this paper.
-
-\end{backmatter}
 \end{document}

### 2016-06-29 tests/hgf_benchmarking/doc/tea_main_text.tex
minor changes to manuscript
added a p-value column
bc reviewer asked for it.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
-% <put your article body there>
 
@@
 %% Background %%
-%%
 \section*{Background}
@@
 GO provides a controlled set of hierarchically ordered terms~\cite{TheGeneOntologyConsortium2000a, TheGeneOntologyConsortium2015} that provide detailed descriptions about the molecular, cellular or biochemical functions of any gene.
-For a given gene list, existing software programs can query whether a particular term is enriched~\cite{Mi2009, McLean2010, Huang2009}.
+For a given gene list, existing software programs can query whether a particular term is enriched~\cite{Mi2009, McLean2010, Huang2009, Pathan2015}.
 One area of biological significance that GO does not include is anatomy.
-One way to address this shortcoming is to use a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `specific cell'), in this case for \emph{C.~elegans}~\cite{Lee2003}.
+One way to address this shortcoming is to use a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `specific cell'), in this case for \emph{C.~elegans}.
+Such an ontology has been described previously for this organism~\cite{Lee2003}.
 Cells and tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation.
@@
 Here we provide a new framework that analyzes a user-input list for enrichment of specific cells and tissues.
-%As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.
 
 Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene.
-%There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.
 DAVID, a common tool for GO enrichment analysis, clusters enriched terms into broad categories~\cite{Huang2007}, whereas PANTHER~\cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (pers.
@@
 To our knowledge, such filtering has not been performed in an algorithmic fashion for a biological ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
-Other pruning methods do exist (see for example~\cite{Kim2007}), but the pruning is query-dependent.
+Other pruning methods do exist (see for example~\cite{Kim2007, Mambretti2013}), but the pruning is query-dependent or generates a brand new `brief ontology'.
 We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.
 
-%Our tool is available within WormBase and provides users with a text-based file of the enrichment results as well as a simple and clear graph of the results that exhibit the largest fold-change enrichment.
-
+We have developed a tool that tests a user-provided list of genes for term enrichment using a nematode-specific tissue ontology.
+This ontology, which is not a module of Gene Ontology, is verbose.
+We trim our ontology using an algorithmic approach, outlined below, that reduces multiple hypothesis testing issues by limiting testing to terms that are well-annotated.
+The results are provided to the user in a GUI that includes a table of results and an automatically generated bar-chart.
+This software addresses a previously unmet need in the \emph{C.~elegans} community for a tool that reliably and specifically links gene expression with changes in specific cells, organs or tissues in the worm.
 \section*{Results}
@@
 \subsubsection*{Reducing term redundancy through a similarity metric}
+For our tool, we employ a previously generated cell and tissue ontology for \emph{C.~elegans}\cite{Lee2003}, which is maintained and curated by WormBase.
+This ontology contains thousands of anatomiy terms, but not every term is equally well-annotated.
 As a first step to generate our tissue enrichment software, we wished to select tissue terms that were reasonably well-annotated, yet specific enough to provide insight and not redundant with other terms.
-%We also wanted to avoid testing tissues at levels where annotation redundancy becomes problematic.
-For example, several left and right neurons have at least 25 annotated genes and we may want to include them for enrichment testing.
 For example, nematodes have a number of neurons that are placed symmetrically along the left/right body axis, and are functionally similar.
-These left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them%.
-We reasoned that when two tissues have almost identical annotations
-, and therefore we cannot have statistical confidence in differentiating between them.
-As a result, testing these sister tissues provides no additional information compared with testing only the parent node to these sisters.
-%We refer to such sisters as `redundant'.
-To identify redundancy, we defined a similarity metric (see \emph{Methods} section and Figure~\ref{fig:simdiagram}) that can be used to identify sisters that have very high similarity between them; alternatively, redundant sisters could be identified if a single sister had a very high similarity score.
-We referred to these two scoring criteria as `avg' and `any' respectively.
+These left/right neuronal pairs (which are sisters in the ontology) have almost identical annotations, with at most one or two gene differences between them, and therefore we cannot have statistical confidence in differentiating between them.
+As a result, testing these sister terms provides no additional information compared with testing only the parent node to these sisters.
+To identify redundancy, we defined two possible similarity metrics (see \emph{Methods} section and Figure~\ref{fig:simdiagram}) that can be used to identify ontology sisters that have very high similarity between them.
+Intuitively, a set of sisters can be considered very similar if they share most gene annotations.
+Within a given set of sisters, we can calculate a similarity score for a single node by counting the number of unique annotations it contains and dividing by the total number of unique annotations in the sister set.
+Having assigned to each sister a similarity score, we can identify the \textbf{average} similarity score for this set of sisters, and if this average value exceeds a threshold, all of the sisters are removed from the ontology.
+An alternative method is check whether \textbf{any} of the scores exceeds a predetermined threshold, and if so remove this sister set from the ontology.
+We referred to these two scoring criteria as `\textbf{avg}' and `\textbf{any}' respectively.
 
@@
 These filters were used to compile a static dictionary that we employ for all analyses (see \emph{Validation of the algorithm and parameter selection} section for details).
+Our trimming pipeline is applied with every new WormBase release.
+This ensures that the ontology database we are using remains up-to-date with regards to both addition or removal of specific terms as well as with regard to gene expression annotations.
 
@@
 Genes that have a q-value less than a given alpha are considered significant.
-Our default setting is an alpha of 0.1.
+Our default setting is an alpha of 0.1, which is a standard threshold broadly agreed upon by the RNA-seq community (see for example \cite{Love2014}).
 
@@
 These names are processed into standard WormBase gene IDs (WBGene IDs).
-The program returns a table containing all the enriched terms and associated  information such as number of terms in gene list and expected number of terms.%with significant tissues, their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change ($\text{observed hits} /\text{expected hits}$)
+The program returns a table containing all the enriched terms and associated  information such as number of terms in gene list and expected number of terms.
 Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values.
@@
 \subsection*{Validation of the algorithm and optimizing parameter selection}
-We wanted to select a dictionary that included enough terms to be specific beyond the most basic %largest -> basic
- \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change.
+We wanted to select a dictionary that included enough terms to be specific beyond the most basic \emph{C.~elegans} tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change.
 Larger tissues are correlated with better annotation, so increasing term specificity is associated with losses in statistical power.
-%The phrase below doesn't seem useful
-%To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.
-% Gold-standard data sets were annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
 To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}.
@@
 To select between dictionaries generated between 50, 33 and 25 annotation cut-offs, and also to ensure the terms that are selected as enriched by our algorithm are reasonable, we looked in detail at the enrichment analysis results.
-%Work well
-% TODO add WBbt codes!
 Most results were comparable and expected.
 For some sets, all dictionaries performed well.
-For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} all terms were neuron-related %including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays
-regardless of the dictionary used (see Table~\ref{tab:interagree}).
+For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} all terms were neuron-related regardless of the dictionary used (see Table~\ref{tab:interagree}).
 On the other hand, for a set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the 50 cutoff dictionary was only able to identify `oocyte WBbt:006797' which is not a germline precursor, although it is germline related; whereas the two smaller dictionaries singled out actual germline precursor cells---at the 33 cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as enriched, and at the 25 gene cutoff the embryonic germline precursor terms `P$_4$',`P$_3$' and `P$_2$' were identified in addition to `Z2' and `Z3'.
@@
 We assessed the internal agreement of our tool by using independent gene sets that we expected to be enriched in the same tissues.
-We used two  pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two PVD sets~\cite{Spencer2011, Smith2010}; and two  GABAergic sets~\cite{Spencer2011, Cinar2005}.%; two  pharyngeal gene sets~\cite{Spencer2011, Gaudet2004a}; and two  intestinal gene sets~\cite{Spencer2011, Pauli2006}.
+We used two  pan-neuronal sets~\cite{Spencer2011, Watson2008a}; two PVD sets~\cite{Spencer2011, Smith2010}; and two  GABAergic sets~\cite{Spencer2011, Cinar2005}.
 Overall, the tool has good internal agreement.
 On most sets, the same terms were enriched, although order was somewhat variable (see Table~\ref{tab:interagree}), and most high-scoring terms were preserved between sets.
-%The intestinal gene-sets and pharyngeal gene sets comparisons were exceptions, since at least one gene set showed no enriched terms in all dictionaries, so we did not consider them as informative for assessing internal agreement.
 All comparisons can be found online in our Github repository (see Availability of data and materials).
@@
 Indeed, several studies~\cite{Meisel2014, Zhang2005} have provided evidence that \emph{C.~elegans} uses chemosensory neurons to identify pathogens.
-%One bacterium did not ellicit the same pattern of down-regulation of neuronal-associated genes.
-\emph{E.~faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E.~faecalis} may have a different pathogenic profile.
 Our results highlight the involvement of various \emph{C.~elegans} neuronal tissues in pathogen defense.
@@
 %Trimming discussion
-We have also presented the first, to our knowledge, ontology trimming algorithm for biomedical ontologies.
+We have also presented the first, to our knowledge, ontology trimming algorithm applied to biomedical ontologies.
 This algorithm, which is very easy to execute, places strong limits on the number of terms selected for testing.
@@
 
-We endeavoured to benchmark our tool well, but our analysis cannot address problems related to %identify good negative controls
- spurious term enrichment.
-% Even for our set of positive controls, the statistical analysis sometimes throws out unexpected results.
-For example, the embryonic germline precursor gene set had the term `AB', which refers to the anterior cell in the 2-cell stage \emph{C.~elegans} embryo, as the most enriched term in the dictionaries with cut off of 25 and 33.
-Is this an error, or does this hint at new biology?
+We endeavoured to benchmark our tool well, but our analysis cannot address problems related to spurious term enrichment.
 Although we were unable to determine false-positive and false-negative rates, we do not believe this should deter scientists from using our tool.
@@
 
+Since nodes can have multiple parents (and therefore multiple sister sets), a complete set of similarity scores was calculated before trimming the ontology, and nodes were removed from the ontology if they exceeded the similarity threshold at least once in any comparison.
+
 \subsubsection*{Implementation}
@@
 \section*{Availability of data and materials}
-Our web implementation is available at \url{https://www.wormbase.org/enrichment/tea}. Our software can also be downloaded using Python's pip installer via the command
+Our web implementation is available at \url{http://www.wormbase.org/tools/enrichment/tea/tea.cgi}. Our software can also be downloaded using Python's pip installer via the command
 
@@
   \begin{subfigure}[b]{0.3\textwidth}
-    % \includegraphics[width=\textwidth]{figures/RedundancyTrimmingOntogeny.png}
+    % \includegraphics[width=\textwidth]{figures/Redundancy_Trimming_Ontology.pdf}
    	 	\caption{}
@@
   \begin{subfigure}[b]{0.3\textwidth}
-    % \includegraphics[width=\textwidth]{figures/FloorTrimmingOntogeny.png}
+    % \includegraphics[width=\textwidth]{figures/Floor_Trimming_Ontogeny.pdf}
 	      \caption{}
@@
   \begin{subfigure}[b]{0.3\textwidth}
-    % \includegraphics[width=\textwidth]{figures/CeilingTrimmingOntogeny.png}
+    % \includegraphics[width=\textwidth]{figures/Ceiling_Trimming_Ontology.pdf}
     	\caption{}
@@
 \begin{figure}[htbp]
-    % \includegraphics[width=0.95\textwidth]{figures/engelmann-GO-tea-comparison.png}
+    % \includegraphics[width=0.95\textwidth]{figures/engelmann-GO-tea-comparison.pdf}
 	\captionsetup{width= 0.95\textwidth}

### 2016-06-29 setup.py

@@ -10,7 +10,7 @@ from setuptools import setup, find_packages
 import os
 import sys
 
-version = '0.13.13'
+version = '0.13.16'
 
 # just type in python setup.py publish and this takes care of publishing to pypi!
 

### 2016-06-30 tests/hgf_benchmarking/doc/cover_letter/cover_letter.tex

%\title{Example letter using the newlfm LaTeX package}
%
% See http://texblog.org/2013/11/11/latexs-alternative-letter-class-newlfm/
% and http://www.ctan.org/tex-archive/macros/latex/contrib/newlfm
% for more information.
%
\documentclass[12pt,stdletter,orderfromtodate,sigleft]{newlfm}
\usepackage{blindtext, xfrac}

 \newcommand{\pointRaised}[2]{\medskip \hrule \noindent 
                \textsl{{\fontseries{b} #1}: #2}} 
 \newcommand{\reply}{\noindent \textbf{Reply}:\ }

\newlfmP{dateskipbefore=50pt}
\newlfmP{sigsize=50pt}
\newlfmP{sigskipbefore=50pt}
 \newlfmP{Headlinewd=0pt,Footlinewd=0pt}

\namefrom{Paul W. Sternberg and David Angeles-Albores}
\addrfrom{
    HHMI and Division of Biology and Biological Engineering,
	Caltech
}
\addrto{%
    Dr. Joao Carlos Setubal,
	BMC Bioinformatics
}
\dateset{June 29, 2016}
\greetto{Dear Dr. Setubal}
\closeline{Sincerely,}
\begin{document}
\begin{newlfm}

 Enclosed is our manuscript, titled `Tissue Enrichment Analysis for \emph{C.~elegans} genomics' to be considered for publication in BMC Bioinformatics. We have previously received two reviews and an editorial comment, which we address below.

\section{Editorial Comment}
 \pointRaised{1}{My own comment is that you may want to do a second check in the literature for previous ontology-trimming methods. For example, Garrido et al. Different approaches to build brief ontologies (Volume 348 of the series Communications in Computer and Information Science pp 232-246, Springer, 2013) presents one such methodology. }

 \reply{We thank Dr. Setubal for pointing us towards the book chapter by Garrido \emph{et al}. We have included the citation in the text for completeness. Originally we had not included it because the book chapter in question, like much of the ontology pruning literature, explains methods for generation of a complete ontological structure that is a logical subset of a greater ontology. Our approach does not generate an ontological structure, but rather identifies terms in the ontology that are balanced with regards to specificity (tree depth) and evidence (gene annotations). Once the terms are selected, these terms are not rebuilt into a formal `brief ontology'. However, we recognize the similarity of our work with the Computer Science literature and have included this citation as suggested as a consequence. 
  }

\section{Review 1}

\pointRaised{1}{Authors haven’t discussed how often the background database for the enrichment analysis will be updated. Every time, when new GO terms are added by the GO consortium does the trimming algorithm takes into account these terms and refine the method?
}

\reply{The background dataset for the enrichment analysis will be updated with every version of WormBase. As part of a new WormBase release, the tissue ontology is annotated with new gene expression data, and our trimming pipeline is used to generate a new background dataset. Due to the tight integration of this tool with WormBase, we can guarantee that our database will continuously remain up to date, which is a drawback other popular tools very often suffer from.

We have updated the text to reflect this by adding the following text at the end of the subsection \textbf{Filtering greatly reduces the number of nodes used for analysis}:

``Our trimming pipeline is applied as part of each new WormBase release. This ensures that the ontology database we are using remains up-to-date with regards to both addition or removal of specific terms as well as with regard to gene expression annotations.''

}

\pointRaised{2}{I am skeptical about the usage of this tool as it is only confined and tested with \emph{C.~elegans} genome. Such tools should be always part of the big enrichment analysis tool in the form of plugins rather than as standalone.\\
For example, the recently published `FunRich', a functional enrichment analysis tools has an option of using custom database. Using this custom database, any species genome can be used as background database for functional enrichment analysis which also serves the purpose of identifying GO terms enriched in the input gene lists. 
}

\reply{While we understand the concerns regarding the usage of standalone software, we would like to point out that the cell and tissue ontology referred to in the text has been developed independently of GO. The ontology was developed and is maintained, annotated and extended independently by WormBase. Moreover, due to the species-specific nature of cell and tissue ontologies, at this moment we cannot allow users to input any species genome as a background database. To do this, we would require cell and tissue ontologies for every species in question (available only for some species); ortholog mapping between species (available) and ontology mapping between species (non-existent).

While we do intend to expand TEA to major model systems in the future, we cannot follow the approach that PANTHER takes of identifying orthologs in a background set and using pre-existing GO annotations to identify enrichment. Doing this would yield uninterpretable results, for example if a researcher input a Drosophila background and enrichment set, he/she would at best receive a list of C. elegans tissues. To add tissue enrichment analysis for other species in the future, more cell and tissue ontologies must be developed. Other tissue ontologies do exist at this point in time, such as a Zebrafish ontology.
}

\pointRaised{3}{Majority of the figures is slightly distorted and needs to be replaced with high resolution clear images.}

\reply{We have remade all of the figures using vector-formatting and saved each figure as a PDF for maximum resolution. We thank the reviewer for pointing this out.}

\pointRaised{4}{Citing ‘FunRich’ tool mainly used for the functional enrichment analysis is worth mentioning in the ‘Background’ section of the manuscript.
}

\reply{We have cited `FunRich' and thank reviewer 1 for making us aware of this tool.}

\pointRaised{5}{In the web GUI (http://www.wormbase.org/tools/enrichment/tea/tea.cgi), please include standard list of C.elegans gene names by default in the search box.}

\reply{We have added example genes in the search box.
}

\pointRaised{6}{In the results table, include Hypergeometric p-values in addition to q-values.}

\reply{We have added p-values to the results table, however, we feel that these values add no relevant information for the user, since the table only contains statistically significant results.}

\section{Review 2}

\pointRaised{1}{Introduction needs to be re-organized and partially re-written because it feels a bit strange that details about pruning and trimming are discussed in the last sentences. Instead, the authors need to outline the ideas behind the software and its implementation. A brief outline of what was achieved with the development of this software should also be included.}

\reply{We have re-written the introduction to reflect this. At the end of our section titled \textbf{Background}, we have included the following paragraph:

``We have developed a tool that tests a user-provided list of genes for term enrichment using a nematode-specific tissue ontology. This ontology, which is not a module of Gene Ontology, is verbose. We trim our ontology using an algorithmic approach, outlined below, that reduces multiple hypothesis testing issues by limiting testing to terms that are well-annotated. The results are provided to the user in a GUI that includes a table of results and an automatically generated bar-chart. This software addresses a previously unmet need in the \emph{C.~elegans} community for a tool that reliably and specifically links gene expression with changes in specific cells, organs or tissues in the worm.''
}

\pointRaised{2}{The sources of data need to be better described. This may be obvious for the worm researchers but may be harder to understand for those outside.}

\reply{We would like to thank Reviewer 2 for pointing this out to us, as it had not ocurred to us that non-worm researchers may find the sources of data confusing. We have added the following paragraph explaining the sources of data more extensively in our section \textbf{Reducing term redundancy through a similarity metric}:

``For our tool, we employ a previously generated cell and tissue ontology for \emph{C.~elegans}[6], which is maintained and curated by WormBase.''

}

\pointRaised{3}{The significance and the way to determine the parameter S in the `avg' vs `any' is pretty hard to understand. More illustrations and explanations need to be provided}

\reply{We have expanded our verbal explanation of the similarity criterions to make it more approachable to readers. We did not include additional figures because we did not know how to graphically show the difference between a threshold based on an average score and one based on a supremum score. We added the following text to the subsection \textbf{Reducing term redundancy through a similarity metric}:

``Intuitively, a set of sisters can be considered very similar if they share most gene annotations. Within a given set of sisters, we can calculate a similarity score for a single node by counting the number of unique annotations it contains and dividing by the total number of unique annotations in the sister set. Having assigned to each sister a similarity score, we can identify the \textbf{average} similarity score for this set of sisters, and if this average value exceeds a threshold, all of the sisters are removed from the ontology. An alternative method is check whether \textbf{any} of the scores exceeds a predetermined threshold, and if so remove this sister set from the ontology. We referred to these two scoring criteria as `\textbf{avg}' and `\textbf{any}' respectively.''

}

\pointRaised{4}{Hypergeometric tests. What was the benefit of setting alpha to 0.1 vs 0.05 or another value? Did you find this using some of the established datasets?}

\reply{An alpha of 0.1 is standard for many applications that employ FDR. We didn’t fine-tune this parameter cutoff. We have added a sentence to explicitly point this out in the text, and we have also made it clear that this parameter can be tuned by users in the command-line, but not the web, version. 
}

\end{newlfm}
\end{document}

### 2016-07-01 tests/hgf_benchmarking/doc/reviewr_rebuttals/reviewer_rebuttals.tex
added cover letter, reviewer rebuttals

%\title{Example letter using the newlfm LaTeX package}
%
% See http://texblog.org/2013/11/11/latexs-alternative-letter-class-newlfm/
% and http://www.ctan.org/tex-archive/macros/latex/contrib/newlfm
% for more information.
%
\documentclass[12pt,stdletter,orderfromtodate,sigleft]{newlfm}
\usepackage{blindtext, xfrac}

 \newcommand{\pointRaised}[2]{\medskip \hrule \noindent 
                \textsl{{\fontseries{b} #1}: #2}} 
 \newcommand{\reply}{\noindent \textbf{Reply}:\ }

\newlfmP{dateskipbefore=50pt}
\newlfmP{sigsize=50pt}
\newlfmP{sigskipbefore=50pt}
 \newlfmP{Headlinewd=0pt,Footlinewd=0pt}

\namefrom{Paul W. Sternberg and David Angeles-Albores}
\addrfrom{
    HHMI and Division of Biology and Biological Engineering,
	Caltech
}
\addrto{%
    Dr. Joao Carlos Setubal,
	BMC Bioinformatics
}
\dateset{June 29, 2016}
\greetto{Dear Dr. Setubal}
\closeline{Sincerely,}
\begin{document}
\begin{newlfm}

 % Enclosed is our manuscript, titled `Tissue Enrichment Analysis for \emph{C.~elegans} genomics' to be considered for publication in BMC Bioinformatics. We have previously received two reviews and an editorial comment, which we address below.

We have attached our detailed responses to the comments and reviews to our paper, titled `Tissue Enrichment Analysis for \emph{C.~elegans} Genomics'. We hope our answers are satisfactory.

\section{Editor's Comment}
 \pointRaised{1}{My own comment is that you may want to do a second check in the literature for previous ontology-trimming methods. For example, Garrido et al. Different approaches to build brief ontologies (Volume 348 of the series Communications in Computer and Information Science pp 232-246, Springer, 2013) presents one such methodology. }

 \reply{We thank Dr. Setubal for pointing us towards the book chapter by Garrido \emph{et al}. We have included the citation in the text for completeness. Originally we had not included it because the book chapter in question, like much of the ontology pruning literature, explains methods for generation of a complete ontological structure that is a logical subset of a greater ontology. Our approach does not generate an ontological structure, but rather identifies terms in the ontology that are balanced with regards to specificity (tree depth) and evidence (gene annotations). Once the terms are selected, these terms are not rebuilt into a formal `brief ontology'. However, we recognize the similarity of our work with the Computer Science literature and have included this citation as suggested as a consequence. 
  }

\section{Review 1}

\pointRaised{1}{Authors haven’t discussed how often the background database for the enrichment analysis will be updated. Every time, when new GO terms are added by the GO consortium does the trimming algorithm takes into account these terms and refine the method?
}

\reply{The background dataset for the enrichment analysis will be updated with every version of WormBase every two months. As part of a new WormBase release, the tissue ontology is annotated with new gene expression data, and our trimming pipeline is used to generate a new background dataset. Due to the tight integration of this tool with WormBase, we can guarantee that our database will continuously remain up to date, which is a drawback other popular tools very often suffer from.

We have updated the text to reflect this by adding the following text at the end of the subsection \textbf{Filtering greatly reduces the number of nodes used for analysis}:

``Our trimming pipeline is applied as part of each new WormBase release. This ensures that the ontology database we are using remains up-to-date with regards to both addition or removal of specific terms as well as with regard to gene expression annotations.''

}

\pointRaised{2}{I am skeptical about the usage of this tool as it is only confined and tested with \emph{C.~elegans} genome. Such tools should be always part of the big enrichment analysis tool in the form of plugins rather than as standalone.\\
For example, the recently published `FunRich', a functional enrichment analysis tools has an option of using custom database. Using this custom database, any species genome can be used as background database for functional enrichment analysis which also serves the purpose of identifying GO terms enriched in the input gene lists. 
}

\reply{While we understand the concerns regarding the usage of standalone software, we would like to point out that the cell and tissue ontology referred to in the text has been developed independently of the GO. The ontology was developed and is maintained, annotated and extended independently by WormBase. Moreover, due to the species-specific nature of cell and tissue ontologies, at this moment we cannot allow users to input any species genome as a background database. To do this, we would require cell and tissue ontologies for every species in question (available only for some species); ortholog mapping between species (available) and ontology mapping between species (non-existent).

While we do intend to expand TEA to major model systems in the future, we cannot follow the approach that PANTHER takes of identifying orthologs in a background set and using pre-existing GO annotations to identify enrichment. Doing this would yield uninterpretable results, for example if a researcher input a Drosophila background and enrichment set, he/she would at best receive a list of C. elegans tissues. To add tissue enrichment analysis for other species in the future, more cell and tissue ontologies must be developed. Other tissue ontologies do exist at this point in time, such as a Zebrafish anatomy ontology.
}

\pointRaised{3}{Majority of the figures is slightly distorted and needs to be replaced with high resolution clear images.}

\reply{We have remade all of the figures using vector-formatting and saved each figure as a PDF for maximum resolution. We thank the reviewer for pointing this out.}

\pointRaised{4}{Citing ‘FunRich’ tool mainly used for the functional enrichment analysis is worth mentioning in the ‘Background’ section of the manuscript.
}

\reply{We have cited `FunRich' and thank reviewer 1 for making us aware of this tool.}

\pointRaised{5}{In the web GUI (http://www.wormbase.org/tools/enrichment/tea/tea.cgi), please include standard list of C.elegans gene names by default in the search box.}

\reply{We have added example genes in the search box.
}

\pointRaised{6}{In the results table, include Hypergeometric p-values in addition to q-values.}

\reply{We have added p-values to the results table. However, we feel that these values add no relevant information for the user, since the table only contains statistically significant results as assessed by a q-value cutoff of 0.1.}

\section{Review 2}

\pointRaised{1}{Introduction needs to be re-organized and partially re-written because it feels a bit strange that details about pruning and trimming are discussed in the last sentences. Instead, the authors need to outline the ideas behind the software and its implementation. A brief outline of what was achieved with the development of this software should also be included.}

\reply{We have re-written the introduction to reflect this. At the end of our section titled \textbf{Background}, we have included the following paragraph:

``We have developed a tool that tests a user-provided list of genes for term enrichment using a nematode-specific tissue ontology. This ontology, which is not a module of Gene Ontology, is verbose. We trim our ontology using an algorithmic approach, outlined below, that reduces multiple hypothesis testing issues by limiting testing to terms that are well-annotated. The results are provided to the user in a GUI that includes a table of results and an automatically generated bar-chart. This software addresses a previously unmet need in the \emph{C.~elegans} community for a tool that reliably and specifically links gene expression with changes in specific cells, organs or tissues in the worm.''
}

\pointRaised{2}{The sources of data need to be better described. This may be obvious for the worm researchers but may be harder to understand for those outside.}

\reply{We would like to thank Reviewer 2 for pointing this out to us, as it had not ocurred to us that non-worm researchers may find the sources of data confusing. We have added the following paragraph explaining the sources of data more extensively in our section \textbf{Reducing term redundancy through a similarity metric}:

``For our tool, we employ a previously generated cell and tissue ontology for \emph{C.~elegans}[6], which is maintained and curated by WormBase.''

}

\pointRaised{3}{The significance and the way to determine the parameter S in the `avg' vs `any' is pretty hard to understand. More illustrations and explanations need to be provided}

\reply{We have expanded our verbal explanation of the similarity criterions to make it more approachable to readers. We did not include additional figures because we did not know how to graphically show the difference between a threshold based on an average score and one based on a supremum score. We added the following text to the subsection \textbf{Reducing term redundancy through a similarity metric}:

``Intuitively, a set of sisters can be considered very similar if they share most gene annotations. Within a given set of sisters, we can calculate a similarity score for a single node by counting the number of unique annotations it contains and dividing by the total number of unique annotations in the sister set. Having assigned to each sister a similarity score, we can identify the \textbf{average} similarity score for this set of sisters, and if this average value exceeds a threshold, all of the sisters are removed from the ontology. An alternative method is check whether \textbf{any} of the scores exceeds a predetermined threshold, and if so remove this sister set from the ontology. We referred to these two scoring criteria as `\textbf{avg}' and `\textbf{any}' respectively.''

}

\pointRaised{4}{Hypergeometric tests. What was the benefit of setting alpha to 0.1 vs 0.05 or another value? Did you find this using some of the established datasets?}

\reply{An alpha of 0.1 is standard for many applications that employ FDR. We didn’t fine-tune this parameter cutoff. We have added a sentence to explicitly point this out in the text, and we have also made it clear that this parameter can be tuned by users in the command-line, but not the web, version. 
}

\end{newlfm}
\end{document}

### 2016-07-01 tests/hgf_benchmarking/doc/cover_letter/cover_letter.tex
added cover letter, reviewer rebuttals

@@
 
- Enclosed is our manuscript, titled `Tissue Enrichment Analysis for \emph{C.~elegans} genomics' to be considered for publication in BMC Bioinformatics.
-We have previously received two reviews and an editorial comment, which we address below.
-
-\section{Editorial Comment}
- \pointRaised{1}{My own comment is that you may want to do a second check in the literature for previous ontology-trimming methods.
-For example, Garrido et al.
-Different approaches to build brief ontologies (Volume 348 of the series Communications in Computer and Information Science pp 232-246, Springer, 2013) presents one such methodology.
-}
-
- \reply{We thank Dr.
-Setubal for pointing us towards the book chapter by Garrido \emph{et al}.
-We have included the citation in the text for completeness.
-Originally we had not included it because the book chapter in question, like much of the ontology pruning literature, explains methods for generation of a complete ontological structure that is a logical subset of a greater ontology.
-Our approach does not generate an ontological structure, but rather identifies terms in the ontology that are balanced with regards to specificity (tree depth) and evidence (gene annotations).
-Once the terms are selected, these terms are not rebuilt into a formal `brief ontology'.
-However, we recognize the similarity of our work with the Computer Science literature and have included this citation as suggested as a consequence.
-  }
-
-\section{Review 1}
-
-\pointRaised{1}{Authors haven’t discussed how often the background database for the enrichment analysis will be updated.
-Every time, when new GO terms are added by the GO consortium does the trimming algorithm takes into account these terms and refine the method?
-}
-
-\reply{The background dataset for the enrichment analysis will be updated with every version of WormBase.
-As part of a new WormBase release, the tissue ontology is annotated with new gene expression data, and our trimming pipeline is used to generate a new background dataset.
-Due to the tight integration of this tool with WormBase, we can guarantee that our database will continuously remain up to date, which is a drawback other popular tools very often suffer from.
-
-We have updated the text to reflect this by adding the following text at the end of the subsection \textbf{Filtering greatly reduces the number of nodes used for analysis}:
-
-``Our trimming pipeline is applied as part of each new WormBase release.
-This ensures that the ontology database we are using remains up-to-date with regards to both addition or removal of specific terms as well as with regard to gene expression annotations.''
-
-}
-
-\pointRaised{2}{I am skeptical about the usage of this tool as it is only confined and tested with \emph{C.~elegans} genome.
-Such tools should be always part of the big enrichment analysis tool in the form of plugins rather than as standalone.\\
-For example, the recently published `FunRich', a functional enrichment analysis tools has an option of using custom database.
-Using this custom database, any species genome can be used as background database for functional enrichment analysis which also serves the purpose of identifying GO terms enriched in the input gene lists.
-}
-
-\reply{While we understand the concerns regarding the usage of standalone software, we would like to point out that the cell and tissue ontology referred to in the text has been developed independently of GO.
-The ontology was developed and is maintained, annotated and extended independently by WormBase.
-Moreover, due to the species-specific nature of cell and tissue ontologies, at this moment we cannot allow users to input any species genome as a background database.
-To do this, we would require cell and tissue ontologies for every species in question (available only for some species); ortholog mapping between species (available) and ontology mapping between species (non-existent).
-
-While we do intend to expand TEA to major model systems in the future, we cannot follow the approach that PANTHER takes of identifying orthologs in a background set and using pre-existing GO annotations to identify enrichment.
-Doing this would yield uninterpretable results, for example if a researcher input a Drosophila background and enrichment set, he/she would at best receive a list of C.
-elegans tissues.
-To add tissue enrichment analysis for other species in the future, more cell and tissue ontologies must be developed.
-Other tissue ontologies do exist at this point in time, such as a Zebrafish ontology.
-}
-
-\pointRaised{3}{Majority of the figures is slightly distorted and needs to be replaced with high resolution clear images.}
-
-\reply{We have remade all of the figures using vector-formatting and saved each figure as a PDF for maximum resolution. We thank the reviewer for pointing this out.}
-
-\pointRaised{4}{Citing ‘FunRich’ tool mainly used for the functional enrichment analysis is worth mentioning in the ‘Background’ section of the manuscript.
-}
-
-\reply{We have cited `FunRich' and thank reviewer 1 for making us aware of this tool.}
-
-\pointRaised{5}{In the web GUI (http://www.wormbase.org/tools/enrichment/tea/tea.cgi), please include standard list of C.elegans gene names by default in the search box.}
-
-\reply{We have added example genes in the search box.
-}
-
-\pointRaised{6}{In the results table, include Hypergeometric p-values in addition to q-values.}
-
-\reply{We have added p-values to the results table, however, we feel that these values add no relevant information for the user, since the table only contains statistically significant results.}
-
-\section{Review 2}
-
-\pointRaised{1}{Introduction needs to be re-organized and partially re-written because it feels a bit strange that details about pruning and trimming are discussed in the last sentences.
-Instead, the authors need to outline the ideas behind the software and its implementation.
-A brief outline of what was achieved with the development of this software should also be included.}
-
-\reply{We have re-written the introduction to reflect this. At the end of our section titled \textbf{Background}, we have included the following paragraph:
-
-``We have developed a tool that tests a user-provided list of genes for term enrichment using a nematode-specific tissue ontology.
-This ontology, which is not a module of Gene Ontology, is verbose.
-We trim our ontology using an algorithmic approach, outlined below, that reduces multiple hypothesis testing issues by limiting testing to terms that are well-annotated.
-The results are provided to the user in a GUI that includes a table of results and an automatically generated bar-chart.
-This software addresses a previously unmet need in the \emph{C.~elegans} community for a tool that reliably and specifically links gene expression with changes in specific cells, organs or tissues in the worm.''
-}
-
-\pointRaised{2}{The sources of data need to be better described. This may be obvious for the worm researchers but may be harder to understand for those outside.}
-
-\reply{We would like to thank Reviewer 2 for pointing this out to us, as it had not ocurred to us that non-worm researchers may find the sources of data confusing.
-We have added the following paragraph explaining the sources of data more extensively in our section \textbf{Reducing term redundancy through a similarity metric}:
-
-``For our tool, we employ a previously generated cell and tissue ontology for \emph{C.~elegans}[6], which is maintained and curated by WormBase.''
-
-}
-
-\pointRaised{3}{The significance and the way to determine the parameter S in the `avg' vs `any' is pretty hard to understand. More illustrations and explanations need to be provided}
-
-\reply{We have expanded our verbal explanation of the similarity criterions to make it more approachable to readers.
-We did not include additional figures because we did not know how to graphically show the difference between a threshold based on an average score and one based on a supremum score.
-We added the following text to the subsection \textbf{Reducing term redundancy through a similarity metric}:
-
-``Intuitively, a set of sisters can be considered very similar if they share most gene annotations.
-Within a given set of sisters, we can calculate a similarity score for a single node by counting the number of unique annotations it contains and dividing by the total number of unique annotations in the sister set.
-Having assigned to each sister a similarity score, we can identify the \textbf{average} similarity score for this set of sisters, and if this average value exceeds a threshold, all of the sisters are removed from the ontology.
-An alternative method is check whether \textbf{any} of the scores exceeds a predetermined threshold, and if so remove this sister set from the ontology.
-We referred to these two scoring criteria as `\textbf{avg}' and `\textbf{any}' respectively.''
-
-}
-
-\pointRaised{4}{Hypergeometric tests. What was the benefit of setting alpha to 0.1 vs 0.05 or another value? Did you find this using some of the established datasets?}
-
-\reply{An alpha of 0.1 is standard for many applications that employ FDR.
-We didn’t fine-tune this parameter cutoff.
-We have added a sentence to explicitly point this out in the text, and we have also made it clear that this parameter can be tuned by users in the command-line, but not the web, version.
-}
+ Enclosed is our revised manuscript, titled `Tissue Enrichment Analysis for \emph{C.~elegans} genomics' to be considered for publication in BMC Bioinformatics.
+We have addressed the reviewer comments to the fullest extent of our abilities.
 

### 2016-07-01 tests/hgf_benchmarking/doc/tea_main_text.tex
added cover letter, reviewer rebuttals

@@
 To our knowledge, such filtering has not been performed in an algorithmic fashion for a biological ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
-Other pruning methods do exist (see for example~\cite{Kim2007, Mambretti2013}), but the pruning is query-dependent or generates a brand new `brief ontology'.
+Other pruning methods do exist (see for example~\cite{Kim2007, Mambretti2013}), but the pruning is query-dependent or generates a brand new `brief ontology' which satisfies a set of logic relationships and has certain connectivity requirements.
+We do not propose to regenerate a new `brief ontology', but instead we use our approach to select those nodes that have sufficient annotated evidence for statistical testing.
 We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.
@@
 This ontology, which is not a module of Gene Ontology, is verbose.
-We trim our ontology using an algorithmic approach, outlined below, that reduces multiple hypothesis testing issues by limiting testing to terms that are well-annotated.
+We select nodes from the ontology for statistical testing using an algorithmic approach, outlined below, that reduces multiple hypothesis testing issues by limiting testing to terms that are well-annotated.
 The results are provided to the user in a GUI that includes a table of results and an automatically generated bar-chart.
 This software addresses a previously unmet need in the \emph{C.~elegans} community for a tool that reliably and specifically links gene expression with changes in specific cells, organs or tissues in the worm.
+
 \section*{Results}
-\subsection*{Generating a Gene-Tissue Dictionary}
+\subsection*{Generating a Gene-Tissue Dictionary by Specific Node Selection}
 \subsubsection*{Reducing term redundancy through a similarity metric}
@@
 Within a given set of sisters, we can calculate a similarity score for a single node by counting the number of unique annotations it contains and dividing by the total number of unique annotations in the sister set.
-Having assigned to each sister a similarity score, we can identify the \textbf{average} similarity score for this set of sisters, and if this average value exceeds a threshold, all of the sisters are removed from the ontology.
+Having assigned to each sister a similarity score, we can identify the \textbf{average} similarity score for this set of sisters, and if this average value exceeds a threshold, these sisters are not considered testable candidates.
 An alternative method is check whether \textbf{any} of the scores exceeds a predetermined threshold, and if so remove this sister set from the ontology.
@@
 Many nodes have 0-10 annotations, which we consider too few to accurately test.
-To solve this issue, we implemented a straightforward trimming strategy.
+To solve this issue, we implemented another straightforward node selection strategy.
 For a given terminal node, we test whether the node has more than a threshold number of annotations.
-If it does not, the node is removed.
+If it does not, the node is not used for statistical testing.
 The next higher node in the branch is tested and removed recursively until a node that satisfies the condition is found.
@@
 Additionally, we reasoned that, for any parent node, if all its daughters were selected for testing, there was no additional benefit to test the parent.
-%In other words, if all the daughter nodes are tested, there is little additional information to be gained by including the parent node.
-To address this issue
 We removed parent nodes from the analysis if all their daughter nodes passed the annotation threshold (see Figure~\ref{fig:trim_roots}).
@@
 Applying these three filters reduced the number of ontology terms by an order of magnitude.
-
-%Finally, we also wanted to remove as many terms as possible from the dictionary with the goals of reducing covariance between terms and decreasing multiple testing.
-Decreasing covariance between terms is important because we employ a frequentist approach that assumes all terms are independent.
-Covariation  between terms means that if one of these tissues tests significant, the other terms are much more likely to pass significance testing as well.
-This makes adequate correction for false positive rates considerably more difficult.
 
@@
 These filters were used to compile a static dictionary that we employ for all analyses (see \emph{Validation of the algorithm and parameter selection} section for details).
-Our trimming pipeline is applied with every new WormBase release.
+Our trimming pipeline is applied as part of each new WormBase release.
 This ensures that the ontology database we are using remains up-to-date with regards to both addition or removal of specific terms as well as with regard to gene expression annotations.
@@
 Our default setting is an alpha of 0.1, which is a standard threshold broadly agreed upon by the RNA-seq community (see for example \cite{Love2014}).
+This threshold cannot be altered in the web GUI, but is user tunable through our command-line implementation.
 
@@
 	The complete ontology is annotated continuously by WormBase curators.
-After each update, the ontology is processed to remove uninformative terms.
-The trimmed ontology is used for statistical testing.
+After each update, the ontology is processed to remove uninformative terms, and the remaining terms are used for statistical testing.
 	Users can select a gene list and input it into our tool using our WormBase portal.

### 2016-08-26 tests/hgf_benchmarking/doc/tea_main_text.tex
remove wbids from marplot
because they are super annoying.

@@
 \RequirePackage{natbib}
+\usepackage[none]{hyphenat}
 % \RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
@@
 Over the last ten years, there has been explosive development in methods
-for measuring gene expression. These methods can identify thousands of altered 
-genes between conditions, but understanding these datasets and forming hypotheses based on them remains challenging.
+for measuring gene expression. These methods can identify thousands of  
+genes altered between conditions, but understanding these datasets and forming hypotheses based on them remains challenging.
 One way to analyze these datasets is to associate ontologies (hierarchical, descriptive vocabularies
 with controlled relations between terms) with genes and to look for enrichment of specific terms.
-Although gene ontology (GO) is available for   \emph{Caenorhabditis~elegans}, it does not include anatomical information.
+Although Gene Ontology (GO) is available for   \emph{Caenorhabditis~elegans}, it does not include anatomical information.
 \parttitle{Results} %if any 
 We have developed a tool for identifying  enrichment of \emph{C.~elegans} tissues among gene sets and generated a website GUI where users can access this tool.
-Since a common drawback to ontology enrichment analyses is result verbosity, we developed a very simple trimming algorithm to reduce the ontology size by an order of magnitude.
-We adjusted these filters and validated our tool using a set of 30  gold standards from Expression Cluster data in WormBase and we show our tool can even discriminate between embryonic and larval tissues and can even identify tissues down to the single-cell level.
+Since a common drawback to ontology enrichment analyses is its verbosity, we developed a very simple filtering algorithm to reduce the ontology size by an order of magnitude.
+We adjusted these filters and validated our tool using a set of 30  gold standards from Expression Cluster data in WormBase.
+We show our tool can even discriminate between embryonic and larval tissues and can even identify tissues down to the single-cell level.
 We used our tool to identify multiple neuronal tissues that are down-regulated due to pathogen infection in \emph{C.~elegans}.
@@
 One area of biological significance that GO does not include is anatomy.
-One way to address this shortcoming is to use a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `specific cell'), in this case for \emph{C.~elegans}.
+One way to address this shortcoming is to use a `tissue ontology' that provides a complete anatomical description for an organism (e.g.`tissue', `organ' or `specific cell'), in this case for \emph{C.~elegans}.
 Such an ontology has been described previously for this organism~\cite{Lee2003}.
@@
 
-Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene.
+Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene (which we refer to as `result verbosity').
 DAVID, a common tool for GO enrichment analysis, clusters enriched terms into broad categories~\cite{Huang2007}, whereas PANTHER~\cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (pers.
@@
 To reduce verbosity, we have filtered our ontology using a small set of well-defined criteria to remove terms that do not contribute additional information.
-To our knowledge, such filtering has not been performed in an algorithmic fashion for a biological ontology before---indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
-Other pruning methods do exist (see for example~\cite{Kim2007, Mambretti2013}), but the pruning is query-dependent or generates a brand new `brief ontology' which satisfies a set of logic relationships and has certain connectivity requirements.
+To our knowledge, such filtering has not been performed in an algorithmic fashion for a biological ontology before; indeed, DAVID does not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms.
+Other pruning methods do exist (see for example~\cite{Kim2007, Garrido2012}), but the pruning is query-dependent or generates a brand new `brief ontology' which satisfies a set of logic relationships and has certain connectivity requirements.
 We do not propose to regenerate a new `brief ontology', but instead we use our approach to select those nodes that have sufficient annotated evidence for statistical testing.
@@
 \subsubsection*{Reducing term redundancy through a similarity metric}
-For our tool, we employ a previously generated cell and tissue ontology for \emph{C.~elegans}\cite{Lee2003}, which is maintained and curated by WormBase.
+For our tool, we employ a previously generated cell and tissue ontology for \emph{C.~elegans}~\cite{Lee2003}, which is maintained and curated by WormBase.
 This ontology contains thousands of anatomiy terms, but not every term is equally well-annotated.
@@
 
-Additionally, we reasoned that, for any parent node, if all its daughters were selected for testing, there was no additional benefit to test the parent.
+Additionally, we reasoned that for any parent node if all its daughters were selected for testing, there was no additional benefit to test the parent.
 We removed parent nodes from the analysis if all their daughter nodes passed the annotation threshold (see Figure~\ref{fig:trim_roots}).
@@
 Genes that have a q-value less than a given alpha are considered significant.
-Our default setting is an alpha of 0.1, which is a standard threshold broadly agreed upon by the RNA-seq community (see for example \cite{Love2014}).
+Our default setting is an alpha of 0.1, which is a standard threshold broadly agreed upon by the scientific community (see for example~\cite{Love2014, Pawitan2005, Storey2003}).
 This threshold cannot be altered in the web GUI, but is user tunable through our command-line implementation.
@@
 For some sets, all dictionaries performed well.
-For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} all terms were neuron-related regardless of the dictionary used (see Table~\ref{tab:interagree}).
-On the other hand, for a set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the 50 cutoff dictionary was only able to identify `oocyte WBbt:006797' which is not a germline precursor, although it is germline related; whereas the two smaller dictionaries singled out actual germline precursor cells---at the 33 cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as enriched, and at the 25 gene cutoff the embryonic germline precursor terms `P$_4$',`P$_3$' and `P$_2$' were identified in addition to `Z2' and `Z3'.
+For example, in our `all neuron enriched sets'~\cite{Spencer2011, Watson2008a} all terms were neuron-related regardless of the dictionary used (see Table~\ref{tab:intraagree}).
+On the other hand, for a set enriched for germline precursor expression in the embryo~\cite{Spencer2011}, the 50 cutoff dictionary was only able to identify `oocyte WBbt:006797', which is not a germline precursor although it is germline related; whereas the two smaller dictionaries singled out actual germline precursor cells---at the 33 cutoff, our tool identified the larval germline precursor cells `Z2' and `Z3' as enriched, and at the 25 gene cutoff the embryonic germline precursor terms `P$_4$',`P$_3$' and `P$_2$' were identified in addition to `Z2' and `Z3'.
 %Differences between dictionaries
@@
 Overall, the tool has good internal agreement.
-On most sets, the same terms were enriched, although order was somewhat variable (see Table~\ref{tab:interagree}), and most high-scoring terms were preserved between sets.
+On most sets, the same terms were enriched, although order was somewhat variable (see Table~\ref{fig:interagree}), and most high-scoring terms were preserved between sets.
 All comparisons can be found online in our Github repository (see Availability of data and materials).
@@
 We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated.
-Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade.
+Initially we noticed that genes that are down-regulated tend to be twice as better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade.
 Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct.
-Three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes.
+Three of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues among the down-regulated genes.
 As an independent verification, we also performed GO analysis using PANTHER on the down-regulated genes for \emph{D.~coniospora}.
@@
 Specifically, we believe an enriched term may broadly suggest one of two hypotheses: if a list is enriched in a particular anatomical region, that anatomical region is affected by the experimental treatment; alternatively, the anatomical regions that are enriched reflect biologically relevant interactions between tissues.
-We believe the first hypothesis is a reasonable hypothesis to make in the case of whole-worm RNA-seq data for example, whereas the second hypothesis may be more plausible in cases where a researcher already knows what tissues a particular gene list came from, as may be the case in single-cell RNA-seq.
+We believe the first hypothesis is a reasonable one to make in the case of whole-worm RNA-seq data for example, whereas the second hypothesis may be more plausible in cases where a researcher already knows what tissues a particular gene list came from, as may be the case in single-cell RNA-seq.
 
@@
 %Trimming discussion
-We have also presented the first, to our knowledge, ontology trimming algorithm applied to biomedical ontologies.
-This algorithm, which is very easy to execute, places strong limits on the number of terms selected for testing.
+We have also presented the first, to our knowledge, ontology term filtering algorithm applied to biomedical ontologies.
+This algorithm, which is very easy to execute, identifies terms that have specificity and statistical power for hypothesis testing.
 Due to the nature of all ontologies as hierarchical, acyclical graphs with term inheritance, term annotations are correlated along any given branch.
@@
 As with any other tool based on statistical sampling, our analysis is most vulnerable to bias in the data set.
-For example, expression reports are negatively biased against germline expression because of the difficulties associated with expressing DNA in this tissue~\cite{Kelly1997}.
+For example, expression reports are negatively biased against germline expression because of the difficulties associated with expressing transgenes in this tissue~\cite{Kelly1997}.
 As time passes, we are certain the accuracy and power of this tool will improve thanks to the continuing efforts of the  worm research community; indeed, without the community reports of tissue expression in the first place, this tool would not be possible.
+
+\section*{Conclusions}
+
+We have built a tissue enrichment tool that employs a tissue ontology previously developed by WormBase.
+We use a simple algorithm to identify the best ontology terms for statistical testing and in this way minimize multiple testing problems.
+Our tool is available within WormBase or can be downloaded for offline use via `pip install'.
 
@@
 	lab for helpful discussion. 
+\section*{Funding}
+	This work was supported by HG002223 from the National Human Genome Research 	Institute to PWS., and H.H.M.I., with which PWS is an investigator.
+The 	funding agencies had no input into the design, execution or interpretation 	of our experiments, nor into the writing of this manuscript.
 
@@
    }
-  \label{fig:intragree}
+  \label{fig:interagree}
 \end{figure}
@@
 	\caption{
-	Comparison of results for a neuronal-enriched gene set from Watson~\cite{Watson2008a}.
-We ran the same genelist on a dictionary with a minimum annotation cutoff of 50, similarity threshold of 0.95 and similarity method `any' versus another with a minimum annotation cutoff of 33, similarity threshold of 0.95 and similarity method `any'.
+	Comparison of results for a GABAergic neuronal-enriched gene set from Watson~\cite{Watson2008a} showing that results are similar regardless of annotation cutoff.
+We ran the same gene list on a dictionary with a minimum annotation cutoff of 50, similarity threshold of 0.95 and similarity method `any' versus another with a minimum annotation cutoff of 33, similarity threshold of 0.95 and similarity method `any'.
 In the table, columns are labeled with their significance value (Q-value) or enrichment fold change followed by a hyphen and a number which indicates which the cutoff for the dictionary that was used for testing.
@@
 Hyphens denote not-applicable values, which occurs when a particular tissue is not present in both dictionaries.
-Full table is available in Additional File 2.
 	}
@@
 	% \end{adjustbox}
-	\label{tab:interagree}
+	\label{tab:intraagree}
 

### 2016-08-26 setup.py
setup.py
corrected a spelling mistake

@@ -10,7 +10,7 @@ from setuptools import setup, find_packages
 import os
 import sys
 
-version = '0.13.16'
+version = '0.13.17'
 
 # just type in python setup.py publish and this takes care of publishing to pypi!
 
@@ -52,7 +52,7 @@ if sys.argv[-1] == 'test':
         modules = map(__import__, test_requirements)
     except ImportError as e:
         err_msg = e.message.replace("No module named ", "")
-        msg = "%s is not installed. Install your test requirments." % err_msg
+        msg = "%s is not installed. Install your test requirements." % err_msg
         raise ImportError(msg)
     os.system('py.test')
     sys.exit()

### 2016-08-26 build/lib/tissue_enrichment_analysis/hypergeometricTests.py
removed WBbts from marplot
previous fix didn’t come thru/// need to get better at doing this whole
git thing!

@@ -297,8 +297,12 @@ def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
     # sort by q value change
     df.sort_values(['Q value', 'Enrichment Fold Change'],
                    ascending=[True, False], inplace=True)
+
+    # added August 26 2016:
+    lenID = 11
+    hr_labels = df['Tissue'][:n_bars].str[:-lenID-1]
     # plot first n_bars
-    ax = sns.barplot(x=df[y][:n_bars], y=df['Tissue'][:n_bars], ax=ax)
+    ax = sns.barplot(x=df[y][:n_bars], y=hr_labels, ax=ax)
 
     # fix the plot to prettify it
     ax.set_ylabel('Tissue', fontsize=15)

### 2016-08-26 setup.py

@@ -10,7 +10,7 @@ from setuptools import setup, find_packages
 import os
 import sys
 
-version = '0.13.17'
+version = '0.13.19'
 
 # just type in python setup.py publish and this takes care of publishing to pypi!
 

### 2016-12-14 pea_paper_docs/src/dictionary_generator.py
started writing PEA
also added gene_ontology.csv to github, since it is probably too large
to host there

@@ -1,0 +1,580 @@
+"""Contains classes solr_query, node and sisters."""
+# -*- coding: utf-8 -*-
+
+from urllib.request import urlopen
+import simplejson
+import json
+import numpy as np
+import pandas as pd
+import contextlib
+# import copy
+
+
+class solr_query():
+    """
+    A solr_query class that stores URLs.
+
+    Attbts:
+    solr_url - the main solr_url
+    """
+
+    def __init__(self, solr_url, query):
+        """Initialize the solr_query object."""
+        self.solr_url = solr_url
+        self.query = query
+
+    def set_solr_url(self, url):
+        """Assign a url to the solr_query object."""
+        self.solr_url = url
+
+    def add_query_url(self, url):
+        """Add a query url to the solr_query object."""
+        self.query = url
+
+    def open_query(self):
+        """
+        Given a query, append it to the main url.
+
+        Open URL and use simplejson to load the results
+        """
+        try:
+            with contextlib.closing(urlopen(self.solr_url +
+                                    self.query)) as conn:
+                return simplejson.load(conn)
+        except:
+            raise Warning('URL is invalid or may have timed out')
+
+
+class node():
+    """
+    A node is intended to be a single ontology term.
+
+    Attributes:
+    name - wbbt id
+    parents
+    genes
+    similarity - no. of genes assoc. with this node divided
+                 by the set of genes of its sister set
+    drop -- whether to drop or not
+    good_name -- human readable plus wbbt
+
+
+    QUERIES FOR RELATIONS AND GENES ARE LAMBDA FUNCTIONS
+    query_relation(x) -- gets families of tissue x
+    query_genes(x) -- gets genes assoc with x
+    query_readable
+    """
+
+    def __init__(self, name):
+        """Initialization function."""
+        self.name = name
+        self.daughters = []
+        self.parents = []
+        self.genes = []
+        self.similarity = 0
+        self.drop = False
+        self.good_name = ''
+
+    def get_name(self, human_readable):
+        """Generate a good name (human readable + WBid)."""
+        if human_readable == '':
+            print('warning, empty human readable name')
+        self.good_name = human_readable + ' ' + self.name
+
+    def add_daughter(self, daughter):
+        """Add a daughter to this node."""
+        self.daughters.append(daughter)
+        self.daughters = list(set(self.daughters))  # prevent redundancy
+
+    def add_parent(self, parent):
+        """Add a parent to this node."""
+        self.parents.append(parent)
+        self.parents = list(set(self.parents))
+
[…]

### 2016-12-14 pea_paper_docs/src/pea_benchmarks.py

@@ -1,0 +1,17 @@
+import tissue_enrichment_analysis as tea
+import pandas as pd
+import matplotlib.pyplot as plt
+import os
+
+phenotype_df = pd.read_csv('../input/phenotype_ontology.csv')
+
+
+phenotype_df.columns
+
+for f in os.listdir('../input/genesets_golden'):
+    if f != '':
+        print(f)
+        genes = pd.read_csv('../input/genesets_golden/'+f)
+        df, _ = tea.enrichment_analysis(genes.gene, phenotype_df,
+                                        show=False)
+        df.to_csv('../output/pea_goldenset_results/' + f + '.csv', index=False)

### 2016-12-14 tea_paper_docs/src/Engelmann_Reanalysis.py

@@ -1,0 +1,161 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Mon Feb 15 14:07:04 2016
+A script to analyze some of the genes identified by Engelmann et al 2011,
+A Comprehensive Analysis of Gene Expression Changes Provoked by Bacterial
+and Fungal Infection in C elegans.
+
+@author: dangeles
+"""
+import tissue_enrichment_analysis as tea  # the library to be used
+import pandas as pd
+import numpy as np
+import seaborn as sns
+import matplotlib.pyplot as plt
+import os
+
+# tissue dictionary to use
+tissue_df = pd.read_csv('../input/final_cutoff33_threshold0.95_methodany.csv')
+
+# rna-seq datasets to reanalyze
+#Bacterial
+dfLum = pd.read_csv('../input/Engelmann/luminescens_Engelmann_2011.csv')
+dfMarc = pd.read_csv('../input/Engelmann/marcescens_Engelmann_2011.csv')
+dfFaec = pd.read_csv('../input/Engelmann/faecalis_Engelmann_2011.csv')
+
+# Fungal
+dfHarp = pd.read_csv('../input/Engelmann/harposporium_Engelmann_2011.csv')
+dfCon = pd.read_csv('../input/Engelmann/coniospora_Engelmann_2011.csv')
+# dfSin= pd.read_csv('../input/Engelmann/sinha_2012.csv')
+
+# dictionary of gene names
+names = pd.read_csv('../input/Engelmann/c_elegans.PRJNA13758.WS241.livegeneIDs.unmaprm.txt',
+                    sep='\t', comment='#')
+
+# make the directories to place the analyses
+dirEngelmann = '../output/Engelmann/'
+dirAnalysis = '../output/Engelmann/Analysis'
+dirGraphs = '../output/Engelmann/Graphs'
+DIRS = [dirEngelmann, dirAnalysis, dirGraphs]
+
+# Make all the necessary dirs if they don't already exist
+for d in DIRS:
+    if not os.path.exists(d):
+        os.makedirs(d)
+
+
+# Selector functions to draw WBIDs from
+f = lambda x, y: (names.HumanReadable.isin(x[y]))
+g = lambda x, y: (names.GeneName.isin(x[y]))
+
+# ==============================================================================
+# ==============================================================================
+# # Engelmann Analysis
+# ==============================================================================
+# ==============================================================================
+
+
+class receptacle(object):
+
+    def __init__(self, name):
+        self.name = name
[…]

### 2016-12-14 tea_paper_docs/src/hgf_benchmark_script.py

@@ -1,0 +1,387 @@
+# -*- coding: utf-8 -*-
+"""
+A script to benchmark TEA.
+
+@david angeles
+dangeles@caltech.edu
+"""
+
+import tissue_enrichment_analysis as tea  # the library to be used
+import pandas as pd
+import os
+import numpy as np
+import seaborn as sns
+import matplotlib.pyplot as plt
+import re
+import matplotlib as mpl
+
+sns.set_context('paper')
+
+# pd.set_option('display.float_format', lambda x:'%f'%x)
+pd.set_option('precision', 3)
+# this script generates a few directories.
+
+
+dirOutput = '../output/'
+dirSummaries = '../output/SummaryInformation/'
+dirHGT25_any = '../output/HGT25_any_Results/'
+dirHGT33_any = '../output/HGT33_any_Results/'
+dirHGT50_any = '../output/HGT50_any_Results/'
+dirHGT100_any = '../output/HGT100_any_Results/'
+dirComp = '../output/comparisons/'
+
+DIRS = [dirOutput, dirSummaries, dirHGT25_any,
+        dirHGT50_any, dirHGT100_any, dirComp]
+# open the relevant file
+path_sets = '../input/genesets_golden/'
+path_dicts = '../input/WS252AnatomyDictionary/'
+
+
+# Make all the necessary dirs if they don't already exist
+for d in DIRS:
+    if not os.path.exists(d):
+        os.makedirs(d)
+
+
+# Make the file that will hold the summaries and make the columns.
+with open(dirSummaries+'ExecutiveSummary.csv', 'w') as fSum:
+    fSum.write('#Summary of results from all benchmarks\n')
+    fSum.write('NoAnnotations,Threshold,Method,EnrichmentSetUsed,TissuesTested,GenesSubmitted,TissuesReturned,GenesUsed,AvgFold,AvgQ,GenesInDict\n')
+
+# ==============================================================================
+# ==============================================================================
+# # Perform the bulk of the analysis, run every single dictionary on every set
+# ==============================================================================
+# ==============================================================================
+i = 0
+# look in the dictionaries
+for folder in os.walk(path_dicts):
+    # open each one
+    for f_dict in folder[2]:
+        if f_dict == '.DS_Store':
+            continue
+        tissue_df = pd.read_csv(path_dicts+f_dict)
+
+        # tobedropped when tissue dictionary is corrected
[…]

### 2016-12-14 tests/hgf_benchmarking/src/Engelmann_Reanalysis.py

@@ -1,161 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Mon Feb 15 14:07:04 2016
-A script to analyze some of the genes identified by Engelmann et al 2011,
-A Comprehensive Analysis of Gene Expression Changes Provoked by Bacterial
-and Fungal Infection in C elegans.
-
-@author: dangeles
-"""
-import tissue_enrichment_analysis as tea  # the library to be used
-import pandas as pd
-import numpy as np
-import seaborn as sns
-import matplotlib.pyplot as plt
-import os
-
-# tissue dictionary to use
-tissue_df = pd.read_csv('../input/final_cutoff33_threshold0.95_methodany.csv')
-
-# rna-seq datasets to reanalyze
-#Bacterial
-dfLum = pd.read_csv('../input/Engelmann/luminescens_Engelmann_2011.csv')
-dfMarc = pd.read_csv('../input/Engelmann/marcescens_Engelmann_2011.csv')
-dfFaec = pd.read_csv('../input/Engelmann/faecalis_Engelmann_2011.csv')
-
-# Fungal
-dfHarp = pd.read_csv('../input/Engelmann/harposporium_Engelmann_2011.csv')
-dfCon = pd.read_csv('../input/Engelmann/coniospora_Engelmann_2011.csv')
-# dfSin= pd.read_csv('../input/Engelmann/sinha_2012.csv')
-
-# dictionary of gene names
-names = pd.read_csv('../input/Engelmann/c_elegans.PRJNA13758.WS241.livegeneIDs.unmaprm.txt',
-                    sep='\t', comment='#')
-
-# make the directories to place the analyses
-dirEngelmann = '../output/Engelmann/'
-dirAnalysis = '../output/Engelmann/Analysis'
-dirGraphs = '../output/Engelmann/Graphs'
-DIRS = [dirEngelmann, dirAnalysis, dirGraphs]
-
-# Make all the necessary dirs if they don't already exist
-for d in DIRS:
-    if not os.path.exists(d):
-        os.makedirs(d)
-
-
-# Selector functions to draw WBIDs from
-f = lambda x, y: (names.HumanReadable.isin(x[y]))
-g = lambda x, y: (names.GeneName.isin(x[y]))
-
-# ==============================================================================
-# ==============================================================================
-# # Engelmann Analysis
-# ==============================================================================
-# ==============================================================================
-
-
-class receptacle(object):
-
-    def __init__(self, name):
-        self.name = name
[…]

### 2016-12-14 tests/hgf_benchmarking/src/clean_dictionary_maker.py

@@ -1,118 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Wed Mar  2 14:24:04 2016
-
-A script to generate 'clean' dictionaries that exclude
-golden standards. 
-
-This script 
-
-@author: dangeles
-"""
-import dictionary_generator as dg
-from dictionary_generator import ontology
-import sys
-x_array= [100, 50, 33, 25]
-thresh_array= [1, .95, .9]
-methods= ['avg', 'any']
-path= '../input/WS252AnatomyDictionary/'
-
-
- #main solr url
-solr_url = 'http://caprica.caltech.edu:8080/solr/anatomy/'
-#queries must be lambda functions
-#query for terms. Finds terms that have x or more annotating genes
-query_terms= lambda x: 'select?qt=standard&indent=on&wt=json&version=2.2&fl=id&start=0&rows=0&q=document_category:bioentity&facet=true&facet.field=regulates_closure&facet.limit=-1&facet.mincount={0}&facet.sort=count&fq=source:%22WB%22&fq=-qualifier:%22not%22'.format(x)
-#query for relationships. given a wbbt ID, find the nodes connected to it. 
-query_relation= lambda x: "select?qt=standard&fl=topology_graph_json&version=2.2&wt=json&indent=on&rows=1&q=id:%22{0}%22&fq=document_category:%22ontology_class%22".format(x)
-#query for number of genes. given a wbbt ID, find genes assoc. with it. 
-query_genes= lambda x: "select?qt=standard&indent=on&wt=json&version=2.2&fl=id&start=0&rows=10000&q=document_category:bioentity&fq=source:%22WB%22&fq=-qualifier:%22not%22&fq=regulates_closure:%22{0}%22".format(x)
-#query for readable names
-query_readable= "select?qt=standard&fl=id,annotation_class_label&version=2.2&wt=json&indent=on&rows=100000&q=id:*&fq=document_category:ontology_class&fq=-is_obsolete:true"
-queries= [query_terms, query_relation, query_genes, query_readable]
-min_annot= 2
-
-for x in x_array:
-    for threshold in thresh_array:
-        for method in methods:    
-            print(x, threshold, method)
-            sys.stdout.flush()
-            print(solr_url)
-            trial1= ontology('tissue_ontology', x, threshold, method, solr_url)
-            trial1.set_min_cutoff(5)
-            trial1.add_nodes(query_terms, query_readable)
-            trial1.find_node_annotations(query_genes)
[…]

### 2016-12-14 tests/hgf_benchmarking/src/dictionary_maker_2.py

@@ -1,425 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Wed Mar  2 14:39:03 2016
-
-@author: dangeles
-"""
-
-from urllib.request import urlopen
-import simplejson
-import json
-import numpy as np
-import pandas as pd
-import contextlib
-
-class solr_query(object):
-    """
-    A solr_query class that stores URLs
-    
-    Attbts:
-    solr_url - the main solr_url
-    """
-    def __init__(self, solr_url, query):
-        self.solr_url= solr_url
-        self.query= query
-        
-    def set_solr_url(self, url):
-        self.solr_url= url
-    def add_query_url(self, url):
-        self.query= url
-        
-    def open_query(self):
-        """
-        Given a query, append it to the main url. Open URL and
-        use simplejson to load the results
-        """
-        try:
-            with contextlib.closing(urlopen(self.solr_url+ self.query)) as conn:
-                return simplejson.load(conn)
-        except:
-            raise Warning('URL is invalid or may have timed out')
-        
-
-class node(object):
-    """
-    A node is intended to be a single ontology term
-    
-    Attributes:
-    name - wbbt id
-    parents
-    genes
-    similarity - no. of genes assoc. with this node divided by the set of genes of its sister set
-    drop -- whether to drop or not
-    good_name -- human readable plus wbbt
-    
-    
-    QUERIES FOR RELATIONS AND GENES ARE LAMBDA FUNCTIONS
-    query_relation(x) -- gets families of tissue x
-    query_genes(x) -- gets genes assoc with x
-    query_readable
-
-    """
-    def __init__(self, name):
-        self.name= name
-        self.daughters= []
-        self.parents= []
-        self.genes= []
-        self.similarity= 0
-        self.drop= False
-        self.good_name= ''
-    
-    def get_name(self, human_readable):        
-        self.good_name= human_readable + ' '+ self.name
-        
-    def add_daughter(self, daughter):
-        self.daughters.append(daughter)
-    
-    def add_parent(self, parent):
-        self.parents.append(parent)
-    
-    def add_annotation(self, gene):
-        self.gemes.append(gene)
-    
-    def throw_away(self):
-        self.drop= True
-    
-    def calc_similarity(self, sim):
-        self.similarity= sim
-
-    def find_family(self, solr_url, query_relation):
-        """
-        query_relation(x) --lambda function
-        """
-        #get the json object
[…]

### 2016-12-14 tests/hgf_benchmarking/src/hgf_benchmark_script.py

@@ -1,387 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-A script to benchmark TEA.
-
-@david angeles
-dangeles@caltech.edu
-"""
-
-import tissue_enrichment_analysis as tea  # the library to be used
-import pandas as pd
-import os
-import numpy as np
-import seaborn as sns
-import matplotlib.pyplot as plt
-import re
-import matplotlib as mpl
-
-sns.set_context('paper')
-
-# pd.set_option('display.float_format', lambda x:'%f'%x)
-pd.set_option('precision', 3)
-# this script generates a few directories.
-
-
-dirOutput = '../output/'
-dirSummaries = '../output/SummaryInformation/'
-dirHGT25_any = '../output/HGT25_any_Results/'
-dirHGT33_any = '../output/HGT33_any_Results/'
-dirHGT50_any = '../output/HGT50_any_Results/'
-dirHGT100_any = '../output/HGT100_any_Results/'
-dirComp = '../output/comparisons/'
-
-DIRS = [dirOutput, dirSummaries, dirHGT25_any,
-        dirHGT50_any, dirHGT100_any, dirComp]
-# open the relevant file
-path_sets = '../input/genesets_golden/'
-path_dicts = '../input/WS252AnatomyDictionary/'
-
-
-# Make all the necessary dirs if they don't already exist
-for d in DIRS:
-    if not os.path.exists(d):
-        os.makedirs(d)
-
-
-# Make the file that will hold the summaries and make the columns.
-with open(dirSummaries+'ExecutiveSummary.csv', 'w') as fSum:
-    fSum.write('#Summary of results from all benchmarks\n')
-    fSum.write('NoAnnotations,Threshold,Method,EnrichmentSetUsed,TissuesTested,GenesSubmitted,TissuesReturned,GenesUsed,AvgFold,AvgQ,GenesInDict\n')
-
-# ==============================================================================
-# ==============================================================================
-# # Perform the bulk of the analysis, run every single dictionary on every set
-# ==============================================================================
-# ==============================================================================
-i = 0
-# look in the dictionaries
-for folder in os.walk(path_dicts):
-    # open each one
-    for f_dict in folder[2]:
-        if f_dict == '.DS_Store':
-            continue
-        tissue_df = pd.read_csv(path_dicts+f_dict)
-
-        # tobedropped when tissue dictionary is corrected
[…]

### 2016-12-14 tissue_enrichment_analysis/hypergeometricTests.py

@@ -76,11 +76,11 @@ def hgf(gene_list, tissue_df):
     tissue_df = tissue_df.set_index('wbid')
 
     # number of tissues in the dictionary
-    sums_of_tissues = tissue_df.sum()[1:]  # this object can be identified by
+    sums_of_tissues = tissue_df.sum()  # this object can be identified by
     # the column names of tissues and excludes gene IDs
 
     # total size of the urn
-    total_genes = tissue_df.shape[0]  # total genes in the dictionary
+    total_genes = tissue_df.sum().sum()  # total labels in the dictionary
 
     # slice out the rows from tissue_dictionary that came from the list
     wanted_dictionary = tissue_df.loc[wanted]

### 2016-12-15 pea_paper_docs/src/dictionary_generator.py
added gwas data, benchmarked all ontologies

@@ -31,18 +31,29 @@ class solr_query():
         """Add a query url to the solr_query object."""
         self.query = url
 
-    def open_query(self):
+    def open_query(self, p=0, timeout=10):
         """
         Given a query, append it to the main url.
 
         Open URL and use simplejson to load the results
         """
-        try:
-            with contextlib.closing(urlopen(self.solr_url +
-                                    self.query)) as conn:
-                return simplejson.load(conn)
-        except:
-            raise Warning('URL is invalid or may have timed out')
+        timer = 10  # don't try more than 10x per website
+        if p:
+            print(self.solr_url + self.query)
+
+        while timer > 0:
+            try:
+                with contextlib.closing(urlopen(self.solr_url +
+                                        self.query, timeout=timeout)) as conn:
+                    return simplejson.load(conn)
+            except:
+                # raise Warning('URL is invalid or may have timed out')
+                timer -= 1
+                pass
+        # raise an error if the timer reached 0
+        if timer == 0:
+            print(self.solr_url + self.query)
+            raise Warning('Url could not be contacted or is invalid')
 
 
 class node():
@@ -104,7 +115,7 @@ class node():
         """Calculate similarity."""
         self.similarity = sim
 
-    def find_family(self, solr_url, query_relation):
+    def find_family(self, solr_url, query_relation, p=0):
         """
         Find the family for this node by using solr_url and query_relation.
 
@@ -114,7 +125,7 @@ class node():
         rsp_rlshp = solr_query(solr_url, query_relation(self.name))
 
         # extract the array with all the right information
-        array_of_rlshps = rsp_rlshp.open_query()['response']['docs'][0]
+        array_of_rlshps = rsp_rlshp.open_query(p=p)['response']['docs'][0]
 
         # go through the array, turning each line into a dictionary
         # these mini-dictionaries contain the edges between nodes
@@ -291,8 +302,10 @@ class ontology():
 
     def find_node_family(self, lambda_query_rlshp):
         """Find the nodes that are related to this one."""
+        p = 1
         for n in iter(self.nodes):
-            self.nodes[n].find_family(self.solr_url, lambda_query_rlshp)
+            self.nodes[n].find_family(self.solr_url, lambda_query_rlshp, p=p)
+            p = 0
 
[…]

### 2016-12-15 pea_paper_docs/src/pea_benchmarks.py

@@ -4,14 +4,124 @@ import matplotlib.pyplot as plt
 import os
 
 phenotype_df = pd.read_csv('../input/phenotype_ontology.csv')
+go_df = pd.read_csv('../input/gene_ontology.csv')
+
+snp = pd.read_csv('../input/known_snps.csv', sep=',', comment='#')
 
 
-phenotype_df.columns
+def run(path, f, output_path, dictionary, **kwargs):
+    """Run and save e.a.."""
+    genes = pd.read_csv(path+f, **kwargs)
+    df, _ = tea.enrichment_analysis(genes.gene, dictionary,
+                                    show=False)
+    df.to_csv(output_path + f + '.csv', index=False)
 
 for f in os.listdir('../input/genesets_golden'):
     if f != '':
-        print(f)
-        genes = pd.read_csv('../input/genesets_golden/'+f)
-        df, _ = tea.enrichment_analysis(genes.gene, phenotype_df,
-                                        show=False)
-        df.to_csv('../output/pea_goldenset_results/' + f + '.csv', index=False)
+        run('../input/genesets_golden/', f,
+            '../output/pea_goldenset_results/', phenotype_df)
+        run('../input/genesets_golden/', f,
+            '../output/go_goldenset_results/', go_df)
+
+for f in os.listdir('../input/fog2'):
+    if f != '':
+        run('../input/fog2/', f, '../output/fog2/pea_', phenotype_df)
+        run('../input/fog2/', f, '../output/fog2/go_', go_df)
+
+for f in os.listdir('../input/hypoxia'):
+    if f != '':
+        run('../input/hypoxia/', f, '../output/hypoxia/pea_', phenotype_df)
+        run('../input/hypoxia/', f, '../output/hypoxia/go_', go_df)
+
+for disease in snp.Disease_Trait.unique():
+    print(disease, snp[snp.Disease_Trait == disease].shape[0])
+
+depression = ['Major depressive disorder', 'Major depressive disorder (broad)']
+celiac = ['Celiac disease', 'Celiac disease and Rheumatoid arthritis']
+macular_degeneration = ['Age-related macular degeneration',
+                        'Age-related macular degeneration (CNV vs. GA)',
+                        'Age-related macular degeneration (CNV)',
+                        'Age-related macular degeneration (GA)',
+                        'Age-related macular degeneration (extreme sampling)',
+                        'Age-related macular degeneration (wet)']
[…]

### 2016-12-15 setup.py
corrected a bug that affected how p-values were calculated
by a bug, I mean a mistake I made. But sure, a bug.
Basically, before the total number of balls in an urn was = number of
genes in dictionary.
Now, n_balls = sum of balls in each tissue.

@@ -12,7 +12,7 @@ from setuptools import setup, find_packages
 import os
 import sys
 
-version = '0.13.21'
+version = '0.14.00'
 
 # just type in python setup.py publish and
 # this takes care of publishing to pypi!

### 2016-12-16 tea_paper_docs/doc/reviewer_rebuttals/reviewer_rebuttals.tex

--- 2016-07-01 tests/hgf_benchmarking/doc/reviewr_rebuttals/reviewer_rebuttals.tex
@@
 %
-\documentclass[12pt,stdletter,orderfromtodate,sigleft]{newlfm}
+\documentclass[12pt]{article}
 \usepackage{blindtext, xfrac}
@@
 
-\newlfmP{dateskipbefore=50pt}
-\newlfmP{sigsize=50pt}
-\newlfmP{sigskipbefore=50pt}
- \newlfmP{Headlinewd=0pt,Footlinewd=0pt}
-
-\namefrom{Paul W. Sternberg and David Angeles-Albores}
-\addrfrom{
-    HHMI and Division of Biology and Biological Engineering,
-	Caltech
-}
-\addrto{%
-    Dr. Joao Carlos Setubal,
-	BMC Bioinformatics
-}
-\dateset{June 29, 2016}
-\greetto{Dear Dr. Setubal}
-\closeline{Sincerely,}
 \begin{document}
-\begin{newlfm}
 
@@
 We didn’t fine-tune this parameter cutoff.
-We have added a sentence to explicitly point this out in the text, and we have also made it clear that this parameter can be tuned by users in the command-line, but not the web, version.
-}
+We have added a sentence to explicitly point this out in the text, and we have also made it clear that this parameter can be tuned by users in the command-line, but not the web, version.}
 
-\end{newlfm}
 \end{document}

### 2016-12-16 tea_paper_docs/doc/SI/FolderStructure.tex

--- 2016-04-19 tests/hgf_benchmarking/doc/SI/FolderStructure.tex
@@
 
-In the SI, we provide the entire output of our tests for the interested reader. The folder is organized as follows:
+In the SI, we provide the entire output of our tests for the interested reader.
+We provide two folders.
+The gene set folder contains all the gene sets used to benchmark our tool.
+Gene sets are named according to a common schema.
+Files are named\\ \texttt{WBPaperID\_AnatomyTermEnriched(HumanReadable)\_AnatomyTermWBbtID\_NumberOfGenesInSet.csv}.
+
+The results folder contains all the results generated by our analysis. The results folder is organized as follows (bolded text refers to a folder, otherwise files are listed):
+
+\begin{itemize}
+	\item \textbf{Engelmann}
+	\item \textbf{HGT25\_any\_Results}
+	\item \textbf{HGT33\_any\_Results}
+	\item \textbf{HGT50\_any\_Results}
+	\item \textbf{HGT100\_any\_Results}
+	\item \textbf{SummaryInformation}
+	\item \textbf{comparisons}
+	\item test\_list\_efaecelis.txt
+\end{itemize}
 
 \subsection*{HGTXX\_any\_Results}
-The folders titled HGTXX\_any\_Results contain the enrichment results of all the 30 golden sets tested with a dictionary with cutoff XX, similarity threshold $=0.95$ and thresholding method `any'.
+The folders titled \textbf{HGTXX\_any\_Results} contain the enrichment results of all the 30 golden sets tested with a dictionary with cutoff \textbf{XX}, similarity threshold$=0.95$ and thresholding method `any'.
+
 The folders contain two kinds of files, `.pdf' and `.csv' files.
@@
 \texttt{WBPaper00013489\_Ray\_Enriched\_WBbt\_0006941\_25.csv}\\
-refers to the \textbf{WormBase Paper 0013489}, which should be enriched in `\textbf{Ray Enriched (WBbt:0006941)}' and contains \textbf{25} genes.
-The gene set that was used for this analysis is contained in the SI folder named `\textbf{golden gene sets}' and is contained in a homologously named csv file.
+refers to the \emph{WormBase Paper 0013489}, which should be enriched in `\emph{Ray Enriched (WBbt:0006941)}' and contains \emph{25} genes.
+The gene set that was used for this analysis is contained in the SI folder named `\emph{golden gene sets}' and is contained in a homologously named csv file.
 
@@
 \subsection*{Comparisons}
-	Contains csv files of the comparisons between gene sets or within dictionaries. The file nomenclature is:\
-neuronal\_comparison\_33\_WBPaper00024970\_with\_WBPaper0037950\_complete.csv
+	Contains csv files of the comparisons between gene sets or within dictionaries. The file nomenclature is:\\
+\texttt{neuronal\_comparison\_33\_WBPaper00024970\_with\_WBPaper0037950\_complete.csv}\\
 Refers to a `neuronal comparison' using dictionary 33, comparing papers 24970 with 37950.
@@
 
-Alternatively, when comparing two dictionaries, the nomenclature is:
-neuronal\_comparison\_GABAergic\_33-50\_WBPaper0037950\_complete.csv
+Alternatively, when comparing two dictionaries, the nomenclature is:\\
+\texttt{neuronal\_comparison\_GABAergic\_33-50\_WBPaper0037950\_complete.csv}\\
 Which refers to a `GABAergic neuronal' lists, comparing dictionaries with cutoffs of 33-50 (all other parameters are the same) using paper 37950.
 
+The complete version of table 2 in the paper is \\
+\texttt{neuronal\_comparison\_33\_WBPaper00024970\_with\_WBPaper0037950\_complete.csv}\\
+
 \subsection*{Summary Info}
+	This folder contains
 
-\subsection*{test\_list\_efaecalis.txt}
+	\begin{itemize}
+		\item ExecutiveSummary.csv --- a table containing summary information ofthe results of all the enrichment analysis we performed.
+The table contains the parameters of the dictionary used to run the analysis (cutoff, similarity threshold and thresholding method), the name of the gene set analysed, the number of terms tested, the number of genes submitted for analysis, the number of terms that tested significant, the average fold change of terms that tested significant, the average q-value of terms that tested significant and the number of genes in the dictionary.
+		\item TissueNumbers.csv --- A table showing the number of terms in each dictionary that was generated. 
+		\item avgFoldChange.pdf --- KDE of the average fold change value for each dictionary
+		\item avgQKDE\_method=any.pdf --- KDE of the average q-value for each dictionary 
+		\item avgQKDE\_method=avg.pdf --- KDE of the average fold change value for each dictionary
+		\item fractissuesKDE\_method=any.pdf --- KDE of the average number of tissues in the dictionary that tested significant for each dictionary. 
+	\end{itemize}
+
+\subsubsection*{test\_list\_efaecalis.txt}
+This file was used to generate figure 2 in the text and is a slightly processed version of Engelmann's downregulated genes for \emph{E.~faecalis} (this list was processed into WBIDs).
 

### 2017-01-04 pea_paper_docs/tex/pea.tex
added linker cell/ciliary neurons analyses

\documentclass[10pt,letterpaper,twocolumn]{article}
% \usepackage[twocolumn]{geometry}
\usepackage{lmodern}% http://ctan.org/pkg/lm

% use Unicode characters - try changing the option if you run into troubles with special characters (e.g. umlauts)
\usepackage[utf8x]{inputenc}
% clean citations
\usepackage{cite}
% hyperref makes references clicky. use \url{www.example.com} or \href{www.example.com}{description} to add a clicky url
\usepackage{nameref}
% line numbers
\usepackage[right]{lineno}
% improves typesetting in LaTeX
% \usepackage{microtype}
% degree symbol
\usepackage{gensymb}
\usepackage{amsmath}
\usepackage{booktabs}
% use adjustwidth environment to exceed text width (see examples in text)
\usepackage{changepage}

% adjust caption style
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}

% remove brackets from references
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% headrule, footrule and page numbers
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
% \fancyheadoffset[L]{2.25in}
% \fancyfootoffset[L]{2.25in}

% use \textcolor{color}{text} for colored text (e.g. highlight to-do areas)
\usepackage{color}

% define custom colors (this one is for figure captions)
\definecolor{Gray}{gray}{.25}

% this is required to include graphics
\usepackage{graphicx}

% use if you want to put caption to the side of the figure - see example in text
\usepackage{sidecap}
% \usepackage[urlcolor  = blue]{hyperref}

% hyperurls packages:
\usepackage{xcolor}
\usepackage[colorlinks = true,
            linkcolor = blue,
            urlcolor  = blue,
            citecolor = blue,
            anchorcolor = blue]{hyperref}

% use for have text wrap around figures
\usepackage{wrapfig}
\usepackage[pscoord]{eso-pic}
\usepackage[fulladjust]{marginnote}
\reversemarginpar{}

% new commands
\newcommand{\cel}{\emph{C.~elegans}}
\newcommand{\fog}{\emph{\mbox{fog-2}}}
\newcommand{\ecol}{\emph{E.~coli}}

\title{
  \Large
  \textbf{Phenotype Enrichment Discovers Phenologues for Disease Modeling in \cel{}}
}

\author{David Angeles-Albores\textsuperscript{1,$\dagger{}$}
\and{}
Paul W. Sternberg\textsuperscript{1,*}
}

\begin{document}

% \vspace*{0.35in}

% title goes here:
\twocolumn[
% title
\maketitle

]
% now start line numbers
\nolinenumbers{}

\section*{Introduction}

\section*{Results}

\subsection*{Developing PEA and GEA}

\subsection*{Applying the WormBase Enrichment Suite to Discover Disease Phenologues}
\subsubsection*{Obesity}
\subsubsection*{Schizophrenia}
\subsubsection*{Type-2 Diabetes}
\subsubsection*{Crohn's Disease}
\subsubsection*{Prostate Cancer}

\subsection*{Ontology Enrichment as an aid for Screen Design}
\subsubsection*{Enrichment in the Linker Cell Transcriptome}
\subsubsection*{Enrichment in a Linker Cell Screen}

\end{document}

### 2017-01-04 pea_paper_docs/src/pea_benchmarks.py

@@ -1,40 +1,67 @@
+# -*- coding: utf-8 -*-
+"""Benchmarking PEA."""
 import tissue_enrichment_analysis as tea
 import pandas as pd
-import matplotlib.pyplot as plt
+# import matplotlib.pyplot as plt
 import os
 
+# import data
 phenotype_df = pd.read_csv('../input/phenotype_ontology.csv')
 go_df = pd.read_csv('../input/gene_ontology.csv')
+tissue_df = tea.fetch_dictionary()
 
 snp = pd.read_csv('../input/known_snps.csv', sep=',', comment='#')
+linker_cell = pd.read_csv('../input/linker_cell.txt', sep='\t')
+lc_screen = pd.read_csv('../input/lc_screen.csv')
 
 
-def run(path, f, output_path, dictionary, **kwargs):
+wbids = pd.read_csv('../input/wormbase_ids.txt', sep='\t', header=None,
+                    names=['wbid', 'human_read', 'target_id'])
+
+ciliated_df = pd.read_excel('~/Downloads/mmc3.xlsx')
+# define a function to run e.a.s quickly
+def run(path, f, output_path, dictionary, column='gene', **kwargs):
     """Run and save e.a.."""
     genes = pd.read_csv(path+f, **kwargs)
-    df, _ = tea.enrichment_analysis(genes.gene, dictionary,
+    df, _ = tea.enrichment_analysis(genes[column], dictionary,
                                     show=False)
+    df = df[df.Observed > 2]
     df.to_csv(output_path + f + '.csv', index=False)
 
+# run the benchmark analyses
 for f in os.listdir('../input/genesets_golden'):
     if f != '':
         run('../input/genesets_golden/', f,
             '../output/pea_goldenset_results/', phenotype_df)
         run('../input/genesets_golden/', f,
             '../output/go_goldenset_results/', go_df)
+        run('../input/genesets_golden/', f,
+            '../output/tea_goldenset_results/', tissue_df)
 
+# run the fog-2 analysis
 for f in os.listdir('../input/fog2'):
     if f != '':
-        run('../input/fog2/', f, '../output/fog2/pea_', phenotype_df)
-        run('../input/fog2/', f, '../output/fog2/go_', go_df)
+        run('../input/fog2/', f, '../output/fog2/pea_', phenotype_df,
+            column='ens_gene')
[…]

### 2017-01-04 tissue_enrichment_analysis/hypergeometricTests.py

@@ -94,7 +94,7 @@ def hgf(gene_list, tissue_df):
     # make a hash with the p-values for enrichment of each tissue.
     p_hash = {}
     exp_hash = {}  # expected number for each tissue
-    for i, name in enumerate(tissue_df.columns.values[1:]):
+    for i, name in enumerate(tissue_df.columns.values):
         # if the total number of genes is zero, return p= 1 for all tissues
         if total == 0:
             p_hash[name] = 1
@@ -107,7 +107,6 @@ def hgf(gene_list, tissue_df):
             if wanted_sum[name] == 0:
                 p_hash[name] = 1
             else:
-
                 # no. of balls of color name picked
                 # total number of balls in urn
                 # total number of balls of color name in urn
@@ -338,7 +337,8 @@ def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
 
 def fetch_dictionary():
     """Fetch the dictionary we want."""
-    url_tissue = 'ftp://caltech.wormbase.org/pub/TissueEnrichmentAnalysis/anat_dict.csv'
+    url_tissue = 'ftp://caltech.wormbase.org/pub/' +\
+                 'TissueEnrichmentAnalysis/anat_dict.csv'
 
     try:
         with contextlib.closing(urlopen(url_tissue)) as conn:

### 2017-01-05 pea_paper_docs/tex/pea.tex
started writing and improved code base
TEA should run many times faster as a result.

Some numbers may have changed? I can’t figure out if that’s because of
the dictionary I’m using or something else, but I’m worried that there
was a bug prior. This new code is considerably cleaner and it looks
fine to me….

@@
 
-% use Unicode characters - try changing the option if you run into troubles with special characters (e.g. umlauts)
+% use Unicode characters - try changing the option
+% if you run into troubles with special characters (e.g. umlauts)
 \usepackage[utf8x]{inputenc}
@@
 \usepackage{cite}
-% hyperref makes references clicky. use \url{www.example.com} or \href{www.example.com}{description} to add a clicky url
+% hyperref makes references clicky. use \url{www.example.com}
+% or \href{www.example.com}{description} to add a clicky url
 \usepackage{nameref}
@@
 % adjust caption style
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}
+\usepackage[aboveskip=1pt,labelfont=bf,
+            labelsep=period,singlelinecheck=off]{caption}
 
@@
 
+\newcommand{\hobesity}{957}
+\newcommand{\wobesity}{614}
+
+\newcommand{\hschizo}{899}
+\newcommand{\wschizo}{433}
+
+\newcommand{\hcrohn}{390}
+\newcommand{\wcrohn}{213}
+
+\newcommand{\hprostate}{273}
+\newcommand{\wprostate}{273}
+
+\newcommand{\hdiabetes}{396}
+\newcommand{\wdiabetes}{224}
+
 \title{
   \Large
-  \textbf{Phenotype Enrichment Discovers Phenologues for Disease Modeling in \cel{}}
+  \textbf{Phenotype Enrichment Discovers Phenologs for Disease Modeling in \cel{}}
 }
@@
 \section*{Introduction}
+The last decade has seen an explosion of techniques capable of genome-wide
+measurements. Some examples of genome-wide tools include RNA-seq~\cite{} to
+measure gene expression, CHIP-seq~\cite{} to measure binding, or genome-wide
+methylation profiling to understand epigenetic changes. These tools are capable
+of generating large quantities of data. Understanding this data, and generating
+hypotheses from it remains challenging. A common approach used to understand
+these datasets is to reduce the dimensionality of the data via enrichment
+analyses of ontologies, which helps researchers understand what terms are
+overrepresented beyond random levels. By analyzing overrepresented terms in
+aggregate, researchers can better understand what biological processes were most
+affected in a given experiment, and form hypotheses about what is
+happening~\cite{}. This approach is limited by what ontologies can be tested for
+enrichment. In \cel{}, only the gene ontology (GO) and tissue ontology are
+available for testing~\cite{} even though curated gene, tissue and phenotype
+ontologies exist. Another limitation is that tissue enrichment testing
+is not offered on the same websites as GO enrichment testing, which requires
+users to test their data on different websites that may or may not use different
+methodologies to detect enrichment.
+
+Another way to use enrichment tools is for evolutionary comparison purposes.
+In molecular biology it is often useful to know when a gene is homologous
+between two species---that is to say, common by descent---because knowledge of
+homology often brings with it knowledge of function~\cite{}. Indeed, many
+important gene regulatory networks (GRNs) are conserved between organisms as
+highly diverged as nematodes and humans (for example, see~\cite{}). While genes
+and GRNs may be conserved between species, their outputs often differ however.
+For example, although homeobox genes are known to be important for limb
+formation in many animals~\cite{}, these genes do not form limbs in nematodes.
+The concept of a phenolog has been put forward to explain relationships between
+phenotypes that have the same underlying genetic regulatory network~\cite{}.
+Formally, two phenotypes are phenologs of each other if the homologs of the
+genes that cause a phenotype in an organism cause a second phenotype in another.
+
+In order to study a clinically relevant disease, an appropriate model has to be
+established. A straightforward method towards establishing a disease model in
+\cel{} is to link a disease to a causal gene, then to identify the homologous
+gene in \cel{} and then to study the function of the genetic homolog to
+extrapolate back to humans. However, this method relies on the existence of
+known disease genes and requires that the homolog have a phenotype that can be
+reliably identified and studied. A fundamentally different way to establish a
+disease model in \cel{} would be to identify the phenologs of the disease to be
+studied in \cel{}, and to use that phenolog as the basis for screens to identify
+genes that are associated with that phenolog. This approach has been
+successfully used in the past to make non-obvious links between phenotypes in
+different species~\cite{}.
+
+In order to facilitate understanding of large datasets, and to make discovery of
+phenologs easier, we have developed a complete enrichment tool suite in WormBase
+that allows users to rapidly perform all analyses on curated \cel{} ontologies
+using the same methodology for each one. They are located at
+% TODO: include links.
+We applied our tools towards the unbiased discovery of phenologues of
+multigenic, complex diseases including schizophrenia, type-2 diabetes,
+crohn's disease and prostate cancer by using genes associated with these
+diseases via genome-wide association studies. We also illustrate the utility of
+the complete enrichment suite for finding new relationships in complex data by
+analyzing a ciliary neuron transcriptome~\cite{}.
+
+\section*{Methods}
 
 \section*{Results}
+\subsection*{Developing the WormBase Enrichment Suite}
+% TODO add numbers instead of XXX
+We developed the dictionaries for PEA and GEA using the same procedure as was
+used for TEA~\cite{}. We generated a dictionary that included terms with at
+least 50 annotating genes or more and had a similarity threshold of 0.95 for PEA
+(the total number of terms in the dictionary is XXX);\@ and we generated a
+dictionary that included terms with at least 100 annotating genes or more and
+had a similarity threshold of 0.95 for GEA (the total number of terms in this
+dictionary is XXX).\@ Next, we benchmarked the dictionaries on the same gene
+sets as TEA and obtained enrichment of all the expected categories. For example,
+on a gene set enriched for embryonic muscle genes~\cite{}, the top two enriched
+phenotype terms by q-value were `muscle system morphology variant' and `body
+wall muscle thick filament variant'; the top two enriched GO terms were
+`developmental process' and `contractile fiber'. For all the benchmarking
+results, see supplementary information. Having generated and validated our
+dictionaries, we proceeded to identify phenologs for several common human diseases.
 
-\subsection*{Developing PEA and GEA}
+\subsection*{Applying the WormBase Enrichment Suite}
+In order to discover phenologs, we first needed to identify genes that
+contribute to a disease in an unbiased manner. One way to discover gene
+associations in an unbiased manner is to perform genome-wide association
+studies (GWAS) in human populations. Therefore, we used the GWAS NHGRI-EBI
+Catalog~\cite{} to identify genes associated with human diseases. We found the
+best nematode candidate homologs for these genes using DIOPT~\cite{} and applied
+our enrichment suite to each of these gene regulatory networks.
 
-\subsection*{Applying the WormBase Enrichment Suite to Discover Disease Phenologues}
-\subsubsection*{Obesity}
+\subsubsection*{Obesity-related traits}
+Human genes in Obesity-related traits: 957\\
+Worm genes in Obesity-related traits: 614
+
+Obesity-related traits is a category within the GWAS NHGRI-EBI catalog that
+pools studies that have measured obesity and other traits associated with
+obesity, such as heart rate, physical activity, hormone levels, body composition
+and cholesterol levels. Since this category includes many parameters,
+we expected there would be many phenologs. GWAS studies have identified
+\hobesity{} genes associated with these traits. Using DIOPT, we found
+\wobesity{} homologs for these genes, of which XXX had a phenotype.
+
+Top results for obesity-related traits included `acetylcholinesterase inhibitor
+response variant', `behavior variant', `neurite morphology variant' and `thin'.
+Terms involving locomotion were significantly enriched, as were terms involving
+body shape and food consumption ($q<0.1$). Concomitant with these phenologs was
+an enrichment in tissues including the \cel{} `tail', `sex organ', and an
+eclectic collection of neuronal tissues and cells.
+
 \subsubsection*{Schizophrenia}
+Human genes in schizophrenia: 899\\
+Worm genes in schizophrenia: 433
+
+Schizophrenia results included `acetylcholinesterase inhibitor response variant',
+multiple terms involving vulval formation and induction, cell division defects
+(`sister chromatid segregation defective early embryo', `cytokinesis variant')
+and terms involving lethality or arrest at multiple stages. Tissue enrichment
+showed that the anal depressor and sphincter muscles were significantly
+overrepresented, as were neuronal tissues. Somatic gonad, including the
+spermatheca and the distal tip cell were also significantly overrepresented in
+this dataset.
+
 \subsubsection*{Type-2 Diabetes}
+Human genes in diabetes: 396\\
+Worm genes in diabetes: 224
+
+The phenolog results for type-2 diabetes included `somatic gonad morphology
+variant', and `Q neuroblast lineage migration variant'.
+
 \subsubsection*{Crohn's Disease}
+Human genes in Crohn's Disease: 390\\
+Worm genes in Crohn's Disease: 213
+
+Crohn's disease in the worm presents traits including `brood size variants',
+`P granule defective', `cell fate transformation', `male mating defective' as
+well as various defects involving the \cel{} gonad. Reflecting this, TEA showed
+that Crohn's associated genes in the worm were enriched in the reproductive
+tract and somatic gonad, as well as the dorsal, ventral and lateral nerve cords
+and the nerve ring.
+
 \subsubsection*{Prostate Cancer}
+Human genes in prostate cancer: 273\\
+Worm genes in prostate cancer: 273
+
+Genes associated with prostate cancer in nematodes were associated with cell fate
+transformations, adult lethal phenotypes and P granule localization defects,
+and enriched tissues included the nerve cords, the sex organs and the tail of the
+animal.
 
 \subsection*{Ontology Enrichment as an aid for Screen Design}
-\subsubsection*{Enrichment in the Linker Cell Transcriptome}
-\subsubsection*{Enrichment in a Linker Cell Screen}
+An additional use for a tool like PEA would be as a tool to help guide and
+design screens to identify genes from an RNA-seq or other genome-wide experiment
+for further study. This would be particularly useful in cases when researchers
+may not know what phenotype to expect, in which case PEA can guide selection of
+a phenotype. Another use case is a scenario where the phenotype under study is
+not easy to screen for. By finding phenologs to the phenotype of interest, the
+researcher can design an easier screen for genes that affect the phenolog in
+question, then re-test genes for the original phenotype of interest.
+
+\subsubsection*{Enrichment in the Ciliary Neuronal Transcriptome}
+As an example of how ontology enrichment can improve our understanding of
+transcriptomes,
+we selected a ciliary neuron dataset~\cite{} and ran the complete WormBase
+Enrichment Suite on it. Ciliary neurons are present in the \cel{} male tail, but
+they are also present in the male cephalic sensillum and hermaphrodites also have
+ciliated neurons. The results are illuminating. PEA reveals that the ciliated
+neuron transcriptome is enriched for genes that are typically associated with
+chromosome segregation, aneuploidy and spindle defects. This makes sense---cilia
+employ microtubules for structural integrity, and microtubules are required for
+chromosome segregation. Whereas PEA suggests that this dataset is enriched in
+microtubule defects, TEA points at the \cel{} gonad primordium, somatic gonad
+and germline precursors as the sites where genes associated with ciliary neurons
+are enriched.  \cel{} has a stereotyped cell lineage~\cite{}, and no cellular
+division happens after a certain point in post-embryonic development, with the
+exception of the germline.
+
+\section*{Conclusions}
 

### 2017-01-05 pea_paper_docs/src/pea_benchmarks.py

@@ -19,6 +19,8 @@ wbids = pd.read_csv('../input/wormbase_ids.txt', sep='\t', header=None,
                     names=['wbid', 'human_read', 'target_id'])
 
 ciliated_df = pd.read_excel('~/Downloads/mmc3.xlsx')
+
+
 # define a function to run e.a.s quickly
 def run(path, f, output_path, dictionary, column='gene', **kwargs):
     """Run and save e.a.."""
@@ -63,130 +65,130 @@ for f in os.listdir('../input/hypoxia'):
 #     print(disease, snp[snp.Disease_Trait == disease].shape[0])
 
 
-depression = ['Major depressive disorder', 'Major depressive disorder (broad)']
-celiac = ['Celiac disease', 'Celiac disease and Rheumatoid arthritis']
-macular_degeneration = ['Age-related macular degeneration',
-                        'Age-related macular degeneration (CNV vs. GA)',
-                        'Age-related macular degeneration (CNV)',
-                        'Age-related macular degeneration (GA)',
-                        'Age-related macular degeneration (extreme sampling)',
-                        'Age-related macular degeneration (wet)']
-alcoholism = ['Alcohol and nictotine co-dependence', 'Alcohol consumption',
-              'Alcohol dependence', 'Alcoholism' +
-              '(12-month weekly alcohol consumption)',
-              'Alcoholism (alcohol dependence factor score)',
-              'Alcoholism (alcohol use disorder factor score)',
-              'Alcoholism (heaviness of drinking)']
-alzheimers = ['Alzheimer\'s disease', 'Alzheimer\'s disease (age of onset)',
-              'Alzheimer\'s disease (cognitive decline)',
-              'Alzheimer\'s disease (late onset)',
-              'Alzheimer\'s disease (neuritic plaque pathology)',
-              'Alzheimer\'s disease biomarkers']
-adhd = ['Attention deficit hyperactivity disorder',
-        'Attention deficit hyperactivity disorder (combined symptoms)',
-        'Attention deficit hyperactivity disorder'
-        '(hyperactivity-impulsivity symptoms)',
-        'Attention deficit hyperactivity disorder (inattention symptoms)',
-        'Attention deficit hyperactivity disorder (time to onset)',
-        'Attention deficit hyperactivity disorder and conduct disorder',
[…]

### 2017-01-05 tissue_enrichment_analysis/hypergeometricTests.py

@@ -27,20 +27,23 @@ def pass_list(user_provided, tissue_dictionary):
     tissue_dictionary is a pandas dataframe that you should know about
     user_provided is a list of gene names
     """
-    length = tissue_dictionary.shape[0]  # how many genes in the dict
-
-    # make an empty dataframe
-    present = pd.DataFrame(index=range(length), columns=['wbid', 'provided'])
-
-    # fill wbid column with all the gene names
-    present.wbid = tissue_dictionary.wbid
-
-    # go through and pass attendance -- 1 if present, 0 otherwise
-    for item in user_provided:
-        present.provided[present.wbid == item] = 1
-
-    # remove NA's and make them 0
-    present.provided = present.provided.fillna(0)
+    # changed 4 Jan 2017
+    # length = tissue_dictionary.shape[0]  # how many genes in the dict
+    #
+    # # make an empty dataframe
+    # present = pd.DataFrame(index=range(length), columns=['wbid', 'provided'])
+    #
+    # # fill wbid column with all the gene names
+    # present.wbid = tissue_dictionary.wbid
+    #
+    # # go through and pass attendance -- 1 if present, 0 otherwise
+    # for item in user_provided:
+    #     present.provided[present.wbid == item] = 1
+    #
+    # # remove NA's and make them 0
+    # present.provided = present.provided.fillna(0)
+    ind = tissue_dictionary.wbid.isin(user_provided)
+    present = tissue_dictionary[ind].wbid
 
     # return df
     return present
@@ -65,17 +68,18 @@ def hgf(gene_list, tissue_df):
     tissue_dictionary should be a pandas df
     """
     # figure out what genes are in the user provided list
-    present = pass_list(gene_list, tissue_df)
+    # present = pass_list(gene_list, tissue_df)
 
-    unused = present[present.provided == 0].wbid
+    # unused = present[present.provided == 0].wbid # 4 Jan 2017
 
     # slice out only the genes that were present from the user-provided list
-    wanted = present.wbid[present.provided == 1]
+    # wanted = present.wbid[present.provided == 1]
+    wanted = pass_list(gene_list, tissue_df)
 
     # re-index the dictionary s.t. the wbid is the index
     tissue_df = tissue_df.set_index('wbid')
 
-    # number of tissues in the dictionary
+    # number of balls per tissue in the dictionary
     sums_of_tissues = tissue_df.sum()  # this object can be identified by
[…]

### 2017-01-05 build/lib/tissue_enrichment_analysis/hypergeometricTests.py
changed the way TEA calculates hypergeometric
used to be the number of balls picked was = number of genes provided.
However, this is false. Number of balls picked = number of BALLS
provided.
so:
picked = sum over all balls

@@ -27,20 +27,23 @@ def pass_list(user_provided, tissue_dictionary):
     tissue_dictionary is a pandas dataframe that you should know about
     user_provided is a list of gene names
     """
-    length = tissue_dictionary.shape[0]  # how many genes in the dict
-
-    # make an empty dataframe
-    present = pd.DataFrame(index=range(length), columns=['wbid', 'provided'])
-
-    # fill wbid column with all the gene names
-    present.wbid = tissue_dictionary.wbid
-
-    # go through and pass attendance -- 1 if present, 0 otherwise
-    for item in user_provided:
-        present.provided[present.wbid == item] = 1
-
-    # remove NA's and make them 0
-    present.provided = present.provided.fillna(0)
+    # changed 4 Jan 2017
+    # length = tissue_dictionary.shape[0]  # how many genes in the dict
+    #
+    # # make an empty dataframe
+    # present = pd.DataFrame(index=range(length), columns=['wbid', 'provided'])
+    #
+    # # fill wbid column with all the gene names
+    # present.wbid = tissue_dictionary.wbid
+    #
+    # # go through and pass attendance -- 1 if present, 0 otherwise
+    # for item in user_provided:
+    #     present.provided[present.wbid == item] = 1
+    #
+    # # remove NA's and make them 0
+    # present.provided = present.provided.fillna(0)
+    ind = tissue_dictionary.wbid.isin(user_provided)
+    present = tissue_dictionary[ind].wbid
 
     # return df
     return present
@@ -65,17 +68,18 @@ def hgf(gene_list, tissue_df):
     tissue_dictionary should be a pandas df
     """
     # figure out what genes are in the user provided list
-    present = pass_list(gene_list, tissue_df)
+    # present = pass_list(gene_list, tissue_df)
 
-    unused = present[present.provided == 0].wbid
+    # unused = present[present.provided == 0].wbid # 4 Jan 2017
 
     # slice out only the genes that were present from the user-provided list
-    wanted = present.wbid[present.provided == 1]
+    # wanted = present.wbid[present.provided == 1]
+    wanted = pass_list(gene_list, tissue_df)
 
     # re-index the dictionary s.t. the wbid is the index
     tissue_df = tissue_df.set_index('wbid')
 
-    # number of tissues in the dictionary
+    # number of balls per tissue in the dictionary
     sums_of_tissues = tissue_df.sum()  # this object can be identified by
[…]

### 2017-01-05 pea_paper_docs/src/pea_benchmarks.py

@@ -10,6 +10,9 @@ phenotype_df = pd.read_csv('../input/phenotype_ontology.csv')
 go_df = pd.read_csv('../input/gene_ontology.csv')
 tissue_df = tea.fetch_dictionary()
 
+tissue_df.shape
+phenotype_df.shape
+
 snp = pd.read_csv('../input/known_snps.csv', sep=',', comment='#')
 linker_cell = pd.read_csv('../input/linker_cell.txt', sep='\t')
 lc_screen = pd.read_csv('../input/lc_screen.csv')
@@ -25,8 +28,8 @@ ciliated_df = pd.read_excel('~/Downloads/mmc3.xlsx')
 def run(path, f, output_path, dictionary, column='gene', **kwargs):
     """Run and save e.a.."""
     genes = pd.read_csv(path+f, **kwargs)
-    df, _ = tea.enrichment_analysis(genes[column], dictionary,
-                                    show=False)
+    df = tea.enrichment_analysis(genes[column], dictionary,
+                                 show=False)
     df = df[df.Observed > 2]
     df.to_csv(output_path + f + '.csv', index=False)
 
@@ -216,7 +219,7 @@ with open('gene_homologs.csv', 'w+') as F:
 candidates = []
 for trait in all_snps.TRAIT.unique():
     if all_snps[all_snps.TRAIT == trait].shape[0] > 200:
-        print(trait)
+        print(trait, all_snps[all_snps.TRAIT == trait].shape[0])
         candidates += [trait]
 
 # what are the pubmedids associated with the list of traits?
@@ -249,17 +252,19 @@ for trait in candidates:
         # one of the traits foolishly has a '/'
         if '/' in trait:
             trait = 'post bronchodilator fev1 fevc ratio'
-        df, _ = tea.enrichment_analysis(worm_genes, phenotype_df, show=False)
-        df = df[df.Observed > 3]
+        df = tea.enrichment_analysis(worm_genes, phenotype_df, show=False)
+        # print(_.head(), _.shape)
+        # break
+        df = df[df.Observed > 10]
         df.to_csv('../output/phenologues_2/pea_' + trait + '.csv', index=False)
 
-        df, _ = tea.enrichment_analysis(worm_genes, tissue_df, show=False)
-        df = df[df.Observed > 3]
+        df = tea.enrichment_analysis(worm_genes, tissue_df, show=False)
+        df = df[df.Observed > 10]
[…]

### 2017-01-05 setup.py

@@ -12,7 +12,7 @@ from setuptools import setup, find_packages
 import os
 import sys
 
-version = '0.14.00'
+version = '0.15.00'
 
 # just type in python setup.py publish and
 # this takes care of publishing to pypi!

### 2017-01-07 pea_paper_docs/tex/pea.tex
fixed command line script, added deconstructing ego paragraph
Wrote about how to deconstruct the egl phenotype in terms of tissues

--- 2017-01-04 pea_paper_docs/tex/pea.tex
@@
 
+\newcommand{\hobesity}{957}
+\newcommand{\wobesity}{614}
+
+\newcommand{\hlupus}{283}
+\newcommand{\wlupus}{135}
+
+\newcommand{\harthritis}{309}
+\newcommand{\warthritis}{124}
+
+\newcommand{\qval}[1]{
+                      \ensuremath{
+                                  q<10^{-#1}
+                                  }
+                      }
+
 \title{
   \Large
-  \textbf{Phenotype Enrichment Discovers Phenologues for Disease Modeling in \cel{}}
+  \textbf{
+  % Phenotype Enrichment Discovers Phenologs for Disease Modeling in \cel{}
+  Phenotype and gene ontology enrichment as guides for disease modeling
+  in \cel{}
+          }
 }
@@
 \section*{Introduction}
+The last decade has seen an explosion of techniques capable of genome-wide
+measurements. Some examples of genome-wide tools include RNA-seq~\cite{} to
+measure gene expression, CHIP-seq~\cite{} to measure binding, or genome-wide
+methylation profiling to understand epigenetic changes~\cite{}. These tools are capable
+of generating large quantities of data. Understanding these data, and generating
+hypotheses from them remains challenging. A common approach used to understand
+these datasets is to reduce the dimensionality of the data via enrichment
+analyses of ontologies~\cite{}, which helps researchers understand what terms are
+overrepresented beyond random levels. By analyzing overrepresented terms in
+aggregate, researchers can better understand what biological processes were most
+affected in a given experiment, and form hypotheses about what is
+happening~\cite{}.
+This approach is limited by what ontologies can be tested for
+enrichment.
+The best-known ontology for biological research is the Gene Ontology (GO), which
+provides a controlled language to describe molecular and cellular functions of
+genes~\cite{}. In \cel{}, curated tissue and phenotype ontologies exist which provide
+controlled languages with which to describe \cel{} anatomy and phenotypes
+respectively~\cite{}. However, enrichment tools only exist for gene and tissue ontologies
+in the community today (see for example~\cite{}). Another limitation is that tissue enrichment testing
+is not offered on the same websites as GO enrichment testing, which requires
+users to test their data on different websites that may or may not use different
+methodologies to detect enrichment.
+
+Another way to use enrichment tools is for evolutionary comparison purposes.
+In molecular biology it is often useful to know when a gene is homologous
+between two species---that is to say, common by descent---because knowledge of
+homology often brings with it knowledge of function~\cite{}. Indeed, many
+important gene regulatory networks (GRNs) are conserved between organisms as
+highly diverged as nematodes and humans (for example, see~\cite{}). While genes
+and GRNs may be conserved between species, their outputs often differ however.
+For example, although homeobox genes are known to be important for limb
+formation in many animals~\cite{}, these genes do not form limbs in nematodes.
+The concept of a phenolog has been put forward to explain relationships between
+phenotypes that have the same underlying genetic regulatory network~\cite{}.
+Formally, two phenotypes are phenologs of each other if the homologs of the
+genes that cause a phenotype in an organism cause a second phenotype in another.
+
+To study a clinically relevant disease in a non-human, an appropriate
+model has to be established. A straightforward method towards establishing a
+disease model in
+\cel{} is to link a disease to a causal gene, then to identify the homologous
+gene in \cel{} and then to study the function of the genetic homolog to
+extrapolate back to humans. However, this method relies on the existence of
+known disease genes and requires that the homolog have a phenotype that can be
+reliably identified and studied. A fundamentally different way to establish a
+disease model in \cel{} would be to identify the phenologs of the disease to be
+studied in \cel{} by identifying disease-associated human genes in an unbiased
+manner through genome-wide association studies (GWAS) and identified candidate
+homolog genes in \cel{}. The homologs can be used to identify \cel{} disease
+phenologs, which can in turn be used as the basis for screens to identify
+genes that are associated with that phenolog. Approaches similar to this have
+been successfully used in the past to make non-obvious links between phenotypes
+in different species~\cite{}.
+
+The concept of a phenolog can also be useful when applied within a species.
+In \cel{}, not all phenotypes are equally easy to study. Although genome-wide
+measurements can help elucidate the genetic network underlying a phenotype,
+devising screens to test which genes are functionally important can be difficult.
+A common strategy to study phenotypes that are difficult to screen is to
+select an easier-to-screen phenolog, and to test positive hits for the true
+phenotype of interest
+afterwards~\cite{}. Currently, selection of screening phenotypes is
+performed based on researcher experience.
+By formalizing phenotype enrichment analysis as a tool with which to analyze
+gene sets, researchers should be able to formally establish phenologs, which has
+consequences for screen design.
+
+An additional problem with genome-wide
+queries of \cel{} states (be they developmental, neuronal, or other) is that
+they do not
+always have a straightforward interpretation in terms of phenotypes. In these
+situations, researchers must rely on intuition to select a phenotype to screen
+for. As a result, many hits may go unexplored that would prove fruitful. The
+question of how to design a screen that is maximally informative is an important
+question that has so far not been addressed within this community.
+
+To facilitate understanding of large datasets, and to make discovery of
+phenologs easier, we have completed an enrichment tool suite in WormBase
+that allows users to rapidly perform phenotype, tissue and gene ontology
+enrichment analyses (PEA, TEA and GEA respectively) on curated \cel{} ontologies
+using the same methodology for each one. They are located at
+% TODO: include links.
+We applied our tools towards the unbiased discovery of phenologues of
+multigenic, complex diseases including
+% Diseases to study:
+systemic lupus erythematosus, obesity and obesity-related traits,  and
+rheumatoid arthritis
+by using genes associated with these
+diseases via genome-wide association studies. We also illustrate the utility of
+the complete enrichment suite for finding new relationships in complex data by
+analyzing a ciliary neuron transcriptome~\cite{}.
+
+\section*{Methods}
+\subsection*{Building the dictionaries}
+
+\subsection*{Human disease phenolog identification}
+We used the GWAS EBI-NHGRI catalog~\cite{} to extract information on all genome-wide
+association studies deposited there. We only selected traits that had $>300$
+associated genes. We identified 24 traits that met our criteria. Next, we used
+DIOPT~\cite{} to identify candidate homologs for the genes associated with these
+traits. Briefly, DIOPT combines a large number of methods for identifying homologs
+and returns homolog candidates associated with a compound score. Depending on the score,
+homologs can be considered `high', `moderate' or `low' rank, reflecting confidence
+in the homology. Many-to-one and one-to-many homology relationships are allowed
+in DIOPT, reflecting a mixture of uncertainty and family expansion/reduction.
+For our study, we only accepted homolog candidates with `high' or `moderate' scores
+and we did not insist on a one-to-one relationship between genes.
+
+After we identified worm homologs for each trait, we reassessed how many traits
+still had $>100$ gene candidates, and dropped all traits that had less than this
+for our analysis. We identified 18 traits that met this criteria. The gene
+lists for each of these 18 traits were then analyzed for gene, tissue and
+phenotype enrichment. Tissue enrichment was performed using the WormBase Tissue
+Enrichment Analysis tool (TEA)~\cite{}.
 
 \section*{Results}
+\subsection*{Developing the WormBase enrichment suite}
+% TODO add numbers instead of XXX
+We developed the dictionaries for PEA and GEA using the same procedure as was
+used for TEA~\cite{}. We generated a dictionary that included terms with at
+least 50 annotating genes or more and had a similarity threshold of 0.95 for PEA
+(the total number of terms in the dictionary was 251, annotated by 9,169 genes
+ for the version XXXXX);\@ and we generated a
+dictionary that included terms with at least 50 annotating genes or more and
+had a similarity threshold of 0.95 for GEA (the total number of terms in this
+dictionary is 271, annotated by 14,636 genes for the version XXXX).
+\@ Next, we benchmarked the dictionaries on the same gene
+sets as TEA and obtained enrichment of all the expected categories. For example,
+on a gene set enriched for embryonic muscle genes~\cite{}, the top two enriched
+phenotype terms by q-value were `muscle system morphology variant' and `body
+wall muscle thick filament variant'; the top two enriched GO terms were
+`myofibril' and `striated muscle dense body'. For all the benchmarking
+results, see supplementary information. Having generated and validated our
+dictionaries, we proceeded to identify phenologs for several common human
+diseases.
 
-\subsection*{Developing PEA and GEA}
+\subsection*{Applying the WormBase enrichment suite}
+To discover phenologs, we first needed to identify genes that
+contribute to a disease in an unbiased manner. One way to discover gene
+associations in an unbiased manner is to perform GWSA in human populations.
+Therefore, we used the GWAS NHGRI-EBI
+Catalog~\cite{} to identify genes associated with human diseases. We found the
+best nematode candidate homologs for these genes using DIOPT~\cite{} and applied
+our enrichment suite to each of these gene regulatory networks.
 
-\subsection*{Applying the WormBase Enrichment Suite to Discover Disease Phenologues}
-\subsubsection*{Obesity}
-\subsubsection*{Schizophrenia}
-\subsubsection*{Type-2 Diabetes}
-\subsubsection*{Crohn's Disease}
-\subsubsection*{Prostate Cancer}
+\subsubsection*{Obesity-related traits}
+Obesity-related traits is a category within the GWAS NHGRI-EBI catalog that
+pools studies that have measured obesity and other traits associated with
+obesity, such as heart rate, physical activity, hormone levels, body composition
+and cholesterol levels. Since this category includes many parameters,
+we expected there would be many phenologs. GWAS studies have identified
+\hobesity{} genes associated with these traits. Using DIOPT, we found
+\wobesity{} homologs for these genes. In total, $341/\wobesity{}$ genes had
+at least one phenotype annotation; $548/\wobesity{}$ had at least one gene ontology term
+annotation; and $427/\wobesity{}$ had at least one tissue term annotation.
 
-\subsection*{Ontology Enrichment as an aid for Screen Design}
-\subsubsection*{Enrichment in the Linker Cell Transcriptome}
-\subsubsection*{Enrichment in a Linker Cell Screen}
+Top results for obesity-related traits included `acetylcholinesterase inhibitor
+response variant' (38 genes, \qval{6}),
+`neurite morphology variant' (21 genes, \qval{2}),
+and `thin' (31 genes, \qval{2}).
+Terms involving locomotion were significantly enriched, as were terms involving
+body shape and food consumption (\qval{1}). Concomitant with these phenologs was
+a tissue enrichment in neuron-related terms. GO enrichment suggested that these
+genes are participating in `iron ion binding' (40 genes, \qval{20}) and
+`tetrapyrrole binding' (37 genes, \qval{13}).
+
+Tissue and phenotype enrichment therefore suggest that obesity-related traits
+may be studied in \cel{} through neuron physiology and function, specifically
+with respect to acetylcholinesterase inhibitors. Moreover, GO enrichment
+implicates iron and tetrapyrrole binding as metabolic components of
+the obesity-related phenologs in \cel{}.
+
+\subsection*{Systemic lupus erythematosus}
+Systemic lupus is an autoimmune disease that is believed to be polygenic in
+nature~\cite{}. It mainly affects women and is characterized by painful
+and swollen joints, hair loss, and fatigue~\cite{}. Since worms do not have a
+cellular immune system, we were interested in what phenologs corresponded to
+this disorder in \cel{}. To establish phenolog candidates, we obtained
+\hlupus{} genes associated with the disease via GWAS studies, and found
+\wlupus{} homolog candidates in \cel{}.
+
+Lupus-associated homologs were reasonably well annotated. Slightly more than half
+of the genes had at least one phenotype annotation ($76/\wlupus{}$) and almost
+all genes were annotated to at least one tissue or gene ontology term
+($104/\wlupus{}$ and $115/\wlupus$ genes respectively).
+We found that Lupus-associated homologs were enriched in `aneuploidy' (7 genes,
+\qval{1}) and `meiotic chromosome segregation' (8 genes, \qval{1}). `Cell fate
+transformation' (6 genes, \qval{1}), and `excess intestinal cells'
+(5 genes, \qval{1}) were also overrepresented, as was `male tail morphology'
+(6 genes, \qval{1}). Finally, the phenotype `nonsense mRNA accumulation' was also
+enriched (5 genes, \qval{1}).
+Meanwhile, TEA suggested that the
+`excretory duct cell' (5 genes, \qval{2}) and the `posterior gonad arm' are
+overrepresented in this dataset. We also found that the Pn.p cells P3.p through
+P8.p were enriched in this dataset (5 genes, \qval{1}). GO enrichment pointed
+at `modification-dependent macromolecule catabolic process' (23 genes,
+\qval{15}) as a molecular function that characterizes this dataset. However,
+this GO term was enriched only due to a single gene family, the \emph{skr} gene
+family. Almost the entire \emph{skr} family was considered a candidate homolog
+to the SKP1 human gene, making the GO enrichment suspect.
+
+Enrichment of the terms for `aneuploidy', `meiotic chromosome segregation',
+and `excess intestinal cells' were largely driven by the same gene group, which
+includes \emph{cki-1}, and several \emph{skr} genes. On the other hand, `cell
+fate transformation' and `male tail morphology' reflected the involvement of
+developmental genes \emph{let-23}, and \emph{lin-12} among others. The term
+`nonsense mRNA accumulation' was the result of \emph{pept-3}, \emph{smg-7},
+\emph{tsr-1}, \emph{dhcr-7} and \emph{F08B4.7}. Therefore, we conclude that
+systemic lupus erythematosus is potentially represented by a combination of three
+phenotypes in \cel{}: A cell proliferation phenotype (either increased or
+decreased), probably marked by increased
+aneuploidy; a developmental phenotype involving cell fate transformation and
+leading to dysmorphias; and a molecular phenotype involving impairment of the
+nonsense-mediated decay pathway.
+% TODO: Is the text below true???
+The results from the tissue enrichment analysis
+highlighted three tissues that are particularly sensitive to \emph{lin} mutations
+(the gonad, the excretory duct cell and the vulval precursor cells),
+and the gonad arms undergo large quantities of nuclear proliferation.
+
+\subsection*{Rheumatoid arthritis}
+Rheumatoid arthritis is an auto-immune disease that is characterized by
+swollen and painful joints that progressively deteriorate~\cite{}. Unlike lupus,
+rheumatoid arthritis is not life-threatening~\cite{}, and comorbidity between rheumatoid
+arthritis and lupus is low~\cite{}, suggesting that they may have at least
+partially distinct genetic causes. We found \harthritis{} genes associated with
+rheumatoid arthritis, for which we found \warthritis{} worm homolog candidates.
+
+% TODO: Is no tissue enrichment observed bc poor annotations, or because
+% all tissues randomly represented?
+The only phenotype that was enriched for these homologs was `short' (10 genes,
+\qval{4}), even though 64 homologs were associated with at least
+one phenotype term. No tissue was enriched in this dataset. Because
+82 genes are annotated to have expression in at least one tissue, the lack of
+enrichment does not reflect ignorance about the sites of expression of these
+genes. GEA showed that enriched
+molecular functions for these genes include `collagen trimer' (22 genes,
+\qval{15}). However, this term was enriched as the result of degeneracy in the
+homolog candidates for the SFTPD gene. Other terms included
+ `glycosylation' (10 genes, \qval{4}) and `Golgi apparatus' (11
+genes, \qval{3}), but this enrichment was also the result of degenerate homolog
+candidates for the human gene B3GNT7 which encodes a
+beta-1,3-N-acetylgalactosaminyltransferase.
+
+The `short' phenotype was the result of the \emph{cat-4}, \emph{dpy-7},
+\emph{rnt-1}, \emph{sem-4}, \emph{unc-116}, \emph{ocrl-1} and some genes in the
+\emph{fat} family. Although these genes are bound by a common phenotype, any
+genetic relationships between these genes are not immediately clear. Some genes,
+like \emph{sem-4} and \emph{rnt-1} are likely transcription factors with roles
+in development (including hypodermal development).
+Others are molecular motors (\emph{unc-116}) that are broadly expressed throughout
+the body of \cel{}. Yet others have known roles in neuron and muscle function,
+such as \emph{ocrl-1} and \emph{cat-4}. The `short' phenotype is a subset of
+the `body length variant' phenotype. Body length in \cel{} can be controlled via
+cell size, shape and number\cite{}; alternatively, cuticle development can alter body
+shape~\cite{}; finally, muscles can alter the effective body length artificially~\cite{}.
+
+\subsection*{Ontology Enrichment as an aid for screen design}
+An additional use for a tool like PEA would be as a tool to help guide and
+design screens to identify genes from an RNA-seq or other genome-wide experiment
+for further study. This would be particularly useful in cases when researchers
+may not know what phenotype to expect, in which case PEA can guide selection of
+a phenotype. Another use case is a scenario where the phenotype under study is
+not easy to screen for. By finding phenologs to the phenotype of interest, the
+researcher can design an easier screen for genes that affect the phenolog in
+question, then re-test genes for the original phenotype of interest.
+
+\subsubsection*{Enrichment in the ciliary neuronal transcriptome}
+As an example of how ontology enrichment can improve our understanding of
+transcriptomes,
+we selected a ciliary neuron dataset~\cite{} and ran the complete WormBase
+Enrichment Suite on it. Ciliary neurons are present in the \cel{} male tail, but
+they are also present in the male cephalic sensillum and hermaphrodites also have
+ciliated neurons. PEA reveals that the ciliated
+neuron transcriptome is enriched for genes that are typically associated with
+`meiotic chromosome segregation' (46 genes, \qval{5}),
+`aneuploidy' (42 genes, \qval{5}) and `spindle defective early embryos' (45
+genes, \qval{2}).
+
+In addition, TEA points at the \cel{} gonad primordium, the somatic gonad
+and early embryonic cells as the sites where genes associated with ciliary neurons
+are enriched. The `male distal tip cell' is a tissue
+that is overrepresented in this dataset, but `distal tip cell' is not enriched,
+which suggests that structures that are present only in the male tissue are
+overrepresented in this dataset.
+
+Although one interpretation of the results would be that microtubule genes
+are driving the enrichment of these terms, another possibility is that
+there are cell-cycle genes that are driving the enrichment of these phenotypes
+and tissues. Indeed, GO enrichment shows terms such as `DNA replication' (29
+genes, \qval{5}), and `purine NTP-dependent helicase activity'
+(15 genes, \qval{1}). Visual inspection of list in question reveals that
+cell-cycle and DNA replication/repair genes are abundant in this transcriptome
+and include genes such as \emph{atm-1}, \emph{dna-2}, or \emph{hpr-17}. This
+analysis reveals that the ciliary neuron transcriptome is enriched in
+genes associated with microtubules, but the cell-cycle machinery is also
+carefully regulated.
+
+\subsection*{Deconstructing the \emph{egl} phenotype}
+How does a phenotype emerge? We realized that with the tools that we have
+developed, it is possible to understand what tissues contribute to a phenotype
+in a probabilistic framework. In other words, we can extract all genes associated
+with a particular phenotype, then search for tissue terms that are enriched to
+understand how a phenotype arises from interactions between anatomical regions.
+As a test of this, we selected the \emph{egg-laying defective} (\emph{egl}) phenotype.
+\emph{A priori}, it is not clear what tissues may contribute the most towards
+an \emph{egl} phenotype. In \cel{}, egg-laying is a complex behavior that involves
+a large number of tissues. The somatic gonad acts as a repository for the eggs,
+the uterine seam cells help protect the uterus, and a variety of muscles help
+contract the uterus and open the vulva to lay an egg. The vulva must be well-formed
+to allow passage of an egg, and the hermaphrodite-specific neuron (HSN) helps the
+animal sense lay eggs, possibly by sensing pressure on the uterus. The
+complexity of the interactions that happen to allow egg-laying make understanding
+the \emph{egl} phenotype in terms of tissues a challenging task.
+
+We extracted all of the \cel{} genes that have been associated with an \emph{egl}
+phenotype and we used TEA to understand what tissues are enriched. The `HSN'
+was the most enriched tissue contributing to \emph{egl} phenotypes (\qval{13}).
+Next, the vulD and vulC vulval cells were most enriched respectively (\qval{7}).
+The somatic gonad was fourth most enriched (\qval{6}). The next biggest contribution
+comes from the Pn.p cells and the vulE, vulF cells (\qval{6}), followed by the
+vulA, vulB2 and vulB1 cells (\qval{5}). The uterine muscles are enriched at lower
+levels (\qval{2}), comparably to the uterine seam cells. Therefore, the \emph{egl}
+phenotype would seem to emerge primarily from defects in sensing (via the HSN),
+secondarily from defects in the vulva, and only sometimes from defects in the
+uterine seam cells or muscles. It is notable that the vulva was not equally
+enriched. Although all the `vul' cells are annotated to a similar degree (between
+50--70 genes for each cell type), the vulD and vulE cells were considerably more
+likely to be associated with an \emph{egl} phenotype than the others. This may
+reflect the fact that vulD and vulE are the site of attachment for four vulval
+muscles, vm1. Perhaps this attachment is particularly fragile, and perturbations
+to these cells prevents adequate function of these muscles. However, vulF is
+also attached to a set of four additional vulval muscles, vm2. Why
+is vulF less associated with an \emph{egl} phenotype?
+
+\section*{Conclusions}
 

### 2017-01-07 build/lib/tissue_enrichment_analysis/hypergeometricTests.py

@@ -69,9 +69,7 @@ def hgf(gene_list, tissue_df):
     """
     # figure out what genes are in the user provided list
     # present = pass_list(gene_list, tissue_df)
-
     # unused = present[present.provided == 0].wbid # 4 Jan 2017
-
     # slice out only the genes that were present from the user-provided list
     # wanted = present.wbid[present.provided == 1]
     wanted = pass_list(gene_list, tissue_df)
@@ -84,7 +82,7 @@ def hgf(gene_list, tissue_df):
     # the column names of tissues and excludes gene IDs
 
     # total size of the urn
-    total_genes = tissue_df.sum().sum()  # total labels in the dictionary
+    total_balls = tissue_df.sum().sum()  # total labels in the dictionary
 
     # slice out the rows from tissue_dictionary that came from the list
     wanted_dictionary = tissue_df.loc[wanted]
@@ -92,22 +90,22 @@ def hgf(gene_list, tissue_df):
     # get the total number of labels from each tissue
     wanted_sum = wanted_dictionary.sum()
 
-    # get the total number of genes provided by the user that are in dictionary
-    total = wanted.shape[0]
+    # get the total number of balls provided by the user that are in dictionary
+    picked = wanted_sum.sum()
 
     # make a hash with the p-values for enrichment of each tissue.
     p_hash = {}
     exp_hash = {}  # expected number for each tissue
     for i, name in enumerate(tissue_df.columns.values):
         # if the total number of genes is zero, return p= 1 for all tissues
-        if total == 0:
+        if picked == 0:
             p_hash[name] = 1
         else:
             # if a certain tissue has never been called, don't test it
             # in certain pathological cases where the list size is small
             # ie. close to 1, probability of never drawing a given label
             # can be close to 1, so survival function can be less than 0.05
-            # although the math is right, this behaviour is clearly undesirable
+            # although the math is right, this behavior is clearly undesirable
             if wanted_sum[name] == 0:
                 p_hash[name] = 1
             else:
@@ -116,14 +114,11 @@ def hgf(gene_list, tissue_df):
                 # total number of balls of color name in urn
                 # total number of balls picked out
                 n_obs = wanted_sum[name]
-                t_dict = total_genes
                 s_tissue = sums_of_tissues[name]
-                t_picked = total
-#                print(n_obs,t_dict,s_tissue,t_picked)
[…]

### 2017-01-07 pea_paper_docs/src/dictionary_generator.py

@@ -300,12 +300,10 @@ class ontology():
                 continue
             self.nodes[val['id']].get_name(val['annotation_class_label'])
 
-    def find_node_family(self, lambda_query_rlshp):
+    def find_node_family(self, lambda_query_rlshp, p=0):
         """Find the nodes that are related to this one."""
-        p = 1
         for n in iter(self.nodes):
             self.nodes[n].find_family(self.solr_url, lambda_query_rlshp, p=p)
-            p = 0
 
     def find_node_annotations(self, lambda_query_genes):
         """Fetch the annotations for this node."""
@@ -500,16 +498,23 @@ if __name__ == '__main__':
                 "%22{0}%22&fq=document_category:%22ontology_class%22"
         return s.format(x)
 
-    def query_genes(x):
+    def query_genes(x, ontology=args.ontology):
         """
         find the genes associated with every node.
 
         given a wbbt ID `x`, open URL that contains genes assoc. with it.
         """
-        s = "select?qt=standard&indent=on&wt=json&version=2.2&" +\
-            "fl=id&start=0&rows=10000&q=document_category:bioentity" +\
-            "&fq=source:%22WB%22&fq=-qualifier:%22not%22&" +\
-            "fq=regulates_closure:%22{0}%22"
+        if ontology != 'go':
+            s = "select?qt=standard&indent=on&wt=json&version=2.2&" +\
+                "fl=id&start=0&rows=10000&q=document_category:bioentity" +\
+                "&fq=source:%22WB%22&fq=-qualifier:%22not%22&" +\
+                "fq=regulates_closure:%22{0}%22"
+        else:
+            s = "select?qt=standard&indent=on&wt=json&version=2.2&" +\
+                "fl=id&start=0&rows=10000&q=document_category:bioentity" +\
+                "&fq=source:%22WB%22&fq=taxon:%22NCBITaxon:6239%22" +\
+                "&fq=-qualifier:%22not%22&" +\
+                "fq=regulates_closure:%22{0}%22"
         return s.format(x)
 
     # query for readable names
@@ -581,7 +586,7 @@ if __name__ == '__main__':
 
     for n in trial1.good:
         tissues.append(n)
-        print(n)
+        # print(n)
         genes = genes+trial1.good[n].genes
     genes = list(set(genes))
 

### 2017-01-07 pea_paper_docs/src/pea_benchmarks.py

@@ -2,26 +2,55 @@
 """Benchmarking PEA."""
 import tissue_enrichment_analysis as tea
 import pandas as pd
-# import matplotlib.pyplot as plt
+import matplotlib.pyplot as plt
 import os
 
-# import data
+# focus diseases for the paper. Only do certain analyses on these:
+diseases = ['Systemic lupus erythematosus', 'Obesity-related traits',
+            'Rheumatoid arthritis']
+
+# import dictionaries
 phenotype_df = pd.read_csv('../input/phenotype_ontology.csv')
-go_df = pd.read_csv('../input/gene_ontology.csv')
+go_df = pd.read_csv('../input/go_dictionary.csv')
 tissue_df = tea.fetch_dictionary()
 
-tissue_df.shape
-phenotype_df.shape
+# figure out the shapes of each thing and print
+print("Tissue Dictionary:")
+print("Genes: {0}\nTerms: {1}".format(tissue_df.shape[0], tissue_df.shape[1]))
+print("\nPhenotype Dictionary:")
+print("Genes: {0}\nTerms: {1}".format(phenotype_df.shape[0],
+                                      phenotype_df.shape[1]))
+print("\nGO Dictionary:")
+print("Genes: {0}\nTerms: {1}".format(go_df.shape[0], go_df.shape[1]))
 
-snp = pd.read_csv('../input/known_snps.csv', sep=',', comment='#')
-linker_cell = pd.read_csv('../input/linker_cell.txt', sep='\t')
-lc_screen = pd.read_csv('../input/lc_screen.csv')
+# TODO: fix this path:
+# from the gwas catalog, identify diseases with more than 100 associated genes
+gwas_catalog = '~/Downloads/gwas_catalog_v1.0-associations_e87_r2016-12-12.tsv'
 
+# SNP/GWAS data
+# snp = pd.read_csv('../input/known_snps.csv', sep=',', comment='#')
+# load the homolog data for the SNP/GWAS data
+homologs = pd.read_csv('../input/gwas_homologs_worm.txt', sep='\t')
 
+# all wormbase IDs
 wbids = pd.read_csv('../input/wormbase_ids.txt', sep='\t', header=None,
                     names=['wbid', 'human_read', 'target_id'])
 
+# Ciliary Neuron Transcriptome (from Juan Wang and Maureen Barr)
 ciliated_df = pd.read_excel('~/Downloads/mmc3.xlsx')
+
+# the minimum number of observations to keep for the GWAS enrich. analysis
[…]

### 2017-01-07 setup.py

@@ -12,7 +12,7 @@ from setuptools import setup, find_packages
 import os
 import sys
 
-version = '0.15.00'
+version = '0.16.04'
 
 # just type in python setup.py publish and
 # this takes care of publishing to pypi!

### 2017-01-07 tissue_enrichment_analysis/hypergeometricTests.py

@@ -208,7 +208,6 @@ def enrichment_analysis(gene_list, tissue_df, alpha=0.05, aname='',
     gene_list: a list of non-redundant WBIDs
     tissue_df: as provided by WormBase (use fetch_dictionary)
     alpha: significance threshold, defaults to 0.05
-    f: filename for the enrichment analysis
     aname= filename to use to save results
     show= Whether to print results or not.
 

### 2017-01-10 pea_paper_docs/tex/pea.tex
added egl to tissue conditional analysis
and wrote the paragraph for it…..

@@
 happening~\cite{}.
-This approach is limited by what ontologies can be tested for
-enrichment.
+This approach is limited by what ontologies can be tested for enrichment.
 The best-known ontology for biological research is the Gene Ontology (GO), which
@@
 
-\subsection*{Deconstructing the \emph{egl} phenotype}
+\subsection*{Deconstructing the Egl phenotype}
 How does a phenotype emerge? We realized that with the tools that we have
@@
 understand how a phenotype arises from interactions between anatomical regions.
-As a test of this, we selected the \emph{egg-laying defective} (\emph{egl}) phenotype.
-\emph{A priori}, it is not clear what tissues may contribute the most towards
-an \emph{egl} phenotype. In \cel{}, egg-laying is a complex behavior that involves
-a large number of tissues. The somatic gonad acts as a repository for the eggs,
+As a test of this, we selected the egg-laying defective (Egl) phenotype.
+In \cel{}, egg-laying is a complex behavior that involves
+a large number of tissues~\cite{}. The somatic gonad acts as a repository for the eggs,
 the uterine seam cells help protect the uterus, and a variety of muscles help
-contract the uterus and open the vulva to lay an egg. The vulva must be well-formed
-to allow passage of an egg, and the hermaphrodite-specific neuron (HSN) helps the
-animal sense lay eggs, possibly by sensing pressure on the uterus. The
-complexity of the interactions that happen to allow egg-laying make understanding
-the \emph{egl} phenotype in terms of tissues a challenging task.
+contract the uterus and open the vulva to lay an egg~\cite{}. The vulva must be well-formed
+to allow passage of an egg, and the hermaphrodite-specific neuron (HSN) is involved
+in the egg-laying control~\cite{}. The complexity of the interactions that
+happen to allow egg-laying make understanding the Egl phenotype in terms of
+tissues a challenging task.
 
-We extracted all of the \cel{} genes that have been associated with an \emph{egl}
-phenotype and we used TEA to understand what tissues are enriched. The `HSN'
-was the most enriched tissue contributing to \emph{egl} phenotypes (\qval{13}).
-Next, the vulD and vulC vulval cells were most enriched respectively (\qval{7}).
-The somatic gonad was fourth most enriched (\qval{6}). The next biggest contribution
-comes from the Pn.p cells and the vulE, vulF cells (\qval{6}), followed by the
-vulA, vulB2 and vulB1 cells (\qval{5}). The uterine muscles are enriched at lower
-levels (\qval{2}), comparably to the uterine seam cells. Therefore, the \emph{egl}
-phenotype would seem to emerge primarily from defects in sensing (via the HSN),
-secondarily from defects in the vulva, and only sometimes from defects in the
-uterine seam cells or muscles. It is notable that the vulva was not equally
+We extracted all of the \cel{} genes that have been associated with an Egl
+phenotype and we used TEA to understand what tissues are enriched. The HSN
+was enriched more than five-fold above background (\qval{7}) as were vulD, vulC,
+vulE and vulF (\qval{6}). The vulA, vulB2 and vulB1 were
+enriched at slightly lower levels (\qval{5}), whereas the uterine muscles and
+uterine seam cells were enriched more than twice above background levels
+(\qval{2}).
+Therefore, the Egl phenotype would seem to emerge primarily from defects in the
+HSN, secondarily from defects in the vulva, and only sometimes from defects in the
+uterine seam cells or muscles. It is notable that all vulval cells were not equally
 enriched. Although all the `vul' cells are annotated to a similar degree (between
-50--70 genes for each cell type), the vulD and vulE cells were considerably more
-likely to be associated with an \emph{egl} phenotype than the others. This may
+50--70 genes for each cell type), the vulD and vulC cells had the largest
+enrichment effect size and the lowest q-values, suggesting that these cells
+are more likely to be associated with an Egl phenotype than the others. This may
 reflect the fact that vulD and vulE are the site of attachment for four vulval
 muscles, vm1. Perhaps this attachment is particularly fragile, and perturbations
-to these cells prevents adequate function of these muscles. However, vulF is
+to these cells prevent adequate function of these muscles. In support of these
+observations, P7.pa had the largest fold-enrichment of any tissue. In \cel{},
+P7.pa gives rise to vulD and vulC. However, vulF is
 also attached to a set of four additional vulval muscles, vm2. Why
-is vulF less associated with an \emph{egl} phenotype?
+is vulF less associated with an Egl phenotype?
 
 \section*{Conclusions}
+The addition of GO and Phenotype Ontology enrichment testing to WormBase marks
+an important step towards a unified set of analyses that can help researchers
+to understand genomic datasets. By offering tissue, phenotype and gene ontology
+enrichment on a single site, researchers will benefit both in the speed at which
+the analysis is completed, but also from the fact that the analyses are all
+carried out with the same methodology and the same underlying algorithms.
 

### 2017-01-10 pea_paper_docs/src/pea_benchmarks.py

@@ -203,11 +203,50 @@ for trait in candidates:
 
 
 for trait in diseases:
+    # get genes associated with trait:
+    g = all_snps[all_snps.TRAIT == trait]['REPORTED GENE(S)'].values
+    current = []
+    for i, gi in enumerate(g):
+        if type(gi) is not str:
+            continue
+        list_of_genes = [x.strip() for x in gi.split(',')]
+        for x in list_of_genes:
+            current += [x]
+
     sel1 = homologs.HID.isin(current)
     worm_genes = homologs[sel1 & sel2].WormBaseID
-
-    if 'Lupus' in trait:
-        print(pheno_traits[pheno_traits.term.str.contains(aneuploidy)].wbid)
+    print(trait)
+    # if 'lupus' in trait:
+        # ind1 = pheno_traits.term.str.contains('nonsense mRNA accumulation')
+        # ind2 = pheno_traits.wbid.isin(worm_genes)
+        # aneuploidy_genes = pheno_traits[ind1 & ind2].wbid
+        # whatever = wbids[wbids.wbid.isin(aneuploidy_genes)]
+        # print(whatever)
+        # ind1 = tissue_traits.term.str.contains('excretory duct cell')
+        # ind2 = tissue_traits.wbid.isin(worm_genes)
+        # aneuploidy_genes = tissue_traits[ind1 & ind2].wbid
+        # whatever = wbids[wbids.wbid.isin(aneuploidy_genes)]
+        # print(whatever)
+        # ind1 = go_traits.term.str.contains('modification-dependent macromolecule catabolic process')
+        # ind2 = go_traits.wbid.isin(worm_genes)
+        # aneuploidy_genes = go_traits[ind1 & ind2].wbid
+        # whatever = wbids[wbids.wbid.isin(aneuploidy_genes)]
+        # print(whatever)
+    if 'Rheumatoid' in trait:
+        # ind1 = pheno_traits.term.str.contains('short WBPhenotype:0000324')
+        # ind2 = pheno_traits.wbid.isin(worm_genes)
+        # aneuploidy_genes = pheno_traits[ind1 & ind2].wbid
+        # whatever = wbids[wbids.wbid.isin(aneuploidy_genes)]
+        # print(whatever)
+        # ind1 = tissue_traits.term.str.contains('excretory duct cell')
+        # ind2 = tissue_traits.wbid.isin(worm_genes)
[…]

### 2017-01-10 tissue_enrichment_analysis/hypergeometricTests.py

@@ -309,6 +309,7 @@ def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
 
     df: dataframe as output by implement_hypergmt_enrichment_tool
     y: One of 'Fold Change', 'Q value' or a user generated column
+    title - Title for the graph, also file name
     n_bars: number of bars to be shown, defaults to 15
     dirGraps: directory to save figures to. if not existent,
     generates a new folder

### 2017-01-10 pea_paper_docs/tex/pea.tex
modified hypergeometric scripts
plotting is difficult since I must remove WB identifiers, yet these are
different lengths depending on the ontology.

I added an ‘analysis’ variable to the plot_enrichment and
fetch_dictionary functions to let the user specify what analysis is
performed.

@@
 
+% more space between rows
+\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}
+
 \title{
@@
 respectively~\cite{}. However, enrichment tools only exist for gene and tissue ontologies
-in the community today (see for example~\cite{}). Another limitation is that tissue enrichment testing
-is not offered on the same websites as GO enrichment testing, which requires
-users to test their data on different websites that may or may not use different
-methodologies to detect enrichment.
+in the community today (see for example~\cite{}). Another limitation is that
+tissue enrichment testing is not offered on the same websites as GO enrichment
+testing, which requires users to test their data on different websites that may
+or may not use different methodologies to detect enrichment.
 
@@
 \section*{Methods}
-\subsection*{Building the dictionaries}
-
 \subsection*{Human disease phenolog identification}
@@
 
+% TODO: SO.... ?
+
 \subsection*{Ontology Enrichment as an aid for screen design}
@@
 
-\subsection*{Deconstructing the Egl phenotype}
+\subsection*{Deconstructing phenotype-tissue relationships}
+\subsubsection*{Tissue enrichment on the Egl gene set reveals cellular
+components of the phenotype}
 How does a phenotype emerge? We realized that with the tools that we have
@@
 
+\subsubsection*{Quantifying the anatomy-phenotype
+                mapping via Bayesian probabilities}
+Another way to understand the phenotype-anatomy mapping is by considering
+how informative a given anatomy term is on a particular phenotype, or vice-versa.
+To this end, we calculated two conditional probabilities that helped us answer
+this question. The first conditional probability,
+
+\begin{equation}
+  P( \text{a gene has Egl annotation} | \text{it is expressed in } X)
+\end{equation}
+
+answers the question: For a gene with an expression pattern that includes the
+tissue term X (i.e., the gene is expressed at least in X), what is the probability
+that this gene has an Egl phenotype (i.e., the phenotype annotations for this
+gene include Egl)? For simplicity, we can re-write this equation more succintly
+by removing a few words. The calculation of this probability is straightforward
+and follows from the definition of conditional probability:
+
+\begin{equation}
+  P(\text{Egl}|X) = \frac{N_\text{genes annotated Egl and X}}
+                         {N_\text{genes annotated with X}}.
+\label{egl_x}
+\end{equation}
+
+Equation~\ref{egl_x} measures how likely a gene is to be annotated with an Egl
+phenotype given that its expression pattern includes the term $X$. A related
+quantity (which is neither the inverse nor the complement) is the conditional
+probability that a gene which is annotated with at least the Egl phenotype is
+expressed in tissue $X$. That is to say,
+
+\begin{equation}
+  P(X|\text{Egl}) = \frac{N_\text{genes annotated Egl and X}}
+                         {N_\text{genes annotated with Egl}}.
+  \label{x_egl}
+\end{equation}
+
+Equation~\ref{x_egl} tells us how probable it is that any given gene that is
+annotated with an Egl phenotype includes $X$ as a tissue term. Taken together,
+equations~\ref{egl_x} and~\ref{x_egl} help us understand how predictive
+anatomic expression is of phenotypes, and how predictive phenotypes are of
+anatomic expression.
+
+\begin{table*}
+  \renewcommand{\familydefault}{\sfdefault}\normalfont{}
+  \centering{}
+  \ra{1.3}
+  \caption{Conditional probabilities for various tissues. The
+  first column shows the conditional probability that a gene has an Egl
+  phenotype given that it has expression in tissue $X$ (given by the row).
+  The second column shows the conditional probability that a gene has expression
+  in the anatomy term X given that it has an Egl phenotype. The first 9 terms
+  are the terms for which $P(\text{Egl}|X)$ is maximized. The last three terms
+  are the terms which have the highest $P(X|\text{Egl})$. For clarity, the Pn.p
+  cells are not shown even though $P(\text{Egl}|\text{Pn.p})\sim 0.24$.}
+
+  \begin{tabular}{@{}lcc@{}}
+  \toprule{}
+  Tissue & $P(\text{Egl}| X)$ & $P(X|\text{Egl})$\\
+  \bottomrule{}\\
+  % Data goes here
+  P7.pa & $0.30$ & $0.04$\\
+  HSN & $0.27$ & $0.11$\\
+  vulC & $0.27$ & $0.06$\\
+  vulD & $0.26$ & $0.07$\\
+  vulE & $0.25$ & $0.06$\\
+  vulF & $0.24$ & $0.06$\\
+  vulA & $0.24$ & $0.05$\\
+  vulB2 & $0.23$ & $0.05$\\
+  vulB1 & $0.22$ & $0.05$\\
+  nervous system & $0.02$ & $0.72$ \\
+  pharynx  & $0.00$ & $0.46$\\
+  sex organ & $0.07$ & $0.41$\\
+  tail & $0.05$ & $0.33$\\
+  \bottomrule{}
+  \end{tabular}
+\label{tab:cond_probs}
+\end{table*}
+
+We calculated the conditional probability that a gene has an Egl phenotype given
+that it's expression pattern includes a tissue term $X$ and we searched for the
+tissue terms that maximized this probability. The list of terms that maximized this
+probability reflected the results from running TEA on the subset of genes that have
+an Egl phenotype. We also calculated the conditional probability that a gene has
+expression in a tissue term $X$ given that it is annotated with an Egl phenotype
+and we searched for terms that maximized this probability. The terms that maximized
+this probability were `nervous system', `pharynx' (a body part with a lot of neurons),
+`sex organ' and `tail' (a body part with neurons and hypodermis). In general, the
+terms that had a high $P(\text{Egl}|X)$ did not have a high $P(X|\text{Egl})$.
+Additionally, the terms that had a high $P(X|\text{Egl})$ are broad terms that
+include a lot of cells, whereas the terms that had a high $P(\text{Egl}|X)$
+were considerably more specific.
+We conclude that the Egl phenotype arises from a small set of tissues. The Egl
+phenotype can be best predicted by genes with expression patterns that include
+at least one of a small number of cells (mainly vul cells, HSN). On the other
+hand, answering whether the expression pattern of a gene includes a particular
+anatomic region or tissue given that the gene has an Egl phenotype is hard to
+do for small tissues or single cells. However, guesses about what functional system
+or broad anatomic region may be affected by an Egl gene can be answered with
+confidence ($\sim70\%$ of the time, the nervous system is the system
+affected by an Egl mutant).
+
 \section*{Conclusions}
+
+We have highlighted three possible uses for phenotype and GO enrichment analyses.
+Taken together, the WormBase Enrichment Suite can help guide researchers as they
+search for a phenologue that is representative of a human disease that has no
+immediately obvious counterpart in the worm. Such phenologues may benefit from
+the fact that they represent an unbiased approach to disease modeling as long as
+the human genes are selected from unbiased screens. GWAS data suggests that certain
+aspects of lupus biology may be best understood from a developmental perspective
+in \cel{}. On the other hand, our results suggest that rheumatoid arthritis
+might not be effectively modeled in the worm, although it is possible that our
+inability to find phenologues more concretely is the result of the GWAS screens:
+If GWAS screens identified factors that lead to worse prognosis of rheumatoid
+arthritis, but not factors that cause onset of the disease, this could explain
+the lack of phenotype enrichment in \cel{}.
+
+% Phenologues
+
+% Screen Design
+
+% Curation and annotation biases
+
 The addition of GO and Phenotype Ontology enrichment testing to WormBase marks
@@
 enrichment on a single site, researchers will benefit both in the speed at which
-the analysis is completed, but also from the fact that the analyses are all
+the analysis is completed and from the fact that the analyses are all
 carried out with the same methodology and the same underlying algorithms.

### 2017-01-10 pea_paper_docs/src/pea_benchmarks.py

@@ -114,9 +114,16 @@ for g in all_snps['REPORTED GENE(S)'].values:
 
 # save the array of genes to a file for translation into DIOPT
 array_of_genes = list(set(array_of_genes))
+print('writing unique genes to gene_homologs.csv...')
+print('The following gene names could not be decoded (silly excel!)')
 with open('gene_homologs.csv', 'w+') as F:
     for g in array_of_genes:
-        F.write(g+'\n')
+        if type(g) is not str:
+            print(g, type(g))
+        try:
+            F.write(g+'\n')
+        except:
+            print(g)
 
 # find all the diseases with more than 200 associated genes and save the
 # list of traits that meet criterion
@@ -190,64 +197,83 @@ for trait in candidates:
         df = tea.enrichment_analysis(worm_genes, phenotype_df, show=False)
         df = df[df.Observed > n_min_obs]
         df.to_csv('../output/phenologues_2/pea_' + trait + '.csv', index=False)
+        if 'lupus' in trait:
+            print('Graphing PEA results for ', trait)
+            fig, ax = plt.subplot()
+            tea.plot_enrichment_results(df, title='../output/lupus_pea',
+                                        save=True)
 
         df = tea.enrichment_analysis(worm_genes, tissue_df, show=False)
         df = df[df.Observed > n_min_obs]
         df.to_csv('../output/disease_tissues_2/tea_' + trait + '.csv',
                   index=False)
+        if 'lupus' in trait:
+            print('Graphing TEA results for ', trait)
+            fig, ax = plt.subplot()
+            tea.plot_enrichment_results(df, title='../output/lupus_tea',
+                                        save=True)
+            tea.plot_enrichment_results(df, title='../output/lupus_tea',
+                                        save=True)
 
         df = tea.enrichment_analysis(worm_genes, go_df, show=False)
         df = df[df.Observed > n_min_obs]
         df.to_csv('../output/disease_go_2/gea_' + trait + '.csv',
                   index=False)
+        if 'lupus' in trait:
+            print('Graphing GEA results for ', trait)
+            fig, ax = plt.subplot()
+            tea.plot_enrichment_results(df, title='../output/lupus_gea',
[…]

### 2017-01-10 tissue_enrichment_analysis/hypergeometricTests.py

@@ -303,13 +303,15 @@ def enrichment_analysis(gene_list, tissue_df, alpha=0.05, aname='',
 
 
 def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
-                            n_bars=15, dirGraphs='', save=True, **kwargs):
+                            analysis='tissue', n_bars=15, dirGraphs='',
+                            save=True, **kwargs):
     """
     A plot function for TEA.
 
     df: dataframe as output by implement_hypergmt_enrichment_tool
     y: One of 'Fold Change', 'Q value' or a user generated column
     title - Title for the graph, also file name
+    analysis - one of `tissue`, `phenotype` or `go`
     n_bars: number of bars to be shown, defaults to 15
     dirGraps: directory to save figures to. if not existent,
     generates a new folder
@@ -325,6 +327,12 @@ def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
         print('dataframe is empty!')
         return
 
+    if analysis.lower() not in ['tissue', 'phenotype', 'go']:
+        raise ValueError('analysis variable must be one of' +
+                         '`tissue`, `phenotype` or `go`')
+
+    analysis = analysis.lower()
+
     ax = kwargs.pop('ax', None)
     ftype = kwargs.pop('ftype', 'svg')
 #
@@ -336,8 +344,17 @@ def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
                    ascending=[True, False], inplace=True)
 
     # added August 26 2016:
-    lenID = 11
-    hr_labels = df['Tissue'][:n_bars].str[:-lenID-1]
+    tissue_ID = 11
+    pheno_ID = 19
+    go_ID = 11
+
+    if analysis == 'phenotype':
+        hr_labels = df['Tissue'][:n_bars].str[:-pheno_ID-1]
+    elif analysis == 'tissue':
+        hr_labels = df['Tissue'][:n_bars].str[:-tissue_ID-1]
+    elif analysis == 'go':
+        hr_labels = df['Tissue'][:n_bars].str[:-go_ID-1]
+
     # plot first n_bars
 
     with sns.axes_style('whitegrid'):
@@ -345,7 +362,7 @@ def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
         ax = sns.barplot(x=df[y][:n_bars], y=hr_labels, ax=ax)
 
     # fix the plot to prettify it
[…]

### 2017-01-10 build/lib/tissue_enrichment_analysis/hypergeometricTests.py
hypergeometric tests had a hardcoded variable that was wrong
within the plot_enrichment function.
Also, forgot that version numbering should have changed more, since
backwards compatibility has been reduced…

@@ -208,7 +208,6 @@ def enrichment_analysis(gene_list, tissue_df, alpha=0.05, aname='',
     gene_list: a list of non-redundant WBIDs
     tissue_df: as provided by WormBase (use fetch_dictionary)
     alpha: significance threshold, defaults to 0.05
-    f: filename for the enrichment analysis
     aname= filename to use to save results
     show= Whether to print results or not.
 
@@ -304,12 +303,15 @@ def enrichment_analysis(gene_list, tissue_df, alpha=0.05, aname='',
 
 
 def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
-                            n_bars=15, dirGraphs='', save=True, **kwargs):
+                            analysis='tissue', n_bars=15, dirGraphs='',
+                            save=True, **kwargs):
     """
     A plot function for TEA.
 
     df: dataframe as output by implement_hypergmt_enrichment_tool
     y: One of 'Fold Change', 'Q value' or a user generated column
+    title - Title for the graph, also file name
+    analysis - one of `tissue`, `phenotype` or `go`
     n_bars: number of bars to be shown, defaults to 15
     dirGraps: directory to save figures to. if not existent,
     generates a new folder
@@ -325,6 +327,12 @@ def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
         print('dataframe is empty!')
         return
 
+    if analysis.lower() not in ['tissue', 'phenotype', 'go']:
+        raise ValueError('analysis variable must be one of' +
+                         '`tissue`, `phenotype` or `go`')
+
+    analysis = analysis.lower()
+
     ax = kwargs.pop('ax', None)
     ftype = kwargs.pop('ftype', 'svg')
 #
@@ -336,8 +344,17 @@ def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
                    ascending=[True, False], inplace=True)
 
     # added August 26 2016:
-    lenID = 11
-    hr_labels = df['Tissue'][:n_bars].str[:-lenID-1]
+    tissue_ID = 11
+    pheno_ID = 19
+    go_ID = 11
+
+    if analysis == 'phenotype':
+        hr_labels = df['Tissue'][:n_bars].str[:-pheno_ID-1]
+    elif analysis == 'tissue':
+        hr_labels = df['Tissue'][:n_bars].str[:-tissue_ID-1]
+    elif analysis == 'go':
+        hr_labels = df['Tissue'][:n_bars].str[:-go_ID-1]
+
     # plot first n_bars
 
[…]

### 2017-01-10 setup.py

@@ -12,7 +12,7 @@ from setuptools import setup, find_packages
 import os
 import sys
 
-version = '0.16.04'
+version = '0.16.10'
 
 # just type in python setup.py publish and
 # this takes care of publishing to pypi!

### 2017-01-12 pea_paper_docs/tex/pea.tex
added figures to paper

@@
 \usepackage{graphicx}
+% set the figures directory
+\graphicspath{{./figs/}}
 
@@
 
+\begin{figure}[htbp]
+  \renewcommand{\familydefault}{\sfdefault}\normalfont{}
+  \centering
+  \includegraphics[width=\linewidth]{gwas-design.pdf}
+  \caption{Experimental design for human-nematode phenologue identification.
+           We used GWAS candidates from the EBI-NHGRI catalog to identify
+           disease associated genes, then used DIOPT to identify candidate
+           homologs in \cel{}. Homologs were used to run phenotype, gene
+           and tissue ontology enrichment analyses to identify disease
+           phenologues.}
+\label{fig:gwas}
+\end{figure}
+
 After we identified worm homologs for each trait, we reassessed how many traits
@@
 lists for each of these 18 traits were then analyzed for gene, tissue and
-phenotype enrichment. Tissue enrichment was performed using the WormBase Tissue
-Enrichment Analysis tool (TEA)~\cite{}.
+phenotype enrichment (see~Fig.\ref{fig:gwas}). Tissue enrichment was
+performed using the WormBase Tissue Enrichment Analysis (TEA) tool~\cite{}.
 
@@
 (6 genes, \qval{1}). Finally, the phenotype `nonsense mRNA accumulation' was also
-enriched (5 genes, \qval{1}).
-Meanwhile, TEA suggested that the
+enriched (5 genes, \qval{1}) (see Fig.~\ref{fig:lupus}).
+
+\begin{figure}[htbp]
+  \renewcommand{\familydefault}{\sfdefault}\normalfont{}
+  \centering
+  \includegraphics[width=\linewidth]{systemic-lupus.pdf}
+  \caption{Phenologue identification for systemic lupus erythematosus.
+           \textbf{A} Phenotype Enrichment Analysis. \textbf{B} GO Enrichment
+           Analysis. \textbf{C} Lupus in \cel{} may be best represented by
+           a combination of three phenotypes: Cell proliferation possibly
+           accompanied by aneuploidy; cell fate transformations that may lead
+           to dysmorphias; and a molecular phenotype involving impairment of
+           the Nonsense-Mediated Decay (NMD) pathway.}
+\label{fig:lupus}
+\end{figure}
+
+TEA suggested that the
 `excretory duct cell' (5 genes, \qval{2}) and the `posterior gonad arm' are
@@
 `aneuploidy' (42 genes, \qval{5}) and `spindle defective early embryos' (45
-genes, \qval{2}).
+genes, \qval{2}) (see Fig.~\ref{fig:cilia}).
+
+\begin{figure}[htbp]
+  \renewcommand{\familydefault}{\sfdefault}\normalfont{}
+  \centering
+  \includegraphics[width=\linewidth]{ciliary-transcriptome.pdf}
+  \caption{PEA shows that the ciliary transcriptome is enriched in phenotypes
+  related to cell division. Although this could reflect enrichment of
+  microtubules and microtubule-related genes, the enrichment is at least
+  partially driven by cell-cycle and DNA repair genes.}
+\label{fig:cilia}
+\end{figure}
 
@@
 uterine seam cells were enriched more than twice above background levels
-(\qval{2}).
+(\qval{2}) (see Fig.~\ref{fig:egl}).
 Therefore, the Egl phenotype would seem to emerge primarily from defects in the
@@
 is vulF less associated with an Egl phenotype?
+
+\begin{figure}[htbp]
+  \renewcommand{\familydefault}{\sfdefault}\normalfont{}
+  \centering
+  \includegraphics[width=\linewidth]{Egl-phenotype.pdf}
+  \caption{The Egl phenotype is a complex phenotype that is the result of
+  interactions between many tissues. To dissect the contributions of individual
+  tissues to generating an Egl phenotype, we obtained all genes annotated with
+  with an Egl term. \textbf{A} We used TEA on the set of Egl genes to identify
+  enriched tissues. \textbf{B} Anatomic diagram showing the tissues that are
+  most enriched in the Egl gene set. Color coding shows the qualitative
+  ordering of enrichment (red-Most enriched, yellow-least enriched). For clarity,
+  not all cells are shown. All vm1 (4 cells) and vm2 (4 cells) muscles are
+  symmetrical arranged around the vulva, but only 2 cells are shown for each.
+  There are two seam cells on the left and right side of the vulva, but only
+  cell on the right is shown. Only cells in the adult are shown.}
+\label{fig:egl}
+\end{figure}
 
@@
 do for small tissues or single cells. However, guesses about what functional system
-or broad anatomic region may be affected by an Egl gene can be answered with
-confidence ($\sim70\%$ of the time, the nervous system is the system
-affected by an Egl mutant).
+or broad anatomic region an Egl gene is expressed in can be answered with
+confidence ($\sim70\%$ of the time, an Egl mutant is expressed in
+the nervous system).
 

### 2017-01-12 build/lib/tissue_enrichment_analysis/hypergeometricTests.py

@@ -346,7 +346,7 @@ def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
     # added August 26 2016:
     tissue_ID = 11
     pheno_ID = 19
-    go_ID = 11
+    go_ID = 10
 
     if analysis == 'phenotype':
         hr_labels = df['Tissue'][:n_bars].str[:-pheno_ID-1]

### 2017-01-12 pea_paper_docs/src/pea_benchmarks.py

@@ -169,7 +169,7 @@ for trait in candidates:
 
     current = list(set(current))
     sel1 = (homologs.HID.isin(current))
-    worm_genes = homologs[sel1 & sel2].WormBaseID
+    worm_genes = homologs[sel1 & sel2].WormBaseID.values
 
     # only run E.A. if there are enough homologs:
     if worm_genes.shape[0] > 100:
@@ -195,40 +195,34 @@ for trait in candidates:
 
         # enrichment analyses:
         df = tea.enrichment_analysis(worm_genes, phenotype_df, show=False)
-        df = df[df.Observed > n_min_obs]
+        df = df[df.Observed > n_min_obs].copy()
         df.to_csv('../output/phenologues_2/pea_' + trait + '.csv', index=False)
         if 'lupus' in trait:
             print('Graphing PEA results for ', trait)
-            fig, ax = plt.subplot()
+            fig, ax = plt.subplots()
             tea.plot_enrichment_results(df, title='../output/lupus_pea',
-                                        save=True)
+                                        save=True, analysis='phenotype')
 
         df = tea.enrichment_analysis(worm_genes, tissue_df, show=False)
-        df = df[df.Observed > n_min_obs]
+        df = df[df.Observed > n_min_obs].copy()
         df.to_csv('../output/disease_tissues_2/tea_' + trait + '.csv',
                   index=False)
         if 'lupus' in trait:
             print('Graphing TEA results for ', trait)
-            fig, ax = plt.subplot()
+            fig, ax = plt.subplots()
             tea.plot_enrichment_results(df, title='../output/lupus_tea',
-                                        save=True)
-            tea.plot_enrichment_results(df, title='../output/lupus_tea',
-                                        save=True)
+                                        save=True, analysis='tissue')
 
         df = tea.enrichment_analysis(worm_genes, go_df, show=False)
-        df = df[df.Observed > n_min_obs]
+        df = df[df.Observed > n_min_obs].copy()
         df.to_csv('../output/disease_go_2/gea_' + trait + '.csv',
                   index=False)
         if 'lupus' in trait:
             print('Graphing GEA results for ', trait)
-            fig, ax = plt.subplot()
+            fig, ax = plt.subplots()
             tea.plot_enrichment_results(df, title='../output/lupus_gea',
[…]

### 2017-01-17 pea_paper_docs/tex/pea.tex
added most citations to pea.

@@
 The last decade has seen an explosion of techniques capable of genome-wide
-measurements. Some examples of genome-wide tools include RNA-seq~\cite{} to
-measure gene expression, CHIP-seq~\cite{} to measure binding, or genome-wide
-methylation profiling to understand epigenetic changes~\cite{}. These tools are capable
+measurements. Some examples of genome-wide tools include
+RNA-seq~\cite{Mortazavi2008} to measure gene expression or
+CHIP-seq~\cite{Johnson2007} to measure protein binding. These tools are capable
 of generating large quantities of data. Understanding these data, and generating
@@
 these datasets is to reduce the dimensionality of the data via enrichment
-analyses of ontologies~\cite{}, which helps researchers understand what terms are
+analyses of ontologies~\cite{TheGeneOntologyConsortium2000a}, which helps
+researchers understand what terms are
 overrepresented beyond random levels. By analyzing overrepresented terms in
@@
 affected in a given experiment, and form hypotheses about what is
-happening~\cite{}.
+happening~\cite{Rhee2008}.
 This approach is limited by what ontologies can be tested for enrichment.
@@
 provides a controlled language to describe molecular and cellular functions of
-genes~\cite{}. In \cel{}, curated tissue and phenotype ontologies exist which provide
-controlled languages with which to describe \cel{} anatomy and phenotypes
-respectively~\cite{}. However, enrichment tools only exist for gene and tissue ontologies
-in the community today (see for example~\cite{}). Another limitation is that
+genes~\cite{TheGeneOntologyConsortium2000a}.
+In \cel{}, gene, tissue and phenotype ontologies exist
+% which provide
+% controlled languages
+with which to describe \cel{} anatomy and phenotypes
+respectively~\cite{Schindelman2011,Lee2003}. These ontologies are curated by
+professional curators at WormBase, which is a repository of all \cel{}
+data~\cite{Howe2016}.
+However, enrichment tools
+only exist for gene and tissue ontologies
+in the community today (see for example~\cite{Chikina2009,Mi2013,Angeles-Albores2016}).
+Another limitation is that
 tissue enrichment testing is not offered on the same websites as GO enrichment
@@
 between two species---that is to say, common by descent---because knowledge of
-homology often brings with it knowledge of function~\cite{}. Indeed, many
+homology often brings with it knowledge of function. Indeed, many
 important gene regulatory networks (GRNs) are conserved between organisms as
-highly diverged as nematodes and humans (for example, see~\cite{}). While genes
+highly diverged as nematodes and humans (for example, see~\cite{Sternberg1998}).
+While genes
 and GRNs may be conserved between species, their outputs often differ however.
 For example, although homeobox genes are known to be important for limb
-formation in many animals~\cite{}, these genes do not form limbs in nematodes.
+formation in many animals~\cite{Zakany2007}, these genes do not form limbs in nematodes.
 The concept of a phenolog has been put forward to explain relationships between
-phenotypes that have the same underlying genetic regulatory network~\cite{}.
+phenotypes that have the same underlying genetic
+regulatory network~\cite{McGary2010,Lehner2013}.
 Formally, two phenotypes are phenologs of each other if the homologs of the
@@
 been successfully used in the past to make non-obvious links between phenotypes
-in different species~\cite{}.
+in different species~\cite{McGary2010}.
 
@@
 An additional problem with genome-wide
-queries of \cel{} states (be they developmental, neuronal, or other) is that
+queries of \cel{} states (be they developmental, such as
+L1, L2, dauer; behavioral states such as awake versus asleep; or other) is that
 they do not
@@
 by using genes associated with these
-diseases via genome-wide association studies. We also illustrate the utility of
+diseases via genome-wide association studies. We illustrate the utility of
 the complete enrichment suite for finding new relationships in complex data by
-analyzing a ciliary neuron transcriptome~\cite{}.
+analyzing a ciliary neuron transcriptome~\cite{Wang2015}. Finally, we show that
+the dictionaries generated for these enrichment analyses can help elucidate the
+contributions of specific tissues to specific phenotypes.
 
 \section*{Methods}
+\subsection*{Implementing the enrichment analyses}
+All scripts were implemented in Python 3.5~\cite{Rossum2011}. We used
+pandas~\cite{McKinney2011} and scipy~\cite{Oliphant2007} to write the
+statistical testing framework. Matplotlib~\cite{Hunter2007} and Seaborn~\cite{Waskom}
+libraries are used to generate all plots. Testing was performed using the
+WormBase version XXXXX. The WormBase Enrichment suite can be installed using
+pip via the command:
+
+\texttt{pip install tissue\_enrichment\_analysis}
+
+A brief tutorial on how to use the command line tool can be found in the SI.
+A brief tutorial on how to interact with the suite using python is available
+as a Jupyter notebook~\cite{Perez2007} in the SI.
+
 \subsection*{Human disease phenolog identification}
-We used the GWAS EBI-NHGRI catalog~\cite{} to extract information on all genome-wide
+We used the GWAS EBI-NHGRI catalog~\cite{MacArthur2016} to extract information
+on all genome-wide
 association studies deposited there. We only selected traits that had $>300$
 associated genes. We identified 24 traits that met our criteria. Next, we used
-DIOPT~\cite{} to identify candidate homologs for the genes associated with these
+DIOPT~\cite{Hu2011} to identify candidate homologs for the genes associated with these
 traits. Briefly, DIOPT combines a large number of methods for identifying homologs
@@
 phenotype enrichment (see~Fig.\ref{fig:gwas}). Tissue enrichment was
-performed using the WormBase Tissue Enrichment Analysis (TEA) tool~\cite{}.
+performed using the WormBase Tissue Enrichment Analysis (TEA)
+tool~\cite{Angeles-Albores2016}.
 
@@
 We developed the dictionaries for PEA and GEA using the same procedure as was
-used for TEA~\cite{}. We generated a dictionary that included terms with at
+used for TEA~\cite{Angeles-Albores2016}. We generated a dictionary that
+included terms with at
 least 50 annotating genes or more and had a similarity threshold of 0.95 for PEA
@@
 \@ Next, we benchmarked the dictionaries on the same gene
-sets as TEA and obtained enrichment of all the expected categories. For example,
-on a gene set enriched for embryonic muscle genes~\cite{}, the top two enriched
+sets as TEA and obtained enrichment of all the expected
+categories~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a,
+Pauli2006, Portman2004, Fox2007, Smith2010}.
+For example,
+on a gene set enriched for embryonic muscle genes~\cite{Watson2008a},
+the top two enriched
 phenotype terms by q-value were `muscle system morphology variant' and `body
@@
 Therefore, we used the GWAS NHGRI-EBI
-Catalog~\cite{} to identify genes associated with human diseases. We found the
-best nematode candidate homologs for these genes using DIOPT~\cite{} and applied
+Catalog~\cite{MacArthur2016} to identify genes associated with human diseases. We found the
+best nematode candidate homologs for these genes using DIOPT~\cite{Hu2011} and applied
 our enrichment suite to each of these gene regulatory networks.
@@
 Systemic lupus is an autoimmune disease that is believed to be polygenic in
-nature~\cite{}. It mainly affects women and is characterized by painful
-and swollen joints, hair loss, and fatigue~\cite{}. Since worms do not have a
+nature~\cite{Mohan2015}. It mainly affects women and is characterized by painful
+and swollen joints, hair loss, and fatigue~\cite{Lisnevskaia2014}. Since worms do not have a
 cellular immune system, we were interested in what phenologs corresponded to
@@
 Rheumatoid arthritis is an auto-immune disease that is characterized by
-swollen and painful joints that progressively deteriorate~\cite{}. Unlike lupus,
-rheumatoid arthritis is not life-threatening~\cite{}, and comorbidity between rheumatoid
-arthritis and lupus is low~\cite{}, suggesting that they may have at least
+swollen and painful joints that progressively deteriorate~\cite{Smolen2016}. Unlike lupus,
+rheumatoid arthritis is not life-threatening, and comorbidity between rheumatoid
+arthritis and lupus has not been described in past comorbidity studies~\cite{Dougados2013},
+suggesting that they may have at least
 partially distinct genetic causes. We found \harthritis{} genes associated with
@@
  `glycosylation' (10 genes, \qval{4}) and `Golgi apparatus' (11
-genes, \qval{3}), but this enrichment was also the result of degenerate homolog
+genes, \qval{3}), but these terms were enriched as the result of degenerate homolog
 candidates for the human gene B3GNT7 which encodes a
@@
 like \emph{sem-4} and \emph{rnt-1} are likely transcription factors with roles
-in development (including hypodermal development).
+in development (including hypodermal development)~\cite{Desai1988,Ji2004}.
 Others are molecular motors (\emph{unc-116}) that are broadly expressed throughout
@@
 transcriptomes,
-we selected a ciliary neuron dataset~\cite{} and ran the complete WormBase
+we selected a ciliary neuron dataset~\cite{Wang2015} and ran the complete WormBase
 Enrichment Suite on it. Ciliary neurons are present in the \cel{} male tail, but
@@
 analysis reveals that the ciliary neuron transcriptome is enriched in
-genes associated with microtubules, but the cell-cycle machinery is also
-carefully regulated.
+genes associated with microtubules, but includes genes that are thought to
+interact with DNA either via repair mechanisms or cell-cycle
+control~\cite{Hofmann2000,Lee2003a,Bailly2010}.
 
@@
 In \cel{}, egg-laying is a complex behavior that involves
-a large number of tissues~\cite{}. The somatic gonad acts as a repository for the eggs,
+a large number of tissues~\cite{Li1990}. The somatic gonad acts as a
+repository for the eggs,
 the uterine seam cells help protect the uterus, and a variety of muscles help
-contract the uterus and open the vulva to lay an egg~\cite{}. The vulva must be well-formed
+contract the uterus and open the vulva to lay an egg~\cite{Sulston1977}. The
+vulva must be well-formed
 to allow passage of an egg, and the hermaphrodite-specific neuron (HSN) is involved
-in the egg-laying control~\cite{}. The complexity of the interactions that
+in the egg-laying control~\cite{Schafer2005}. The complexity of the interactions that
 happen to allow egg-laying make understanding the Egl phenotype in terms of
@@
 \section*{Conclusions}
-
-We have highlighted three possible uses for phenotype and GO enrichment analyses.
-Taken together, the WormBase Enrichment Suite can help guide researchers as they
-search for a phenologue that is representative of a human disease that has no
-immediately obvious counterpart in the worm. Such phenologues may benefit from
-the fact that they represent an unbiased approach to disease modeling as long as
-the human genes are selected from unbiased screens. GWAS data suggests that certain
-aspects of lupus biology may be best understood from a developmental perspective
-in \cel{}. On the other hand, our results suggest that rheumatoid arthritis
-might not be effectively modeled in the worm, although it is possible that our
-inability to find phenologues more concretely is the result of the GWAS screens:
-If GWAS screens identified factors that lead to worse prognosis of rheumatoid
-arthritis, but not factors that cause onset of the disease, this could explain
-the lack of phenotype enrichment in \cel{}.
-
-% Phenologues
-
-% Screen Design
-
-% Curation and annotation biases
-
 The addition of GO and Phenotype Ontology enrichment testing to WormBase marks
 an important step towards a unified set of analyses that can help researchers
-to understand genomic datasets. By offering tissue, phenotype and gene ontology
-enrichment on a single site, researchers will benefit both in the speed at which
-the analysis is completed and from the fact that the analyses are all
-carried out with the same methodology and the same underlying algorithms.
+to understand genomic datasets. These enrichment analyses will allow the community to
+fully benefit from the data curation ongoing at WormBase. By using the same
+algorithms to generate enrichment dictionaries for testing and using the
+same model to test for term enrichment, these tools provide a coherent framework
+with which to analyze genomic data. In particular, it is our hope that
+phenotype enrichment will be of use to geneticists performing genome-wide
+analyses, because they are familiar with the ontological terms that are tested
+and with their biological meaning. Ideally, PEA could allow researchers to
+design better screens to maximize target gene identification by quantifying
+the phenotypes that are most overrepresented. An intriguing new direction of
+research would be to create a controlled language for screening methods. With
+such a language, we should be able to computationally suggest to a researcher
+what screens one may wish to perform given a dataset.
+
+\bibliography{pea-citations}
+\bibliographystyle{pnas2011}
 

### 2017-02-03 pea_paper_docs/tex/pea.tex
wrote a bit for the manuscript

@@
 
-\newcommand{\qval}[1]{
-                      \ensuremath{
-                                  q<10^{-#1}
-                                  }
-                      }
+\newcommand{\qval}[1]{\ensuremath{q<10^{-#1}}}
 
@@
 RNA-seq~\cite{Mortazavi2008} to measure gene expression or
-CHIP-seq~\cite{Johnson2007} to measure protein binding. These tools are capable
+CHIP-seq~\cite{Johnson2007} to measure protein binding to chromatin. These tools are capable
 of generating large quantities of data. Understanding these data, and generating
@@
 highly diverged as nematodes and humans (for example, see~\cite{Sternberg1998}).
-While genes
-and GRNs may be conserved between species, their outputs often differ however.
-For example, although homeobox genes are known to be important for limb
-formation in many animals~\cite{Zakany2007}, these genes do not form limbs in nematodes.
+While genes and GRNs may be conserved between species, their outputs often differ.
+For example, the gene Pax6 (Eyeless in \emph{Drosophila melanogaster})
+is involved in eye formation in humans and fruit-flies~\cite{Quiring1994}. Although
+nematodes have conserved this gene, they do not have eyes~\cite{Zhang1995,Chisholm1995}.
 The concept of a phenolog has been put forward to explain relationships between
@@
 regulatory network~\cite{McGary2010,Lehner2013}.
-Formally, two phenotypes are phenologs of each other if the homologs of the
+Formally, two phenotypes are phenologs of each other if the orthologs of the
 genes that cause a phenotype in an organism cause a second phenotype in another.
@@
 manner through genome-wide association studies (GWAS) and identified candidate
-homolog genes in \cel{}. The homologs can be used to identify \cel{} disease
+homolog genes in \cel{}. The orthologs can be used to identify \cel{} disease
 phenologs, which can in turn be used as the basis for screens to identify
@@
 select an easier-to-screen phenolog, and to test positive hits for the true
-phenotype of interest
-afterwards~\cite{}. Currently, selection of screening phenotypes is
-performed based on researcher experience.
+phenotype of interest afterwards~\cite{}. Currently, selection of screening
+phenotypes is performed based on researcher experience.
 By formalizing phenotype enrichment analysis as a tool with which to analyze
@@
 always have a straightforward interpretation in terms of phenotypes. In these
-situations, researchers must rely on intuition to select a phenotype to screen
-for. As a result, many hits may go unexplored that would prove fruitful. The
+situations, researchers must rely on intuition to select a phenotype for which to
+screen. As a result, many hits may go unexplored that would prove fruitful. The
 question of how to design a screen that is maximally informative is an important
@@
 associated genes. We identified 24 traits that met our criteria. Next, we used
-DIOPT~\cite{Hu2011} to identify candidate homologs for the genes associated with these
-traits. Briefly, DIOPT combines a large number of methods for identifying homologs
+DIOPT~\cite{Hu2011} to identify candidate orthologs for the genes associated with these
+traits. Briefly, DIOPT combines a large number of methods for identifying orthologs
 and returns homolog candidates associated with a compound score. Depending on the score,
-homologs can be considered `high', `moderate' or `low' rank, reflecting confidence
+orthologs can be considered `high', `moderate' or `low' rank, reflecting confidence
 in the homology. Many-to-one and one-to-many homology relationships are allowed
@@
            disease associated genes, then used DIOPT to identify candidate
-           homologs in \cel{}. Homologs were used to run phenotype, gene
+           orthologs in \cel{}. Orthologs were used to run phenotype, gene
            and tissue ontology enrichment analyses to identify disease
@@
 
-After we identified worm homologs for each trait, we reassessed how many traits
+After we identified worm orthologs for each trait, we reassessed how many traits
 still had $>100$ gene candidates, and dropped all traits that had less than this
@@
 Catalog~\cite{MacArthur2016} to identify genes associated with human diseases. We found the
-best nematode candidate homologs for these genes using DIOPT~\cite{Hu2011} and applied
+best nematode candidate orthologs for these genes using DIOPT~\cite{Hu2011} and applied
 our enrichment suite to each of these gene regulatory networks.
-
-\subsubsection*{Obesity-related traits}
-Obesity-related traits is a category within the GWAS NHGRI-EBI catalog that
-pools studies that have measured obesity and other traits associated with
-obesity, such as heart rate, physical activity, hormone levels, body composition
-and cholesterol levels. Since this category includes many parameters,
-we expected there would be many phenologs. GWAS studies have identified
-\hobesity{} genes associated with these traits. Using DIOPT, we found
-\wobesity{} homologs for these genes. In total, $341/\wobesity{}$ genes had
-at least one phenotype annotation; $548/\wobesity{}$ had at least one gene ontology term
-annotation; and $427/\wobesity{}$ had at least one tissue term annotation.
-
-Top results for obesity-related traits included `acetylcholinesterase inhibitor
-response variant' (38 genes, \qval{6}),
-`neurite morphology variant' (21 genes, \qval{2}),
-and `thin' (31 genes, \qval{2}).
-Terms involving locomotion were significantly enriched, as were terms involving
-body shape and food consumption (\qval{1}). Concomitant with these phenologs was
-a tissue enrichment in neuron-related terms. GO enrichment suggested that these
-genes are participating in `iron ion binding' (40 genes, \qval{20}) and
-`tetrapyrrole binding' (37 genes, \qval{13}).
-
-Tissue and phenotype enrichment therefore suggest that obesity-related traits
-may be studied in \cel{} through neuron physiology and function, specifically
-with respect to acetylcholinesterase inhibitors. Moreover, GO enrichment
-implicates iron and tetrapyrrole binding as metabolic components of
-the obesity-related phenologs in \cel{}.
 
@@
 
-Lupus-associated homologs were reasonably well annotated. Slightly more than half
+Lupus-associated orthologs were reasonably well annotated. Slightly more than half
 of the genes had at least one phenotype annotation ($76/\wlupus{}$) and almost
@@
 ($104/\wlupus{}$ and $115/\wlupus$ genes respectively).
-We found that Lupus-associated homologs were enriched in `aneuploidy' (7 genes,
+We found that Lupus-associated orthologs were enriched in `aneuploidy' (7 genes,
 \qval{1}) and `meiotic chromosome segregation' (8 genes, \qval{1}). `Cell fate
@@
 % all tissues randomly represented?
-The only phenotype that was enriched for these homologs was `short' (10 genes,
-\qval{4}), even though 64 homologs were associated with at least
+The only phenotype that was enriched for these orthologs was `short' (10 genes,
+\qval{4}), even though 64 orthologs were associated with at least
 one phenotype term. No tissue was enriched in this dataset. Because
@@
 the `body length variant' phenotype. Body length in \cel{} can be controlled via
-cell size, shape and number\cite{}; alternatively, cuticle development can alter body
-shape~\cite{}; finally, muscles can alter the effective body length artificially~\cite{}.
+cell size or shape\cite{}; alternatively, cuticle development can alter body
+shape~\cite{}; finally, muscles can alter the effective body length
+due to their contraction state~\cite{}.
 
@@
 
-\subsection*{Ontology Enrichment as an aid for screen design}
+\subsubsection*{Obesity-related traits}
+Obesity-related traits is a category within the GWAS NHGRI-EBI catalog that
+pools studies that have measured obesity and other traits associated with
+obesity, such as heart rate, physical activity, hormone levels, body composition
+and cholesterol levels. Since this category includes many parameters,
+we expected there would be many phenologs. GWAS studies have identified
+\hobesity{} genes associated with these traits. Using DIOPT, we found
+\wobesity{} orthologs for these genes. In total, $341/\wobesity{}$ genes had
+at least one phenotype annotation; $548/\wobesity{}$ had at least one gene ontology term
+annotation; and $427/\wobesity{}$ had at least one tissue term annotation.
+
+Top results for obesity-related traits included `acetylcholinesterase inhibitor
+response variant' (38 genes, \qval{6}),
+`neurite morphology variant' (21 genes, \qval{2}),
+and `thin' (31 genes, \qval{2}).
+Terms involving locomotion were significantly enriched, as were terms involving
+body shape and food consumption (\qval{1}). Concomitant with these phenologs was
+a tissue enrichment in neuron-related terms. GO enrichment suggested that these
+genes are participating in `iron ion binding' (40 genes, \qval{20}) and
+`tetrapyrrole binding' (37 genes, \qval{13}).
+
+Tissue and phenotype enrichment therefore suggest that obesity-related traits
+may be studied in \cel{} through neuron physiology and function, specifically
+with respect to acetylcholinesterase inhibitors. Moreover, GO enrichment
+implicates iron and tetrapyrrole binding as metabolic components of
+the obesity-related phenologs in \cel{}.
+
+\subsection*{Ontology enrichment as an aid for screen design}
 An additional use for a tool like PEA would be as a tool to help guide and
@@
 are enriched. The `male distal tip cell' is a tissue
-that is overrepresented in this dataset, but `distal tip cell' is not enriched,
-which suggests that structures that are present only in the male tissue are
-overrepresented in this dataset.
+that is overrepresented in this dataset, but `distal tip cell' is not enriched.
+In \cel{} the hermaphrodite distal tip cells (DTCs) and the male DTC are very
+similar to each other in their biological functions (both maintain a stem cell
+niche in the distal gonad). However, the male DTC is non-migratory, whereas the
+hermaphrodite DTCs are migratory. Therefore, the term `male distal tip cell'
+may reflect cellular aspects that are correlated with the non-migratory aspects
+of the DTC biology, such as maintenance of proliferation.
 
@@
   There are two seam cells on the left and right side of the vulva, but only
-  cell on the right is shown. Only cells in the adult are shown.}
+  cell on the right is shown. Only terminally differentiated cells are shown.}
 \label{fig:egl}
@@
 expression in a tissue term $X$ given that it is annotated with an Egl phenotype
-and we searched for terms that maximized this probability. The terms that maximized
-this probability were `nervous system', `pharynx' (a body part with a lot of neurons),
+and we searched for terms that maximized this probability (see
+Table~\ref{tab:cond_probs}). The terms that maximized this probability were
+`nervous system', `pharynx' (a body part with a lot of neurons),
 `sex organ' and `tail' (a body part with neurons and hypodermis). In general, the

### 2017-02-28 pea_paper_docs/tex/pea.tex

@@
 \usepackage{changepage}
-
 % adjust caption style
@@
             labelsep=period,singlelinecheck=off]{caption}
-
 % remove brackets from references
@@
 % \fancyfootoffset[L]{2.25in}
-
 % use \textcolor{color}{text} for colored text (e.g. highlight to-do areas)
 \usepackage{color}
-
 % define custom colors (this one is for figure captions)
 \definecolor{Gray}{gray}{.25}
-
 % this is required to include graphics
@@
 \graphicspath{{./figs/}}
-
 % use if you want to put caption to the side of the figure - see example in text
@@
 \newcommand{\ecol}{\emph{E.~coli}}
-
 \newcommand{\hobesity}{957}
 \newcommand{\wobesity}{614}
-
 \newcommand{\hlupus}{283}
 \newcommand{\wlupus}{135}
-
 \newcommand{\harthritis}{309}
@@
 
-\author{David Angeles-Albores\textsuperscript{1,$\dagger{}$}
+\author{David Angeles-Albores\textsuperscript{1}
 \and{}
-Paul W. Sternberg\textsuperscript{1,*}
+Raymond YN Lee\textsuperscript{1}
+\and{}
+Juancarlos Chan\textsuperscript{1}
+\and{}
+Paul W Sternberg\textsuperscript{1,*}
 }
@@
 
+\textbf{1} Department of Biology and Biological Engineering,
+and Howard Hughes Medical Institute, Caltech, Pasadena, CA, 91125, USA\\
+\textbf{*} Corresponding author. Contact:pws@caltech.edu
+
+\section*{Abstract}
+\bf{}
+Genome-wide experiments have the capacity to generate massive amounts of unbiased
+data about an organism. In order to interpret this data, dimensionality reduction
+techniques are required. One approach is to annotate genes using controlled languages
+and to test experimental datasets for term enrichment using probabilistic methods.
+Although gene, phenotype and anatomy ontologies exist for \cel{}, no unified
+software offers enrichment analyses of all the ontologies using the same methodology.
+Here, we present the WormBase Enrichment Suite, which offers users the
+ability to test all nematode ontologies simultaneously. We show that
+the WormBase Enrichment Suite provides valuable insight into different
+biological problems. Briefly, we show that phenotype enrichment analysis (PEA) can
+help researchers identify disease phenologs, phenotypes that are homologous across species,
+which can inform disease modeling in \cel{}. The WormBase Enrichment Suite
+analysis can also shed light on RNA-seq datasets by showing what
+molecular functions are enriched, which phenotypes these functions are implicated
+in and what tissues are overrepresented in the dataset. Finally, we explore the
+phenotype-anatomy relationship, showing that a small subset of highly specific
+tissues are disproportionately likely to cause an Egl phenotype, but inferring
+tissue expression from an Egl phenotype is limited to the largest tissues.
 ]
+
 % now start line numbers
@@
 select an easier-to-screen phenolog, and to test positive hits for the true
-phenotype of interest afterwards~\cite{}. Currently, selection of screening
-phenotypes is performed based on researcher experience.
-By formalizing phenotype enrichment analysis as a tool with which to analyze
-gene sets, researchers should be able to formally establish phenologs, which has
-consequences for screen design.
+phenotype of interest afterwards. For example, candidate genes to extend the \cel{}
+lifespan can be first screened for using heat shock survival
+genes involved in \cel{} aging
+sensitivity can be identified using stress assays~\cite{Kim2007a,Mehta2009}.
+Currently, selection of screening phenotypes is performed based on researcher
+experience. By formalizing phenotype enrichment analysis as a tool with which to
+analyze gene sets, researchers should be able to formally establish phenologs,
+which has consequences for screen design.
+% eating defects in an ablated strain
 
@@
 enrichment analyses (PEA, TEA and GEA respectively) on curated \cel{} ontologies
-using the same methodology for each one. They are located at
-% TODO: include links.
+using the same methodology for each one.
+% They are located at
+% % TODO: include links.
 We applied our tools towards the unbiased discovery of phenologues of
@@
 libraries are used to generate all plots. Testing was performed using the
-WormBase version XXXXX. The WormBase Enrichment suite can be installed using
+WormBase version WS256. The WormBase Enrichment suite can be installed using
 pip via the command:
@@
 
-A brief tutorial on how to use the command line tool can be found in the SI.
-A brief tutorial on how to interact with the suite using python is available
-as a Jupyter notebook~\cite{Perez2007} in the SI.
+% A brief tutorial on how to use the command line tool can be found in the SI.
+% A brief tutorial on how to interact with the suite using python is available
+% as a Jupyter notebook~\cite{Perez2007} in the SI.
 
@@
 (the total number of terms in the dictionary was 251, annotated by 9,169 genes
- for the version XXXXX);\@ and we generated a
+ for the version WS256);\@ and we generated a
 dictionary that included terms with at least 50 annotating genes or more and
 had a similarity threshold of 0.95 for GEA (the total number of terms in this
-dictionary is 271, annotated by 14,636 genes for the version XXXX).
+dictionary is 271, annotated by 14,636 genes for the version WS256).
 \@ Next, we benchmarked the dictionaries on the same gene
@@
 the `body length variant' phenotype. Body length in \cel{} can be controlled via
-cell size or shape\cite{}; alternatively, cuticle development can alter body
-shape~\cite{}; finally, muscles can alter the effective body length
-due to their contraction state~\cite{}.
-
-% TODO: SO.... ?
+cell size or shape\cite{Wang2002,Nakano2004}; alternatively, cuticle development can alter body
+shape~\cite{Cox1980,Kramer1988}; finally, muscles can alter the effective body
+length due to their contraction state~\cite{Lewis1980}.
 

### 2017-03-01 tissue_enrichment_analysis/hypergeometricTests.py
changed default behavior for tea

@@ -304,7 +304,7 @@ def enrichment_analysis(gene_list, tissue_df, alpha=0.05, aname='',
 
 def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
                             analysis='tissue', n_bars=15, dirGraphs='',
-                            save=True, **kwargs):
+                            save=False, **kwargs):
     """
     A plot function for TEA.
 
@@ -382,8 +382,6 @@ def plot_enrichment_results(df, y='Enrichment Fold Change', title='',
             plt.savefig('{0}.{1}'.format(title, ftype), dpi=1200)
 
     return ax
-#    plt.show()
-#    plt.close()
 
 # ==============================================================================
 #
@@ -503,7 +501,7 @@ if __name__ == '__main__':
     df_results = enrichment_analysis(gene_list, tissue_df, alpha=q,
                                      show=False)
 
-    dfname = title+'.csv'
+    dfname = title + '.csv'
     df_results.to_csv(dfname, index=False)
 
     if prnt:

### 2017-04-17 pea_paper_docs/tex/pea.tex
added corrections suggested by raymond

@@
 data about an organism. In order to interpret this data, dimensionality reduction
-techniques are required. One approach is to annotate genes using controlled languages
-and to test experimental datasets for term enrichment using probabilistic methods.
-Although gene, phenotype and anatomy ontologies exist for \cel{}, no unified
-software offers enrichment analyses of all the ontologies using the same methodology.
+techniques are required. One approach is to annotate genes using unified,
+unambiguous vocabularies and to test experimental datasets for term enrichment
+using probabilistic methods.
+Although gene, phenotype and anatomy ontologies exist for \cel{}, no one
+software offers enrichment analyses of all three ontologies.
 Here, we present the WormBase Enrichment Suite, which offers users the
@@
 the WormBase Enrichment Suite provides valuable insight into different
-biological problems. Briefly, we show that phenotype enrichment analysis (PEA) can
-help researchers identify disease phenologs, phenotypes that are homologous across species,
-which can inform disease modeling in \cel{}. The WormBase Enrichment Suite
+biological problems, and that phenotype enrichment analysis (PEA) can
+help researchers identify disease phenologs, phenotypes that are homologous across
+species, which can inform disease modeling in \cel{}. The WormBase Enrichment Suite
 analysis can also shed light on RNA-seq datasets by showing what
@@
 
-% A brief tutorial on how to use the command line tool can be found in the SI.
-% A brief tutorial on how to interact with the suite using python is available
-% as a Jupyter notebook~\cite{Perez2007} in the SI.
+Enrichment analyses are performed as described in~\cite{Angeles-Albores2016}.
+Briefly, a static dictionary is constructed for each ontology using the
+most recent WormBase release annotation information. This dictionary is generated
+using a trimming algorithm described in~\cite{Angeles-Albores2016} to remove
+terms that are too general and uninformative as well as terms that are very
+specific but do not have many annotations. When the user provides a set of genes
+to test against, we use a hypergeometric model to test whether a given term within
+an ontology is occurring at a greater frequency than expected by random chance.
 

### 2017-08-21 pea_paper_docs/tex/pea.tex
fixed URL in README.md

@@
-\documentclass[10pt,letterpaper,twocolumn]{article}
-% \usepackage[twocolumn]{geometry}
+\documentclass[10pt, onecolumn]{article}
+\usepackage[margin=1in]{geometry}
 \usepackage{lmodern}% http://ctan.org/pkg/lm
+\usepackage{authblk} % adds affiliations
 
-% use Unicode characters - try changing the option
-% if you run into troubles with special characters (e.g. umlauts)
 \usepackage[utf8x]{inputenc}
-% clean citations
-\usepackage{cite}
-% hyperref makes references clicky. use \url{www.example.com}
-% or \href{www.example.com}{description} to add a clicky url
 \usepackage{nameref}
-% line numbers
 \usepackage[right]{lineno}
-% improves typesetting in LaTeX
-% \usepackage{microtype}
-% degree symbol
-\usepackage{gensymb}
 \usepackage{amsmath}
 \usepackage{booktabs}
-% use adjustwidth environment to exceed text width (see examples in text)
+\usepackage[numbers,super]{natbib}
 \usepackage{changepage}
+
 % adjust caption style
@@
 
-\author{David Angeles-Albores\textsuperscript{1}
-\and{}
-Raymond YN Lee\textsuperscript{1}
-\and{}
-Juancarlos Chan\textsuperscript{1}
-\and{}
-Paul W Sternberg\textsuperscript{1,*}
-}
+\author[1,2]{David Angeles-Albores}
+\author[1]{Raymond Y. N. Lee}
+\author[1]{Juancarlos Chan}
+\author[1,2,*]{Paul W. Sterberg}
+\affil[1]{Division of Biology and Biological Engineering, Caltech,
+Pasadena, CA, 91125, USA}
+\affil[2]{Howard Hughes Medical Institute, Caltech, Pasadena, CA, 91125, USA}
+\renewcommand\Affilfont{\itshape\small{}}
 
@@
 % title goes here:
-\twocolumn[
 % title
@@
 \section*{Abstract}
-\bf{}
+\textbf{
 Genome-wide experiments have the capacity to generate massive amounts of unbiased
-data about an organism. In order to interpret this data, dimensionality reduction
+data about an organism. To interpret these data, dimensionality reduction
 techniques are required. One approach is to annotate genes using unified,
@@
 tissues are disproportionately likely to cause an Egl phenotype, but inferring
-tissue expression from an Egl phenotype is limited to the largest tissues.
-]
+tissue expression from an egg-laying (Egl) phenotype is limited to the largest tissues.
+}
+
+\vspace{10mm}
 
@@
 RNA-seq~\cite{Mortazavi2008} to measure gene expression or
-CHIP-seq~\cite{Johnson2007} to measure protein binding to chromatin. These tools are capable
-of generating large quantities of data. Understanding these data, and generating
-hypotheses from them remains challenging. A common approach used to understand
-these datasets is to reduce the dimensionality of the data via enrichment
-analyses of ontologies~\cite{TheGeneOntologyConsortium2000a}, which helps
-researchers understand what terms are
-overrepresented beyond random levels. By analyzing overrepresented terms in
-aggregate, researchers can better understand what biological processes were most
-affected in a given experiment, and form hypotheses about what is
-happening~\cite{Rhee2008}.
-This approach is limited by what ontologies can be tested for enrichment.
-The best-known ontology for biological research is the Gene Ontology (GO), which
-provides a controlled language to describe molecular and cellular functions of
-genes~\cite{TheGeneOntologyConsortium2000a}.
-In \cel{}, gene, tissue and phenotype ontologies exist
-% which provide
-% controlled languages
-with which to describe \cel{} anatomy and phenotypes
-respectively~\cite{Schindelman2011,Lee2003}. These ontologies are curated by
-professional curators at WormBase, which is a repository of all \cel{}
-data~\cite{Howe2016}.
-However, enrichment tools
-only exist for gene and tissue ontologies
-in the community today (see for example~\cite{Chikina2009,Mi2013,Angeles-Albores2016}).
-Another limitation is that
-tissue enrichment testing is not offered on the same websites as GO enrichment
-testing, which requires users to test their data on different websites that may
-or may not use different methodologies to detect enrichment.
+CHIP-seq~\cite{Johnson2007} to measure protein binding to chromatin. These tools
+are capable of generating large quantities of data. Understanding these data,
+and generating hypotheses from them remains challenging. A common approach used
+to understand these datasets is to reduce the dimensionality of the data via
+enrichment analyses of ontologies~\cite{TheGeneOntologyConsortium2000a}, which
+helps researchers understand what terms are overrepresented beyond random
+levels. By analyzing overrepresented terms in aggregate, researchers can better
+understand what biological processes were most affected in a given experiment,
+and form hypotheses about what is happening~\cite{Rhee2008}. This approach is
+limited by what ontologies can be tested for enrichment. The best-known ontologies
+for biological research are the Gene Ontologies (GO), which provide a controlled
+language to describe the biological, cellular and molecular aspects and functions of
+genes~\cite{TheGeneOntologyConsortium2000a}. In \cel{}, gene, tissue and
+phenotype ontologies exist with which to describe \cel{} anatomy and phenotypes
+respectively~\cite{Schindelman2011,Lee2003}. These latter ontologies are developed by
+professional curators at WormBase, which is a repository of \cel{}
+data~\cite{Howe2016}. However, enrichment tools currently only exist for gene and tissue
+ontologies in the community today (see for
+example~\cite{Chikina2009,Mi2013,Angeles-Albores2016}). Another limitation is
+that tissue enrichment testing is not offered on the same websites as GO
+enrichment testing, which requires users to test their data on different
+websites that may or may not use different methodologies to detect enrichment.
 
@@
 
-To study a clinically relevant disease in a non-human, an appropriate
-model has to be established. A straightforward method towards establishing a
-disease model in
-\cel{} is to link a disease to a causal gene, then to identify the homologous
+To study a clinically relevant disease in a non-human, an appropriate model has
+to be established. A straightforward method towards establishing a disease model
+in \cel{} is to link a disease to a causal gene, then to identify the homologous
 gene in \cel{} and then to study the function of the genetic homolog to
@@
 homolog genes in \cel{}. The orthologs can be used to identify \cel{} disease
-phenologs, which can in turn be used as the basis for screens to identify
-genes that are associated with that phenolog. Approaches similar to this have
-been successfully used in the past to make non-obvious links between phenotypes
-in different species~\cite{McGary2010}.
+phenologs, which can in turn be used as the basis for screens to identify genes
+that are associated with that phenolog. Approaches similar to this have been
+successfully used in the past to make non-obvious links between phenotypes in
+different species~\cite{McGary2010}.
 
-The concept of a phenolog can also be useful when applied within a species.
-In \cel{}, not all phenotypes are equally easy to study. Although genome-wide
+The concept of a phenolog can also be useful when applied within a species. In
+\cel{}, not all phenotypes are equally easy to study. Although genome-wide
 measurements can help elucidate the genetic network underlying a phenotype,
-devising screens to test which genes are functionally important can be difficult.
-A common strategy to study phenotypes that are difficult to screen is to
-select an easier-to-screen phenolog, and to test positive hits for the true
-phenotype of interest afterwards. For example, candidate genes to extend the \cel{}
-lifespan can be first screened for using heat shock survival
-genes involved in \cel{} aging
-sensitivity can be identified using stress assays~\cite{Kim2007a,Mehta2009}.
-Currently, selection of screening phenotypes is performed based on researcher
-experience. By formalizing phenotype enrichment analysis as a tool with which to
-analyze gene sets, researchers should be able to formally establish phenologs,
-which has consequences for screen design.
-% eating defects in an ablated strain
+devising screens to test which genes are functionally important can be
+difficult. A common strategy to study phenotypes that are difficult to screen is
+to select an easier-to-screen phenolog, and to test positive hits for the true
+phenotype of interest afterwards. For example, candidate genes to extend the
+\cel{} lifespan can be first screened for using heat shock survival genes
+involved in \cel{} aging sensitivity can be identified using stress
+assays~\cite{Kim2007a,Mehta2009}. Currently, selection of screening phenotypes
+is performed based on researcher experience. By formalizing phenotype enrichment
+analysis as a tool with which to analyze gene sets, researchers should be able
+to formally establish phenologs, which has consequences for screen design. %
+eating defects in an ablated strain
 
-An additional problem with genome-wide
-queries of \cel{} states (be they developmental, such as
-L1, L2, dauer; behavioral states such as awake versus asleep; or other) is that
-they do not
-always have a straightforward interpretation in terms of phenotypes. In these
-situations, researchers must rely on intuition to select a phenotype for which to
-screen. As a result, many hits may go unexplored that would prove fruitful. The
-question of how to design a screen that is maximally informative is an important
-question that has so far not been addressed within this community.
+An additional problem with genome-wide queries of \cel{} states (be they
+developmental, such as L1, L2, dauer; behavioral states such as awake versus
+asleep; or other) is that they do not always have a straightforward
+interpretation in terms of phenotypes. In these situations, researchers must
+rely on intuition to select a phenotype for which to screen. As a result, many
+hits may go unexplored that would prove fruitful. The question of how to design
+a screen that is maximally informative is an important question that has so far
+not been addressed within this community.
 
-To facilitate understanding of large datasets, and to make discovery of
-phenologs easier, we have completed an enrichment tool suite in WormBase
-that allows users to rapidly perform phenotype, tissue and gene ontology
-enrichment analyses (PEA, TEA and GEA respectively) on curated \cel{} ontologies
-using the same methodology for each one.
-% They are located at
-% % TODO: include links.
-We applied our tools towards the unbiased discovery of phenologues of
-multigenic, complex diseases including
-% Diseases to study:
-systemic lupus erythematosus, obesity and obesity-related traits,  and
-rheumatoid arthritis
-by using genes associated with these
-diseases via genome-wide association studies. We illustrate the utility of
-the complete enrichment suite for finding new relationships in complex data by
-analyzing a ciliary neuron transcriptome~\cite{Wang2015}. Finally, we show that
-the dictionaries generated for these enrichment analyses can help elucidate the
-contributions of specific tissues to specific phenotypes.
+To facilitate understanding of large datasets, we have developed the WormBase
+Enrichment Suite (URL:\
+\url{http://www.wormbase.org/tools/enrichment/tea/tea.cgi}) that allows users to
+rapidly perform phenotype, tissue and gene ontology enrichment analyses (PEA,
+TEA and GEA respectively) on curated \cel{} ontologies using the same
+methodology for each one. As examples of use cases, we applied our tools towards
+the unbiased discovery of phenologs of multigenic, complex diseases including
+systemic lupus erythematosus, obesity and obesity-related traits, and rheumatoid
+arthritis by using genes associated with these diseases via genome-wide
+association studies. We illustrate the utility of the complete enrichment suite
+for finding new relationships in complex data by analyzing a ciliary neuron
+transcriptome~\cite{Wang2015}. Finally, we show that the dictionaries generated
+for these enrichment analyses can help elucidate the contributions of specific
+tissues to specific phenotypes.
 
@@
 pandas~\cite{McKinney2011} and scipy~\cite{Oliphant2007} to write the
-statistical testing framework. Matplotlib~\cite{Hunter2007} and Seaborn~\cite{Waskom}
-libraries are used to generate all plots. Testing was performed using the
-WormBase version WS256. The WormBase Enrichment suite can be installed using
-pip via the command:
+statistical testing framework. Matplotlib~\cite{Hunter2007} and
+Seaborn~\cite{Waskom} libraries are used to generate all plots. Testing was
+performed using the WormBase version WS256. The WormBase Enrichment suite can be
+installed using pip via the command:
 
@@
 Enrichment analyses are performed as described in~\cite{Angeles-Albores2016}.
-Briefly, a static dictionary is constructed for each ontology using the
-most recent WormBase release annotation information. This dictionary is generated
+Briefly, a static dictionary is constructed for each ontology using the most
+recent WormBase release annotation information. This dictionary is generated
 using a trimming algorithm described in~\cite{Angeles-Albores2016} to remove
@@
 specific but do not have many annotations. When the user provides a set of genes
-to test against, we use a hypergeometric model to test whether a given term within
-an ontology is occurring at a greater frequency than expected by random chance.
+to test against, we use a hypergeometric model to test whether a given term
+within an ontology is occurring at a greater frequency than expected by random
+chance.
 
@@
 We used the GWAS EBI-NHGRI catalog~\cite{MacArthur2016} to extract information
-on all genome-wide
-association studies deposited there. We only selected traits that had $>300$
-associated genes. We identified 24 traits that met our criteria. Next, we used
-DIOPT~\cite{Hu2011} to identify candidate orthologs for the genes associated with these
-traits. Briefly, DIOPT combines a large number of methods for identifying orthologs
-and returns homolog candidates associated with a compound score. Depending on the score,
-orthologs can be considered `high', `moderate' or `low' rank, reflecting confidence
-in the homology. Many-to-one and one-to-many homology relationships are allowed
-in DIOPT, reflecting a mixture of uncertainty and family expansion/reduction.
-For our study, we only accepted homolog candidates with `high' or `moderate' scores
-and we did not insist on a one-to-one relationship between genes.
+on all genome-wide association studies deposited there. We only selected traits
+that had $>300$ associated genes. We identified 24 traits that met our criteria.
+Next, we used DIOPT~\cite{Hu2011} to identify candidate orthologs for the genes
+associated with these traits. Briefly, DIOPT combines a large number of methods
+for identifying orthologs and returns homolog candidates associated with a
+compound score. Depending on the score, orthologs can be considered `high',
+`moderate' or `low' rank, reflecting confidence in the homology. Many-to-one and
+one-to-many homology relationships are allowed in DIOPT, reflecting a mixture of
+uncertainty and family expansion/reduction. For our study, we only accepted
+homolog candidates with `high' or `moderate' scores and we did not insist on a
+one-to-one relationship between genes.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{gwas-design.pdf}
-  \caption{Experimental design for human-nematode phenologue identification.
+  \includegraphics[width=.5\linewidth]{gwas-design.pdf}
+  \caption{Experimental design for human-nematode phenolog identification.
            We used GWAS candidates from the EBI-NHGRI catalog to identify
@@
            and tissue ontology enrichment analyses to identify disease
-           phenologues.}
+           phenologs.}
 \label{fig:gwas}
@@
 still had $>100$ gene candidates, and dropped all traits that had less than this
-for our analysis. We identified 18 traits that met this criteria. The gene
-lists for each of these 18 traits were then analyzed for gene, tissue and
-phenotype enrichment (see~Fig.\ref{fig:gwas}). Tissue enrichment was
-performed using the WormBase Tissue Enrichment Analysis (TEA)
-tool~\cite{Angeles-Albores2016}.
+for our analysis. We identified 18 traits that met this criteria. The gene lists
+for each of these 18 traits were then analyzed for gene, tissue and phenotype
+enrichment (see~Fig.\ref{fig:gwas}). Tissue enrichment was performed using the
+WormBase Tissue Enrichment Analysis (TEA) tool~\cite{Angeles-Albores2016}.
 
@@
 We developed the dictionaries for PEA and GEA using the same procedure as was
-used for TEA~\cite{Angeles-Albores2016}. We generated a dictionary that
-included terms with at
-least 50 annotating genes or more and had a similarity threshold of 0.95 for PEA
-(the total number of terms in the dictionary was 251, annotated by 9,169 genes
- for the version WS256);\@ and we generated a
-dictionary that included terms with at least 50 annotating genes or more and
-had a similarity threshold of 0.95 for GEA (the total number of terms in this
-dictionary is 271, annotated by 14,636 genes for the version WS256).
-\@ Next, we benchmarked the dictionaries on the same gene
-sets as TEA and obtained enrichment of all the expected
-categories~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a,
-Pauli2006, Portman2004, Fox2007, Smith2010}.
-For example,
-on a gene set enriched for embryonic muscle genes~\cite{Watson2008a},
-the top two enriched
-phenotype terms by q-value were `muscle system morphology variant' and `body
-wall muscle thick filament variant'; the top two enriched GO terms were
-`myofibril' and `striated muscle dense body'. For all the benchmarking
-results, see supplementary information. Having generated and validated our
-dictionaries, we proceeded to identify phenologs for several common human
-diseases.
+used for TEA~\cite{Angeles-Albores2016}. We generated a dictionary that included
+terms with at least 50 annotating genes or more and had a similarity threshold
+of 0.95 for PEA (the total number of terms in the dictionary was 251, annotated
+by 9,169 genes for the version WS256);\@ and we generated a dictionary that
+included terms with at least 50 annotating genes or more and had a similarity
+threshold of 0.95 for GEA (the total number of terms in this dictionary is 271,
+annotated by 14,636 genes for the version WS256). \@ Next, we benchmarked the
+dictionaries on the same gene sets as TEA and obtained enrichment of all the
+expected categories~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a,
+Pauli2006, Portman2004, Fox2007, Smith2010}. For example, on a gene set enriched
+for embryonic muscle genes~\cite{Watson2008a}, the top two enriched phenotype
+terms by q-value were `muscle system morphology variant' and `body wall muscle
+thick filament variant'; the top two enriched GO terms were `myofibril' and
+`striated muscle dense body'. For all the benchmarking results, see
+supplementary information. Having generated and validated our dictionaries, we
+proceeded to identify phenologs for several common human diseases.
 
 \subsection*{Applying the WormBase enrichment suite}
-To discover phenologs, we first needed to identify genes that
-contribute to a disease in an unbiased manner. One way to discover gene
-associations in an unbiased manner is to perform GWSA in human populations.
-Therefore, we used the GWAS NHGRI-EBI
-Catalog~\cite{MacArthur2016} to identify genes associated with human diseases. We found the
-best nematode candidate orthologs for these genes using DIOPT~\cite{Hu2011} and applied
-our enrichment suite to each of these gene regulatory networks.
+To discover phenologs, we first needed to identify genes that contribute to a
+disease in an unbiased manner. One way to discover gene associations in an
+unbiased manner is to perform GWSA in human populations. Therefore, we used the
+GWAS NHGRI-EBI Catalog~\cite{MacArthur2016} to identify genes associated with
+human diseases. We found the best nematode candidate orthologs for these genes
+using DIOPT~\cite{Hu2011} and applied our enrichment suite to each of these gene
+regulatory networks.
 
@@
 nature~\cite{Mohan2015}. It mainly affects women and is characterized by painful
-and swollen joints, hair loss, and fatigue~\cite{Lisnevskaia2014}. Since worms do not have a
-cellular immune system, we were interested in what phenologs corresponded to
-this disorder in \cel{}. To establish phenolog candidates, we obtained
-\hlupus{} genes associated with the disease via GWAS studies, and found
+and swollen joints, hair loss, and fatigue~\cite{Lisnevskaia2014}. Since worms
+do not have a cellular immune system, we were interested in what phenologs
+corresponded to this disorder in \cel{}. To establish phenolog candidates, we
+obtained \hlupus{} genes associated with the disease via GWAS studies, and found
 \wlupus{} homolog candidates in \cel{}.
 
-Lupus-associated orthologs were reasonably well annotated. Slightly more than half
-of the genes had at least one phenotype annotation ($76/\wlupus{}$) and almost
-all genes were annotated to at least one tissue or gene ontology term
-($104/\wlupus{}$ and $115/\wlupus$ genes respectively).
-We found that Lupus-associated orthologs were enriched in `aneuploidy' (7 genes,
-\qval{1}) and `meiotic chromosome segregation' (8 genes, \qval{1}). `Cell fate
-transformation' (6 genes, \qval{1}), and `excess intestinal cells'
-(5 genes, \qval{1}) were also overrepresented, as was `male tail morphology'
-(6 genes, \qval{1}). Finally, the phenotype `nonsense mRNA accumulation' was also
-enriched (5 genes, \qval{1}) (see Fig.~\ref{fig:lupus}).
+Lupus-associated orthologs were reasonably well annotated. Slightly more than
+half of the genes had at least one phenotype annotation ($76/\wlupus{}$) and
+almost all genes were annotated to at least one tissue or gene ontology term
+($104/\wlupus{}$ and $115/\wlupus$ genes respectively). We found that
+Lupus-associated orthologs were enriched in `aneuploidy' (7 genes, \qval{1}) and
+`meiotic chromosome segregation' (8 genes, \qval{1}). `Cell fate transformation'
+(6 genes, \qval{1}), and `excess intestinal cells' (5 genes, \qval{1}) were also
+overrepresented, as was `male tail morphology' (6 genes, \qval{1}). Finally, the
+phenotype `nonsense mRNA accumulation' was also enriched (5 genes, \qval{1})
+(see Fig.~\ref{fig:lupus}).
 
@@
   \centering
-  \includegraphics[width=\linewidth]{systemic-lupus.pdf}
-  \caption{Phenologue identification for systemic lupus erythematosus.
+  \includegraphics[width=.5\linewidth]{systemic-lupus.pdf}
+  \caption{phenolog identification for systemic lupus erythematosus.
            \textbf{A} Phenotype Enrichment Analysis. \textbf{B} GO Enrichment
@@
 
-TEA suggested that the
-`excretory duct cell' (5 genes, \qval{2}) and the `posterior gonad arm' are
-overrepresented in this dataset. We also found that the Pn.p cells P3.p through
-P8.p were enriched in this dataset (5 genes, \qval{1}). GO enrichment pointed
-at `modification-dependent macromolecule catabolic process' (23 genes,
-\qval{15}) as a molecular function that characterizes this dataset. However,
-this GO term was enriched only due to a single gene family, the \emph{skr} gene
-family. Almost the entire \emph{skr} family was considered a candidate homolog
-to the SKP1 human gene, making the GO enrichment suspect.
+TEA suggested that the `excretory duct cell' (5 genes, \qval{2}) and the
+`posterior gonad arm' are overrepresented in this dataset. We also found that
+the Pn.p cells P3.p through P8.p were enriched in this dataset (5 genes,
+\qval{1}). GO enrichment pointed at `modification-dependent macromolecule
+catabolic process' (23 genes, \qval{15}) as a molecular function that
+characterizes this dataset. However, this GO term was enriched only due to a
+single gene family, the \emph{skr} gene family. Almost the entire \emph{skr}
+family was considered a candidate homolog to the SKP1 human gene, making the GO
+enrichment suspect.
 
-Enrichment of the terms for `aneuploidy', `meiotic chromosome segregation',
-and `excess intestinal cells' were largely driven by the same gene group, which
+Enrichment of the terms for `aneuploidy', `meiotic chromosome segregation', and
+`excess intestinal cells' were largely driven by the same gene group, which
 includes \emph{cki-1}, and several \emph{skr} genes. On the other hand, `cell
@@
 \emph{tsr-1}, \emph{dhcr-7} and \emph{F08B4.7}. Therefore, we conclude that
-systemic lupus erythematosus is potentially represented by a combination of three
-phenotypes in \cel{}: A cell proliferation phenotype (either increased or
-decreased), probably marked by increased
-aneuploidy; a developmental phenotype involving cell fate transformation and
-leading to dysmorphias; and a molecular phenotype involving impairment of the
-nonsense-mediated decay pathway.
-% TODO: Is the text below true???
-The results from the tissue enrichment analysis
-highlighted three tissues that are particularly sensitive to \emph{lin} mutations
-(the gonad, the excretory duct cell and the vulval precursor cells),
+systemic lupus erythematosus is potentially represented by a combination of
+three phenotypes in \cel{}: A cell proliferation phenotype (either increased or
+decreased), probably marked by increased aneuploidy; a developmental phenotype
+involving cell fate transformation and leading to dysmorphias; and a molecular
+phenotype involving impairment of the nonsense-mediated decay pathway. % TODO:
+Is the text below true??? The results from the tissue enrichment analysis
+highlighted three tissues that are particularly sensitive to \emph{lin}
+mutations (the gonad, the excretory duct cell and the vulval precursor cells),
 and the gonad arms undergo large quantities of nuclear proliferation.
@@
 \subsection*{Rheumatoid arthritis}
-Rheumatoid arthritis is an auto-immune disease that is characterized by
-swollen and painful joints that progressively deteriorate~\cite{Smolen2016}. Unlike lupus,
-rheumatoid arthritis is not life-threatening, and comorbidity between rheumatoid
-arthritis and lupus has not been described in past comorbidity studies~\cite{Dougados2013},
-suggesting that they may have at least
-partially distinct genetic causes. We found \harthritis{} genes associated with
-rheumatoid arthritis, for which we found \warthritis{} worm homolog candidates.
+Rheumatoid arthritis is an auto-immune disease that is characterized by swollen
+and painful joints that progressively deteriorate~\cite{Smolen2016}. Unlike
+lupus, rheumatoid arthritis is not life-threatening, and comorbidity between
+rheumatoid arthritis and lupus has not been described in past comorbidity
+studies~\cite{Dougados2013}, suggesting that they may have at least partially
+distinct genetic causes. We found \harthritis{} genes associated with rheumatoid
+arthritis, for which we found \warthritis{} worm homolog candidates.
 
@@
 The only phenotype that was enriched for these orthologs was `short' (10 genes,
-\qval{4}), even though 64 orthologs were associated with at least
-one phenotype term. No tissue was enriched in this dataset. Because
-82 genes are annotated to have expression in at least one tissue, the lack of
-enrichment does not reflect ignorance about the sites of expression of these
-genes. GEA showed that enriched
+\qval{4}), even though 64 orthologs were associated with at least one phenotype
+term. No tissue was enriched in this dataset. Because 82 genes are annotated to
+have expression in at least one tissue, the lack of enrichment does not reflect
+ignorance about the sites of expression of these genes. GEA showed that enriched
 molecular functions for these genes include `collagen trimer' (22 genes,
 \qval{15}). However, this term was enriched as the result of degeneracy in the
-homolog candidates for the SFTPD gene. Other terms included
- `glycosylation' (10 genes, \qval{4}) and `Golgi apparatus' (11
-genes, \qval{3}), but these terms were enriched as the result of degenerate homolog
-candidates for the human gene B3GNT7 which encodes a
-beta-1,3-N-acetylgalactosaminyltransferase.
+homolog candidates for the SFTPD gene. Other terms included `glycosylation' (10
+genes, \qval{4}) and `Golgi apparatus' (11 genes, \qval{3}), but these terms
+were enriched as the result of degenerate homolog candidates for the human gene
+B3GNT7 which encodes a beta-1,3-N-acetylgalactosaminyltransferase.
 
@@
 in development (including hypodermal development)~\cite{Desai1988,Ji2004}.
-Others are molecular motors (\emph{unc-116}) that are broadly expressed throughout
-the body of \cel{}. Yet others have known roles in neuron and muscle function,
-such as \emph{ocrl-1} and \emph{cat-4}. The `short' phenotype is a subset of
-the `body length variant' phenotype. Body length in \cel{} can be controlled via
-cell size or shape\cite{Wang2002,Nakano2004}; alternatively, cuticle development can alter body
-shape~\cite{Cox1980,Kramer1988}; finally, muscles can alter the effective body
-length due to their contraction state~\cite{Lewis1980}.
+Others are molecular motors (\emph{unc-116}) that are broadly expressed
+throughout the body of \cel{}. Yet others have known roles in neuron and muscle
+function, such as \emph{ocrl-1} and \emph{cat-4}. The `short' phenotype is a
+subset of the `body length variant' phenotype. Body length in \cel{} can be
+controlled via cell size or shape\cite{Wang2002,Nakano2004}; alternatively,
+cuticle development can alter body shape~\cite{Cox1980,Kramer1988}; finally,
+muscles can alter the effective body length due to their contraction
+state~\cite{Lewis1980}.
 
@@
 obesity, such as heart rate, physical activity, hormone levels, body composition
-and cholesterol levels. Since this category includes many parameters,
-we expected there would be many phenologs. GWAS studies have identified
-\hobesity{} genes associated with these traits. Using DIOPT, we found
-\wobesity{} orthologs for these genes. In total, $341/\wobesity{}$ genes had
-at least one phenotype annotation; $548/\wobesity{}$ had at least one gene ontology term
-annotation; and $427/\wobesity{}$ had at least one tissue term annotation.
+and cholesterol levels. Since this category includes many parameters, we
+expected there would be many phenologs. GWAS studies have identified \hobesity{}
+genes associated with these traits. Using DIOPT, we found \wobesity{} orthologs
+for these genes. In total, $341/\wobesity{}$ genes had at least one phenotype
+annotation; $548/\wobesity{}$ had at least one gene ontology term annotation;
+and $427/\wobesity{}$ had at least one tissue term annotation.
 
 Top results for obesity-related traits included `acetylcholinesterase inhibitor
-response variant' (38 genes, \qval{6}),
-`neurite morphology variant' (21 genes, \qval{2}),
-and `thin' (31 genes, \qval{2}).
-Terms involving locomotion were significantly enriched, as were terms involving
-body shape and food consumption (\qval{1}). Concomitant with these phenologs was
-a tissue enrichment in neuron-related terms. GO enrichment suggested that these
-genes are participating in `iron ion binding' (40 genes, \qval{20}) and
-`tetrapyrrole binding' (37 genes, \qval{13}).
+response variant' (38 genes, \qval{6}), `neurite morphology variant' (21 genes,
+\qval{2}), and `thin' (31 genes, \qval{2}). Terms involving locomotion were
+significantly enriched, as were terms involving body shape and food consumption
+(\qval{1}). Concomitant with these phenologs was a tissue enrichment in
+neuron-related terms. GO enrichment suggested that these genes are participating
+in `iron ion binding' (40 genes, \qval{20}) and `tetrapyrrole binding' (37
+genes, \qval{13}).
 
@@
 with respect to acetylcholinesterase inhibitors. Moreover, GO enrichment
-implicates iron and tetrapyrrole binding as metabolic components of
-the obesity-related phenologs in \cel{}.
+implicates iron and tetrapyrrole binding as metabolic components of the
+obesity-related phenologs in \cel{}.
 
@@
 As an example of how ontology enrichment can improve our understanding of
-transcriptomes,
-we selected a ciliary neuron dataset~\cite{Wang2015} and ran the complete WormBase
-Enrichment Suite on it. Ciliary neurons are present in the \cel{} male tail, but
-they are also present in the male cephalic sensillum and hermaphrodites also have
-ciliated neurons. PEA reveals that the ciliated
-neuron transcriptome is enriched for genes that are typically associated with
-`meiotic chromosome segregation' (46 genes, \qval{5}),
-`aneuploidy' (42 genes, \qval{5}) and `spindle defective early embryos' (45
-genes, \qval{2}) (see Fig.~\ref{fig:cilia}).
+transcriptomes, we selected a ciliary neuron dataset~\cite{Wang2015} and ran the
+complete WormBase Enrichment Suite on it. Ciliary neurons are present in the
+\cel{} male tail, but they are also present in the male cephalic sensillum and
+hermaphrodites also have ciliated neurons. PEA reveals that the ciliated neuron
+transcriptome is enriched for genes that are typically associated with `meiotic
+chromosome segregation' (46 genes, \qval{5}), `aneuploidy' (42 genes, \qval{5})
+and `spindle defective early embryos' (45 genes, \qval{2}) (see
+Fig.~\ref{fig:cilia}).
 
@@
   \centering
-  \includegraphics[width=\linewidth]{ciliary-transcriptome.pdf}
+  \includegraphics[width=.5\linewidth]{ciliary-transcriptome.pdf}
   \caption{PEA shows that the ciliary transcriptome is enriched in phenotypes
@@
 
-In addition, TEA points at the \cel{} gonad primordium, the somatic gonad
-and early embryonic cells as the sites where genes associated with ciliary neurons
-are enriched. The `male distal tip cell' is a tissue
-that is overrepresented in this dataset, but `distal tip cell' is not enriched.
-In \cel{} the hermaphrodite distal tip cells (DTCs) and the male DTC are very
-similar to each other in their biological functions (both maintain a stem cell
-niche in the distal gonad). However, the male DTC is non-migratory, whereas the
-hermaphrodite DTCs are migratory. Therefore, the term `male distal tip cell'
-may reflect cellular aspects that are correlated with the non-migratory aspects
-of the DTC biology, such as maintenance of proliferation.
+In addition, TEA points at the \cel{} gonad primordium, the somatic gonad and
+early embryonic cells as the sites where genes associated with ciliary neurons
+are enriched. The `male distal tip cell' is a tissue that is overrepresented in
+this dataset, but `distal tip cell' is not enriched. In \cel{} the hermaphrodite
+distal tip cells (DTCs) and the male DTC are very similar to each other in their
+biological functions (both maintain a stem cell niche in the distal gonad).
+However, the male DTC is non-migratory, whereas the hermaphrodite DTCs are
+migratory. Therefore, the term `male distal tip cell' may reflect cellular
+aspects that are correlated with the non-migratory aspects of the DTC biology,
+such as maintenance of proliferation.
 
-Although one interpretation of the results would be that microtubule genes
-are driving the enrichment of these terms, another possibility is that
-there are cell-cycle genes that are driving the enrichment of these phenotypes
-and tissues. Indeed, GO enrichment shows terms such as `DNA replication' (29
-genes, \qval{5}), and `purine NTP-dependent helicase activity'
-(15 genes, \qval{1}). Visual inspection of list in question reveals that
-cell-cycle and DNA replication/repair genes are abundant in this transcriptome
-and include genes such as \emph{atm-1}, \emph{dna-2}, or \emph{hpr-17}. This
-analysis reveals that the ciliary neuron transcriptome is enriched in
-genes associated with microtubules, but includes genes that are thought to
-interact with DNA either via repair mechanisms or cell-cycle
+Although one interpretation of the results would be that microtubule genes are
+driving the enrichment of these terms, another possibility is that there are
+cell-cycle genes that are driving the enrichment of these phenotypes and
+tissues. Indeed, GO enrichment shows terms such as `DNA replication' (29 genes,
+\qval{5}), and `purine NTP-dependent helicase activity' (15 genes, \qval{1}).
+Visual inspection of list in question reveals that cell-cycle and DNA
+replication/repair genes are abundant in this transcriptome and include genes
+such as \emph{atm-1}, \emph{dna-2}, or \emph{hpr-17}. This analysis reveals that
+the ciliary neuron transcriptome is enriched in genes associated with
+microtubules, but includes genes that are thought to interact with DNA either
+via repair mechanisms or cell-cycle
 control~\cite{Hofmann2000,Lee2003a,Bailly2010}.
@@
 developed, it is possible to understand what tissues contribute to a phenotype
-in a probabilistic framework. In other words, we can extract all genes associated
-with a particular phenotype, then search for tissue terms that are enriched to
-understand how a phenotype arises from interactions between anatomical regions.
-As a test of this, we selected the egg-laying defective (Egl) phenotype.
-In \cel{}, egg-laying is a complex behavior that involves
-a large number of tissues~\cite{Li1990}. The somatic gonad acts as a
-repository for the eggs,
-the uterine seam cells help protect the uterus, and a variety of muscles help
-contract the uterus and open the vulva to lay an egg~\cite{Sulston1977}. The
-vulva must be well-formed
-to allow passage of an egg, and the hermaphrodite-specific neuron (HSN) is involved
-in the egg-laying control~\cite{Schafer2005}. The complexity of the interactions that
-happen to allow egg-laying make understanding the Egl phenotype in terms of
-tissues a challenging task.
+in a probabilistic framework. In other words, we can extract all genes
+associated with a particular phenotype, then search for tissue terms that are
+enriched to understand how a phenotype arises from interactions between
+anatomical regions. As a test of this, we selected the egg-laying defective
+(Egl) phenotype. In \cel{}, egg-laying is a complex behavior that involves a
+large number of tissues~\cite{Li1990}. The somatic gonad acts as a repository
+for the eggs, the uterine seam cells help protect the uterus, and a variety of
+muscles help contract the uterus and open the vulva to lay an
+egg~\cite{Sulston1977}. The vulva must be well-formed to allow passage of an
+egg, and the hermaphrodite-specific neuron (HSN) is involved in the egg-laying
+control~\cite{Schafer2005}. The complexity of the interactions that happen to
+allow egg-laying make understanding the Egl phenotype in terms of tissues a
+challenging task.
 
 We extracted all of the \cel{} genes that have been associated with an Egl
-phenotype and we used TEA to understand what tissues are enriched. The HSN
-was enriched more than five-fold above background (\qval{7}) as were vulD, vulC,
-vulE and vulF (\qval{6}). The vulA, vulB2 and vulB1 were
-enriched at slightly lower levels (\qval{5}), whereas the uterine muscles and
-uterine seam cells were enriched more than twice above background levels
-(\qval{2}) (see Fig.~\ref{fig:egl}).
-Therefore, the Egl phenotype would seem to emerge primarily from defects in the
-HSN, secondarily from defects in the vulva, and only sometimes from defects in the
-uterine seam cells or muscles. It is notable that all vulval cells were not equally
-enriched. Although all the `vul' cells are annotated to a similar degree (between
-50--70 genes for each cell type), the vulD and vulC cells had the largest
-enrichment effect size and the lowest q-values, suggesting that these cells
-are more likely to be associated with an Egl phenotype than the others. This may
-reflect the fact that vulD and vulE are the site of attachment for four vulval
-muscles, vm1. Perhaps this attachment is particularly fragile, and perturbations
-to these cells prevent adequate function of these muscles. In support of these
-observations, P7.pa had the largest fold-enrichment of any tissue. In \cel{},
-P7.pa gives rise to vulD and vulC. However, vulF is
-also attached to a set of four additional vulval muscles, vm2. Why
-is vulF less associated with an Egl phenotype?
+phenotype and we used TEA to understand what tissues are enriched. The HSN was
+enriched more than five-fold above background (\qval{7}) as were vulD, vulC,
+vulE and vulF (\qval{6}). The vulA, vulB2 and vulB1 were enriched at slightly
+lower levels (\qval{5}), whereas the uterine muscles and uterine seam cells were
+enriched more than twice above background levels (\qval{2}) (see
+Fig.~\ref{fig:egl}). Therefore, the Egl phenotype would seem to emerge primarily
+from defects in the HSN, secondarily from defects in the vulva, and only
+sometimes from defects in the uterine seam cells or muscles. It is notable that
+all vulval cells were not equally enriched. Although all the `vul' cells are
+annotated to a similar degree (between 50--70 genes for each cell type), the
+vulD and vulC cells had the largest enrichment effect size and the lowest
+q-values, suggesting that these cells are more likely to be associated with an
+Egl phenotype than the others. This may reflect the fact that vulD and vulE are
+the site of attachment for four vulval muscles, vm1. Perhaps this attachment is
+particularly fragile, and perturbations to these cells prevent adequate function
+of these muscles. In support of these observations, P7.pa had the largest
+fold-enrichment of any tissue. In \cel{}, P7.pa gives rise to vulD and vulC.
+However, vulF is also attached to a set of four additional vulval muscles, vm2.
+Why is vulF less associated with an Egl phenotype?
 
@@
   \centering
-  \includegraphics[width=\linewidth]{Egl-phenotype.pdf}
+  \includegraphics[width=.6\linewidth]{Egl-phenotype.pdf}
   \caption{The Egl phenotype is a complex phenotype that is the result of
@@
                 mapping via Bayesian probabilities}
-Another way to understand the phenotype-anatomy mapping is by considering
-how informative a given anatomy term is on a particular phenotype, or vice-versa.
-To this end, we calculated two conditional probabilities that helped us answer
-this question. The first conditional probability,
-
+Another way to understand the phenotype-anatomy mapping is by considering how
+informative a given anatomy term is on a particular phenotype, or vice-versa. To
+this end, we calculated two conditional probabilities that helped us answer this
+question. The first conditional probability,
 \begin{equation}
@@
 \end{equation}
-
 answers the question: For a gene with an expression pattern that includes the
-tissue term X (i.e., the gene is expressed at least in X), what is the probability
-that this gene has an Egl phenotype (i.e., the phenotype annotations for this
-gene include Egl)? For simplicity, we can re-write this equation more succintly
-by removing a few words. The calculation of this probability is straightforward
-and follows from the definition of conditional probability:
-
+tissue term X (i.e., the gene is expressed at least in X), what is the
+probability that this gene has an Egl phenotype (i.e., the phenotype annotations
+for this gene include Egl)? For simplicity, we can re-write this equation more
+succintly by removing a few words. The calculation of this probability is
+straightforward and follows from the definition of conditional probability:
 \begin{equation}
@@
 \end{equation}
-
 Equation~\ref{egl_x} measures how likely a gene is to be annotated with an Egl
@@
 expressed in tissue $X$. That is to say,
-
 \begin{equation}
@@
 \end{equation}
-
 Equation~\ref{x_egl} tells us how probable it is that any given gene that is
 annotated with an Egl phenotype includes $X$ as a tissue term. Taken together,
-equations~\ref{egl_x} and~\ref{x_egl} help us understand how predictive
-anatomic expression is of phenotypes, and how predictive phenotypes are of
-anatomic expression.
+equations~\ref{egl_x} and~\ref{x_egl} help us understand how predictive anatomic
+expression is of phenotypes, and how predictive phenotypes are of anatomic
+expression.
 
@@
   \ra{1.3}
-  \caption{Conditional probabilities for various tissues. The
-  first column shows the conditional probability that a gene has an Egl
-  phenotype given that it has expression in tissue $X$ (given by the row).
-  The second column shows the conditional probability that a gene has expression
-  in the anatomy term X given that it has an Egl phenotype. The first 9 terms
-  are the terms for which $P(\text{Egl}|X)$ is maximized. The last three terms
-  are the terms which have the highest $P(X|\text{Egl})$. For clarity, the Pn.p
-  cells are not shown even though $P(\text{Egl}|\text{Pn.p})\sim 0.24$.}
-
   \begin{tabular}{@{}lcc@{}}
-  \toprule{}
-  Tissue & $P(\text{Egl}| X)$ & $P(X|\text{Egl})$\\
-  \bottomrule{}\\
-  % Data goes here
-  P7.pa & $0.30$ & $0.04$\\
+  \toprule{}Tissue & $P(\text{Egl}| X)$ & $P(X|\text{Egl})$\\
+  \bottomrule{}P7.pa & $0.30$ & $0.04$\\
   HSN & $0.27$ & $0.11$\\
@@
   \end{tabular}
+  \caption{Conditional probabilities for various tissues. The
+  first column shows the conditional probability that a gene has an Egl
+  phenotype given that it has expression in tissue $X$ (given by the row).
+  The second column shows the conditional probability that a gene has expression
+  in the anatomy term X given that it has an Egl phenotype. The first 9 terms
+  are the terms for which $P(\text{Egl}|X)$ is maximized. The last three terms
+  are the terms which have the highest $P(X|\text{Egl})$. For clarity, the Pn.p
+  cells are not shown even though $P(\text{Egl}|\text{Pn.p})\sim 0.24$.}
 \label{tab:cond_probs}
@@
 that it's expression pattern includes a tissue term $X$ and we searched for the
-tissue terms that maximized this probability. The list of terms that maximized this
-probability reflected the results from running TEA on the subset of genes that have
-an Egl phenotype. We also calculated the conditional probability that a gene has
-expression in a tissue term $X$ given that it is annotated with an Egl phenotype
-and we searched for terms that maximized this probability (see
+tissue terms that maximized this probability. The list of terms that maximized
+this probability reflected the results from running TEA on the subset of genes
+that have an Egl phenotype. We also calculated the conditional probability that
+a gene has expression in a tissue term $X$ given that it is annotated with an
+Egl phenotype and we searched for terms that maximized this probability (see
 Table~\ref{tab:cond_probs}). The terms that maximized this probability were
-`nervous system', `pharynx' (a body part with a lot of neurons),
-`sex organ' and `tail' (a body part with neurons and hypodermis). In general, the
-terms that had a high $P(\text{Egl}|X)$ did not have a high $P(X|\text{Egl})$.
-Additionally, the terms that had a high $P(X|\text{Egl})$ are broad terms that
-include a lot of cells, whereas the terms that had a high $P(\text{Egl}|X)$
-were considerably more specific.
-We conclude that the Egl phenotype arises from a small set of tissues. The Egl
-phenotype can be best predicted by genes with expression patterns that include
-at least one of a small number of cells (mainly vul cells, HSN). On the other
-hand, answering whether the expression pattern of a gene includes a particular
-anatomic region or tissue given that the gene has an Egl phenotype is hard to
-do for small tissues or single cells. However, guesses about what functional system
-or broad anatomic region an Egl gene is expressed in can be answered with
-confidence ($\sim70\%$ of the time, an Egl mutant is expressed in
-the nervous system).
+`nervous system', `pharynx' (a body part with a lot of neurons), `sex organ' and
+`tail' (a body part with neurons and hypodermis). In general, the terms that had
+a high $P(\text{Egl}|X)$ did not have a high $P(X|\text{Egl})$. Additionally,
+the terms that had a high $P(X|\text{Egl})$ are broad terms that include a lot
+of cells, whereas the terms that had a high $P(\text{Egl}|X)$ were considerably
+more specific. We conclude that the Egl phenotype arises from a small set of
+tissues. The Egl phenotype can be best predicted by genes with expression
+patterns that include at least one of a small number of cells (mainly vul cells,
+HSN). On the other hand, answering whether the expression pattern of a gene
+includes a particular anatomic region or tissue given that the gene has an Egl
+phenotype is hard to do for small tissues or single cells. However, guesses
+about what functional system or broad anatomic region an Egl gene is expressed
+in can be answered with confidence ($\sim70\%$ of the time, an Egl-causing gene
+is expressed in the nervous system).
 
@@
 The addition of GO and Phenotype Ontology enrichment testing to WormBase marks
-an important step towards a unified set of analyses that can help researchers
-to understand genomic datasets. These enrichment analyses will allow the community to
-fully benefit from the data curation ongoing at WormBase. By using the same
-algorithms to generate enrichment dictionaries for testing and using the
-same model to test for term enrichment, these tools provide a coherent framework
-with which to analyze genomic data. In particular, it is our hope that
-phenotype enrichment will be of use to geneticists performing genome-wide
-analyses, because they are familiar with the ontological terms that are tested
-and with their biological meaning. Ideally, PEA could allow researchers to
-design better screens to maximize target gene identification by quantifying
-the phenotypes that are most overrepresented. An intriguing new direction of
-research would be to create a controlled language for screening methods. With
-such a language, we should be able to computationally suggest to a researcher
-what screens one may wish to perform given a dataset.
+an important step towards a unified set of analyses that can help researchers to
+understand genomic datasets. These enrichment analyses will allow the community
+to fully benefit from the data curation ongoing at WormBase. By using the same
+algorithms to generate enrichment dictionaries for testing and using the same
+model to test for term enrichment, these tools provide a coherent framework with
+which to analyze genomic data. In particular, it is our hope that phenotype
+enrichment will be of use to geneticists performing genome-wide analyses,
+because they are familiar with the ontological terms that are tested and with
+their biological meaning. Ideally, PEA could allow researchers to design better
+screens to maximize target gene identification by quantifying the phenotypes
+that are most overrepresented. An intriguing new direction of research would be
+to create a controlled language for screening methods. With such a language, we
+should be able to computationally suggest to a researcher what screens one may
+wish to perform given a dataset.
 

### 2018-02-06 pea_paper_docs/src/pea_benchmarks.py
important upgrades
barplot now plots logQ, enrichment analysis calculations should be
slightly faster now, reduced code verbosity and re-wrote PEA as a upub.

@@ -37,7 +37,8 @@ wbids = pd.read_csv('../input/wormbase_ids.txt', sep='\t', header=None,
                     names=['wbid', 'human_read', 'target_id'])
 
 # Ciliary Neuron Transcriptome (from Juan Wang and Maureen Barr)
-ciliated_df = pd.read_excel('~/Downloads/mmc3.xlsx')
+ciliated_df = pd.read_excel('../input/ciliary_neuron_transcriptome_DEseq2.xlsx'
+                            )
 
 # the minimum number of observations to keep for the GWAS enrich. analysis
 # this ensures that only high-scoring results are kept.

### 2018-02-06 tissue_enrichment_analysis/hypergeometricTests.py

@@ -12,6 +12,7 @@ An experimental list of gene names
 # -*- coding: utf-8 -*-
 import pandas as pd
 from scipy import stats
+import numpy as np
 import os
 import sys
 from urllib.request import urlopen
@@ -21,35 +22,10 @@ import contextlib
 def pass_list(user_provided, tissue_dictionary):
     """
     A function to check which genes in provided list are in the dictionary.
-
-    Takes two vectors, user_provided and tissue_dictionary in that order
-    tissue_dictionary is a pandas dataframe that you should know about
-    user_provided is a list of gene names
     """
-    # changed 4 Jan 2017
-    # length = tissue_dictionary.shape[0]  # how many genes in the dict
-    #
-    # # make an empty dataframe
-    # present = pd.DataFrame(index=range(length), columns=['wbid', 'provided'])
-    #
-    # # fill wbid column with all the gene names
-    # present.wbid = tissue_dictionary.wbid
-    #
-    # # go through and pass attendance -- 1 if present, 0 otherwise
-    # for item in user_provided:
-    #     present.provided[present.wbid == item] = 1
-    #
-    # # remove NA's and make them 0
-    # present.provided = present.provided.fillna(0)
     ind = tissue_dictionary.wbid.isin(user_provided)
     present = tissue_dictionary[ind].wbid
-
-    # return df
     return present
-
-# ==============================================================================
-# hgf is short for hypergeometric function
-# ==============================================================================
 
 
 def hgf(gene_list, tissue_df):
@@ -67,24 +43,19 @@ def hgf(gene_list, tissue_df):
     tissue_dictionary should be a pandas df
     """
     # figure out what genes are in the user provided list
-    # present = pass_list(gene_list, tissue_df)
-    # unused = present[present.provided == 0].wbid # 4 Jan 2017
-    # slice out only the genes that were present from the user-provided list
-    # wanted = present.wbid[present.provided == 1]
-    wanted = pass_list(gene_list, tissue_df)
+    # wanted = pass_list(gene_list, tissue_df)
 
     # re-index the dictionary s.t. the wbid is the index
     tissue_df = tissue_df.set_index('wbid')
 
     # number of balls per tissue in the dictionary
-    sums_of_tissues = tissue_df.sum()  # this object can be identified by
-    # the column names of tissues and excludes gene IDs
+    sums_of_tissues = tissue_df.sum()
 
-    # total size of the urn
-    total_balls = tissue_df.sum().sum()  # total labels in the dictionary
[…]

### 2018-02-06 pea_paper_docs/tex/pea.tex

@@
 \usepackage{changepage}
+\usepackage{listings}
+\usepackage{color}
 
@@
 
+\definecolor{codegreen}{rgb}{0,0.6,0}
+\definecolor{codegray}{rgb}{0.5,0.5,0.5}
+\definecolor{codepurple}{rgb}{0.58,0,0.82}
+\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
+
+\lstdefinestyle{mystyle}{
+    backgroundcolor=\color{backcolour},
+    commentstyle=\color{codegreen},
+    keywordstyle=\color{magenta},
+    numberstyle=\tiny\color{codegray},
+    stringstyle=\color{codepurple},
+    basicstyle=\footnotesize,
+    breakatwhitespace=false,
+    breaklines=true,
+    captionpos=b,
+    keepspaces=true,
+    numbers=left,
+    numbersep=5pt,
+    showspaces=false,
+    showstringspaces=false,
+    showtabs=false,
+    tabsize=2
+}
+
+\lstset{style=mystyle}
+
 % new commands
@@
 \author[1]{Juancarlos Chan}
-\author[1,2,*]{Paul W. Sterberg}
+\author[1,2,*]{Paul W. Sternberg}
 \affil[1]{Division of Biology and Biological Engineering, Caltech,
@@
 \affil[2]{Howard Hughes Medical Institute, Caltech, Pasadena, CA, 91125, USA}
+\affil[*]{Corresponding author. Contact: pws@caltech.edu}
 \renewcommand\Affilfont{\itshape\small{}}
@@
 
-\textbf{1} Department of Biology and Biological Engineering,
-and Howard Hughes Medical Institute, Caltech, Pasadena, CA, 91125, USA\\
-\textbf{*} Corresponding author. Contact:pws@caltech.edu
+% \textbf{1} Department of Biology and Biological Engineering, Pasadena, CA,
+% 91125, USA\\
+% \textbf{2} Howard Hughes Medical Institute, Caltech, Pasadena, CA, 91125, USA
+% \textbf{*} Corresponding author. Contact:pws@caltech.edu
 
-\section*{Abstract}
-\textbf{
-Genome-wide experiments have the capacity to generate massive amounts of unbiased
+\section*{ }
+\textbf{Genome-wide experiments have the capacity to generate massive amounts of unbiased
 data about an organism. To interpret these data, dimensionality reduction
@@
 Here, we present the WormBase Enrichment Suite, which offers users the
-ability to test all nematode ontologies simultaneously. We show that
+ability to test all nematode ontologies in parallel. We show that
 the WormBase Enrichment Suite provides valuable insight into different
 biological problems, and that phenotype enrichment analysis (PEA) can
-help researchers identify disease phenologs, phenotypes that are homologous across
-species, which can inform disease modeling in \cel{}. The WormBase Enrichment Suite
-analysis can also shed light on RNA-seq datasets by showing what
-molecular functions are enriched, which phenotypes these functions are implicated
-in and what tissues are overrepresented in the dataset. Finally, we explore the
-phenotype-anatomy relationship, showing that a small subset of highly specific
-tissues are disproportionately likely to cause an Egl phenotype, but inferring
-tissue expression from an egg-laying (Egl) phenotype is limited to the largest tissues.
+help researchers identify disease phenologs, phenotypes that arise from the
+action of orthologous genes across species, which can inform disease modeling in
+\cel{}. The WormBase Enrichment Suite analysis can also shed light on
+RNA-seq datasets by showing what molecular functions are enriched, which
+phenotypes these functions are implicated in and what tissues are
+overrepresented in the dataset. Finally, we explore the phenotype-anatomy
+relationship, showing that a small subset of highly specific tissues are
+disproportionately likely to affect the Egl phenotype, but inferring gene tissue
+expression from an egg-laying (Egl) phenotype is limited to the largest tissues.
 }
@@
 language to describe the biological, cellular and molecular aspects and functions of
-genes~\cite{TheGeneOntologyConsortium2000a}. In \cel{}, gene, tissue and
+genes~\cite{TheGeneOntologyConsortium2000a}. In \cel{}, anatomy and
 phenotype ontologies exist with which to describe \cel{} anatomy and phenotypes
@@
 enrichment testing, which requires users to test their data on different
-websites that may or may not use different methodologies to detect enrichment.
+websites that use different methodologies and databases to detect enrichment.
 
-Another way to use enrichment tools is for evolutionary comparison purposes.
-In molecular biology it is often useful to know when a gene is homologous
+Another use of enrichment tools is for evolutionary comparison purposes.
+In molecular biology it is often useful to know when a gene is orthologous
 between two species---that is to say, common by descent---because knowledge of
-homology often brings with it knowledge of function. Indeed, many
+orthology often brings with it knowledge of function. Indeed, many
 important gene regulatory networks (GRNs) are conserved between organisms as
@@
 regulatory network~\cite{McGary2010,Lehner2013}.
-Formally, two phenotypes are phenologs of each other if the orthologs of the
-genes that cause a phenotype in an organism cause a second phenotype in another.
+% Formally, two phenotypes are phenologs of each other if the orthologs of the
+% genes that cause a phenotype in an organism cause a second phenotype in another.
 
 To study a clinically relevant disease in a non-human, an appropriate model has
-to be established. A straightforward method towards establishing a disease model
-in \cel{} is to link a disease to a causal gene, then to identify the homologous
-gene in \cel{} and then to study the function of the genetic homolog to
+to be established. One method to establish a disease model
+in \cel{} is to link a disease to a causal gene, then to identify the
+\cel{} ortholog, study the function of the ortholog in order to
 extrapolate back to humans. However, this method relies on the existence of
-known disease genes and requires that the homolog have a phenotype that can be
+known disease genes and requires that the ortholog have a phenotype that can be
 reliably identified and studied. A fundamentally different way to establish a
@@
 manner through genome-wide association studies (GWAS) and identified candidate
-homolog genes in \cel{}. The orthologs can be used to identify \cel{} disease
+ortholog genes in \cel{}. The orthologs can be used to identify \cel{} disease
 phenologs, which can in turn be used as the basis for screens to identify genes
@@
 difficult. A common strategy to study phenotypes that are difficult to screen is
-to select an easier-to-screen phenolog, and to test positive hits for the true
-phenotype of interest afterwards. For example, candidate genes to extend the
-\cel{} lifespan can be first screened for using heat shock survival genes
-involved in \cel{} aging sensitivity can be identified using stress
-assays~\cite{Kim2007a,Mehta2009}. Currently, selection of screening phenotypes
-is performed based on researcher experience. By formalizing phenotype enrichment
+to select an easier-to-screen phenolog, and to test positive hits for the original
+phenotype of interest afterwards. For example, candidate genes involved in \cel{}
+aging control can be more easily identified using stress
+assays~\cite{Kim2007a,Mehta2009}. Currently, which phenotypes are screened for
+depends on the researcher's prior experience. By formalizing phenotype enrichment
 analysis as a tool with which to analyze gene sets, researchers should be able
-to formally establish phenologs, which has consequences for screen design. %
-eating defects in an ablated strain
+to formally establish phenologs, which has consequences for screen design.
 
@@
 rely on intuition to select a phenotype for which to screen. As a result, many
-hits may go unexplored that would prove fruitful. The question of how to design
+loci may go unexplored that would prove relevant. The question of how to design
 a screen that is maximally informative is an important question that has so far
@@
 Enrichment Suite (URL:\
-\url{http://www.wormbase.org/tools/enrichment/tea/tea.cgi}) that allows users to
+\url{http://www.wormbase.org/tools/enrichment/tea/tea.cgi}), which allows users to
 rapidly perform phenotype, tissue and gene ontology enrichment analyses (PEA,
-TEA and GEA respectively) on curated \cel{} ontologies using the same
-methodology for each one. As examples of use cases, we applied our tools towards
+TEA and GEA respectively) using a unified methodology. As examples of use cases,
+we applied our tools towards
 the unbiased discovery of phenologs of multigenic, complex diseases including
@@
 for finding new relationships in complex data by analyzing a ciliary neuron
-transcriptome~\cite{Wang2015}. Finally, we show that the dictionaries generated
-for these enrichment analyses can help elucidate the contributions of specific
-tissues to specific phenotypes.
+transcriptome~\cite{Wang2015} as a test case. Finally, we show that the
+dictionaries generated for these enrichment analyses can help elucidate the
+contributions of specific tissues to specific phenotypes.
 
@@
 
-Enrichment analyses are performed as described in~\cite{Angeles-Albores2016}.
+Enrichment analyses are performed as described in Angeles-Albores
+\emph{et al}~\cite{Angeles-Albores2016}.
 Briefly, a static dictionary is constructed for each ontology using the most
 recent WormBase release annotation information. This dictionary is generated
-using a trimming algorithm described in~\cite{Angeles-Albores2016} to remove
+using a previously described trimming algorithm~\cite{Angeles-Albores2016} to remove
 terms that are too general and uninformative as well as terms that are very
 specific but do not have many annotations. When the user provides a set of genes
-to test against, we use a hypergeometric model to test whether a given term
-within an ontology is occurring at a greater frequency than expected by random
-chance.
+to test against, we use a hypergeometric model to test whether a given ontology
+term is occurring at a greater frequency than expected by random chance.
 
@@
 that had $>300$ associated genes. We identified 24 traits that met our criteria.
-Next, we used DIOPT~\cite{Hu2011} to identify candidate orthologs for the genes
+Next, we used DIOPT~\cite{Hu2011} to identify candidate \cel{} orthologs for the genes
 associated with these traits. Briefly, DIOPT combines a large number of methods
-for identifying orthologs and returns homolog candidates associated with a
+for identifying orthologs and returns ortholog candidates associated with a
 compound score. Depending on the score, orthologs can be considered `high',
-`moderate' or `low' rank, reflecting confidence in the homology. Many-to-one and
-one-to-many homology relationships are allowed in DIOPT, reflecting a mixture of
+`moderate' or `low' rank, reflecting confidence in the orthology. Many-to-one and
+one-to-many orthology relationships are allowed in DIOPT, reflecting a mixture of
 uncertainty and family expansion/reduction. For our study, we only accepted
-homolog candidates with `high' or `moderate' scores and we did not insist on a
+ortholog candidates with `high' or `moderate' scores and we did not insist on a
 one-to-one relationship between genes.
@@
            disease associated genes, then used DIOPT to identify candidate
-           orthologs in \cel{}. Orthologs were used to run phenotype, gene
-           and tissue ontology enrichment analyses to identify disease
+           orthologs in \cel{}. Orthologs were used to run tissue, phenotype,
+           gene ontology enrichment analyses to identify disease
            phenologs.}
@@
 used for TEA~\cite{Angeles-Albores2016}. We generated a dictionary that included
-terms with at least 50 annotating genes or more and had a similarity threshold
+terms which were annotated to at least 50 genes or more and had a similarity threshold
 of 0.95 for PEA (the total number of terms in the dictionary was 251, annotated
-by 9,169 genes for the version WS256);\@ and we generated a dictionary that
-included terms with at least 50 annotating genes or more and had a similarity
+to 9,169 genes for the version WS256);\@ and we generated a dictionary that
+included terms which are annotated to at least 50 genes or more and had a similarity
 threshold of 0.95 for GEA (the total number of terms in this dictionary is 271,
-annotated by 14,636 genes for the version WS256). \@ Next, we benchmarked the
-dictionaries on the same gene sets as TEA and obtained enrichment of all the
+annotated to 14,636 genes for the WormBase version WS256). \@ Next, we benchmarked the
+dictionaries on the same training gene sets as TEA and obtained enrichment of all the
 expected categories~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a,
@@
 thick filament variant'; the top two enriched GO terms were `myofibril' and
-`striated muscle dense body'. For all the benchmarking results, see
-supplementary information. Having generated and validated our dictionaries, we
-proceeded to identify phenologs for several common human diseases.
+`striated muscle dense body'. Having generated and validated our dictionaries,
+we proceeded to identify phenologs for several common human diseases.
 
@@
 disease in an unbiased manner. One way to discover gene associations in an
-unbiased manner is to perform GWSA in human populations. Therefore, we used the
+unbiased manner is to perform GWAS in human populations. Therefore, we used the
 GWAS NHGRI-EBI Catalog~\cite{MacArthur2016} to identify genes associated with
@@
 obtained \hlupus{} genes associated with the disease via GWAS studies, and found
-\wlupus{} homolog candidates in \cel{}.
+\wlupus{} ortholog candidates in \cel{}.
 
@@
 the Pn.p cells P3.p through P8.p were enriched in this dataset (5 genes,
-\qval{1}). GO enrichment pointed at `modification-dependent macromolecule
+\qval{1}). GO enrichment pointed to `modification-dependent macromolecule
 catabolic process' (23 genes, \qval{15}) as a molecular function that
@@
 single gene family, the \emph{skr} gene family. Almost the entire \emph{skr}
-family was considered a candidate homolog to the SKP1 human gene, making the GO
+family was considered a candidate ortholog to the SKP1 human gene, making the GO
 enrichment suspect.
 
-Enrichment of the terms for `aneuploidy', `meiotic chromosome segregation', and
+Enrichment of the terms `aneuploidy', `meiotic chromosome segregation', and
 `excess intestinal cells' were largely driven by the same gene group, which
@@
 involving cell fate transformation and leading to dysmorphias; and a molecular
-phenotype involving impairment of the nonsense-mediated decay pathway. % TODO:
-Is the text below true??? The results from the tissue enrichment analysis
+phenotype involving impairment of the nonsense-mediated decay pathway.
+The results from the tissue enrichment analysis
 highlighted three tissues that are particularly sensitive to \emph{lin}
-mutations (the gonad, the excretory duct cell and the vulval precursor cells),
-and the gonad arms undergo large quantities of nuclear proliferation.
+mutations (the gonad, the excretory duct cell and the vulval precursor cells).
+Gonad development also requires large quantities of nuclear proliferation.
 
@@
 distinct genetic causes. We found \harthritis{} genes associated with rheumatoid
-arthritis, for which we found \warthritis{} worm homolog candidates.
+arthritis, for which we found \warthritis{} worm ortholog candidates.
 
@@
 \qval{15}). However, this term was enriched as the result of degeneracy in the
-homolog candidates for the SFTPD gene. Other terms included `glycosylation' (10
+ortholog candidates for the SFTPD gene. Other terms included `glycosylation' (10
 genes, \qval{4}) and `Golgi apparatus' (11 genes, \qval{3}), but these terms
-were enriched as the result of degenerate homolog candidates for the human gene
+were enriched as the result of degenerate ortholog candidates for the human gene
 B3GNT7 which encodes a beta-1,3-N-acetylgalactosaminyltransferase.
@@
 transcriptomes, we selected a ciliary neuron dataset~\cite{Wang2015} and ran the
-complete WormBase Enrichment Suite on it. Ciliary neurons are present in the
+complete WormBase Enrichment Suite on it (see Listing~\ref{code} for a brief
+tutorial on how to do this in Python). Ciliary neurons are present in the
 \cel{} male tail, but they are also present in the male cephalic sensillum and
@@
 Fig.~\ref{fig:cilia}).
+
+\begin{lstlisting}[language=Python, caption=Brief programmatic example of how
+to perform a complete phentoype enrichment analysis in Python., label={code}]
+  import pandas as pd
+  import tissue_enrichment_analysis as tea
+  import matplotlib.pyplot as plt
+
+  # read in the file to be analyzed
+  # this file has two columns of interest:
+  # `gene_name', contains WormBase IDs
+  # `pval', which contains the p-value of each ORF
+  df = pd.read_csv('path/to/data_file.csv')
+
+  # import the phenotype dictionary to load into tea
+  phenotype_dictionary = tea.fetch('phenotype')
+
+  # get the gene IDs to be analyzed
+  # we only want those genes with pval < 0.05
+  gene_names = df[df.pval < 0.05].gene_IDs
+
+  # run the enrichment analysis.
+  # the line below generates a new pandas dataframe with the results
+  results = tea.enrichment_analysis(wbid_list, tissue_df=phenotype_dictionary)
+
+  # save the results to a file:
+  results.to_csv('path/to/analysis_results.csv', index=False)
+
+  # plot the results
+  # The `kind' variable depends on the enrichment analysis:
+  # Tissue, Phenotype or GO.
+  tea.plot_enrichment_results(results, kind='Phenotype')
+
+  # save the plot
+  plt.savefig('path/to/graph.pdf')
+\end{lstlisting}
 
@@
 \qval{5}), and `purine NTP-dependent helicase activity' (15 genes, \qval{1}).
-Visual inspection of list in question reveals that cell-cycle and DNA
-replication/repair genes are abundant in this transcriptome and include genes
-such as \emph{atm-1}, \emph{dna-2}, or \emph{hpr-17}. This analysis reveals that
-the ciliary neuron transcriptome is enriched in genes associated with
-microtubules, but includes genes that are thought to interact with DNA either
-via repair mechanisms or cell-cycle
+Visual inspection the genes driving enrichment of these terms reveals that
+cell-cycle and DNA replication/repair genes are abundant in this transcriptome
+and include genes such as \emph{atm-1}, \emph{dna-2}, or \emph{hpr-17}. This
+analysis reveals that the ciliary neuron transcriptome is enriched in genes
+associated with microtubules, but includes genes that are thought to interact
+with DNA either via repair mechanisms or cell-cycle
 control~\cite{Hofmann2000,Lee2003a,Bailly2010}.
@@
 components of the phenotype}
-How does a phenotype emerge? We realized that with the tools that we have
+How does a phenotype emerge? With the tools that we have
 developed, it is possible to understand what tissues contribute to a phenotype
@@
 egg~\cite{Sulston1977}. The vulva must be well-formed to allow passage of an
-egg, and the hermaphrodite-specific neuron (HSN) is involved in the egg-laying
-control~\cite{Schafer2005}. The complexity of the interactions that happen to
-allow egg-laying make understanding the Egl phenotype in terms of tissues a
-challenging task.
+egg, and the hermaphrodite-specific neuron (HSN) is involved in the control of
+egg-laying~\cite{Schafer2005}. The complexity of the interactions required for
+egg-laying makes understanding the Egl phenotype in terms of tissue interactions
+a challenging task.
 
@@
 However, vulF is also attached to a set of four additional vulval muscles, vm2.
-Why is vulF less associated with an Egl phenotype?
+% Why is vulF less associated with an Egl phenotype?
 
@@
 \begin{equation}
-  P( \text{a gene has Egl annotation} | \text{it is expressed in } X)
+  P( \text{a gene has Egl annotation} | \text{gene is expressed in } X)
 \end{equation}
@@
 We calculated the conditional probability that a gene has an Egl phenotype given
-that it's expression pattern includes a tissue term $X$ and we searched for the
+that its expression pattern includes a tissue term $X$ and we searched for the
 tissue terms that maximized this probability. The list of terms that maximized
@@
 `tail' (a body part with neurons and hypodermis). In general, the terms that had
-a high $P(\text{Egl}|X)$ did not have a high $P(X|\text{Egl})$. Additionally,
+a high $P(\text{Egl}|X)$ did not have a high $P(X|\text{Egl})$. In addition,
 the terms that had a high $P(X|\text{Egl})$ are broad terms that include a lot
@@
 
+\section*{Acknowledgments}
+This work was supported by HHMI with whom PWS is an investigator and by
+the NIH grant U41 HG002223.
+
 \bibliography{pea-citations}

### 2018-02-06 tissue_enrichment_analysis/hypergeometricTests.py
fixed bugs
fixed bugs where logq values would not be plotted if zero; fixed a bug
where labels appeared because of a bad column name; fixed bug where
plot_enrichment modified data frame each time it was called.

@@ -236,15 +236,14 @@ def plot_enrichment_results(df, y='logq', title='', analysis='tissue',
     ftype = kwargs.pop('ftype', 'svg')
 #
     if ax is None:
-        ax = plt.gca()
+        fig, ax = plt.subplots(figsize=(14, 8))
 
     # sort by q value change
     df.sort_values(['Q value', 'Enrichment Fold Change'],
                    ascending=[True, False], inplace=True)
 
     # make a logq bar:
-    if y.lower() == 'logq':
-        df['logq'] = -df['Q value'].apply(np.log10)
+    logq = -df['Q value'].apply(np.log10)
 
     # added August 26 2016:
     tissue_ID = 11
@@ -252,22 +251,22 @@ def plot_enrichment_results(df, y='logq', title='', analysis='tissue',
     go_ID = 10
 
     if analysis == 'phenotype':
-        df.Tissue = df.Term.str[:-pheno_ID-1]
+        yvals = df.Term.str[:-pheno_ID-1]
     elif analysis == 'tissue':
-        df.Tissue = df.Term.str[:-tissue_ID-1]
+        yvals = df.Term.str[:-tissue_ID-1]
     elif analysis == 'go':
-        df.Tissue = df.Term.str[:-go_ID-1]
+        yvals = df.Term.str[:-go_ID-1]
 
     # plot first n_bars
     with sns.axes_style('whitegrid'):
-        ax = sns.barplot(x=df[y][:n_bars], y=df.Term[:n_bars], ax=ax)
+        ax = sns.barplot(x=logq[:n_bars], y=yvals[:n_bars], ax=ax)
 
     # fix the plot to prettify it
     ax.set_ylabel('Terms', fontsize=15)
     if y.lower() != 'logq':
         ax.set_xlabel(y, fontsize=15)
     else:
-        ax.set_xlabel('$\log_{10}{q}$')
+        ax.set_xlabel('$\log_{10}{q}$', fontsize=15)
     ax.tick_params(axis='x', labelsize=13)
     ax.tick_params(axis='y', labelsize=13)
     plt.tight_layout()
